### 2015-09-14 paper.tex
...

% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{mathtools}
\everymath{\displaystyle}

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

\newcommand{\ST}{\1\xrightarrow[~n~]{}\1}
\newcommand{\BT}{\xRightarrow[(i,E)]{}}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}
\newcommand{\rr}[1] {{\textbf{\scriptsize{#1}}}}

\newcommand{\1}{\;}
\newcommand{\2}{\;\;}
\newcommand{\3}{\;\;\;}
\newcommand{\5}{\;\;\;\;\;}
\newcommand{\ten}{\5\5}
\newcommand{\twenty}{\ten\ten}

\usepackage{color}
\definecolor{light}{gray}{0.87}
\definecolor{dark}{gray}{0.30}
%\definecolor{light}{rgb}{.90,.90,.90}
\definecolor{darkgreen}{rgb}{0,.50,0}
\definecolor{darkblue}{rgb}{0,0,.50}
\definecolor{darkred}{rgb}{.50,0,0}
\definecolor{darkpur}{rgb}{.50,0,.50}

\usepackage{listings}
%\usepackage{textcomp}
\usepackage{url}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

% Metadata Information
%\acmVolume{9}
%\acmNumber{4}
%\acmArticle{39}
%\acmYear{2010}
%\acmMonth{3}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\doi{0000001.0000001}

%ISSN
%\issn{1234-56789}

% Document starts
\begin{document}

% Page heads
%\markboth{G. Zhou et al.}{A Multifrequency MAC Specially Designed for WSN 
%Applications}

% Title portion
\title{The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}

\author{
Francisco Sant'Anna
\affil{PUC--Rio}
Adriano Branco
\affil{PUC--Rio}
Roberto Ierusalimschy
\affil{PUC--Rio}
Noemi Rodriguez
\affil{PUC--Rio}
Silvana Rossetto
\affil{UFRJ}
}

% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms.
%
Its synchronous programming model allows for a simple reasoning about 
shared-memory concurrency.
Furthermore, its restricted semantics enables deterministic and memory-safe 
programs.

In this work, we propose a formal semantics for \CEU focusing on its particular 
control mechanisms, such as parallel compositions, finalization, and 
stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with C, and another for 
code dissemination in sensor networks.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
%\begin{CCSXML}
%<ccs2012>
 %<concept>
  %<concept_id>10010520.10010553.10010562</concept_id>
  %<concept_desc>Computer systems organization~Embedded systems</concept_desc>
  %<concept_significance>500</concept_significance>
 %</concept>
 %<concept>
  %<concept_id>10010520.10010575.10010755</concept_id>
  %<concept_desc>Computer systems organization~Redundancy</concept_desc>
  %<concept_significance>300</concept_significance>
 %</concept>
 %<concept>
  %<concept_id>10010520.10010553.10010554</concept_id>
  %<concept_desc>Computer systems organization~Robotics</concept_desc>
  %<concept_significance>100</concept_significance>
 %</concept>
 %<concept>
  %<concept_id>10003033.10003083.10003095</concept_id>
  %<concept_desc>Networks~Network reliability</concept_desc>
  %<concept_significance>100</concept_significance>
 %</concept>
%</ccs2012>
%\end{CCSXML}

%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

% We no longer use \terms command
%\terms{Design, Algorithms, Performance}

\keywords{
Concurrency, Determinism, Embedded Systems,
Esterel, Synchronous, Reactivity
}

%\acmformat{Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A.  
%Stankovic,
%and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially
%designed for  wireless sensor network applications.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

%\begin{bottomstuff}
%This work is supported by the National Science Foundation, under
%grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.

%Author's addresses: G. Zhou, Computer Science Department,
%College of William and Mary; Y. Wu  {and} J. A. Stankovic,
%Computer Science Department, University of Virginia; T. Yan,
%Eaton Innovation Center; T. He, Computer Science Department,
%University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
%(Current address) NASA Ames Research Center, Moffett Field, California 94035.
%\end{bottomstuff}

\maketitle

\section{Introduction}

- Esterel

- Ceu
    - first WSN
    - other domains, such as games
    - here, only static semantics
    - uses, classes, gsoc,

- synchronous limitations

%As a trade-off, our design imposes limitations on the language expressiveness, 
%such as doing computationally-intensive operations and meeting hard real-time 
%responsiveness.

\section{Overview of \CEU}
\label{sec.ceu}

\CEU is a synchronous reactive language based on Esterel~\cite{esterel.ieee91} 
with support for multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events that are broadcast to all awaiting trails.
By synchronous, we mean that all trails at any given time are either reacting 
to the current event or are awaiting another event;
in other words, trails are never reacting to different events.

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel and 
\CEU for the following control specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification is translated almost identically in the 
two languages, as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent (lines 4--9).
%
For the second phrase (i.e., the reset behavior), the Esterel version uses a 
\code{abort-when}, which serves the same purpose of \CEU's \code{par/or}:
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The same specification in Esterel and \CEU. %\newline
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\CEU (like Esterel) has a strong imperative flavor, with explicit control flow 
through sequences, loops, and also assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
Programs advance in sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

In the sections that follow, we discuss the four basic differences between \CEU 
and Esterel: deterministic execution for operations with side-effects
(Section~\ref{sec.ceu.det}), safe abortion for lines of execution 
(Section~\ref{sec.ceu.abrt}), first-class timers 
(Section~\ref{sec.ceu.timers}), and stack-based execution for internal events 
(Section~\ref{sec.ceu.ints}).

%\newpage %TTT
\subsection{Deterministic Execution}
\label{sec.ceu.det}

Esterel is deterministic with respect to reactive control: ``the same sequence 
of inputs always produces the same sequence of outputs''~\cite{esterel.primer}.
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in ``\code{call 
f1() || call f2()}'', the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
In \CEU, when multiple trails are active at a time, as in
``\code{par/and~do~\_f1()~with~\_f2()~end}'', they are scheduled in the order 
they appear in the program source code (i.e., \code{\_f1} executes first).
%
This way, \CEU is deterministic also with respect to the order of execution of 
side effects within a reaction.

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency possible.
In contrast, Esterel does not support shared memory: \emph{``if a variable is 
written by some thread, then it can neither be read nor be written by 
concurrent threads''}~\cite{esterel.primer}.
%
For embedded development, we believe that deterministic shared-memory 
concurrency is beneficial, given the extensive use of memory mapped ports for 
I/O and lack of hardware support for real parallelism.
Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The execution order between \code{f1} and \code{f2} is unspecified in 
Esterel, whereas in \CEU, \code{\_f1} always executes before \code{\_f2}.
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\subsection{Safe Abortion}
\label{sec.ceu.abrt}

The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution (i.e., awaiting \code{A} and 
\code{B}) without tweaking them with synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.

The code fragments of Figure~\ref{lst.abortion} shows a corner case for 
abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
%
For instance, it is not clear in \emph{code~a} in Esterel if the call to 
\code{f} should execute or not after \code{A}, given that the body and abortion 
events are the same.
%
For this reason, Esterel provides \emph{weak} and \emph{strong} variations for 
the \code{abort} statement.
With \emph{strong} abortion (the default), the body is aborted immediately and 
the call does not execute.
%
In \CEU, given the deterministic scheduling rules, strong and weak abortions 
can be chosen by reordering trails inside a \code{par/or}, e.g., in 
\emph{code~b}, the second trail is strongly aborted by the first trail and the 
call to \code{\_f} never executes.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
// ESTEREL
abort
    await A;
    call f();
when A;

\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
// CEU
par/or do
    await A;
with
    await A;
    _f();
end
\end{lstlisting}
\end{minipage}
%
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Thread abortion in Esterel and \CEU. %\newline
\label{lst.abortion}
}
\end{figure}

\subsection{First-class Timers}
\label{sec.ceu.timers}

% TODO: ana

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of WSNs, such as timeouts and sensor sampling.
However, support for wall-clock time is somewhat low-level in existing 
languages, usually through timer callbacks or blocking calls with ``sleep''.
%
In any concrete system implementation, however, a requested timeout does not 
expire precisely with zero-delay, a fact that is usually ignored in the 
development process.
We define the difference between the requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in 
sequence (or in a loop) may accumulate a considerable amount of deltas that can 
lead to incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
As an example, consider the following program:

\nopagebreak
\noindent
\begin{minipage}{\linewidth}
\begin{lstlisting}[xleftmargin=2em]
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;
\end{lstlisting}
\end{minipage}

Suppose that after the first \code{await} request, the underlying system gets 
busy and takes 15ms to check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program that follows, although the scheduler cannot 
guarantee that the first trail terminates exactly in 11ms, it can at least 
ensure that the program always terminates with \code{v=1}:

\nopagebreak
\noindent
%\begin{figure}[t]
%\rule{8.5cm}{0.37pt}
\begin{minipage}{\linewidth}
\begin{lstlisting}[xleftmargin=2em]
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
%\caption{ The first trail always terminates the program.
%\label{lst.time}
%}
%\end{figure}

Remember that any non-awaiting sequence is considered to take no time in the 
synchronous model.
Hence, the first trail is guaranteed to terminate before the second trail, 
because $10+1 < 12$.
A similar program in a language without first-class support for timers, would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.
%
The importance of synchronized timers becomes more evident in the presence of 
loops, like in the introductory example of Figure~\ref{lst.all} in which the 
first trail is guaranteed to execute exactly 21 times before being aborted by 
the timer in the second trail.

Note that in extreme scenarios, small timers in sequence (or in a loop) may 
never ``catch up'' with the external clock, resulting in a \emph{delta} that
increases indefinitely.
%There is nothing \CEU can do about that, as it is do
To deal with such cases, the current \emph{delta} is always returned from an 
\code{await} and can be used in programs:

\begin{lstlisting}
loop do
    var int late = await 1ms;
    if late < 1000 then
        <...>   // normal behavior
    else
        <...>   // abnormal behavior
    end
end
\end{lstlisting}

% TODO: uses in ports
% trickle?

\subsection{Internal Events}
\label{sec.ceu.ints}

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during the entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, however, internal events follow a stack-based execution policy, 
similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
For the version in Esterel, given that there is no control dependency between 
the calls to \code{f}, they may execute in any order after \code{A} and 
\code{B} (internally emitted).
%
For the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st]}
\item 2nd trail awakes (line 9), calls \code{\_f(1)}, and terminates.\\
    \emph{stack: [1st]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f(2)}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}[!t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f("2");
$\|$
    await B;
    call f("1");
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external
event void b;  // internal
par/and do
    await A;
    emit b;
    _f("2");
with
    await b;
    _f("1");
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (lines 3-6) that continuously 
awaits its identifying event (line 4), incrementing the value passed as 
reference (line 5).
A trail in parallel (lines 8-11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (line 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (line 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (lines 3-6), in parallel 
with the caller (lines 8-11).
\label{lst.sub}
}
\end{figure}

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself will always be 
ignored, given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring runtime stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, just like coroutines~\cite{lua.coroutines}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.
%
In Section~\ref{sec.adv.frp}, we take advantage of the lack of recursion to 
properly describe mutual dependency among trails in parallel.

\begin{comment}
\vspace{5pt}
\CEU provides no support for standard functions for a number of reasons:
\begin{itemize}
\item The interaction with other \CEU control primitives is not obvious (e.g., 
executing an $await$ or a $par/or$ inside a function).
\item They would still be restricted in some ways given the embedded context 
(e.g.  no recursion or closures).
\item Programs can always recur to $C$ functions for low-level operations.
%\item A dedicated primitive would behave just as described, being a matter of 
%syntactic sugar.
\end{itemize}

Regardless of the limitations, this form of subroutines is widely adopted in 
\CEU programs, given that they were designed to work with the other control 
mechanisms.
Keep in mind that the typical reactive organization of programs (awaiting an 
external stimulus, reacting to it, and going back to awaiting) does not demand 
unrestricted subroutines.
\end{comment}

\section{Formal Semantics}

%\begin{document}

The disciplined synchronous execution of \CEU, together with broadcast 
communication and stacked execution for internal events, may raise doubts about 
the precise execution of programs.
%
In this chapter, we introduce a reduced syntax of \CEU and propose an 
operational semantics in order to formally describe the language, eliminating 
imprecisions with regard to how a program reacts to an external event.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in any conventional 
imperative language).

%The full syntax and semantic rules can be obtained in 
%Appendix~\ref{app.formal}.

%In section~\ref{sec.formal.ceu}, we show how to translate \CEU statements into

\subsection{Abstract Syntax}
\label{sec.sem.syntax}

\begin{figure}[h]
%\rule{8.5cm}{0.37pt}
{\small
\begin{verbatim}
                                   // primary expressions
  p ::= mem(id)                    (any memory access to `id')
      | await(id)                  (await event `id')
      | emit(id)                   (emit event `id')
      | break                      (loop escape)
                                   // compound expressions
      | if mem(id) then p else p   (conditional)
      | p ; p                      (sequence)
      | loop p                     (repetition)
      | p and p                    (par/and)
      | p or p                     (par/or)
      | fin p                      (finalization)
                                   // derived by semantic rules
      | awaiting(id,n)             (awaiting `id' since sequence number `n')
      | emitting(n)                (emitting on stack level `n')
      | p @ loop p                 (unwinded loop)
\end{verbatim}
}%
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

Figure~\ref{lst.formal.syntax} shows the BNF-like syntax for a subset of \CEU 
that is sufficient to describe all semantic peculiarities of the language.
%
The $mem(id)$ primitive represents all accesses, assignments, and $C$ function 
calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Except for the $fin$ and semantic-derived expressions, which are discussed 
further, the other expressions map to their counterparts in the concrete 
language in Figure~\ref{lst.syntax}.
%
Note that $mem$ expressions cannot share identifiers with $await$/$emit$ 
expressions.

\subsection{Operational Semantics}
\label{sec.sem}

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
$$
\LL S, p \RR
    \xrightarrow[~~n~~]{}
\LL S', p' \RR
$$
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_1]) \\
p, p' &\in P
    && (as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (univocally~identifies~a~reaction~chain)
\end{align*}

At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).

%Therefore, there is a single order of execution for side effects (represented 
%by \code{mem} operations), thus, reaction chains are always deterministic.
%
% TODO
%\footnote{We could extend the semantics to describe the full execution of a 
%program by holding new incoming external events in a queue and processing them 
%in consecutive reaction chains that never overlap.}
%
%The semantics requires an explicit stack to properly nest the emission of 
%internal events.
%to prevent that $awaiting$ expressions awake in the same reaction they are 
%reached.

We describe this relation with a set of \emph{small-step} structural semantics 
rules, which are built in such a way that at most one transition is possible at 
any time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<n
    & \textbf{(awaiting)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(emitting)}
\end{align*}
}

An $await$ is simply transformed into an $awaiting$ that remembers the current 
external sequence number $n$ (rule \textbf{await}).
An $awaiting$ can only transit to a $nop$ (rule \textbf{awaiting}) if its 
referred event $id$ matches the top of the stack and its sequence number is 
smaller than the current one ($m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (in 
rule \textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{emitting} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}

Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives the memory identifier and current 
reaction sequence number, returning the current memory value.
%
Although the value is arbitrary, it is unique in a reaction chain, because a 
given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

The rules for loops are analogous to sequences, but use \code{`@'} as 
separators to properly bind breaks to their enclosing loops:

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (loop~p) \RR \ST \LL S, (p~@~loop~p) \RR
    & \textbf{(loop-expd)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
% -----------------------------------------------------------
    {\DS \LL S, (p~@~loop~q) \RR \ST \LL S', (p'~@~loop~q) \RR }
    & \textbf{(loop-adv)}    \\
%%%
& \LL S, (mem(id)~@~loop~p) \RR \ST \LL S, loop~p \RR
    & \textbf{(loop-nop)}    \\
%%%
& \LL S, (break~@~loop~p) \RR \ST \LL S, nop \RR
    & \textbf{(loop-brk)}
\end{eqnarray*}
}

When a program first encounters a $loop$, it first expands its body in sequence 
with itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$mem(id)$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

The rules for parallel $and$ compositions force transitions on the left branch 
$p$ to occur before transitions on the right branch $q$ (rules 
\textbf{and-adv1} and \textbf{and-adv2}).
Then, if one of the sides terminates, the composition is simply substituted by 
the other side (rules \textbf{and-nop1} and \textbf{and-nop2}):

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~mem(id)) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
\end{eqnarray*}
}

The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, defined in Figure~\ref{fig.isBlocked} and used in rule 
\textbf{and-adv2}, requiring the left branch $p$ to be blocked in order to 
allow the right transition from $q$ to $q'$.
%
An expression becomes blocked when all of its trails in parallel hang in 
$awaiting$ and $emitting$ expressions.

\begin{figure}[t]
{\small
\begin{align*}
  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m = n)   \\
  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, \_)             &= false \2  (mem,await,      \\
                                  &    \5\5\5\2 emit,break,if,loop)   %\\
\end{align*}
}%
\rule{14cm}{0.37pt}
\caption{
The recursive predicate $isBlocked$ is true only if all branches in parallel 
are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
\label{fig.isBlocked}
}
\end{figure}

\begin{figure}[t]
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( \_ )          &= mem(id)
\end{align*}
}%
\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

The last two rules \textbf{and-brk1} and \textbf{and-brk2} deal with a $break$ 
in each of the sides in parallel.
A $break$ should terminate the whole composition in order to escape the 
innermost loop (\emph{aborting} the other side).
%
The $clear$ function in the rules, defined in Figure~\ref{fig.formal.clear}, 
concatenates all active $fin$ bodies of the side being aborted (to execute 
before the $and$ rejoins).
Note that there are no transition rules for $fin$ expressions.
This is because once reached, an $fin$ expression only executes when it is 
aborted by a trail in parallel.
In Section~\ref{sec.formal.fins}, we show how an $fin$ is mapped to a 
finalization block in the concrete language.
%
Note that there is a syntactic restriction that an $fin$ body cannot $emit$ or 
$await$---they are guaranteed to completely execute within a reaction chain.

Most rules for parallel $or$ compositions are similar to $and$ compositions:

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   \\
%%%
& \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}

For a parallel $or$, the rules \textbf{or-nop1} and \textbf{or-nop2} must 
terminate the composition, and also apply the function $clear$ to the aborted 
side, in order to properly finalize it.

A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction chain.
%
However, $emitting$ expressions should resume their continuations of previous 
$emit$ in the ongoing reaction, they are just hanged in lower stack indexes 
(see rule \textbf{emit}).
%
Therefore, we define another relation that behaves as the previous if the 
program is not blocked, and, otherwise, pops the stack:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
$$
%
To describe a \emph{reaction chain} in \CEU, i.e., how a program behaves in 
reaction to a single external event, we use the reflexive transitive closure of 
this relation:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we need multiple 
``invocations'' of reaction chains, incrementing the sequence number:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
& ...
\end{align*}
%
Each invocation starts with an external event at the top of the stack and 
finishes with a modified program and an empty stack.
After each invocation, the sequence number is incremented.

\subsection{Concrete Language Mapping}

Although the reduced syntax presented in Figure~\ref{lst.formal.syntax} is 
similar to the concrete language in Figure~\ref{lst.syntax}, there are some 
significant mismatches between \CEU and the formal semantics that require some 
clarification.
In this section, we describe an informal mapping between the two.

Most statements from \CEU map directly to the formal semantics, e.g.,
$
    \code{if}      \mapsto if   ,\2
    \code{';'}     \mapsto ~';' ,\2
    \code{loop}    \mapsto loop ,\2
    \code{par/and} \mapsto and  ,\2
    \code{par/or}  \mapsto or
$.
(Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.)

\subsubsection{await and emit}

The \code{await} and \code{emit} primitives of \CEU are slightly more complex 
in comparison to the formal semantics, as they support communication of values 
between emits and awaits.
In the two-step translation below, we start with the program in \CEU, which 
communicates the value $1$ between the \code{emit} and \code{await} in parallel 
(left-most code).
In the intermediate translation, we include the shared variable \code{e\_} to 
hold the value being communicated between the two trails in order to simplify 
the \code{emit}.
Finally, we convert the program into the equivalent in the formal semantics, 
translating side-effect statements into $mem$ expressions:

\begin{figure}[h!]
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.34\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem
\end{lstlisting}
\end{minipage}
\end{figure}

Note that a similar translation is required for external events, i.e., each 
external event has a corresponding variable that is explicitly set by the 
environment before each reaction chain.

\subsubsection{First-class Timers}

To encompass first-class timers, we need a special \code{TICK} event that 
should be intercalated with each other event occurrence in an application (e.g.  
\emph{e1, e2}):

\begin{align*}
\LL [TICK], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [    ], p' \RR
\\
\LL [e1], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [TICK], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [    ], p''' \RR
\\
\LL [e2], p''' \RR
    & \xRightarrow[~~4~~]{*}
\LL [  ], p'''' \RR
\\
& ...
\end{align*}

The \code{TICK} event has an associated variable \code{TICK\_} (as illustrated 
in the previous section) with the time elapsed between the two occurrences of 
external events.

The translation in two steps from a timer await to the semantics is as follows:

\noindent
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
dt = await 10ms;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}
var int tot = 10000;
loop do
    await TICK;
    tot = tot - TICK_;
    if tot <= 0 then
        dt = tot;
        break;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
mem;
loop(
    await(TICK);
    mem;
    if mem then
        mem;
        break
    else
        nop
)
\end{lstlisting}
\end{minipage}

\subsubsection{Finalization Blocks}
\label{sec.formal.fins}

The biggest mismatch between \CEU and the formal semantics is regarding 
finalization blocks, which require more complex modifications in the program
for a proper mapping using the $fin$ semantic construct.
The code that follows uses a \code{finalize} to safely \code{\_release} the 
reference to \code{ptr} kept after the call to \code{\_hold}:

\begin{lstlisting}
do
    var int* ptr = <...>;
    await A;
    finalize
        _hold(ptr);
    with
        _release(ptr);
    end
    await B;
end
\end{lstlisting}

In the translation to the semantics, the first required modification is to 
catch the \code{do-end} termination to run the finalization code.
For this, we translate the block into a \code{par/or} with the original body in 
parallel with a $fin$ to run the finalization code:

\begin{lstlisting}
par/or do
    var int* ptr = <...>;
    await A;
    _hold(ptr);
    await B;
with
    { fin
        _release(ptr); }
end
\end{lstlisting}

In this intermediate code (mixing the syntaxes), the $fin$ body will execute
whenever the \code{par/or} terminates, either normally (after the \code{await 
B}) or aborted from an outer composition (rules \textbf{and-brk1}, 
\textbf{and-brk2}, \textbf{or-nop1}, \textbf{or-nop2}, \textbf{or-brk1}, and 
\textbf{or-brk2} in the semantics).
%
%Note the choice for a \code{par/or} which terminates only with the first 
%trail.
%
However, the $fin$ will also (incorrectly) execute even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%It may happen that the call to \code{\_release} occurs without a previous call 
%to \code{\_hold}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized:

\begin{lstlisting}[numbers=left,xleftmargin=2em]
f_ = 0;
par/or do
    var int* ptr = <...>;
    await A;
    _hold(ptr);
    f_ = 1;
    await B;
with
    { fin
        if f_ then
            _release(ptr);
        end }
end
\end{lstlisting}

The flag is initially set to false (line 1), avoiding the finalization code to 
execute (lines 9-12).
Only after the call to \code{\_hold} (line 5) that we set the flag to true 
(line 6) and enable the $fin$ body to execute.
%
The complete translation from the original example in \CEU is as follows:

\begin{lstlisting}
mem;    // f_ = 0
(
   mem;         // ptr = <...>
   await(A);
   mem;         // _hold(ptr)
   mem;         // f_ = 1
   await(B);
or
   fin
     if mem then    // if f_
        mem         // release _ptr
     else
        nop
)
\end{lstlisting}

\begin{comment}

There are also some restriction on valid programs in \CEU, which the semantic 
allows:
- loops
- awaits in fins
- local scopes

handled in separate, in a parsing phase that is not related to the execution

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The semantic rules are continuously applied (note the `$*$' klenee operator) 
until
consider only the top of the stack and continuously apply transformations to 
the program until it blocks and no rules can be applied.
The \emph{isBlocked} predicate of Figure~\ref{fig.isBlocked} identifies a 
blocked program based on its structure and event at the top of the stack.
xxx
A program becomes blocked when all parallel branches are hanged in $awaiting$, 
$stacked$, and/or $emitting$ primitives, as defined in 
Figure~\ref{fig:isBlocked}.
an \code{awaiting} is unblocked only if its event matches the top of the stack
the \code{emitting} primitive only proceeds once its stack level is restored.

\begin{itemize}
\item The program is awaiting in all trails, i.e., function $pop$ returns 
$(0,\{\})$.
\item The program terminates, i.e., the small-step rules transform the whole 
program into a $mem(id)$.
\end{itemize}
%
In Section~\ref{XXX} we show that by imposing syntactic restrictions to 
programs, reaction chains always reach one of these conditions in a finite 
number of steps, meaning that reactions to the environment always execute in 
bounded time.

To be compliant with the reactive nature of \CEU, we assume that all programs 
start awaiting the main event ``$\$$'', which is emitted once by the 
environment on startup, i.e., $(i,E)=(1,\{\$\})$ for the very first big step.

As briefly introduced, small-step rules continuously apply transformations to 
unblocked trails.
A program becomes blocked when all parallel branches are hanged in $awaiting$, 
$stacked$, and/or $emitting$ primitives, as defined in 
Figure~\ref{fig:isBlocked}.

All small-step rules are associated with the current (deepest) stack depth 
level $i$ acquired from the previous big step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{comment}
\section{Implementation}

\begin{comment}
As a static language, much of the complexity in the implementation of \CEU{} 
resides in the compile phase.
Nonetheless, some complexity is left to the runtime phase, which has to manage 
first-class timers, finalization blocks, and all bookkeeping related to trails.
\end{comment}

The compilation process of a program in \CEU is composed of three main phases, 
as illustrated in Figure~\ref{fig.impl}:

\begin{figure}[ht]
\centering
\includegraphics[scale=0.20]{impl}
\caption{ Compilation process: from the source code in \CEU to the final 
binary.
\label{fig.impl}
}
\end{figure}

\begin{comment}
The program below is used as our guiding example for this chapter:

\begin{lstlisting}[numbers=left,xleftmargin=2em]
input int A, B, C;
var int ret;
loop do
   par/or do
      do                            // 1st trail
          var int a = await A;
          ret = ret + a;
      end
      do
          var int b = await B;
          ret = ret + b;
      end
      break;
   with
      par/and do                    // 2nd trail
         finalize with
            ret = ret + 10;
         end
         await C;
      with
         await A;
      end
   end
end
...                                 // code after the loop
\end{lstlisting}
\end{comment}

\begin{description}

\item[Parsing]

The parser of \CEU is written in \emph{LPeg}~\cite{lua.lpeg}, a pattern 
matching library that also recognize grammars, making it possible to write the 
tokenizer and grammar with the same tool.
%
The source code is then converted to an \emph{abstract syntax tree (AST)} to be 
used in further phases.
%
This phase may be aborted due to syntax errors in the \CEU source file.

\item[Temporal analysis]

This phase detects inconsistencies in \CEU programs, such as unbounded loops 
and the forms of non-determinism.
%
It also makes some ``classical'' semantic analysis, such as building a symbol 
table for checking variable declarations.
However, most of type checking is delayed to the last phase to take advantage 
of GCC's error handling.
Therefore, this phase needs to annotate the $C$ output with \code{\#line} 
pragmas that match the original file in \CEU.
%
This phase must output code in $C$, given how tied \CEU is to $C$ by design.

\item[Final generation]

The final phase packs the generated $C$ file with the \CEU runtime and 
platform-dependent functionality, compiling them with \emph{gcc} and generating 
the final binary.
%
The \CEU runtime includes the scheduler, timer management, and the external $C$ 
API.
%
The platform files include libraries for I/O and bindings to invoke the \CEU 
scheduler on external events.

\end{description}

In the sections that follow, we discuss the most sensible parts of the compiler 
considering our design, such as the temporal analysis, runtime scheduler, and 
the external API.

\subsection{Temporal Analysis}

As introduced, the \emph{temporal analysis} phase detects inconsistencies in 
\CEU programs.
Here, we focus on the algorithm that detects non-deterministic access to 
variables, as presented in Section~\ref{sec.ceu.shared}.

For each node representing a statement in the program AST, we keep the set of 
events $I$ (for \emph{incoming}) that can lead to the execution of the node, 
and also the set of events $O$ (for \emph{outgoing}) that can terminate the 
node.

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, $C$ calls, and 
declarations simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminate, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from the sequence parent, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition (\code{par/and} / \code{par/or}) may terminate 
from any of its branches, hence $O = O_1 \cup ... \cup O_n$.
\end{itemize}

With all sets calculated, any two nodes that perform side effects and are in 
parallel branches can have their $I$ sets compared for intersections.
If the intersection is not the empty set, they are marked as suspicious (see 
Section~\ref{sec.ceu.shared}).

Figure~\ref{lst.impl.ast} reproduces the second code of Figure~\ref{lst.det} 
and shows the corresponding $AST$ with the sets $I$ and $O$ for each node.
The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (lines 5,8 in the code) have an empty 
intersection of $I$ (lines 6,9 in the AST), hence, they do not conflict.
Note that although the accesses in lines 5, 11 in the code (lines 6,11 in the 
AST) do have an intersection, they are not in parallel and are also safe.

\begin{figure}[h]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\end{minipage}
%
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\end{minipage}
%
\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

\begin{comment}
It is also responsible for setting the priorities for trails (see further) and 
determining the sizes of the queues that are used during runtime.

The program AST is first converted into a graph that represents the execution 
flow.
Figure~\ref{fig:nfa} shows the corresponding graph for our example.

\begin{figure}[ht]
\centering
\includegraphics[scale=0.40]{nfa.png}
\caption{ Flow graph for our guiding example
\label{fig:nfa}
}
\end{figure}

By default, all nodes in a flow graph have priority $0$ (highest).
However, as the figure shows, nodes that represent the termination of 
\emph{par/ors} and loops have lower priorities (the outer, the lower).
The priority scheme is needed to avoid glitches during runtime, and is 
equivalent to traversing a dependency graph in topological order, as employed 
in functional reactive programming implementations.~\cite{frtime.embedding}

The flow graph is then converted to a DFA, as exemplified in 
Section~\ref{sec:ceu:det}.

From its starting node, the flow graph is traversed until reaching await 
nodes---every visited node is inserted into a new DFA state.
Then, every set of awaiting nodes for a given external event starts another DFA 
state.
\end{comment}

\subsection{Memory Layout}
\label{sec:impl:memory}

\CEU{} favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and flags needed during runtime.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        u8 _and_3: 1;
        u8 _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
memory for blocks in sequence are packed in a \code{struct}, while blocks in 
parallel, in a \code{union}.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory address.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

\subsection{Trail Allocation}
\label{sec:impl:gates}

The compiler extracts the maximum number of trails a program can have at the 
same time and creates a static vector to hold runtime information about them.
Again, trails that cannot be active at the same time can share memory slots in 
the static vector.

At any given moment, a trail can be awaiting in one of the following states: 
\code{INACTIVE}, \code{STACKED}, \code{FIN}, or in any event defined in the 
program:

\begin{lstlisting}
enum {
    INACTIVE = 0,
    STACKED,
    FIN,
    EVT_A,      // input void A;
    EVT_e,      // event int e;
    <...>       // other events
}
\end{lstlisting}

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds its associated stack level and is delayed until 
the scheduler runtime level reaches that value again.
%
A \code{FIN} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the sequence number (\emph{seqno}) in which it started awaiting.
%
A trail is represented by the following \code{struct}:

\begin{lstlisting}
struct trail_t {
    state_t evt;
    label_t lbl;
    union {
        unsigned char seqno;
        stack_t       stk;
    };
};
\end{lstlisting}

The field \code{evt} holds the state of the trail (or the event it is 
awaiting); the field \code{lbl} holds the entry point in the code to execute 
when the trail is scheduled; the third field depends on the \code{evt} field 
and may hold the \code{seqno} for an event, or the stack level \code{stk} for a
\code{STACKED} state.

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two and 
requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} will eventually overflow during execution (every 256 
reactions).
However, given that the scheduler traverses all trails in each reaction, it can 
adjust them to properly handle overflows (actually 2 bits to hold the 
\code{seqno} would be already enough).
%
The stack size depends on the maximum depth of nested emissions and is bounded 
to the maximum number of trails, e.g., a trail emits an event that awakes 
another trail, which emits an event that awakes another trail, and so on---the 
last trail cannot awake any trail, because they will be all hanged in a 
\code{STACKED} state.
%
In WSNs applications, the size of \code{trail\_t} is typically only 3 bytes (1 
byte for each field).

\subsubsection{Code Generation}

\begin{figure}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
input void A;
event void e;
// TRAIL 0 - lbl Main
par/and do
  // TRAIL 0 - lbl Main
  await e;
  // TRAIL 0 - lbl Awake_e
  // TRAIL 0 - lbl ParAnd_chk
with
  // TRAIL 1 - lbl ParAnd_sub_2
  await A;
  // TRAIL 1 - lbl Awake_A_1
  emit e;
  // TRAIL 1 - lbl Emit_e_cont
  // TRAIL 1 - lbl ParAnd_chk
end
// TRAIL 0 - lbl ParAnd_out
await A;
// TRAIL 0 - lbl Awake_A_2
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
enum {
  Main = 1,     // ln  3
  Awake_e,      // ln  7
  ParAnd_chk,   // ln  8, 15
  ParAnd_sub_2, // ln 10
  Awake_A_1,    // ln 12
  Emit_e_cont,  // ln 14
  ParAnd_out,   // ln 17
  Awake_A_2     // ln 19
};
\end{lstlisting}
\end{minipage}
\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
Static allocation of trails and entry-point labels.
%}%}
\label{lst.impl.trails}
}
\end{figure}

The example in Figure~\ref{lst.impl.trails} illustrates how trails and labels 
are statically allocated in a program.
%
The program has a maximum of 2 trails, because the \code{par/and} (line 4) can 
reuse \emph{TRAIL 0}, and the join point (line 16) can reuse both \emph{TRAIL 
0} and \emph{TRAIL 1}.
%
Each label is associated with a unique identifier in the \code{enum}.
%
The static vector to hold the two trails in the example is defined as

\begin{lstlisting}
trail_t TRLS[2];
\end{lstlisting}

In the final generated $C$ code, each label becomes a \emph{switch case} 
working as the entry point to execute its associated code.
%
Figure~\ref{lst.impl.code} shows the corresponding code for the program of 
Figure~\ref{lst.impl.trails}.
%
The program is initialized with all trails set to \code{INACTIVE}.
Then, the scheduler executes the \emph{Main} label in the first trail.
%
When the \emph{Main} label reaches the \code{par/and}, it ``stacks'' the 2nd 
trail of the \code{par/and} to run on \emph{TRAIL 1} (line 5-8) and proceeds to 
the code in the 1st trail (lines 10-15), respecting the deterministic execution 
order.
%
The code sets the running \emph{TRAIL 0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of the scheduler takes \emph{TRAIL 1} and executes its 
registered label \code{ParAnd\_sub\_2} (lines 17-22), which sets \emph{TRAIL 1} 
to await \code{EVT\_A} and also halts.

\begin{figure}[t]
\begin{lstlisting}[numbers=left,xleftmargin=2em]
while (<...>) {             // scheduler main loop
   trail_t* trail = <...>   // choose next trail
   switch (trail->lbl) {
      case Main:
         // activate TRAIL 1 to run next
         TRLS[1].evt = STACKED;
         TRLS[1].lbl = ParAnd_sub_2;  // 2nd trail of par/and
         TRLS[1].stk = current_stack;

         // code in the 1st trail of par/and
         // await e;
         TRLS[0].evt = EVT_e;
         TRLS[0].lbl = Awake_e;
         TRLS[0].seq = current_seqno;
         break;

      case ParAnd_sub_2:
         // await A;
         TRLS[1].evt = EVT_A;
         TRLS[1].lbl = Awake_A_1;
         TRLS[1].seq = current_seqno;
            break;

        <...>   // other labels
    }
}
\end{lstlisting}
\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
Generated code for the program of Figure~\ref{lst.impl.trails}.
%}%}
\label{lst.impl.code}
}
\end{figure}

Regarding cancellation, trails in parallel are always allocated in subsequent 
slots in the static vector \code{TRLS}.
Therefore, when a \code{par/or} terminates, the scheduler sequentially searches 
and executes \code{FIN} trails within the range of the \code{par/or}, and then 
clears all of them to \code{INACTIVE} at once.
Given that finalization blocks cannot contain \code{await} statements, the 
whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also trigger the 
same process.

\subsection{The External $C$ API}

As a reactive language, the execution of a program in \CEU is guided entirely 
by the occurrence of external events.
From the implementation perspective, there are three external sources of input 
into programs, which are all exposed as functions in a $C$ API:

\begin{description}
\item[{\textbf\code{ceu\_go\_init()}}:] initializes the program (e.g. trails) 
and executes the ``boot'' reaction (i.e., the \code{Main} label).

\item[{\textbf\code{ceu\_go\_event(id,param)}}:] executes the reaction for the 
received event id and associated parameter.

\item[{\textbf\code{ceu\_go\_wclock(us)}}:] increments the current time in 
microseconds and runs a reaction if any timer expires.

%\item[{\textbf\code{ceu\_go\_async}}:] executes a single loop iteration for 
%the next \code{async}, switching among them in a \emph{round robin} policy.
\end{description}

Given the semantics of \CEU, the functions are guaranteed to take a bounded 
time to execute.
They also return a status code that says if the \CEU{} program has terminated 
after the reactions.
Further calls to the API have no effect on terminated programs.

\begin{comment}
Note that \CEU{} code running from a call to \code{ceu\_go\_async} may emit an 
input event or the passage of time.
In this case, the $C$ implementation makes a tail call to the corresponding 
handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as synchronous 
code has higher priority.

The API reflects the \emph{global asynchronous} part of \CEU{}, as discussed in 
Section~\ref{sec:ceu:gals}.
A simple and opaque API hides local state from the environment, suggesting that 
the execution varies entirely according to the sequence (and parameters) of API 
calls.
\end{comment}

The bindings for the specific platforms are responsible for calling the 
functions in the API in the order that better suit their requirements.
As an example, it is possible to set different priorities for events that occur 
concurrently (i.e. while a reaction chain is running).
However, a binding must never interleave or run multiple functions in parallel.
This would break the \CEU sequential/discrete semantics of time.
%, as discussed in Section~\ref{sec:ceu}.

% TODO:, allowing \CEU{} to be easily embedded in platforms:

\begin{figure}[t]
\begin{lstlisting}[numbers=left,xleftmargin=2em]
implementation
{
    #include "ceu.h"
    #include "ceu.c"

    event void Boot.booted () {
        ceu_go_init();
#ifdef CEU_WCLOCKS
        call Timer.startPeriodic(10);
#endif
    }

#ifdef CEU_WCLOCKS
    event void Timer.fired () {
        ceu_go_wclock(10000);
    }
#endif

#ifdef _EVT_PHOTO_READDONE
    event void Photo.readDone (uint16_t val) {
        ceu_go_event(EVT_PHOTO_READDONE, (void*)val);
    }
#endif

#ifdef _EVT_RADIO_SENDDONE
    event void RadioSend.sendDone (message_t* msg) {
        ceu_go_event(EVT_RADIO_SENDDONE, msg);
    }
#endif

#ifdef _EVT_RADIO_RECEIVE
    event message_t* RadioReceive.receive (message_t* msg) {
        ceu_go_event(EVT_RADIO_RECEIVE, msg);
        return msg;
    }
#endif

    <...>   // other events
}
\end{lstlisting}
\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
The \emph{TinyOS} binding for \CEU.
%}%}
\label{lst.impl.tinyos}
}
\end{figure}

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for \emph{TinyOS} 
which maps \emph{nesC} callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6-11) is called by TinyOS on mote 
startup, so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8-10) 
that triggers callback \code{Timer.fired} (lines 13-17) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 19-23) and the radio transceiver 
(lines 25-36).

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{my,other}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

% History dates
%\received{February 2007}{March 2009}{June 2009}

% Electronic Appendix
%\elecappendix

%\medskip

\end{document}
% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-09-15 paper.tex
...

@@
 Francisco Sant'Anna
-\affil{PUC--Rio}
+\affil{Departamento de Inform\'atica, PUC--Rio}
 Adriano Branco
-\affil{PUC--Rio}
+\affil{Departamento de Inform\'atica, PUC--Rio}
 Roberto Ierusalimschy
-\affil{PUC--Rio}
+\affil{Departamento de Inform\'atica, PUC--Rio}
 Noemi Rodriguez
-\affil{PUC--Rio}
+\affil{Departamento de Inform\'atica, PUC--Rio}
 Silvana Rossetto
-\affil{UFRJ}
+\affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
 }
@@
 %
-Its synchronous programming model allows for a simple reasoning about 
-shared-memory concurrency.
+Employing the synchronous programming model, it allows for a simple reasoning 
+about shared-memory concurrency.
 Furthermore, its restricted semantics enables deterministic and memory-safe 
 programs.
-
+%
+% TODO: some uses?
+%
 In this work, we propose a formal semantics for \CEU focusing on its particular 
@@
 
-- Esterel
-
-- Ceu
-    - first WSN
-    - other domains, such as games
-    - here, only static semantics
-    - uses, classes, gsoc,
-
-- synchronous limitations
+\begin{itemize}
+\item Relevancia de Esterel
+\item Ceu
+    \begin{itemize}
+        \item Foco inicial em "constrained WSNs"
+        \item Movendo para outros dominios? (jogos, artigo Mod'15)
+        \item Usos: SenSys, Terra, sala de aula, GSoC
+    \end{itemize}
+\item Limitacoes do modelo sincrono
+\item Semantica e Implementacao (nesse artigo, somente subset estatico)
+\end{itemize}
 
@@
 %
-The first phrase of the specification is translated almost identically in the 
-two languages, as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
-equivalent (lines 4--9).
-%
-For the second phrase (i.e., the reset behavior), the Esterel version uses a 
-\code{abort-when}, which serves the same purpose of \CEU's \code{par/or}:
+The first phrase of the specification, awaiting and emitting the events, is 
+translated almost identically in the two languages (lines 4--9, in both 
+implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
+equivalent.
+%
+For the second phrase, the reset behavior, the Esterel version uses a 
+\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
+\code{par/or} (lines 3--12):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
@@
 \begin{figure}[t]
-\begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[mathescape=true]
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
 // ESTEREL
@@
 %
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3.5em]
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ The same specification in Esterel and \CEU. %\newline
+\caption{ A control specification implemented in Esterel and \CEU:
+\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
+%\newline
 {\small
@@
 
-\CEU (like Esterel) has a strong imperative flavor, with explicit control flow 
-through sequences, loops, and also assignments.
+\CEU, following the Esterel mindset, has a strong imperative flavor, with 
+explicit control flow through sequences, loops, parallels, and also 
+assignments.
 Being designed for control-intensive applications, it provides support for 
@@
 %
-Programs advance in sequence of discrete reactions to external events.
+In the synchronous model, programs advance in a sequence of discrete reactions 
+to external events.
 Internal computations within a reaction (e.g. expressions, assignments, and 
@@
 
-In the sections that follow, we discuss the four basic differences between \CEU 
-and Esterel: deterministic execution for operations with side-effects
-(Section~\ref{sec.ceu.det}), safe abortion for lines of execution 
+In the sections that follow, we review the main differences between \CEU and 
+Esterel~\cite{ceu.sensys13}: deterministic execution 
+(Section~\ref{sec.ceu.det}), safe abortion with finalization 
 (Section~\ref{sec.ceu.abrt}), first-class timers 
-(Section~\ref{sec.ceu.timers}), and stack-based execution for internal events 
+(Section~\ref{sec.ceu.timers}), and stack-based internal events 
 (Section~\ref{sec.ceu.ints}).
@@
 
-Esterel is deterministic with respect to reactive control: ``the same sequence 
-of inputs always produces the same sequence of outputs''~\cite{esterel.primer}.
+Esterel is only deterministic with respect to reactive control: ``the same 
+sequence of inputs always produces the same sequence of 
+outputs''~\cite{esterel.primer}.
 However, the execution order for operations with side-effects within a reaction 
-is non-deterministic: ``if there is no control dependency, as in ``\code{call 
-f1() || call f2()}'', the order is unspecified and it would be an error to rely 
+is non-deterministic: ``if there is no control dependency, as in \code{"call 
+f1() || call f2()"}, the order is unspecified and it would be an error to rely 
 on it''~\cite{esterel.primer}.
 %
-In \CEU, when multiple trails are active at a time, as in
-``\code{par/and~do~\_f1()~with~\_f2()~end}'', they are scheduled in the order 
-they appear in the program source code (i.e., \code{\_f1} executes first).
-%
-This way, \CEU is deterministic also with respect to the order of execution of 
+In \CEU, when multiple trails are active at a time, they are scheduled in the 
+order they appear in the program source code.
+%
+Hence, \CEU is deterministic also with respect to the order of execution of 
 side effects within a reaction.
+%
+Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
+in Esterel and \CEU to illustrate the semantic difference regarding 
+(non-)determinism.
 
@@
 On the other hand, it provides a priority scheme for trails, and makes 
-shared-memory concurrency possible.
-In contrast, Esterel does not support shared memory: \emph{``if a variable is 
-written by some thread, then it can neither be read nor be written by 
+shared-memory concurrency more tractable.
+In contrast, Esterel precludes support for shared memory: \emph{``if a variable 
+is written by some thread, then it can neither be read nor be written by 
 concurrent threads''}~\cite{esterel.primer}.
 %
-For embedded development, we believe that deterministic shared-memory 
-concurrency is beneficial, given the extensive use of memory mapped ports for 
-I/O and lack of hardware support for real parallelism.
+For constrained embedded development, we believe that deterministic 
+shared-memory concurrency is beneficial, given the extensive use of memory 
+mapped ports for I/O and the lack of hardware support for real parallelism.
 Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
@@
 
-\begin{figure}[t]
-\begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[mathescape=true]
+\begin{figure}[b]
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
 // ESTEREL
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ The execution order between \code{f1} and \code{f2} is unspecified in 
-Esterel, whereas in \CEU, \code{\_f1} always executes before \code{\_f2}.
+\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
+unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
 {\small
@@
 }
-\label{lst.abro}
+\label{lst.det}
 }
@@
 
-\subsection{Safe Abortion}
+\subsection{Safe Abortion with Finalization}
 \label{sec.ceu.abrt}
@@
 The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
-languages can abort awaiting lines of execution (i.e., awaiting \code{A} and 
-\code{B}) without tweaking them with synchronization primitives.
+languages can abort awaiting lines of execution without tweaking them with 
+synchronization primitives.
 In contrast, traditional (asynchronous) multi-threaded languages cannot express 
 thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
-
-The code fragments of Figure~\ref{lst.abortion} shows a corner case for 
+However, handling abortion when dealing with external resources can be 
+challenging, given that external entities are not subject to the same 
+synchronous execution discipline.
+
+\begin{figure}[t]
+\begin{minipage}[t]{0.55\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+input void STOP, RETRANSMIT, SENDACK;
+par/or do
+    await STOP;
+with
+    loop do
+        par/or do
+            await RETRANSMIT;
+        with
+            par/and do
+                await 1min;
+            with
+                var _pkt_t buffer;
+                <fill-buffer-info>
+                _send_enqueue(&buffer);
+                await SENDACK;
+            end
+        end
+    end
+end
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.41\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
+<...>
+    var _pkt_t buffer;
+    <fill-buffer-info>
+    finalize
+        _send_enqueue(&buffer);
+    with
+        _send_cancel(&buffer);
+    end
+    await SENDACK;
+\end{lstlisting}
+\end{minipage}
+%\rule{8.5cm}{0.37pt}
+\caption{A network protocol in \CEU extended with a finalization clause to 
+cancel a transmission.
+%\caption{ Parallel compositions can describe complex state machines.
+\label{lst.fin}
+}
+\end{figure}
+
+The example in the left of Figure~\ref{lst.fin} relies on hierarchical 
+\code{par/or} and \code{par/and} compositions to describe the state machine of 
+a data collection protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
+%
+The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
+represent the external interface of the protocol with a client application.
+%
+The protocol has two trails:
+one monitors the stopping event (line 3);
+the other periodically transmits a packet (lines 5--18).
+%
+The periodic transmission is a loop that starts two other trails (lines 6--17):
+one handles an immediate retransmission request (line 7);
+the other transmits the packet awaiting for a confirmation (lines 9--16).
+%
+The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
+takes at least one minute before looping, to avoid flooding the network with 
+packets.
+%
+At any time, the client may request a retransmission (line 7), which terminates 
+the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
+idle), and restarts the loop (line 5).
+%
+The client may also request to stop the whole protocol at any time (line 3).
+%
+Note that the \code{buffer} packet (line 12) is a local variable whose address 
+is passed to function \code{\_send\_enqueue} for transmission.
+The call enqueues the pointer in the radio driver, which holds it up to the 
+emission of \code{SENDACK} acknowledging the packet transmission.
+%
+In the meantime, if the sending trail is be aborted by the \code{STOP} or 
+\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
+a \emph{dangling pointer} in the radio driver.
+%
+%\subsubsection{Finalization}
+%\label{sec.ceu.fin}
+
+%Programs in \CEU can access \emph{C} libraries available in the underlying 
+%platform by prefixing symbols with an underscore (e.g., 
+%\code{\_printf(<...>)}).
+%
+The \CEU compiler tracks the interaction of \code{par/or} compositions with 
+local variables and stateful \emph{C} functions (e.g., device drivers) in order 
+to preserve safe abortion of trails.
+%
+%Finalization clauses are fundamental to preserve the orthogonality of 
+%\code{par/or} compositions in SSRP.
+%
+For instance, \CEU refuses to compile the program in the left of 
+Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
+clause to accompany the stateful \emph{C} call~\cite{ceu.sensys13}.
+The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
+transmission when the block of \code{buffer} goes out of scope, i.e., the 
+finalization clause (after the \code{with}) executes automatically on external 
+abortion.%
+\footnote{
+Note that the compiler only enforces the programmer to write the finalization 
+clause, but cannot check if it actually handles the resource properly.
+}
+%
+%Note also that \CEU environments rely on \emph{C} libraries that only provide 
+%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.
+
+\begin{comment}
+The code fragments of Figure~\ref{lst.abortion} shows a corner case regarding
 abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
 %
-For instance, it is not clear in \emph{code~a} in Esterel if the call to 
-\code{f} should execute or not after \code{A}, given that the body and abortion 
-events are the same.
+For instance, it is not clear in the code in Esterel if the call to \code{f} 
+should execute or not after \code{A}, given that the body and abortion events 
+are the same.
 %
@@
 can be chosen by reordering trails inside a \code{par/or}, e.g., in 
-\emph{code~b}, the second trail is strongly aborted by the first trail and the 
-call to \code{\_f} never executes.
+Figure~\ref{lst.abortion}, the second trail is strongly aborted by the first 
+trail and the call to \code{\_f} never executes.
 
 \begin{figure}[t]
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}
@@
 %
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}
@@
 %
-\end{lstlisting}
-\end{minipage}
 %\rule{8.4cm}{0.37pt}
-\caption{ Thread abortion in Esterel and \CEU. %\newline
+\caption{ Strong abortion in Esterel and \CEU. %\newline
 \label{lst.abortion}
@@
 \end{figure}
+\end{comment}
 
@@
 \label{sec.ceu.timers}
-
-% TODO: ana
 
@@
 }
-appear in typical patterns of WSNs, such as timeouts and sensor sampling.
+appear in typical patterns of embedded development, such as timeouts and sensor 
+sampling.
 However, support for wall-clock time is somewhat low-level in existing 
-languages, usually through timer callbacks or blocking calls with ``sleep''.
-%
-In any concrete system implementation, however, a requested timeout does not 
+languages, usually through timer callbacks or ``sleep'' blocking calls.
+%
+However, in any concrete system implementation, a requested timeout does not 
 expire precisely with zero-delay, a fact that is usually ignored in the 
@@
 automatically, resulting in more robust applications.
-As an example, consider the following program:
-
-\nopagebreak
-\noindent
-\begin{minipage}{\linewidth}
-\begin{lstlisting}[xleftmargin=2em]
+For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
+first \code{await} request, the underlying system gets busy and takes 15ms to 
+check for expiring awaits.
+The \CEU scheduler will notice that the \code{await 10ms} has not only already 
+expired, but is delayed with \code{delta=5ms}.
+Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
+As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
+the trail is rescheduled for execution, now with \code{delta=4ms}.
+
+\begin{figure}[t]
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}
 var int v;
@@
 \end{minipage}
-
-Suppose that after the first \code{await} request, the underlying system gets 
-busy and takes 15ms to check for expiring awaits.
-The \CEU scheduler will notice that the \code{await 10ms} has not only already 
-expired, but delayed with \code{delta=5ms}.
-Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
-As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
-the trail is rescheduled for execution, now with \code{delta=4ms}.
-
-\CEU also takes into account the fact that time is a physical quantity that can 
-be added and compared.
-For instance, for the program that follows, although the scheduler cannot 
-guarantee that the first trail terminates exactly in 11ms, it can at least 
-ensure that the program always terminates with \code{v=1}:
-
-\nopagebreak
-\noindent
-%\begin{figure}[t]
-%\rule{8.5cm}{0.37pt}
-\begin{minipage}{\linewidth}
-\begin{lstlisting}[xleftmargin=2em]
+%
+\begin{minipage}[t]{0.51\linewidth}
+\begin{lstlisting}
 par/or do
@@
 %\rule{8.5cm}{0.37pt}
-%\caption{ The first trail always terminates the program.
-%\label{lst.time}
-%}
-%\end{figure}
-
-Remember that any non-awaiting sequence is considered to take no time in the 
-synchronous model.
-Hence, the first trail is guaranteed to terminate before the second trail, 
-because $10+1 < 12$.
+\caption{ First-class timers in \CEU.
+\label{lst.timers}
+}
+\end{figure}
+
+\CEU also takes into account the fact that time is a physical quantity that can 
+be added and compared.
+For instance, for the program in the right of Figure~\ref{lst.timers}, although 
+the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
+it can at least ensure that the program always terminates with \code{v=1}:
+%
+Given that any non-awaiting sequence is considered to take no time in the 
+synchronous model, the first trail is guaranteed to terminate before the second 
+trail, because $10+1 < 12$.
+%
 A similar program in a language without first-class support for timers, would 
@@
 reasoning about the execution behavior more difficult.
-%
-The importance of synchronized timers becomes more evident in the presence of 
-loops, like in the introductory example of Figure~\ref{lst.all} in which the 
-first trail is guaranteed to execute exactly 21 times before being aborted by 
-the timer in the second trail.
-
-Note that in extreme scenarios, small timers in sequence (or in a loop) may 
-never ``catch up'' with the external clock, resulting in a \emph{delta} that
-increases indefinitely.
-%There is nothing \CEU can do about that, as it is do
-To deal with such cases, the current \emph{delta} is always returned from an 
-\code{await} and can be used in programs:
-
-\begin{lstlisting}
-loop do
-    var int late = await 1ms;
-    if late < 1000 then
-        <...>   // normal behavior
-    else
-        <...>   // abnormal behavior
-    end
-end
-\end{lstlisting}
-
-% TODO: uses in ports
-% trickle?
 
@@
 %
-For the version in Esterel, given that there is no control dependency between 
-the calls to \code{f}, they may execute in any order after \code{A} and 
-\code{B} (internally emitted).
-%
-For the version in \CEU, the occurrence of \code{A} makes the program behave as 
+In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
+and they are both become active, resulting in the invocation of \code{f()} and 
+\code{g()} in no particular order.
+%
+In the version in \CEU, the occurrence of \code{A} makes the program behave as 
 follows (with the stack contents in italics):
@@
 
-\begin{figure}[!t]
-\begin{minipage}[t]{0.45\linewidth}
+\begin{figure}[b]
+\begin{minipage}[t]{0.43\linewidth}
 \begin{lstlisting}[mathescape=true]
@@
     emit B;
-    call f("2");
+    call f();
 $\|$
     await B;
-    call f("1");
+    call g();
 ]]
@@
 %
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.53\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3.1em]
 // CEU
-input void A;  // external
-event void b;  // internal
+input void A;  // external (in uppercase)
+event void b;  // internal (in lowercase)
 par/and do
@@
     emit b;
-    _f("2");
+    _f();
 with
     await b;
-    _f("1");
+    _g();
 end
@@
 On the one hand, this form of subroutines has a significant limitation that it 
-cannot express recursive calls: an \code{emit} to itself will always be 
-ignored, given that a running body cannot be awaiting itself.
+cannot express recursive calls: an \code{emit} to itself is always ignored, 
+given that a running body cannot be awaiting itself.
 %
@@
 first, they are guaranteed to react in bounded time;
-second, memory for locals is also bounded, not requiring runtime stacks.
+second, memory for locals is also bounded, not requiring data stacks.
 %
@@
 %exceptions.
-%
-In Section~\ref{sec.adv.frp}, we take advantage of the lack of recursion to 
-properly describe mutual dependency among trails in parallel.
 
@@
 
-The disciplined synchronous execution of \CEU, together with broadcast 
-communication and stacked execution for internal events, may raise doubts about 
-the precise execution of programs.
-%
-In this chapter, we introduce a reduced syntax of \CEU and propose an 
-operational semantics in order to formally describe the language, eliminating 
-imprecisions with regard to how a program reacts to an external event.
+%The disciplined synchronous execution of \CEU, together with broadcast 
+%communication and stacked execution for internal events, may raise doubts 
+%%about the precise execution of programs.
+%
+In this section, we introduce a reduced syntax of \CEU and propose an 
+operational semantics in order to formally describe the language.
+%, eliminating imprecisions with regard to how a program reacts to an external 
+%event.
 %
@@
 \begin{figure}[h]
-\begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
 input void A, B;
@@
 %
-\begin{minipage}[t]{0.55\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=2.5em]
@@
 \begin{figure}[t]
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}
@@
 %
-\begin{minipage}[t]{0.55\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}
@@
 \begin{figure}[t]
-\begin{minipage}[t]{0.55\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=2em]

### 2015-09-18 paper.tex
(*) review

@@
         \item Foco inicial em "constrained WSNs"
-        \item Movendo para outros dominios? (jogos, artigo Mod'15)
+        \item Estendendo para outros dominios? (jogos, artigo Mod'15)
         \item Usos: SenSys, Terra, sala de aula, GSoC
@@
 \item Semantica e Implementacao (nesse artigo, somente subset estatico)
+% Artigo de Rust, nao resolve os problemas
 \end{itemize}
@@
 %
-In the synchronous model, programs advance in a sequence of discrete reactions 
-to external events.
+In addition, \CEU also employs the synchronous model, in whicn programs advance 
+in a sequence of discrete reactions to external events.
 Internal computations within a reaction (e.g. expressions, assignments, and 
@@
 Hence, \CEU is deterministic also with respect to the order of execution of 
-side effects within a reaction.
+side effects within a reaction~\cite{ceu.sensys13}.
 %
@@
 shared-memory concurrency more tractable.
-In contrast, Esterel precludes support for shared memory: \emph{``if a variable 
-is written by some thread, then it can neither be read nor be written by 
-concurrent threads''}~\cite{esterel.primer}.
+In contrast, Esterel forbids sharing memory: \emph{``if a variable is written 
+by some thread, then it can neither be read nor be written by concurrent 
+threads''}~\cite{esterel.primer}.
 %
@@
     await SENDACK;
+<...>
 \end{lstlisting}
@@
 
-The example in the left of Figure~\ref{lst.fin} relies on hierarchical 
-\code{par/or} and \code{par/and} compositions to describe the state machine of 
-a data collection protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
+To illustrate threats related to abortion, consider the example in the left of 
+Figure~\ref{lst.fin} with hierarchical \code{par/or} and \code{par/and} 
+compositions describing the state machine of a data collection protocol for 
+sensor networks~\cite{wsn.ctp,ceu.sensys13}.
 %
@@
 The protocol has two trails:
-one monitors the stopping event (line 3);
+one simply monitors the stopping event (line 3);
 the other periodically transmits a packet (lines 5--18).
@@
 one handles an immediate retransmission request (line 7);
-the other transmits the packet awaiting for a confirmation (lines 9--16).
+the other transmits the packet%
+\footnote{
+The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
+with external \emph{C} libraries explicit and are required in \CEU.
+}
+and waits for a confirmation (lines 9--16).
+%
 %
@@
 
-%Programs in \CEU can access \emph{C} libraries available in the underlying 
-%platform by prefixing symbols with an underscore (e.g., 
-%\code{\_printf(<...>)}).
-%
 The \CEU compiler tracks the interaction of \code{par/or} compositions with 
 local variables and stateful \emph{C} functions (e.g., device drivers) in order 
-to preserve safe abortion of trails.
+to preserve safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
 %
@@
 Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
-clause to accompany the stateful \emph{C} call~\cite{ceu.sensys13}.
+clause to accompany the stateful \emph{C} call.
 The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
@@
 }
-appear in typical patterns of embedded development, such as timeouts and sensor 
-sampling.
+appear in typical patterns of embedded development, such as timeout watchdogs 
+and sensor samplings.
 However, support for wall-clock time is somewhat low-level in existing 
@@
 %
-However, in any concrete system implementation, a requested timeout does not 
-expire precisely with zero-delay, a fact that is usually ignored in the 
+Furthermore, in any concrete timer implementation, a requested timeout does not 
+expire precisely without delays, a fact that is usually ignored in the 
 development process.
@@
 The \code{await} statement of \CEU supports wall-clock time and handles deltas 
-automatically, resulting in more robust applications.
+automatically~\cite{ceu.sensys13}, resulting in more robust applications.
 For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
@@
 In \CEU, however, internal events follow a stack-based execution policy, 
-similar to subroutine calls in typical programming languages.
+similar to subroutine calls in typical programming 
+languages~\cite{ceu.sensys13}, contrasting with queue-based event handling 
+adopted for external events.
 %
@@
 In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
-and they are both become active, resulting in the invocation of \code{f()} and 
+and they both become active, resulting in the invocation of \code{f()} and 
 \code{g()} in no particular order.
@@
     \emph{stack: [1st]}
-\item 2nd trail awakes (line 9), calls \code{\_f(1)}, and terminates.\\
+\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
     \emph{stack: [1st]}
-\item 1st trail (on top of the stack) resumes, calls \code{\_f(2)}, and 
+\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
     terminates.\\
@@
 
-\begin{figure}[b]
+\begin{figure}%[b]
 \begin{minipage}[t]{0.43\linewidth}
@@
 In particular, they await keeping context information such as locals and the 
-program counter, just like coroutines~\cite{lua.coroutines}.
+program counter, similarly to coroutines~\cite{lua.coroutines}.
 %In Section~\ref{sec.adv.excpt} we show how to use them to implement 
@@
 
-%The disciplined synchronous execution of \CEU, together with broadcast 
-%communication and stacked execution for internal events, may raise doubts 
-%%about the precise execution of programs.
-%
 In this section, we introduce a reduced syntax of \CEU and propose an 
-operational semantics in order to formally describe the language.
-%, eliminating imprecisions with regard to how a program reacts to an external 
-%event.
+operational semantics to formally describe the language.
+We describe a small synchronous kernel with broadcast communication
+highlighting the main differences to Esterel, in particular the stack-based
+execution for internal events.
 %
@@
 
-%The full syntax and semantic rules can be obtained in 
-%Appendix~\ref{app.formal}.
-
-%In section~\ref{sec.formal.ceu}, we show how to translate \CEU statements into
-
 \subsection{Abstract Syntax}
@@
 \begin{figure}[h]
-%\rule{8.5cm}{0.37pt}
-{\small
-\begin{verbatim}
+\begin{lstlisting}[mathescape=true]
                                    // primary expressions
   p ::= mem(id)                   (any memory access to `id')
-      | await(id)                  (await event `id')
-      | emit(id)                   (emit event `id')
-      | break                      (loop escape)
+      $|$ await(id)                  (await event `id')
+      $|$ emit(id)                   (emit event `id')
+      $|$ break                      (loop escape)
                                    // compound expressions
-      | if mem(id) then p else p   (conditional)
-      | p ; p                      (sequence)
-      | loop p                     (repetition)
-      | p and p                    (par/and)
-      | p or p                     (par/or)
-      | fin p                      (finalization)
+      $|$ if mem(id) then p else p   (conditional)
+      $|$ p ; p                      (sequence)
+      $|$ loop p                     (repetition)
+      $|$ p and p                    (par/and)
+      $|$ p or p                     (par/or)
+      $|$ fin p                      (finalization)
                                    // derived by semantic rules
-      | awaiting(id,n)             (awaiting `id' since sequence number `n')
-      | emitting(n)                (emitting on stack level `n')
-      | p @ loop p                 (unwinded loop)
-\end{verbatim}
-}%
+      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
+      $|$ emitting(n)                (emitting on stack level `n')
+      $|$ p @ loop p                 (unwinded loop)
+\end{lstlisting}
 \caption{
@@
 %
+Except for $fin$ and all semantic-derived expressions (i.e., $awating$, 
+$emitting$, and $p~@~loop; p$), which are discussed further, all other 
+expressions map to their counterparts in the concrete language.
+
 The $mem(id)$ primitive represents all accesses, assignments, and $C$ function 
@@
 %
-Except for the $fin$ and semantic-derived expressions, which are discussed 
-further, the other expressions map to their counterparts in the concrete 
-language in Figure~\ref{lst.syntax}.
-%
-Note that $mem$ expressions cannot share identifiers with $await$/$emit$ 
-expressions.
+Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
+an identifier is either a variable or an event.
 
@@
 %
-$$
-\LL S, p \RR
-    \xrightarrow[~~n~~]{}
+\begin{align*}
+\LL S, p \RR &\ST
 \LL S', p' \RR
-$$
+    & \textbf{(relation-inner)}
+\end{align*}
 %
@@
 S, S' &\in id^*
-    &&(sequence~of~event~identifiers: [id_{top}, ..., id_1]) \\
+    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
 p, p' &\in P
-    && (as~described~in~Figure~\ref{lst.formal.syntax}) \\
+    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
 n     &\in \mathds{N}
-    && (univocally~identifies~a~reaction~chain)
+    && (unique~identifier~for~the~reaction~chain)
 \end{align*}
@@
 events on top of it (we discuss how they are popped further).
+The ever-increasing sequence number $n$ will prevent that awaiting expressions 
+awake in the same reaction they are reached (ensuring that the program can 
+progress).
 
@@
 \LL S,\1await(id) \RR &\ST
-\LL S,\1awaiting(id,n) \RR
-    & \textbf{(await)}      \\
+\LL S,\1awaiting(id,n+1) \RR
+    & \textbf{(awaiting)}      \\
 %%%
 \LL id:S,\1awaiting(id,m) \RR &\ST
-\LL id:S,\1nop \RR, \2m<n
-    & \textbf{(awaiting)}   \\
+\LL id:S,\1nop \RR, \2m<=n
+    & \textbf{(awake)}   \\
 %%%
@@
 \LL id:S,\1emitting(|S|) \RR
-    & \textbf{(emit)}       \\
+    & \textbf{(emitting)}       \\
 %%%
@@
 \LL S,\1nop \RR
-    & \textbf{(emitting)}
+    & \textbf{(pop)}
 \end{align*}
@@
 
-An $await$ is simply transformed into an $awaiting$ that remembers the current 
-external sequence number $n$ (rule \textbf{await}).
-An $awaiting$ can only transit to a $nop$ (rule \textbf{awaiting}) if its 
-referred event $id$ matches the top of the stack and its sequence number is 
-smaller than the current one ($m<n$).
+An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
+an artifact to remember the external sequence number $n$ it can awake.
+An $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
+event $id$ matches the top of the stack and its sequence number is smaller or 
+equal than the current one ($m<n$).
 %
@@
 An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
-stands for the stack length), and pushing the referred event on the stack (in 
-rule \textbf{emit}).
+stands for the stack length), and pushing the referred event on the stack (rule 
+\textbf{emitting}).
 With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
-rule \textbf{emitting} expects its parameter to match the current stack level.
+rule \textbf{pop} expects its parameter to match the current stack level.
 This trick provides the desired stack-based semantics for internal events.
@@
 %
-Although the value is arbitrary, it is unique in a reaction chain, because a 
-given expression can execute only once within it (remember that $loops$ must 
-contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
+Although the value here is arbitrary, it is unique in a reaction chain, because 
+a given expression can execute only once within it (remember that $loops$ must 
+contain $awaits$ which, from rule \textbf{awaiting}, cannot awake in the same 
 reaction they are reached).
@@
 
-When a program first encounters a $loop$, it first expands its body in sequence 
-with itself (rule \textbf{loop-expd}).
+When a program encounters a $loop$, it first expands its body in sequence with 
+itself (rule \textbf{loop-expd}).
 Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
@@
 
-The rules for parallel $and$ compositions force transitions on the left branch 
-$p$ to occur before transitions on the right branch $q$ (rules 
-\textbf{and-adv1} and \textbf{and-adv2}).
-Then, if one of the sides terminates, the composition is simply substituted by 
-the other side (rules \textbf{and-nop1} and \textbf{and-nop2}):
+Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
+always force transitions on their left branches $p$ to occur before their right 
+branches $q$:
 
@@
 %%%
+& \frac
+    {\DS \LL S,p \RR \ST \LL S',p' \RR }
+%   -----------------------------------------------------------
+    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
+    & \textbf{(or-adv1)}   \\
+%%%
+& \frac
+    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
+%   -----------------------------------------------------------
+    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
+    & \textbf{(or-adv2)}   \\
+\end{eqnarray*}
+}
+
+The deterministic behavior of the semantics relies on the \emph{isBlocked} 
+predicate, defined in Figure~\ref{fig.isBlocked} and used in rules 
+\textbf{and-adv2} and \textbf{or-adv2}, requiring the left branch $p$ to be 
+blocked in order to allow the right transition from $q$ to $q'$.
+%
+Basically, the \emph{isBlocked} predicate determines that an expression becomes 
+blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
+expressions.
+
+\begin{figure}[t]
+{\small
+\begin{align*}
+  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m = n)   \\
+  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
+  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
+  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
+  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
+                                    isBlocked(n,S,q)             \\
+  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
+                                    isBlocked(n,S,q)             \\
+  isBlocked(n,S, \_)             &= false \2  (mem,await,      \\
+                                  &    \5\5\5\2 emit,break,if,loop)   %\\
+\end{align*}
+}%
+%\rule{14cm}{0.37pt}
+\caption{
+The recursive predicate $isBlocked$ is true only if all branches in parallel 
+are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
+\label{fig.isBlocked}
+}
+\end{figure}
+
+For a parallel $and$, if one of the sides terminates, the composition is simply 
+substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}).
+The last two rules \textbf{and-brk1} and \textbf{and-brk2} deal with a $break$ 
+in each of the sides in parallel.
+A $break$ should terminate the whole composition in order to escape the 
+innermost loop (\emph{aborting} the other side):
+
+{ \setlength{\jot}{20pt}
+\begin{eqnarray*}
 & \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
@@
 
-The deterministic behavior of the semantics relies on the \emph{isBlocked} 
-predicate, defined in Figure~\ref{fig.isBlocked} and used in rule 
-\textbf{and-adv2}, requiring the left branch $p$ to be blocked in order to 
-allow the right transition from $q$ to $q'$.
-%
-An expression becomes blocked when all of its trails in parallel hang in 
-$awaiting$ and $emitting$ expressions.
-
-\begin{figure}[t]
-{\small
-\begin{align*}
-  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m = n)   \\
-  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
-  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
-  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
-  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
-                                    isBlocked(n,S,q)             \\
-  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
-                                    isBlocked(n,S,q)             \\
-  isBlocked(n,S, \_)             &= false \2  (mem,await,      \\
-                                  &    \5\5\5\2 emit,break,if,loop)   %\\
-\end{align*}
-}%
-\rule{14cm}{0.37pt}
-\caption{
-The recursive predicate $isBlocked$ is true only if all branches in parallel 
-are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
-\label{fig.isBlocked}
-}
-\end{figure}
+%
+The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
+all active $fin$ bodies of the side being aborted (to execute before the $and$ 
+rejoins).
+Note that there are no transition rules for $fin$ expressions.
+This is because once reached, a $fin$ expression halts and will only execute 
+when it is aborted by a trail in parallel and is expanded by the $clear$ 
+function.
+In Section~\ref{sec.formal.fins}, we show how a $fin$ is mapped to a 
+finalization block in the concrete language.
+%
+Note that there is a syntactic restriction that a $fin$ body can only contain 
+$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
+reaction chain.
 
@@
 }%
-\rule{14cm}{0.37pt}
+%\rule{14cm}{0.37pt}
 \caption{
@@
 
-The last two rules \textbf{and-brk1} and \textbf{and-brk2} deal with a $break$ 
-in each of the sides in parallel.
-A $break$ should terminate the whole composition in order to escape the 
-innermost loop (\emph{aborting} the other side).
-%
-The $clear$ function in the rules, defined in Figure~\ref{fig.formal.clear}, 
-concatenates all active $fin$ bodies of the side being aborted (to execute 
-before the $and$ rejoins).
-Note that there are no transition rules for $fin$ expressions.
-This is because once reached, an $fin$ expression only executes when it is 
-aborted by a trail in parallel.
-In Section~\ref{sec.formal.fins}, we show how an $fin$ is mapped to a 
-finalization block in the concrete language.
-%
-Note that there is a syntactic restriction that an $fin$ body cannot $emit$ or 
-$await$---they are guaranteed to completely execute within a reaction chain.
-
-Most rules for parallel $or$ compositions are similar to $and$ compositions:
+For a parallel $or$, if one of the sides terminates, the whole composition 
+terminates and applies the function $clear$ to the aborted side in order to 
+properly finalize it (rules \textbf{or-nop1} and \textbf{or-nop2}).
+Finally, a $break$ (rules \textbf{or-brk1} and \textbf{or-brk2}) behaves in the 
+same way as in a parallel $and$:
 
@@
 \begin{eqnarray*}
-& \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
-    & \textbf{(or-adv1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
-    & \textbf{(or-adv2)}   \\
-%%%
 & \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
@@
 
-For a parallel $or$, the rules \textbf{or-nop1} and \textbf{or-nop2} must 
-terminate the composition, and also apply the function $clear$ to the aborted 
-side, in order to properly finalize it.
-
 A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
@@
 $emit$ in the ongoing reaction, they are just hanged in lower stack indexes 
-(see rule \textbf{emit}).
-%
-Therefore, we define another relation that behaves as the previous if the 
-program is not blocked, and, otherwise, pops the stack:
+(see rule \textbf{pop}).
+%
+Therefore, we define another relation that behaves as \textbf{relation-inner} 
+(presented above) if the program is not blocked, and, otherwise, pops the 
+stack:
 %
@@
     { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
+\5\5\5\textbf{(relation-outer)}
 $$
@@
 reaction to a single external event, we use the reflexive transitive closure of 
-this relation:
+\textbf{relation-outer}:
 %
@@
 
-Although the reduced syntax presented in Figure~\ref{lst.formal.syntax} is 
-similar to the concrete language in Figure~\ref{lst.syntax}, there are some 
-significant mismatches between \CEU and the formal semantics that require some 
-clarification.
-In this section, we describe an informal mapping between the two.
-
-Most statements from \CEU map directly to the formal semantics, e.g.,
+Most statements from \CEU map directly to those presented in the reduced syntax 
+of Figure~\ref{lst.formal.syntax}, namely, the \code{if}, \code{`;'}, 
+\code{loop}, \code{par/and}, and \code{par/or}.
+For instance, the \code{if} in the concrete language behaves exactly like the 
+$if$ in the described semantics.
+However, there are some significant mismatches between \CEU and the formal 
+semantics, which we provide a mapping in this section.
+
+\begin{comment}
 $
@@
 $.
-(Again, we are not considering side-effects, which are all mapped to the $mem$ 
-semantic construct.)
+\end{comment}
+Again, we are not considering side-effects, which are all mapped to the $mem$ 
+semantic construct.
 
@@
 
-The \code{await} and \code{emit} primitives of \CEU are slightly more complex 
-in comparison to the formal semantics, as they support communication of values 
-between emits and awaits.
-In the two-step translation below, we start with the program in \CEU, which 
-communicates the value $1$ between the \code{emit} and \code{await} in parallel 
-(left-most code).
+The concrete \code{await} and \code{emit} primitives of \CEU support 
+communication of values between them.
+In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
+the concrete program in \CEU, which communicates the value $1$ between the 
+\code{emit} and \code{await} in parallel (left-most code).
 In the intermediate translation, we include the shared variable \code{e\_} to 
@@
 Finally, we convert the program into the equivalent in the formal semantics, 
-translating side-effect statements into $mem$ expressions:
-
-\begin{figure}[h!]
+translating side-effect statements into $mem$ expressions.
+External events require a similar translation, i.e., each external event has a 
+corresponding variable that is explicitly set by the environment before each 
+reaction chain.
+
+\begin{figure}%[h!]
 \begin{minipage}[t]{0.32\linewidth}
@@
 %
-\begin{minipage}[t]{0.34\linewidth}
+\begin{minipage}[t]{0.32\linewidth}
 \begin{lstlisting}
@@
 \end{minipage}
+\caption{
+Two-step translation from concrete to abstract \code{emit} and \code{await} 
+expressions.
+\label{lst.map.emit.await}
+}
 \end{figure}
 
-Note that a similar translation is required for external events, i.e., each 
-external event has a corresponding variable that is explicitly set by the 
-environment before each reaction chain.
-
 \subsubsection{First-class Timers}
 
-To encompass first-class timers, we need a special \code{TICK} event that 
-should be intercalated with each other event occurrence in an application (e.g.  
-\emph{e1, e2}):
+To encompass first-class timers, we introduce a special external \code{TICK} 
+event that is intercalated with each other event occurrence in an application 
+(e.g.  \emph{e1, e2}):
 
@@
 
-The \code{TICK} event has an associated variable \code{TICK\_} (as illustrated 
-in the previous section) with the time elapsed between the two occurrences of 
-external events.
-
-The translation in two steps from a timer await to the semantics is as follows:
-
-\noindent
+The \code{TICK} event has an associated variable \code{TICK\_} carrying the 
+wall-clock time elapsed between the two occurrences, as depicted by the 
+two-step translation of Figure~\ref{lst.map.timers.2}.
+
+TODO: continuar explicacao
+
+\begin{figure}%[h!]
 \begin{minipage}[t]{0.30\linewidth}
@@
 \begin{lstlisting}
-var int tot = 10000;
+var int tot = 10000; // 10ms
 loop do
@@
 \end{minipage}
+\caption{
+Two-step translation from concrete to abstract timer.
+\label{lst.map.timers.2}
+}
+\end{figure}
 
@@
 for a proper mapping using the $fin$ semantic construct.
-The code that follows uses a \code{finalize} to safely \code{\_release} the 
-reference to \code{ptr} kept after the call to \code{\_hold}:
-
+In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
+concrete program (\code{CODE-1}) that uses a \code{finalize} to safely 
+\code{\_release} the reference to \code{ptr} kept after the call to 
+\code{\_hold}.
+In the translation to the semantics, we first need to catch the \code{do-end} 
+termination to run the finalization code.
+For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
+original body in parallel with a $fin$ to run the finalization code.
+This way, the $fin$ body executes whenever the \code{par/or} terminates, either 
+normally (after the \code{await B}) or aborted from an outer composition.
+%
+However, the $fin$ still (incorrectly) executes even if the call to 
+\code{\_hold} is not reached in the body due to an abort before awaking from 
+the \code{await A}.
+%
+To deal with this issue, for each $fin$ we need a corresponding flag to keep 
+track of code that needs to be finalized (\code{CODE-3}).
+%
+The flag is initially set to false, avoiding the finalization code to execute.
+Only after the call to \code{\_hold} that we set the flag to true and enable 
+the $fin$ body to execute.
+%
+The complete translation substitutes the side-effect operations with $mem$ 
+expressions (\code{CODE-4}).
+
+\begin{figure}%[h!]
+\begin{minipage}[t]{0.26\linewidth}
 \begin{lstlisting}
@@
 end
-\end{lstlisting}
-
-In the translation to the semantics, the first required modification is to 
-catch the \code{do-end} termination to run the finalization code.
-For this, we translate the block into a \code{par/or} with the original body in 
-parallel with a $fin$ to run the finalization code:
-
+
+// CODE-1
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.26\linewidth}
 \begin{lstlisting}
@@
 end
-\end{lstlisting}
-
-In this intermediate code (mixing the syntaxes), the $fin$ body will execute
-whenever the \code{par/or} terminates, either normally (after the \code{await 
-B}) or aborted from an outer composition (rules \textbf{and-brk1}, 
-\textbf{and-brk2}, \textbf{or-nop1}, \textbf{or-nop2}, \textbf{or-brk1}, and 
-\textbf{or-brk2} in the semantics).
-%
-%Note the choice for a \code{par/or} which terminates only with the first 
-%trail.
-%
-However, the $fin$ will also (incorrectly) execute even if the call to 
-\code{\_hold} is not reached in the body due to an abort before awaking from 
-the \code{await A}.
-%It may happen that the call to \code{\_release} occurs without a previous call 
-%to \code{\_hold}.
-%
-To deal with this issue, for each $fin$ we need a corresponding flag to keep 
-track of code that needs to be finalized:
-
-\begin{lstlisting}[numbers=left,xleftmargin=2em]
+
+// CODE-2
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.26\linewidth}
+\begin{lstlisting}
 f_ = 0;
@@
 end
-\end{lstlisting}
-
-The flag is initially set to false (line 1), avoiding the finalization code to 
-execute (lines 9-12).
-Only after the call to \code{\_hold} (line 5) that we set the flag to true 
-(line 6) and enable the $fin$ body to execute.
-%
-The complete translation from the original example in \CEU is as follows:
-
+
+// CODE-3
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.18\linewidth}
 \begin{lstlisting}
-mem;    // f_ = 0
+mem;
 (
-   mem;         // ptr = <...>
+   mem;
    await(A);
-   mem;         // _hold(ptr)
-   mem;         // f_ = 1
+   mem;
+   mem;
    await(B);
@@
    fin
-     if mem then    // if f_
-        mem         // release _ptr
+   if mem then
+     mem
    else
@@
 )
-\end{lstlisting}
+// CODE-4
+\end{lstlisting}
+\end{minipage}
+%
+\caption{
+Three-step translation from concrete to abstract finalization.
+\label{lst.map.fin}
+}
+\end{figure}
 
@@
 
-\item[Temporal analysis]
-
-This phase detects inconsistencies in \CEU programs, such as unbounded loops 
-and the forms of non-determinism.
-%
-It also makes some ``classical'' semantic analysis, such as building a symbol 
-table for checking variable declarations.
-However, most of type checking is delayed to the last phase to take advantage 
-of GCC's error handling.
-Therefore, this phase needs to annotate the $C$ output with \code{\#line} 
-pragmas that match the original file in \CEU.
-%
-This phase must output code in $C$, given how tied \CEU is to $C$ by design.
-
-\item[Final generation]
+\item[Analysis]
+
+This phase detects inconsistencies in \CEU programs, such as unbounded loops,
+and also makes ``classical'' semantic analysis, such as building a symbol table 
+for checking variable declarations.
+%
+This phase outputs code in $C$, given the way \CEU is tied to $C$ by design.
+%
+Some type checking is delayed to the last phase to take advantage of 
+\code{gcc}'s error handling.
+Therefore, we annotate the $C$ output file with \code{\#line} pragmas matching 
+the original file in \CEU.
+
+\item[Code Generation]
 
@@
 %
-The \CEU runtime includes the scheduler, timer management, and the external $C$ 
-API.
+The \CEU runtime comprehends the scheduler, timer management, and the external 
+$C$ API.
 %
@@
 
-In the sections that follow, we discuss the most sensible parts of the compiler 
-considering our design, such as the temporal analysis, runtime scheduler, and 
-the external API.
-
+%In the sections that follow, we discuss the most sensible parts of the 
+%\emph{code generation} phase.
+%compiler considering our design, such as the temporal analysis, runtime 
+%scheduler, and the external API.
+
+\begin{comment}
 \subsection{Temporal Analysis}
@@
 %
-\rule{14cm}{0.37pt}
+%\rule{14cm}{0.37pt}
 \caption{ A program with a corresponding AST describing the sets $I$ and $O$.
@@
 
-\begin{comment}
+begin{comment}
 It is also responsible for setting the priorities for trails (see further) and 
@@
 state.
+end{comment}
+
 \end{comment}
@@
 
-\CEU{} favors a fine-grained use of trails, being common the use of trails that 
+\CEU favors a fine-grained use of trails, being common the use of trails that 
 await a single event.
@@
 \end{minipage}
-\rule{14cm}{0.37pt}
+%\rule{14cm}{0.37pt}
 \caption{
@@
 their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
-\code{MEM.b\_2} can safely share the same memory address.
+\code{MEM.b\_2} can safely share the same memory slot.
 %
@@
 
-The compiler extracts the maximum number of trails a program can have at the 
-same time and creates a static vector to hold runtime information about them.
-Again, trails that cannot be active at the same time can share memory slots in 
-the static vector.
+Each line of execution in \CEU carries associated data, such as which event it 
+is awaiting and which code to execute when it awakes.
+%
+The compiler statically infers the maximum number of trails a program can have 
+at the same time and creates a static vector to hold the runtime information 
+about them.
+%
+Trails that cannot be active at the same time can share memory slots in the 
+static vector.
 
 At any given moment, a trail can be awaiting in one of the following states: 
-\code{INACTIVE}, \code{STACKED}, \code{FIN}, or in any event defined in the 
-program:
+\code{INACTIVE}, \code{STACKED}, \code{FIN}, or in any of the events defined in 
+the program:
 
@@
 A trail waiting for an event stays in the state of the corresponding event, 
-also holding the sequence number (\emph{seqno}) in which it started awaiting.
+also holding the minimum sequence number (\emph{seqno}) in which it can awake.
 %
@@
 adjust them to properly handle overflows (actually 2 bits to hold the 
-\code{seqno} would be already enough).
+\code{seqno} is already enough).
 %
@@
 %
-In WSNs applications, the size of \code{trail\_t} is typically only 3 bytes (1 
-byte for each field).
+In general, in constrained sensor network applications, the size of 
+\code{trail\_t} is typically only 3 bytes (1 byte for each field), imposing a 
+negligible overhead even for trails that only await a single event and 
+terminate.
 
@@
 \begin{figure}[t]
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=2em]
+\begin{minipage}[t]{0.31\linewidth}
+\begin{lstlisting}
 input void A;
@@
   // TRAIL 0 - lbl Awake_e
-  // TRAIL 0 - lbl ParAnd_chk
+  // TRAIL 0 - lbl And_chk
 with
-  // TRAIL 1 - lbl ParAnd_sub_2
+  // TRAIL 1 - lbl And_sub_2
   await A;
@@
   emit e;
-  // TRAIL 1 - lbl Emit_e_cont
-  // TRAIL 1 - lbl ParAnd_chk
-end
-// TRAIL 0 - lbl ParAnd_out
+  // TRAIL 1 - lbl Emit_cont
+  // TRAIL 1 - lbl And_chk
+end
+// TRAIL 0 - lbl And_out
 await A;
@@
 %
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}
+\begin{minipage}[t]{0.31\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
 enum {
@@
   Awake_e,   // ln  7
-  ParAnd_chk,   // ln  8, 15
-  ParAnd_sub_2, // ln 10
+  And_chk,   // ln  8,15
+  And_sub_2, // ln 10
   Awake_A_1, // ln 12
-  Emit_e_cont,  // ln 14
-  ParAnd_out,   // ln 17
+  Emit_cont, // ln 14
+  And_out,   // ln 17
   Awake_A_2  // ln 19
@@
 \end{minipage}
-\rule{14cm}{0.37pt}
+%
+\begin{minipage}[t]{0.36\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
+while (<...>) {
+  trail_t* trail = <...>
+  switch (trail->lbl) {
+  case Main:
+    // activate TRAIL 1
+    TRLS[1].evt = STACKED;
+    TRLS[1].lbl = And_sub_2;
+    TRLS[1].stk = cur_stack;
+
+    // code in the 1st trail
+    // await e;
+    TRLS[0].evt = EVT_e;
+    TRLS[0].lbl = Awake_e;
+    TRLS[0].seq = cur_seqno;
+    break;
+
+  case And_sub_2:
+    // await A;
+    TRLS[1].evt = EVT_A;
+    TRLS[1].lbl = Awake_A_1;
+    TRLS[1].seq = cur_seqno;
+      break;
+
+    <...>  // other labels
+  }
+}
+\end{lstlisting}
+\end{minipage}
 \caption{
 %{\small %\textmd{
-Static allocation of trails and entry-point labels.
+Static allocation of trails, entry-point labels, and generated code.
 %}%}
@@
 The program is initialized with all trails set to \code{INACTIVE}.
-Then, the scheduler executes the \emph{Main} label in the first trail.
-%
-When the \emph{Main} label reaches the \code{par/and}, it ``stacks'' the 2nd 
-trail of the \code{par/and} to run on \emph{TRAIL 1} (line 5-8) and proceeds to 
-the code in the 1st trail (lines 10-15), respecting the deterministic execution 
-order.
-%
-The code sets the running \emph{TRAIL 0} to await \code{EVT\_e} on label 
+Then, the scheduler executes the \emph{Main} label in \code{TRAIL-0}.
+%
+When the \emph{Main} label reaches the \code{par/and}, it ``stacks'' 
+\code{TRAIL-1} (line 5--8) and proceeds to the code in \code{TRAIL-0} (lines 
+10--15), respecting the deterministic execution order.
+%
+The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
 \code{Awake\_e}, and then halts with a \code{break}.
 %
-The next iteration of the scheduler takes \emph{TRAIL 1} and executes its 
-registered label \code{ParAnd\_sub\_2} (lines 17-22), which sets \emph{TRAIL 1} 
+The next iteration of the scheduler takes \code{TRAIL-1} and executes its 
+registered label \code{ParAnd\_sub\_2} (lines 17-22), which sets \code{TRAIL-1} 
 to await \code{EVT\_A} and also halts.
-
-\begin{figure}[t]
-\begin{lstlisting}[numbers=left,xleftmargin=2em]
-while (<...>) {             // scheduler main loop
-   trail_t* trail = <...>   // choose next trail
-   switch (trail->lbl) {
-      case Main:
-         // activate TRAIL 1 to run next
-         TRLS[1].evt = STACKED;
-         TRLS[1].lbl = ParAnd_sub_2;  // 2nd trail of par/and
-         TRLS[1].stk = current_stack;
-
-         // code in the 1st trail of par/and
-         // await e;
-         TRLS[0].evt = EVT_e;
-         TRLS[0].lbl = Awake_e;
-         TRLS[0].seq = current_seqno;
-         break;
-
-      case ParAnd_sub_2:
-         // await A;
-         TRLS[1].evt = EVT_A;
-         TRLS[1].lbl = Awake_A_1;
-         TRLS[1].seq = current_seqno;
-            break;
-
-        <...>   // other labels
-    }
-}
-\end{lstlisting}
-\rule{14cm}{0.37pt}
-\caption{
-%{\small %\textmd{
-Generated code for the program of Figure~\ref{lst.impl.trails}.
-%}%}
-\label{lst.impl.code}
-}
-\end{figure}
 
@@
 Escaping a \code{loop} that contains parallel compositions also trigger the 
-same process.
+same finalization process.
+
+TODO internal emit?
 
@@
 \end{lstlisting}
-\rule{14cm}{0.37pt}
+%\rule{14cm}{0.37pt}
 \caption{
@@
 
+\section{Related Work}
+
+- Kiel Esterel compiler
+
+- We are not concerned with WCET:
+Predictable system design is concerned with the challenge o
+f building sys-
+tems in such a way that requirements can be guaranteed from th
+e design. This
+means that an off-line analysis should demonstrate satisfa
+ction of timing require-
+ments, subject to assumptions made on operating conditions
+foreseen for the sys-
+tem [Stankovic and Ramamritham 1990]
+
+PRET-C
+http://embedded.cs.uni-saarland.de/publications/BuildingTimingPredictableEmbeddedSystems.pdf
+
+\section{Final Remarks}
+
 % Bibliography

### 2015-09-19 paper.tex
revisao da implementacao

@@
 
+TODO: names (other embedded languages)
+
 \begin{figure}[b]
@@
 Figure~\ref{lst.sub}.
-The subroutine \code{inc} is defined as a loop (lines 3-6) that continuously 
+The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
 awaits its identifying event (line 4), incrementing the value passed as 
 reference (line 5).
-A trail in parallel (lines 8-11) invokes the subroutine in reaction to event 
+A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
 \code{A} through an \code{emit} (line 10).
@@
 \end{lstlisting}
-\caption{ Subroutine \code{inc} is defined in a loop (lines 3-6), in parallel 
-with the caller (lines 8-11).
+\caption{ Subroutine \code{inc} is defined in a loop (lines 3--6), in parallel 
+with the caller (lines 8--11).
 \label{lst.sub}
@@
 \code{STACKED} state.
-%
-In general, in constrained sensor network applications, the size of 
-\code{trail\_t} is typically only 3 bytes (1 byte for each field), imposing a 
-negligible overhead even for trails that only await a single event and 
-terminate.
-
-\subsubsection{Code Generation}
+
+In the context of embedded systems, the size of \code{trail\_t} is typically 
+only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
+even for trails that only await a single event and terminate.
+%
+For instance, the \emph{CTP} collection protocol ported to \CEU has at most 
+eight simultaneous lines of execution with an overhead of 2\% in comparison to 
+the original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
+programming)~\cite{ceu.sensys13}.
+
+\subsection{Code Generation and Scheduling}
 
@@
 // TRAIL 0 - lbl Awake_A_2
+
+.
 \end{lstlisting}
@@
 };
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.36\linewidth}
+
+trail_t TRLS[2] = {
+  { STACKED,  Main, 0 };
+  { INACTIVE, 0,    0 };
+};
+
+.
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.37\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=2.5em]
-while (<...>) {
-  trail_t* trail = <...>
-  switch (trail->lbl) {
+void dispatch (trail_t* t) {
+  switch (t->lbl) {
     case Main:
@@
 %{\small %\textmd{
-Static allocation of trails, entry-point labels, and generated code.
+Static allocation of trails, entry-point labels, and dispatch loop.
 %}%}
@@
 
-The example in Figure~\ref{lst.impl.trails} illustrates how trails and labels 
-are statically allocated in a program.
-%
-The program has a maximum of 2 trails, because the \code{par/and} (line 4) can 
-reuse \emph{TRAIL 0}, and the join point (line 16) can reuse both \emph{TRAIL 
-0} and \emph{TRAIL 1}.
-%
-Each label is associated with a unique identifier in the \code{enum}.
-%
-The static vector to hold the two trails in the example is defined as
-
-\begin{lstlisting}
-trail_t TRLS[2];
-\end{lstlisting}
-
-In the final generated $C$ code, each label becomes a \emph{switch case} 
-working as the entry point to execute its associated code.
-%
-Figure~\ref{lst.impl.code} shows the corresponding code for the program of 
-Figure~\ref{lst.impl.trails}.
-%
-The program is initialized with all trails set to \code{INACTIVE}.
-Then, the scheduler executes the \emph{Main} label in \code{TRAIL-0}.
-%
-When the \emph{Main} label reaches the \code{par/and}, it ``stacks'' 
-\code{TRAIL-1} (line 5--8) and proceeds to the code in \code{TRAIL-0} (lines 
-10--15), respecting the deterministic execution order.
+In the final generated code in \emph{C}, each trail label representing an entry 
+point becomes a \emph{switch case} with the associated code to execute.
+%
+Figure~\ref{lst.impl.trails} illustrates the generation process.
+For the program in the left of the figure, the compiler extracts the entry 
+points and associated trails.
+%
+The entry points translate to an \code{enum} in the generated code (lines 
+1--10, in the middle of the figure).
+The trails translate to a vector of \code{trail\_t} elements with the maximum 
+number of simultaneous trails (lines 12--16).
+Initially, \code{TRAIL-0} is set to execute the \code{Main} entry point, while 
+all others are set to \code{INACTIVE}.
+
+The scheduler executes in two passes:
+%
+In the \emph{broadcast} pass, the scheduler sets to \code{STACKED} all trails 
+that are waiting for the current event to execute in the current stack level.
+%
+In the \emph{dispatch} pass, the scheduler executes each trail that is 
+\code{STACKED} to run in the current level, setting it immediately to 
+\code{INACTIVE}.
+
+During the dispatch pass, if a trail executes and emits and internal event, the 
+scheduler increments the stack level and re-executes the two passes.
+After all trails are properly dispatched, the scheduler decrements the stack 
+level and resumes the previous execution.
+For the first reaction chain, the scheduler starts from the \emph{dispatch} 
+pass, given that the \code{Main} label is the only one that can be active at 
+the stack level 0 (line 13, in the middle of Figure~\ref{lst.impl.trails}).
+
+The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
+according to its current label to execute.
+%
+For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
+%
+When the \code{Main} label reaches the \code{par/and}, it ``stacks'' 
+\code{TRAIL-1} (lines 4--8) and proceeds to the code in \code{TRAIL-0} (lines 
+9--14), respecting the deterministic execution order.
 %
@@
 %
-The next iteration of the scheduler takes \code{TRAIL-1} and executes its 
-registered label \code{ParAnd\_sub\_2} (lines 17-22), which sets \code{TRAIL-1} 
+The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
+registered label \code{And\_sub\_2} (lines 16--21), which sets \code{TRAIL-1} 
 to await \code{EVT\_A} and also halts.
 
-Regarding cancellation, trails in parallel are always allocated in subsequent 
-slots in the static vector \code{TRLS}.
-Therefore, when a \code{par/or} terminates, the scheduler sequentially searches 
-and executes \code{FIN} trails within the range of the \code{par/or}, and then 
-clears all of them to \code{INACTIVE} at once.
-Given that finalization blocks cannot contain \code{await} statements, the 
-whole process is guaranteed to terminate in bounded time.
+Regarding abortion and finalization, when a \code{par/or} terminates, the 
+scheduler makes a \emph{broadcast} pass for the \code{FIN} event, but limited 
+to the range of trails covered by the terminating \code{par/or}.
+Trails that do not match the \code{FIN} are set to \code{INACTIVE}, as they 
+have to be aborted.
+Given that trails in parallel are allocated in subsequent slots in the static 
+vector \code{TRLS}, this pass only aborts the desirable trails.
+The subsequent \emph{dispatch} pass executes the finalization code.
+%Given that finalization blocks cannot contain \code{await} statements, the 
+%whole process is guaranteed to terminate in bounded time.
 Escaping a \code{loop} that contains parallel compositions also trigger the 
-same finalization process.
-
-TODO internal emit?
+same abortion process.
 
@@
 
+TODO: refs/cite WSNs
+
 As an example, Figure~\ref{lst.impl.tinyos} shows our binding for \emph{TinyOS} 
@@
 pointing to the labels defined in the program.
-The callback \code{Boot.booted} (lines 6-11) is called by TinyOS on mote 
-startup, so we initialize \CEU inside it (line 7).
-If the \CEU program uses timers, we also start a periodic timer (lines 8-10) 
-that triggers callback \code{Timer.fired} (lines 13-17) every 10 milliseconds 
+The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
+so we initialize \CEU inside it (line 7).
+If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
+that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
 and advances the wall-clock time of \CEU (line 15)%
@@
 The remaining lines map pre-defined TinyOS events that can be used in \CEU 
-programs, such as the light sensor (lines 19-23) and the radio transceiver 
-(lines 25-36).
+programs, such as the light sensor (lines 19--23) and the radio transceiver 
+(lines 25--36).
 

### 2015-09-19 paper.tex
Revisoes do Roberto: - reforcar diferencas a Esterel (em vez de semelhancas) ja no inicio da secao 2 - incluir analise estatica de shared-memory concurrency - reescrever a discussao do exemplo de finalizacao, deixando claro que ele e' "unsafe" no inicio, e que a chamada a enqueue executa em "background"

@@
 
-Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel and 
-\CEU for the following control specification~\cite{esterel.primer}:
+In the sections that follow, we review the main differences between \CEU and 
+Esterel~\cite{ceu.sensys13}:
+shared-memory concurrency (Section~\ref{sec.ceu.det}),
+safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
+first-class timers (Section~\ref{sec.ceu.timers}),
+and stack-based internal events (Section~\ref{sec.ceu.ints}).
+
+Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
+implementations in Esterel and \CEU for the following control 
+specification~\cite{esterel.primer}:
 %
@@
 
-\begin{figure}[t]
+\begin{figure}
 \begin{minipage}[t]{0.48\linewidth}
@@
 %
-In addition, \CEU also employs the synchronous model, in whicn programs advance 
+In addition, \CEU also employs the synchronous model, in which programs advance 
 in a sequence of discrete reactions to external events.
@@
 
-In the sections that follow, we review the main differences between \CEU and 
-Esterel~\cite{ceu.sensys13}: deterministic execution 
-(Section~\ref{sec.ceu.det}), safe abortion with finalization 
-(Section~\ref{sec.ceu.abrt}), first-class timers 
-(Section~\ref{sec.ceu.timers}), and stack-based internal events 
-(Section~\ref{sec.ceu.ints}).
-
 %\newpage %TTT
-\subsection{Deterministic Execution}
+\subsection{Shared-Memory Concurrency}
 \label{sec.ceu.det}
 
+Embedded applications make extensive use of shared memory, such as for sharing 
+resources through memory-mapped registers.
+Hence, an important goal of \CEU is to ensure a reliable execution for 
+concurrent programs that share memory.
+%
 Esterel is only deterministic with respect to reactive control: ``the same 
@@
 outputs''~\cite{esterel.primer}.
+%
 However, the execution order for operations with side-effects within a reaction 
@@
 %
-In \CEU, when multiple trails are active at a time, they are scheduled in the 
-order they appear in the program source code.
-%
-Hence, \CEU is deterministic also with respect to the order of execution of 
-side effects within a reaction~\cite{ceu.sensys13}.
-%
-Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
-in Esterel and \CEU to illustrate the semantic difference regarding 
-(non-)determinism.
-
+Therefore, Esterel forbids sharing memory: \emph{``if a variable is written by 
+some thread, then it can neither be read nor be written by concurrent 
+threads''}~\cite{esterel.primer}.
+
+Concurrency in \CEU is characterized when two or more trail segments in 
+parallel execute during the same reaction chain.
+A trail segment is a sequence of statements followed by an \code{await} (or 
+termination).
+%
+In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
+\code{x} can never run concurrently, because each trail segment reacts to a 
+different input event and, according to the synchronous model, each reaction is 
+atomic.
+%
+However, the program in the right is non-deterministic, because the two 
+assignments to \code{y} occur in the same reaction to input \code{A}.
+
+\begin{figure}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}
+input void A, B;
+var int x = 1;
+par/and do
+    await A;
+    x = x + 1;
+with
+    await B;
+    x = x * 2;
+end
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}
+input void A;
+var int y = 1;
+par/and do
+    await A;
+    y = y + 1;
+with
+    await A;
+    y = y * 2;
+end
+\end{lstlisting}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Shared-memory concurrency in \CEU:
+the code in the left is safe because the trails access \code{x} atomically in 
+different reactions;
+the code in the right is unsafe because both trails access \code{y} in the same 
+reaction.
+%\newline
+{\small
+%TODO.
+}
+\label{lst.shared}
+}
+\end{figure}
+
+\CEU performs a temporal analysis at compile time and detects concurrent 
+accesses to shared variables, as follows~\cite{ceu.sensys13}:
+\emph{if a variable is written in a trail segment, then a concurrent trail 
+segment cannot read or write to that variable, nor dereference a pointer of 
+that variable type.}
+An analogous policy is applied for pointers \emph{vs} variables and pointers 
+\emph{vs} pointers, as well as for system calls with side effects (e.g., 
+\code{printf}).
+The algorithm for the analysis holds the set of all events in preceding 
+\code{await} statements for each variable access.
+Then, the sets for all accesses in parallel trails are compared to assert that 
+no events are shared among them.
+Otherwise the compiler warns about the suspicious accesses.
+
+Regardless of the static analysis, when multiple trails are active at a time in 
+\CEU, they are scheduled in the order they appear in the program source code.
+%
+Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
+suspicious, the assignments to \code{y} both atomic and deterministic, i.e., 
+after \code{A} occurs, \code{y} is $4$ (\code{(1+1)*2}).
+%
 On the one hand, enforcing an execution order for concurrent operations may 
@@
 shared-memory concurrency more tractable.
-In contrast, Esterel forbids sharing memory: \emph{``if a variable is written 
-by some thread, then it can neither be read nor be written by concurrent 
-threads''}~\cite{esterel.primer}.
 %
@@
 mapped ports for I/O and the lack of hardware support for real parallelism.
-Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
+Other synchronous embedded languages, such as SOL~\cite{wsn.sol} and 
+PRET-C\cite{rp.pretc}, made a similar design choice.
 %For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
@@
 
-TODO: names (other embedded languages)
-
-\begin{figure}[b]
+Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
+in Esterel and \CEU to summarize the semantic difference regarding 
+(non-)determinism.
+%
+Even though the program in \CEU executes deterministically, the compiler still 
+issues a warning, because an apparently innocuous reordering of trails modifies 
+the semantics of the program.
+
+\begin{figure}
 \begin{minipage}[t]{0.48\linewidth}
@@
 
-\begin{figure}[t]
+\begin{figure}
 \begin{minipage}[t]{0.55\linewidth}
@@
 %\rule{8.5cm}{0.37pt}
-\caption{A network protocol in \CEU extended with a finalization clause to 
-cancel a transmission.
+\caption{The unsafe network protocol in the left is extended with a 
+finalization clause in the right to handle abortion properly.
 %\caption{ Parallel compositions can describe complex state machines.
@@
 
-To illustrate threats related to abortion, consider the example in the left of 
-Figure~\ref{lst.fin} with hierarchical \code{par/or} and \code{par/and} 
-compositions describing the state machine of a data collection protocol for 
-sensor networks~\cite{wsn.ctp,ceu.sensys13}.
+To illustrate threats related to abortion, consider the unsafe example in the 
+left of Figure~\ref{lst.fin}, describing the state machine of a data collection 
+protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
 %
@@
 %
-The protocol has two trails:
+The protocol has to transmit a packet every minute, unless it receives a 
+\code{RETRANSMIT} request to immediately re-transmit, or a \code{STOP} request 
+to terminate.
+%
+The protocol is composed of two main trails:
 one simply monitors the stopping event (line 3);
@@
 The periodic transmission is a loop that starts two other trails (lines 6--17):
-one handles an immediate retransmission request (line 7);
+one handles the immediate retransmission request (line 7);
 the other transmits the packet%
@@
 packets.
+Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
+the radio driver a pointer to the lexically-scoped packet.
+The driver will make the transmission in the background, holding the packet 
+until it signals the application with the \code{SENDACK} to acknowledge 
+completion.
 %
@@
 %
-Note that the \code{buffer} packet (line 12) is a local variable whose address 
-is passed to function \code{\_send\_enqueue} for transmission.
-The call enqueues the pointer in the radio driver, which holds it up to the 
-emission of \code{SENDACK} acknowledging the packet transmission.
-%
 In the meantime, if the sending trail is be aborted by the \code{STOP} or 
 \code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
-a \emph{dangling pointer} in the radio driver.
+a \emph{dangling pointer} in the radio driver, which will possibly transmit 
+wrong data.
 %
@@
 
-\begin{figure}[t]
+\begin{figure}
 \begin{minipage}[t]{0.45\linewidth}
@@
 %
-A similar program in a language without first-class support for timers, would 
+A similar program in a language without first-class support for timers would 
 depend on the execution timings for the code marked as \code{<...>}, making the 
@@
 
-\begin{figure}%[b]
+\begin{figure}
 \begin{minipage}[t]{0.43\linewidth}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \begin{lstlisting}[mathescape=true]
@@
 \end{align*}
-
+%
 At the beginning of a reaction chain, the stack is initialized with the 
@@
 }
-
+%
 An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
@@
 are straightforward:
-
+%
 { \setlength{\jot}{20pt}
@@
 }
-
+%
 Given that our semantics focuses on control, rules \textbf{if-true} and 
@@
 separators to properly bind breaks to their enclosing loops:
-
+%
 { \setlength{\jot}{20pt}
@@
 }
-
+%
 When a program encounters a $loop$, it first expands its body in sequence with 
@@
 branches $q$:
-
+%
 { \setlength{\jot}{20pt}
@@
 }
-
+%
 The deterministic behavior of the semantics relies on the \emph{isBlocked} 
@@
 
-\begin{figure}[t]
+\begin{figure}
 {\small
@@
 innermost loop (\emph{aborting} the other side):
-
+%
 { \setlength{\jot}{20pt}
@@
 }
-
 %
@@
 same way as in a parallel $and$:
-
+%
 { \setlength{\jot}{20pt}
@@
 }
-
+%
 A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
@@
 
-\begin{figure}%[h!]
+\begin{figure}
 \begin{minipage}[t]{0.32\linewidth}
@@
 
-\begin{figure}%[h!]
+\begin{figure}
 \begin{minipage}[t]{0.30\linewidth}
@@
 
-\begin{figure}%[h!]
+\begin{figure}
 \begin{minipage}[t]{0.26\linewidth}
@@
 
-\begin{figure}[ht]
+\begin{figure}
 \centering
@@
 
-\begin{figure}[h]
+\begin{figure}
 \begin{minipage}[t]{0.48\linewidth}
@@
 
-\begin{figure}[t]
+\begin{figure}
 \begin{minipage}[t]{0.31\linewidth}
@@
 
-\begin{figure}[t]
+\begin{figure}
 \begin{lstlisting}[numbers=left,xleftmargin=2em]

### 2015-09-19 paper.tex
"Stack-Based Internal Events" como primeira subsecao de 2.

@@
 Esterel~\cite{ceu.sensys13}:
+stack-based internal events (Section~\ref{sec.ceu.ints}),
 shared-memory concurrency (Section~\ref{sec.ceu.det}),
 safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
-first-class timers (Section~\ref{sec.ceu.timers}),
-and stack-based internal events (Section~\ref{sec.ceu.ints}).
+and first-class timers (Section~\ref{sec.ceu.timers}).
 
@@
 all possible paths~\cite{ceu.sensys13,esterel.primer}.
+
+\subsection{Stack-Based Internal Events}
+\label{sec.ceu.ints}
+
+Esterel makes no semantic distinctions between internal and external signals, 
+both having only the notion of either presence or absence during the entire 
+reaction~\cite{esterel.preemption}.
+%
+In \CEU, however, internal events follow a stack-based execution policy, 
+similar to subroutine calls in typical programming 
+languages~\cite{ceu.sensys13}, contrasting with queue-based event handling 
+adopted for external events.
+%
+Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
+Esterel and \CEU.
+%
+In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
+and they both become active, resulting in the invocation of \code{f()} and 
+\code{g()} in no particular order.
+%
+In the version in \CEU, the occurrence of \code{A} makes the program behave as 
+follows (with the stack contents in italics):
+%
+{\small
+\begin{enumerate}
+\setlength{\itemsep}{0pt}
+\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
+    \emph{stack: [1st-trail]}
+\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
+    \emph{stack: [1st-trail]}
+\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
+    terminates.\\
+    \emph{stack: []}
+\item Both trails have terminated, so the \code{par/and} rejoins, and the 
+program also terminates;
+\end{enumerate}
+}
+
+\begin{figure}
+\begin{minipage}[t]{0.43\linewidth}
+\begin{lstlisting}[mathescape=true]
+// ESTEREL
+input A;    // external
+signal B;   // internal
+[[
+    await A;
+    emit B;
+    call f();
+$\|$
+    await B;
+    call g();
+]]
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.53\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
+// CEU
+input void A;  // external (in uppercase)
+event void b;  // internal (in lowercase)
+par/and do
+    await A;
+    emit b;
+    _f();
+with
+    await b;
+    _g();
+end
+\end{lstlisting}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Internal signals (events) in Esterel and \CEU. \newline
+\label{lst.prints}
+}
+\end{figure}
+
+Internal events bring support for a limited form of subroutines, as depicted in 
+Figure~\ref{lst.sub}.
+The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
+awaits its identifying event (line 4), incrementing the value passed as 
+reference (line 5).
+A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
+\code{A} through an \code{emit} (line 10).
+Given the stacked execution for internal events, the calling trail pauses, the 
+subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
+awaits the next ``call'' (line 4, again).
+Only after this sequence that the calling trail resumes and passes the 
+assertion test.
+
+\begin{figure}
+\begin{lstlisting}[numbers=left,xleftmargin=2em]
+event int* inc; // subroutine `inc'
+par/or do
+    loop do     // definitions are loops
+        var int* p = await inc;
+        *p = *p + 1;
+    end
+with
+    var int v = 1;
+    await A;
+    emit inc => &v; // call `inc'
+    _assert(v==2);  // after return
+end
+\end{lstlisting}
+\caption{ Subroutine \code{inc} is defined in a loop (lines 3--6), in parallel 
+with the caller (lines 8--11).
+\label{lst.sub}
+}
+\end{figure}
+
+On the one hand, this form of subroutines has a significant limitation that it 
+cannot express recursive calls: an \code{emit} to itself is always ignored, 
+given that a running body cannot be awaiting itself.
+%
+On the other hand, this very same limitation brings some important safety 
+properties to subroutines:
+first, they are guaranteed to react in bounded time;
+second, memory for locals is also bounded, not requiring data stacks.
+%
+Also, this form of subroutines can use the other primitives of \CEU, such as 
+parallel compositions and the \code{await} statement.
+In particular, they await keeping context information such as locals and the 
+program counter, similarly to coroutines~\cite{lua.coroutines}.
+%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
+%exceptions.
+
+\begin{comment}
+\vspace{5pt}
+\CEU provides no support for standard functions for a number of reasons:
+\begin{itemize}
+\item The interaction with other \CEU control primitives is not obvious (e.g., 
+executing an $await$ or a $par/or$ inside a function).
+\item They would still be restricted in some ways given the embedded context 
+(e.g.  no recursion or closures).
+\item Programs can always recur to $C$ functions for low-level operations.
+%\item A dedicated primitive would behave just as described, being a matter of 
+%syntactic sugar.
+\end{itemize}
+
+Regardless of the limitations, this form of subroutines is widely adopted in 
+\CEU programs, given that they were designed to work with the other control 
+mechanisms.
+Keep in mind that the typical reactive organization of programs (awaiting an 
+external stimulus, reacting to it, and going back to awaiting) does not demand 
+unrestricted subroutines.
+\end{comment}
 
@@
 
-\subsection{First-class Timers}
+\subsection{First-Class Timers}
 \label{sec.ceu.timers}
@@
 reasoning about the execution behavior more difficult.
-
-\subsection{Internal Events}
-\label{sec.ceu.ints}
-
-Esterel makes no semantic distinctions between internal and external signals, 
-both having only the notion of either presence or absence during the entire 
-reaction~\cite{esterel.preemption}.
-%
-In \CEU, however, internal events follow a stack-based execution policy, 
-similar to subroutine calls in typical programming 
-languages~\cite{ceu.sensys13}, contrasting with queue-based event handling 
-adopted for external events.
-%
-Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
-Esterel and \CEU.
-%
-In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
-and they both become active, resulting in the invocation of \code{f()} and 
-\code{g()} in no particular order.
-%
-In the version in \CEU, the occurrence of \code{A} makes the program behave as 
-follows (with the stack contents in italics):
-%
-{\small
-\begin{enumerate}
-\setlength{\itemsep}{0pt}
-\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
-    \emph{stack: [1st]}
-\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
-    \emph{stack: [1st]}
-\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
-    terminates.\\
-    \emph{stack: []}
-\item Both trails have terminated, so the \code{par/and} rejoins, and the 
-program also terminates;
-\end{enumerate}
-}
-
-\begin{figure}
-\begin{minipage}[t]{0.43\linewidth}
-\begin{lstlisting}[mathescape=true]
-// ESTEREL
-input A;    // external
-signal B;   // internal
-[[
-    await A;
-    emit B;
-    call f();
-$\|$
-    await B;
-    call g();
-]]
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.53\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
-// CEU
-input void A;  // external (in uppercase)
-event void b;  // internal (in lowercase)
-par/and do
-    await A;
-    emit b;
-    _f();
-with
-    await b;
-    _g();
-end
-\end{lstlisting}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Internal signals (events) in Esterel and \CEU. \newline
-\label{lst.prints}
-}
-\end{figure}
-
-Internal events bring support for a limited form of subroutines, as depicted in 
-Figure~\ref{lst.sub}.
-The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
-awaits its identifying event (line 4), incrementing the value passed as 
-reference (line 5).
-A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
-\code{A} through an \code{emit} (line 10).
-Given the stacked execution for internal events, the calling trail pauses, the 
-subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
-awaits the next ``call'' (line 4, again).
-Only after this sequence that the calling trail resumes and passes the 
-assertion test.
-
-\begin{figure}
-\begin{lstlisting}[numbers=left,xleftmargin=2em]
-event int* inc; // subroutine `inc'
-par/or do
-    loop do     // definitions are loops
-        var int* p = await inc;
-        *p = *p + 1;
-    end
-with
-    var int v = 1;
-    await A;
-    emit inc => &v; // call `inc'
-    _assert(v==2);  // after return
-end
-\end{lstlisting}
-\caption{ Subroutine \code{inc} is defined in a loop (lines 3--6), in parallel 
-with the caller (lines 8--11).
-\label{lst.sub}
-}
-\end{figure}
-
-On the one hand, this form of subroutines has a significant limitation that it 
-cannot express recursive calls: an \code{emit} to itself is always ignored, 
-given that a running body cannot be awaiting itself.
-%
-On the other hand, this very same limitation brings some important safety 
-properties to subroutines:
-first, they are guaranteed to react in bounded time;
-second, memory for locals is also bounded, not requiring data stacks.
-%
-Also, this form of subroutines can use the other primitives of \CEU, such as 
-parallel compositions and the \code{await} statement.
-In particular, they await keeping context information such as locals and the 
-program counter, similarly to coroutines~\cite{lua.coroutines}.
-%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
-%exceptions.
-
-\begin{comment}
-\vspace{5pt}
-\CEU provides no support for standard functions for a number of reasons:
-\begin{itemize}
-\item The interaction with other \CEU control primitives is not obvious (e.g., 
-executing an $await$ or a $par/or$ inside a function).
-\item They would still be restricted in some ways given the embedded context 
-(e.g.  no recursion or closures).
-\item Programs can always recur to $C$ functions for low-level operations.
-%\item A dedicated primitive would behave just as described, being a matter of 
-%syntactic sugar.
-\end{itemize}
-
-Regardless of the limitations, this form of subroutines is widely adopted in 
-\CEU programs, given that they were designed to work with the other control 
-mechanisms.
-Keep in mind that the typical reactive organization of programs (awaiting an 
-external stimulus, reacting to it, and going back to awaiting) does not demand 
-unrestricted subroutines.
-\end{comment}
 

### 2015-09-19 paper.tex
Nome da secao 2.2

@@
 stack-based internal events (Section~\ref{sec.ceu.ints}),
-shared-memory concurrency (Section~\ref{sec.ceu.det}),
+shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
 safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
@@
 %\newpage %TTT
-\subsection{Shared-Memory Concurrency}
+\subsection{Shared-Memory Concurrency and Determinism}
 \label{sec.ceu.det}

### 2015-09-19 paper.tex
Nova Secao 4.1 para descrever a analise temporal para acessso a memoria compratilhada.

@@
 
+% TODO: par/or vs abort, strong/weak abortion
+
 \subsection{Stack-Based Internal Events}
@@
 
-\item[Analysis]
+\item[Temporal Analysis]
 
 This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
-and also makes ``classical'' semantic analysis, such as building a symbol table 
-for checking variable declarations.
+suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
+such as building a symbol table for checking variable declarations.
 %
@@
 
-\begin{comment}
-\subsection{Temporal Analysis}
-
-As introduced, the \emph{temporal analysis} phase detects inconsistencies in 
-\CEU programs.
-Here, we focus on the algorithm that detects non-deterministic access to 
-variables, as presented in Section~\ref{sec.ceu.shared}.
+\subsection{Temporal Analysis for Shared-Memory Concurrency}
+
+The compile-time \emph{temporal analysis} phase detects inconsistencies in \CEU 
+programs.
+Here, we focus on the algorithm that detects suspicious access to shared 
+variables, as discussed in Section~\ref{sec.ceu.det}.
 
@@
 \item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
-\code{loop} and never terminate, i.e., never proceeds to the statement 
+\code{loop} and never terminates, i.e., never proceeds to the statement 
 immediately following it (see also \code{loop} below);
@@
 parallel branches can have their $I$ sets compared for intersections.
-If the intersection is not the empty set, they are marked as suspicious (see 
-Section~\ref{sec.ceu.shared}).
-
-Figure~\ref{lst.impl.ast} reproduces the second code of Figure~\ref{lst.det} 
-and shows the corresponding $AST$ with the sets $I$ and $O$ for each node.
+If the intersection is not the empty set, they are marked as suspicious.
+
+The example in the left of Figure~\ref{lst.impl.ast} has a corresponding $AST$, 
+in the right of the figure, with the sets $I$ and $O$ for each node.
 The event $.$ (dot) represents the ``boot'' reaction.
@@
 intersection of $I$ (lines 6,9 in the AST), hence, they do not conflict.
-Note that although the accesses in lines 5, 11 in the code (lines 6,11 in the 
+Note that although the accesses in lines 5,11 in the code (lines 6,11 in the 
 AST) do have an intersection, they are not in parallel and are also safe.
@@
 
-begin{comment}
+\begin{comment}
 It is also responsible for setting the priorities for trails (see further) and 
@@
 state.
-end{comment}
-
 \end{comment}

### 2015-09-23 paper.tex
Trabalhos relacionados (incompleto)

@@
         \item Usos: SenSys, Terra, sala de aula, GSoC
+        \item ideia de tamanhos, ROM,RAM,CPU
     \end{itemize}
@@
 % Artigo de Rust, nao resolve os problemas
+%https://blog.skcript.com/asynchronous-io-in-rust-36b623e7b965
+%state machines everywhere
 \end{itemize}
+
+%Our work focuses on \emph{concurrency safety}, rather than \emph{type 
+%safety}~\cite{wsn.safety}.%
+%\footnote{
+%We consider both safety aspects to be complimentary and orthogonal, i.e., 
+%type-safety techniques could also be applied to \CEU.
+%}
 
@@
 Esterel makes no semantic distinctions between internal and external signals, 
-both having only the notion of either presence or absence during the entire 
+both having only the notion of either presence or absence during an entire 
 reaction~\cite{esterel.preemption}.
 %
-In \CEU, however, internal events follow a stack-based execution policy, 
-similar to subroutine calls in typical programming 
-languages~\cite{ceu.sensys13}, contrasting with queue-based event handling 
-adopted for external events.
+In \CEU, external input events are unique within reactions, with
+intrinsic queue-based handling, and programs cannot emit them.
+%
+In contrast, internal events follow a stack-based execution policy, similar to 
+subroutine calls in typical programming languages~\cite{ceu.sensys13}.
 %
@@
 \code{x} can never run concurrently, because each trail segment reacts to a 
-different input event and, according to the synchronous model, each reaction is 
-atomic.
+different input event and, according to the semantics of \CEU, cannot occur 
+simultaneously.
 %
@@
 mapped ports for I/O and the lack of hardware support for real parallelism.
-Other synchronous embedded languages, such as SOL~\cite{wsn.sol} and 
-PRET-C\cite{rp.pretc}, made a similar design choice.
+Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
+\emph{PRET-C}\cite{rp.pretc}, made a similar design choice.
 %For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
@@
 the semantics of the program.
+%
+Note that in Esterel multiple external events can coexist in the same reaction, 
+which disallows a similar temporal analysis.
 
@@
 \end{comment}
+
 \section{Implementation}
@@
 For the program in the left of the figure, the compiler extracts the entry 
-points and associated trails.
+points and associated trails, e.g., the label \code{Awake\_e} will execute on 
+\code{TRAIL-0} (line 7).
+%
+For each statement that yields (\code{emit} and \code{await}), resumes 
+(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
+and \code{break}), the compiler splits the trail into segments with associated 
+entry points.
 %
@@
 
+\section{Terra and the \CEU-VM}
+
 \section{Related Work}
 
+\emph{(As subsecoes serao retiradas do texto ao final...)}
+
+\subsection{Semantics}
+
+\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
+disciplined synchronous-reactive model with support for lexical composition of 
+lines of execution.
+%
+However, there are fundamental semantic differences that prevents the design of 
+\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
+%
+In particular, Esterel has a notion of time similar to that of digital circuits 
+in which multiple signals can be active at a clock tick (Esterel is also used 
+in hardware design).
+In \CEU, instead of clock ticks, the occurrence of a single external event that 
+defines a time unit.
+We believe that for software design, this approach simplifies the reasoning 
+about concurrency.
+For instance, the uniqueness of external events in \CEU is a prerequisite for 
+the temporal analysis that enables safe shared-memory concurrency.
+%
+Furthermore, \CEU also distinguishes external events from stack-based internal 
+events, which provides a limited form of subroutines that supports \CEU's 
+reactive statements.
+
+A number of synchronous languages have been designed to interoperate with C, 
+such as \emph{Reactive C}~\cite{rp.rc}, Protothreads \cite{wsn.protothreads}, 
+\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
+%
+They offer Esterel-like parallel compositions with communication via shared 
+variables, relying on deterministic semantics to preserve determinism.
+%
+However, it is the responsibility of the programmer to specify the execution 
+order for threads, based on either explicit priorities or source code lexical 
+order (like \CEU).
+%
+These languages have a tick-based notion of time similar to Esterel, also 
+disallowing the event-based temporal analysis of \CEU.
+
+\emph{URBI}~\cite{rp.urbi} is a reactive scripting language with a rich set of 
+control constructs for time management, event-driven communication, and 
+concurrency.
+Concurrency is based on stackful coroutines, diverging from our goals regarding 
+resource efficiency and static bounds for memory and execution time.
+
+\begin{comment}
+More recently, Wireless Sensor Networks (WSNs) emerged as an active research 
+area for highly constrained embedded concurrency, resulting in the development 
+of many synchronous languages~\cite{wsn.protothreads,wsn.sol,wsn.osm}.
+%
+offer lightweight cooperative multi-threading for embedded systems.
+Its stackless implementation reduces memory consumption but precludes support 
+for local variables.
+\CEU also avoids the use of stacks for trails, but preserves support for locals 
+by calculating the required memory at compile time.
+%
+SOL~\cite{wsn.sol} and OSM~\cite{wsn.osm} provide parallel state machines for 
+WSNs, offering a formal and mature model for programming embedded systems.
+However, the main contributions of \CEU, stacked execution for internal 
+events and safe support for shared-memory concurrency, do not directly adapt to 
+the state-machine formalism.
+
+In common among the referred works is the agreement in providing low-level 
+access (e.g., systems calls and shared-memory) and lock-free concurrency that 
+precludes race conditions on programs.
+However, they do not specify an execution order for tasks reacting to the same 
+external stimulus~\cite{esterel.primer,wsn.protothreads,wsn.osm}.
+This way, if two tasks access the same resource concurrently, even if the 
+accesses are race free, the final outcome is nondeterministic.
+As discussed in Section~\ref{sec.safety.det}, \CEU refuses programs with such 
+behavior.
+
+On the opposite side of the spectrum of concurrency models, asynchronous 
+languages for embedded systems~\cite{wsn.mantisos,arduino.occam}
+assume time independence among processes and are more appropriate for 
+applications with a low synchronization rate or for those involving
+algorithmic-intensive problems.
+\end{comment}
+
+\subsection{Implementation}
+
+Esterel has different compilation backends that synthesizes to software and 
+also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
+Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
+the closest to our implementation for trail allocation and scheduling:
+The compiler slices programs into ``control points'' (analogous to our ``entry 
+points'') and rearranges them into a directed acyclic graph respecting the 
+constructive semantics of Esterel.
+Then, it flattens the graph into sequential code in \emph{C} suitable for 
+static scheduling.
+
+\subsection{Virtual Machines}
+
+\begin{comment}
 - Kiel Esterel compiler
@@
 
-PRET-C
 http://embedded.cs.uni-saarland.de/publications/BuildingTimingPredictableEmbeddedSystems.pdf
+survey of timeing-predictable embedded systems,
+synchronous languages do some at the lang semantic level
+\cite{pret.building}
+\end{comment}
 

### 2015-09-24 paper.tex
Restricao de "delayed awaits"

@@
 
-\begin{comment}
-\vspace{5pt}
-\CEU provides no support for standard functions for a number of reasons:
-\begin{itemize}
-\item The interaction with other \CEU control primitives is not obvious (e.g., 
-executing an $await$ or a $par/or$ inside a function).
-\item They would still be restricted in some ways given the embedded context 
-(e.g.  no recursion or closures).
-\item Programs can always recur to $C$ functions for low-level operations.
-%\item A dedicated primitive would behave just as described, being a matter of 
-%syntactic sugar.
-\end{itemize}
-
-Regardless of the limitations, this form of subroutines is widely adopted in 
-\CEU programs, given that they were designed to work with the other control 
-mechanisms.
-Keep in mind that the typical reactive organization of programs (awaiting an 
-external stimulus, reacting to it, and going back to awaiting) does not demand 
-unrestricted subroutines.
-\end{comment}
+Another distinction regarding event handling is that Esterel supports 
+same-cycle bi-directional communication~\cite{esterel.compiling}, i.e., two 
+threads can react to one another during the same cycle due to mutual signal 
+dependency.
+%
+\CEU imposes a restriction that an \code{await} is only valid for the next 
+reaction, i.e., if an \code{await} and \code{emit} for the same event occur 
+simultaneously in parallel trails, the \code{await} does not awake.
+%
+These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
+re-execution of statements in the same reaction inside loops (known as 
+\emph{schizophrenic statements})~\cite{esterel.schizo}.
+%
+The example of Figure~\ref{lst.delay} relies on this restriction to avoid 
+infinite execution:
+Although both sides of the \code{par/or} have an \code{await} statement, the 
+\code{loop} would restart instantaneously if the \code{await} could awake in 
+the same reaction that reaches it.
+%
+In an atypical scenario requiring immediate awake, delayed awaits can be 
+circumvented by judiciously copying the code to execute on awake.
+
+\begin{figure}
+\begin{lstlisting}
+event void e,f;
+loop do
+    par/or do
+        await e;
+    with
+        emit e;
+        await f;
+    end
+end
+\end{lstlisting}
+\caption{ The \code{await} occurs simultaneously with the \code{emit} and does 
+not awake, preventing the \code{par/or} to terminate and execute the 
+\code{loop} indefinitely.
+\label{lst.delay}
+}
+\end{figure}
 
@@
 reaction.
-%\newline
-{\small
-%TODO.
-}
 \label{lst.shared}
@@
 unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
-{\small
-%TODO.
-}
 \label{lst.det}
@@
 events on top of it (we discuss how they are popped further).
-The ever-increasing sequence number $n$ will prevent that awaiting expressions 
-awake in the same reaction they are reached (ensuring that the program can 
-progress).
+The ever-increasing sequence number $n$ prevents that awaiting expressions 
+awake in the same reaction they are reached (the \emph{delayed awaits} as 
+explained in Section~\ref{sec.ceu.ints}).
 

### 2015-09-24 paper.tex
Cenarios onde a restricao de "delayed await" atrapalha.

@@
 %
-The example of Figure~\ref{lst.delay} relies on this restriction to avoid 
-infinite execution:
+The example in the left of Figure~\ref{lst.delay} (\code{CODE-1}) relies on 
+this restriction to avoid infinite execution:
 Although both sides of the \code{par/or} have an \code{await} statement, the 
@@
 %
-In an atypical scenario requiring immediate awake, delayed awaits can be 
-circumvented by judiciously copying the code to execute on awake.
+In atypical scenarios requiring immediate awake, delayed awaits can be 
+circumvented by manually copying or making some transformation in the code to 
+execute on awake.
+%
+From our experience, in some cases we need to execute a block of code 
+periodically, \emph{including the current reaction}, from requests of an 
+internal event, as illustrated in the middle of Figure~\ref{lst.delay} 
+(\code{CODE-2}).
+In this case, moving the \code{await} to the end of the loop suffices to 
+execute the periodic code immediately (and then in reaction to each \code{emit} 
+request).
+If the periodic \code{emit} depends on a condition, the code becomes more 
+intricate, requiring to copy the condition test to the periodic code, as 
+illustrated in the right of Figure~\ref{lst.delay} (\code{CODE-3}).
 
 \begin{figure}
+\begin{minipage}[t]{0.24\linewidth}
 \begin{lstlisting}
@@
 end
-\end{lstlisting}
-\caption{ The \code{await} occurs simultaneously with the \code{emit} and does 
-not awake, preventing the \code{par/or} to terminate and execute the 
-\code{loop} indefinitely.
+
+/* CODE-1 */
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.36\linewidth}
+\begin{lstlisting}
+event void e;
+par do
+  loop do
+    <...>
+    emit e;  // periodic request
+  end
+with
+  loop do
+    <...>    // code to execute
+    await e; // await after
+  end
+end
+
+/* CODE-2 */
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.36\linewidth}
+\begin{lstlisting}
+event void e;
+var bool execute_now = false;
+par do
+  loop do
+    <...>
+    if <...> then
+        execute_now = true;
+        emit e;
+    end
+  end
+with
+  loop do
+    if execute_now then
+      <...> // code to execute
+    end
+    await e;
+  end
+end
+
+/* CODE-3 */
+\end{lstlisting}
+\end{minipage}
+%
+\caption{ In \code{CODE-1}, the \code{await} occurs simultaneously with the 
+\code{emit} and does not awake, preventing the \code{par/or} to terminate and 
+execute the \code{loop} indefinitely.
+\code{CODE-2} and \code{CODE-3} illustrate how to circumvent the \emph{delayed 
+await} restriction.
 \label{lst.delay}

### 2015-09-24 paper.tex
Revisao Secao 2.1

@@
 \emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
-%\newline
-{\small
-%TODO.
-}
 \label{lst.abro}
@@
 
-Another distinction regarding event handling is that Esterel supports 
-same-cycle bi-directional communication~\cite{esterel.compiling}, i.e., two 
-threads can react to one another during the same cycle due to mutual signal 
-dependency.
-%
-\CEU imposes a restriction that an \code{await} is only valid for the next 
-reaction, i.e., if an \code{await} and \code{emit} for the same event occur 
+Another distinction regarding event handling in comparison to \CEU is that 
+Esterel supports same-cycle bi-directional 
+communication~\cite{esterel.compiling}, i.e., two threads can react to one 
+another during the same cycle due to mutual signal dependency.
+%
+\CEU imposes a restriction that an \code{await} to an internal event is only 
+valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
 simultaneously in parallel trails, the \code{await} does not awake.
@@
 From our experience, in some cases we need to execute a block of code 
-periodically, \emph{including the current reaction}, from requests of an 
-internal event, as illustrated in the middle of Figure~\ref{lst.delay} 
+periodically, \emph{including the current reaction}, starting from requests of 
+an internal event, as illustrated in the middle of Figure~\ref{lst.delay} 
 (\code{CODE-2}).
@@
 explained in Section~\ref{sec.ceu.ints}).
-
-%Therefore, there is a single order of execution for side effects (represented 
-%by \code{mem} operations), thus, reaction chains are always deterministic.
-%
-% TODO
-%\footnote{We could extend the semantics to describe the full execution of a 
-%program by holding new incoming external events in a queue and processing them 
-%in consecutive reaction chains that never overlap.}
-%
-%The semantics requires an explicit stack to properly nest the emission of 
-%internal events.
-%to prevent that $awaiting$ expressions awake in the same reaction they are 
-%reached.
 
@@
 
-% TODO:, allowing \CEU{} to be easily embedded in platforms:
-
 \begin{figure}

### 2015-09-28 paper.tex
Renomeando Sexao 2.1 para "Queue-Based External Events and Stack-Based Internal Events".

@@
 input events that are broadcast to all awaiting trails.
-By synchronous, we mean that all trails at any given time are either reacting 
+By synchronous, we mean that all trails at any given moment are either reacting 
 to the current event or are awaiting another event;
@@
 Esterel~\cite{ceu.sensys13}:
-stack-based internal events (Section~\ref{sec.ceu.ints}),
+queue-based external events and stack-based internal events 
+(Section~\ref{sec.ceu.ints}),
 shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
@@
 
-\subsection{Stack-Based Internal Events}
+\subsection{Queue-Based External Events and Stack-Based Internal Events}
 \label{sec.ceu.ints}
@@
 %
-In \CEU, external input events are unique within reactions, with
-intrinsic queue-based handling, and programs cannot emit them.
-%
-In contrast, internal events follow a stack-based execution policy, similar to 
-subroutine calls in typical programming languages~\cite{ceu.sensys13}.
+In \CEU, external input events are unique within reactions and programs cannot 
+emit them, resulting in an intrinsic queue-based handling.
+%
+In contrast, programs can emit internal events but these follow a stack-based 
+execution policy, similar to subroutine calls in typical programming 
+languages~\cite{ceu.sensys13}.
 %
@@
     \emph{stack: [1st-trail]}
-\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
+\item 2nd trail awakes (line 9), calls \code{\_g()}, and terminates.\\
     \emph{stack: [1st-trail]}
-\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
+\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
     terminates.\\
@@
 Only after this sequence that the calling trail resumes and passes the 
-assertion test.
+assertion test (line 11).
 
@@
 In particular, Esterel has a notion of time similar to that of digital circuits 
-in which multiple signals can be active at a clock tick (Esterel is also used 
-in hardware design).
+in which multiple signals can be active at a clock tick.
+In fact, Esterel is also used in hardware design.
+%
 In \CEU, instead of clock ticks, the occurrence of a single external event that 
 defines a time unit.
-We believe that for software design, this approach simplifies the reasoning 
-about concurrency.
+The event-driven approach is popular in many 
+communities~\cite{TOS,games,GUI,callbacks,nodejs}, and is essentially 
+synchronous, even though it is not commonly/formally 
+noticed/acknowledged/recognized as such.
+%
+We believe that for software design, this approach is more familiar to 
+programmers and simplifies the reasoning about concurrency.
 For instance, the uniqueness of external events in \CEU is a prerequisite for 
@@
 %
-Furthermore, \CEU also distinguishes external events from stack-based internal 
-events, which provides a limited form of subroutines that supports \CEU's 
+Finally, \CEU also distinguishes external events from stack-based internal 
+events, which provides a limited form of coroutines that supports \CEU's 
 reactive statements.
@@
 They offer Esterel-like parallel compositions with communication via shared 
-variables, relying on deterministic semantics to preserve determinism.
+variables, relying on deterministic scheduling to preserve determinism.
 %
 However, it is the responsibility of the programmer to specify the execution 
-order for threads, based on either explicit priorities or source code lexical 
+order for threads, based on either explicit priorities, or source code lexical 
 order (like \CEU).
 %
-These languages have a tick-based notion of time similar to Esterel, also 
-disallowing the event-based temporal analysis of \CEU.
+These languages have a tick-based notion of time similar to Esterel, which 
+prevents the event-based temporal analysis of \CEU.
 
@@
 Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
-the closest to our implementation for trail allocation and scheduling:
+the closest to our implementation with respect to trail allocation and 
+scheduling:
 The compiler slices programs into ``control points'' (analogous to our ``entry 

### 2015-09-28 paper.tex
Revisao da Secao 2.1: figura 4 separada em duas

@@
 %
-The example in the left of Figure~\ref{lst.delay} (\code{CODE-1}) relies on 
-this restriction to avoid infinite execution:
-Although both sides of the \code{par/or} have an \code{await} statement, the 
-\code{loop} would restart instantaneously if the \code{await} could awake in 
-the same reaction that reaches it.
-%
-In atypical scenarios requiring immediate awake, delayed awaits can be 
-circumvented by manually copying or making some transformation in the code to 
-execute on awake.
-%
-From our experience, in some cases we need to execute a block of code 
-periodically, \emph{including the current reaction}, starting from requests of 
-an internal event, as illustrated in the middle of Figure~\ref{lst.delay} 
-(\code{CODE-2}).
-In this case, moving the \code{await} to the end of the loop suffices to 
-execute the periodic code immediately (and then in reaction to each \code{emit} 
-request).
-If the periodic \code{emit} depends on a condition, the code becomes more 
-intricate, requiring to copy the condition test to the periodic code, as 
-illustrated in the right of Figure~\ref{lst.delay} (\code{CODE-3}).
-
-\begin{figure}
-\begin{minipage}[t]{0.24\linewidth}
+The example that follows relies on this restriction to avoid infinite 
+execution:
+
 \begin{lstlisting}
@@
     with
-    emit e;
+        emit e;     // awakes 1s trail and restarts the loop
         await f;
@@
 end
-
-/* CODE-1 */
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.36\linewidth}
+\end{lstlisting}
+
+Even though both sides of the \code{par/or} have an \code{await} statement to 
+avoid instantaneous termination, the \code{loop} would restart instantaneously 
+if the \code{await} could awake in the same reaction that reaches it.
+
+In atypical scenarios requiring immediate awake, delayed awaits can be 
+circumvented by manually copying or making transformations in the code to 
+execute on awake.
+%
+From our experience, in some cases we need to execute a block of code 
+periodically from internal event requests, \emph{including the current 
+reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
+In this case, moving the \code{await} to the end of the loop makes the periodic 
+code to also execute immediately (and then in reactions to each \code{emit} 
+request).
+If the periodic \code{emit} depends on a condition, the code becomes more 
+intricate because we need to copy the condition test to the periodic code, as 
+illustrated in the right of Figure~\ref{lst.delay}.
+
+\begin{figure}
+\begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}
@@
 end
-
-/* CODE-2 */
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.36\linewidth}
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}
 event void e;
-var bool execute_now = false;
+var bool should_execute = false;
 par do
@@
         if <...> then
-        execute_now = true;
+            should_execute = true;
             emit e;
@@
     loop do
-    if execute_now then
+        if should_execute then
             <...> // code to execute
@@
 end
-
-/* CODE-3 */
-\end{lstlisting}
-\end{minipage}
-%
-\caption{ In \code{CODE-1}, the \code{await} occurs simultaneously with the 
-\code{emit} and does not awake, preventing the \code{par/or} to terminate and 
-execute the \code{loop} indefinitely.
-\code{CODE-2} and \code{CODE-3} illustrate how to circumvent the \emph{delayed 
-await} restriction.
+\end{lstlisting}
+\end{minipage}
+%
+\caption{ Examples that circumvent the \emph{delayed await} restriction by 
+post-fixing the \code{await} inside the \code{loop} (in the left), and by 
+copying the condition test (in the right).
 \label{lst.delay}

### 2015-09-28 paper.tex
Revisao da Secao 2.2.

@@
 resources through memory-mapped registers.
-Hence, an important goal of \CEU is to ensure a reliable execution for 
-concurrent programs that share memory.
+Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
+with concurrent lines of execution sharing memory.
 %
@@
 However, the execution order for operations with side-effects within a reaction 
-is non-deterministic: ``if there is no control dependency, as in \code{"call 
-f1() || call f2()"}, the order is unspecified and it would be an error to rely 
+is non-deterministic: ``if there is no control dependency, as in \code{<<call 
+f1() || call f2()>>}, the order is unspecified and it would be an error to rely 
 on it''~\cite{esterel.primer}.
 %
-Therefore, Esterel forbids sharing memory: \emph{``if a variable is written by 
-some thread, then it can neither be read nor be written by concurrent 
-threads''}~\cite{esterel.primer}.
+Therefore, Esterel forbids sharing memory between lines of execution: 
+\emph{``if a variable is written by some thread, then it can neither be read 
+nor be written by concurrent threads''}~\cite{esterel.primer}.
 
@@
 \code{printf}).
-The algorithm for the analysis holds the set of all events in preceding 
-\code{await} statements for each variable access.
-Then, the sets for all accesses in parallel trails are compared to assert that 
-no events are shared among them.
-Otherwise the compiler warns about the suspicious accesses.
-
-Regardless of the static analysis, when multiple trails are active at a time in 
-\CEU, they are scheduled in the order they appear in the program source code.
+%The algorithm for the analysis holds the set of all events in preceding 
+%\code{await} statements for each variable access.
+%Then, the sets for all accesses in parallel trails are compared to assert that 
+%no events are shared among them.
+%Otherwise the compiler warns about the suspicious accesses.
+
+Regardless of the temporal analysis of \CEU, when multiple trails are active 
+during the same reaction, they are scheduled in the order they appear in the 
+program source code.
 %
 Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
-suspicious, the assignments to \code{y} both atomic and deterministic, i.e., 
-after \code{A} occurs, \code{y} is $4$ (\code{(1+1)*2}).
+suspicious, the assignments to \code{y} are both atomic and deterministic, 
+i.e., after the reaction to \code{A} terminates, the value of \code{y} is $4$ 
+(\code{(1+1)*2}).
 %
@@
 \begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
+\begin{lstlisting}[mathescape=true]
 // ESTEREL
+input A;
 [
+    await A;
     call f1();
 $\|$
+    await A;
     call f2();
@@
 \begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+\begin{lstlisting}
 // CEU
+input void A;
 par/and do
+    await A;
     _f1();
 with
+    await A;
     _f2();
@@
 \caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
-unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
+unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
+deterministic scheduling based on lexical order.
 \label{lst.det}

### 2015-09-28 paper.tex
Revisao Secao 2.3

@@
 thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
-However, handling abortion when dealing with external resources can be 
-challenging, given that external entities are not subject to the same 
-synchronous execution discipline.
+Still, handling abortion when dealing with external resources is challenging 
+because they are not subject to the same synchronous execution discipline.
 
@@
 %\rule{8.5cm}{0.37pt}
-\caption{The unsafe network protocol in the left is extended with a 
-finalization clause in the right to handle abortion properly.
+\caption{The unsafe network protocol in the left, which does not compile, is 
+extended with a finalization clause in the right to handle abortion properly.
 %\caption{ Parallel compositions can describe complex state machines.
@@
 
-To illustrate threats related to abortion, consider the unsafe example in the 
-left of Figure~\ref{lst.fin}, describing the state machine of a data collection 
-protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
+To illustrate threats related to abortion, consider the unsafe example, which 
+does not compile, in the left of Figure~\ref{lst.fin}, describing the state 
+machine of a data collection protocol for sensor 
+networks~\cite{wsn.ctp,ceu.sensys13}.
 %
 The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
-represent the external interface of the protocol with a client application.
+represent the external interface of the protocol.
 %
 The protocol has to transmit a packet every minute, unless it receives a 
-\code{RETRANSMIT} request to immediately re-transmit, or a \code{STOP} request 
-to terminate.
+\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
+request to terminate.
 %
@@
 one simply monitors the stopping event (line 3);
-the other periodically transmits a packet (lines 5--18).
+the other periodically transmits the packet (lines 5--18).
 %
@@
 The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
-takes at least one minute before looping, to avoid flooding the network with 
-packets.
+takes at least one minute before looping, in accordance with the specification.
 Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
 the radio driver a pointer to the lexically-scoped packet.
-The driver will make the transmission in the background, holding the packet 
-until it signals the application with the \code{SENDACK} to acknowledge 
-completion.
+The driver makes the transmission in the background, holding the packet until 
+it signals the application with the \code{SENDACK} to acknowledge completion.
 %
@@
 %
-In the meantime, if the sending trail is be aborted by the \code{STOP} or 
+Therefore, if the sending trail is aborted by the \code{STOP} or 
 \code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
 a \emph{dangling pointer} in the radio driver, which will possibly transmit 
-wrong data.
+corrupted data.
 %
@@
 
-The \CEU compiler tracks the interaction of \code{par/or} compositions with 
-local variables and stateful \emph{C} functions (e.g., device drivers) in order 
-to preserve safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
+The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
+\CEU tracks the interaction of \code{par/or} compositions with local variables 
+and stateful \emph{C} functions (e.g., device drivers) in order to preserve 
+safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
 %
@@
 %
-For instance, \CEU refuses to compile the program in the left of 
-Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
-clause to accompany the stateful \emph{C} call.
+In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
+accompany the stateful \emph{C} call.
 The code in the right of Figure~\ref{lst.fin} properly cancels the packet 

### 2015-09-28 paper.tex
Revisao Secao 3.

@@
 For the sake of simplicity, we focus on the control aspects of the language, 
-leaving out side effects and $C$ calls (which behave like in any conventional 
-imperative language).
+leaving out side effects and $C$ calls (which behave like in conventional 
+imperative languages).
 
@@
 
-We describe this relation with a set of \emph{small-step} structural semantics 
-rules, which are built in such a way that at most one transition is possible at 
-any time, resulting in deterministic reaction chains.
+We describe this relation with a set of small-step structural semantics rules, 
+which are built in such a way that at most one transition is possible at any 
+time, resulting in deterministic reaction chains.
 %
@@
 \LL S,\1awaiting(id,n+1) \RR
-    & \textbf{(awaiting)}      \\
+    & \textbf{(await)}      \\
 %%%
@@
 \LL id:S,\1emitting(|S|) \RR
-    & \textbf{(emitting)}       \\
+    & \textbf{(emit)}       \\
 %%%
@@
 %
-An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
-an artifact to remember the external sequence number $n$ it can awake.
-An $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
-event $id$ matches the top of the stack and its sequence number is smaller or 
-equal than the current one ($m<n$).
+An $await$ is simply transformed into an $awaiting$ (rule \textbf{await}) as an 
+artifact to remember the external sequence number $n+1$ it can awake:
+an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
+event $id$ matches the top of the stack and it was reached in a previous 
+reaction (i.e., sequence number $m<n$).
 %
@@
 stands for the stack length), and pushing the referred event on the stack (rule 
-\textbf{emitting}).
+\textbf{emit}).
 With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
@@
 %
-The ``magical'' function $val$ receives the memory identifier and current 
+The ``magical'' function $val$ receives a memory identifier and the current 
 reaction sequence number, returning the current memory value.
@@
 a given expression can execute only once within it (remember that $loops$ must 
-contain $awaits$ which, from rule \textbf{awaiting}, cannot awake in the same 
+contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
 reaction they are reached).
@@
 The deterministic behavior of the semantics relies on the \emph{isBlocked} 
-predicate, defined in Figure~\ref{fig.isBlocked} and used in rules 
-\textbf{and-adv2} and \textbf{or-adv2}, requiring the left branch $p$ to be 
-blocked in order to allow the right transition from $q$ to $q'$.
+predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
+\textbf{and-adv2} and \textbf{or-adv2}.
+These rules require the left branch $p$ to be blocked in order to allow the 
+right transition from $q$ to $q'$.
 %
@@
 blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
-expressions.
+expressions that cannot advance.
 
@@
 in each of the sides in parallel.
-A $break$ should terminate the whole composition in order to escape the 
-innermost loop (\emph{aborting} the other side):
+To escape the innermost loop, a $break$ aborts the other side by applying the 
+$clear$ function (to be described further):
 %
@@
 %
-The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
-all active $fin$ bodies of the side being aborted (to execute before the $and$ 
-rejoins).
-Note that there are no transition rules for $fin$ expressions.
-This is because once reached, a $fin$ expression halts and will only execute 
-when it is aborted by a trail in parallel and is expanded by the $clear$ 
-function.
-In Section~\ref{sec.formal.fins}, we show how a $fin$ is mapped to a 
-finalization block in the concrete language.
-%
-Note that there is a syntactic restriction that a $fin$ body can only contain 
-$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
-reaction chain.
-
-\begin{figure}
-{\small
-\begin{align*}
-  clear( fin~p )       &= p                   \\
-  clear( p~;~q )       &= clear(p)            \\
-  clear( p~@~loop~q) ) &= clear(p)            \\
-  clear( p~and~q )     &= clear(p)~;~clear(q) \\
-  clear( p~or~q )      &= clear(p)~;~clear(q) \\
-  clear( \_ )          &= mem(id)
-\end{align*}
-}%
-%\rule{14cm}{0.37pt}
-\caption{
-The function $clear$ extracts $fin$ expressions in parallel and put their 
-bodies in sequence.
-\label{fig.formal.clear}
-}
-\end{figure}
-
 For a parallel $or$, if one of the sides terminates, the whole composition 
-terminates and applies the function $clear$ to the aborted side in order to 
-properly finalize it (rules \textbf{or-nop1} and \textbf{or-nop2}).
+terminates, also applying the $clear$ function to the aborted side (rules 
+\textbf{or-nop1} and \textbf{or-nop2}).
 Finally, a $break$ (rules \textbf{or-brk1} and \textbf{or-brk2}) behaves in the 
@@
 %
+The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
+all active $fin$ bodies of the side being aborted, so that they execute before 
+the composition rejoins.
+Note that there are no transition rules for $fin$ expressions.
+This is because once reached, a $fin$ expression halts and will only execute 
+when it is aborted by a trail in parallel and is expanded by the $clear$ 
+function.
+In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
+the concrete language to a $fin$ in the formal semantics.
+%
+Note that there is a syntactic restriction that a $fin$ body can only contain 
+$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
+reaction chain.
+
+\begin{figure}
+{\small
+\begin{align*}
+  clear( fin~p )       &= p                   \\
+  clear( p~;~q )       &= clear(p)            \\
+  clear( p~@~loop~q) ) &= clear(p)            \\
+  clear( p~and~q )     &= clear(p)~;~clear(q) \\
+  clear( p~or~q )      &= clear(p)~;~clear(q) \\
+  clear( mem(id) )     &= mem(id)             \\
+  clear( \_ )          &= nop
+\end{align*}
+}%
+%\rule{14cm}{0.37pt}
+\caption{
+The function $clear$ extracts $fin$ expressions in parallel and put their 
+bodies in sequence.
+\label{fig.formal.clear}
+}
+\end{figure}
+
 A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
@@
 %
-However, $emitting$ expressions should resume their continuations of previous 
-$emit$ in the ongoing reaction, they are just hanged in lower stack indexes 
-(see rule \textbf{pop}).
+However, $emitting$ expressions are pending in lower stack indexes and should 
+eventualy resume in the ongoing reaction (see rule \textbf{pop}).
 %
 Therefore, we define another relation that behaves as \textbf{relation-inner} 
-(presented above) if the program is not blocked, and, otherwise, pops the 
-stack:
+(presented above) if the program can advance, and, otherwise, pops the stack:
 %
@@
 %
-Finally, to describe the complete execution of a program, we need multiple 
-``invocations'' of reaction chains, incrementing the sequence number:
+Finally, to describe the complete execution of a program, we trigger multiple 
+``invocations'' of reaction chains in sequence:
 %
@@
 \\
+\LL [e3], p'' \RR
+    & \xRightarrow[~~3~~]{*}
+\LL [  ], p''' \RR
+\\
 & ...
@@
 finishes with a modified program and an empty stack.
-After each invocation, the sequence number is incremented.
+After each invocation, we increment the sequence number.
 

### 2015-09-28 paper.tex
(>) Revisao Secao 3.3

@@
   clear( p~or~q )      &= clear(p)~;~clear(q) \\
-  clear( mem(id) )     &= mem(id)             \\
   clear( \_ )          &= nop
@@
 
-Most statements from \CEU map directly to those presented in the reduced syntax 
-of Figure~\ref{lst.formal.syntax}, namely, the \code{if}, \code{`;'}, 
-\code{loop}, \code{par/and}, and \code{par/or}.
+Most statements from \CEU (``concrete \CEU'') map directly to those presented 
+in the reduced syntax of Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
+%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
+%\code{par/or}.
 For instance, the \code{if} in the concrete language behaves exactly like the 
-$if$ in the described semantics.
-However, there are some significant mismatches between \CEU and the formal 
-semantics, which we provide a mapping in this section.
+$if$ in the formal semantics.
+However, there are some significant mismatches between the concrete and 
+abstract \CEU, and we discuss appropriate mappings in this section.
+%
+Again, we are not considering side-effects, which are all mapped to the $mem$ 
+semantic construct.
 
@@
 \end{comment}
-Again, we are not considering side-effects, which are all mapped to the $mem$ 
-semantic construct.
 
@@
 
-The concrete \code{await} and \code{emit} primitives of \CEU support 
-communication of values between them.
+The concrete \code{await} and \code{emit} primitives support communication of 
+values between them.
 In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
@@
 the \code{emit}.
-Finally, we convert the program into the equivalent in the formal semantics, 
+Finally, we convert the program into the equivalent in the abstract syntax, 
 translating side-effect statements into $mem$ expressions.
@@
 two-step translation of Figure~\ref{lst.map.timers.2}.
-
-TODO: continuar explicacao
+In the concrete program in the left, the variable \code{dt} will hold the 
+residual delta time (as described in Section~\ref{sec.ceu.timers}) after 
+awaking from the timer.
+In the first step of the translation, we expand the \code{await 10ms} to a 
+\code{loop} that decrements the elapsed number of microseconds for each 
+occurrence of \code{TICK}.
+When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
+the variable \code{dt} to contain the appropriate delta.
+In the last step, we convert the program to the abstract syntax.
 
@@
 
-The biggest mismatch between \CEU and the formal semantics is regarding 
+The biggest mismatch between concrete and abstract \CEU is regarding 
 finalization blocks, which require more complex modifications in the program
-for a proper mapping using the $fin$ semantic construct.
+for a proper mapping using the $fin$ expressions.
 In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
@@
 \code{\_hold}.
-In the translation to the semantics, we first need to catch the \code{do-end} 
-termination to run the finalization code.
+In the translation, we first need to catch the \code{do-end} termination to run 
+the finalization code.
 For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 

### 2015-09-28 paper.tex
Revisao Secao 4.1, 4.2

@@
 $I_n=O_{n-1}$.
-The first child inherits $I$ from the sequence parent, and the set $O$ for the 
+The first child inherits $I$ from its parent node, and the set $O$ for the 
 sequence node is copied from its last child, i.e., $O=O_n$.
@@
 %
-\item A parallel composition (\code{par/and} / \code{par/or}) may terminate 
-from any of its branches, hence $O = O_1 \cup ... \cup O_n$.
+\item A parallel composition may terminate from any of its branches, hence $O = 
+O_1 \cup ... \cup O_n$.
 \end{itemize}
 
-With all sets calculated, any two nodes that perform side effects and are in 
-parallel branches can have their $I$ sets compared for intersections.
-If the intersection is not the empty set, they are marked as suspicious.
+With all sets calculated, we take all pairs of nodes that perform side effects 
+and are in parallel branches, comparing their sets $I$ for intersections.
+For each pair, if the intersection is not the empty set, we mark both nodes as 
+suspicious.
 
@@
 resides in fixed memory slots---this is true for the program variables as well 
-as for temporary values and flags needed during runtime.
+as for temporary values and runtime flags.
 %For instance, the first trail in the guiding example requires temporary slots 
@@
 reuse it.
+%
+Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
+memory for blocks in sequence are packed in a \code{struct}, while blocks in 
+parallel, in a \code{union}.
 %In the example, the code following the loop (identified as \code{...}) reuses 
@@
 during runtime.
+%
+As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
+memory layout.
+%
+Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
+with the same name can be distinguished.
+%
+The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
+their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
+\code{MEM.b\_2} can safely share the same memory slot.
+%
+The example also illustrates the presence of runtime flags related to the 
+parallel composition, which also reside in reusable slots in the static memory.
 
@@
     struct {        //   par/and
-        u8 _and_3: 1;
-        u8 _and_4: 1;
+        int _and_3: 1;
+        int _and_4: 1;
     };
@@
 \end{figure}
-
-Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
-memory for blocks in sequence are packed in a \code{struct}, while blocks in 
-parallel, in a \code{union}.
-%
-As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
-memory layout.
-%
-Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
-with the same name can be distinguished.
-%
-The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
-their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
-\code{MEM.b\_2} can safely share the same memory slot.
-%
-The example also illustrates the presence of runtime flags related to the 
-parallel composition, which also reside in reusable slots in the static memory.
 
@@
 noticed/acknowledged/recognized as such.
+(TODO: rewrite)
 %

### 2015-09-28 paper.tex
Revisao Secao 4.3, 4.4

@@
 
-Each line of execution in \CEU carries associated data, such as which event it 
-is awaiting and which code to execute when it awakes.
+Each line of execution in \CEU needs to carry associated data, such as which 
+event it is awaiting and which code to execute when it awakes.
 %
@@
 %
-Trails that cannot be active at the same time can share memory slots in the 
-static vector.
+Like normal variables, trails that cannot be active at the same time can share 
+slots in the static memory vector.
 
@@
 %
-A \code{STACKED} trail holds its associated stack level and is delayed until 
-the scheduler runtime level reaches that value again.
+A \code{STACKED} trail holds an associated stack level and is delayed until the 
+scheduler runtime reaches that level again.
 %
@@
 %
-A trail is represented by the following \code{struct}:
+In concrete terms, a trail is represented by the following \code{struct}:
 
@@
 The size of \code{label\_t} depends primarily on the number of \code{await} 
-statements in the application---each \code{await} splits the code in two and 
-requires a unique entry point in the code for its continuation.
+statements in the application---each \code{await} splits the code in two 
+segments and requires a unique entry point in the code for its continuation.
 Additionally, split \& join points for parallel compositions, \code{emit} 
@@
 %
-The \code{seqno} will eventually overflow during execution (every 256 
+The \code{seqno} could eventually overflow during execution (i.e., every 256 
 reactions).
-However, given that the scheduler traverses all trails in each reaction, it can 
-adjust them to properly handle overflows (actually 2 bits to hold the 
+However, given that the scheduler traverses all trails on every reaction, it 
+can adjust them to properly handle overflows (actually, 2 bits to hold the 
 \code{seqno} is already enough).
 %
-The stack size depends on the maximum depth of nested emissions and is bounded 
-to the maximum number of trails, e.g., a trail emits an event that awakes 
-another trail, which emits an event that awakes another trail, and so on---the 
-last trail cannot awake any trail, because they will be all hanged in a 
+The size of \code{stack\_t} depends on the maximum depth of nested emissions 
+and is bounded to the maximum number of trails.
+In the worst case, a trail emits an event that awakes another trail, which 
+emits an event that awakes another trail, and so on.
+The last trail cannot awake any trail, because they are all hanged in the 
 \code{STACKED} state.
@@
 %
-For instance, the \emph{CTP} collection protocol ported to \CEU has at most 
-eight simultaneous lines of execution with an overhead of 2\% in comparison to 
-the original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
+For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
+simultaneous lines of execution with an overhead of 2\% in comparison to the 
+original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
 programming)~\cite{ceu.sensys13}.
@@
 %
-For each statement that yields (\code{emit} and \code{await}), resumes 
+For each statement that pauses (\code{emit} and \code{await}), resumes 
 (\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
@@
 1--10, in the middle of the figure).
-The trails translate to a vector of \code{trail\_t} elements with the maximum 
-number of simultaneous trails (lines 12--16).
-Initially, \code{TRAIL-0} is set to execute the \code{Main} entry point, while 
-all others are set to \code{INACTIVE}.
+The state of trails translate to a vector of type \code{trail\_t} with the 
+maximum number of simultaneous trails (lines 12--15).
+On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
+(line 13), while all others are set to \code{INACTIVE} (in the example, only 
+one, in line 14).
 
@@
 %
-In the \emph{broadcast} pass, the scheduler sets to \code{STACKED} all trails 
-that are waiting for the current event to execute in the current stack level.
+In the \emph{broadcast} pass, the scheduler sets all trails that are waiting 
+for the current event to \code{STACKED} in the current stack level.
 %
@@
 
-During the dispatch pass, if a trail executes and emits and internal event, the 
+During the dispatch pass, if a trail executes and emits an internal event, the 
 scheduler increments the stack level and re-executes the two passes.
@@
 The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
-according to its current label to execute.
+according to the current label to execute.
 %
@@
 %
-When the \code{Main} label reaches the \code{par/and}, it ``stacks'' 
-\code{TRAIL-1} (lines 4--8) and proceeds to the code in \code{TRAIL-0} (lines 
-9--14), respecting the deterministic execution order.
+When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
+(lines 4--7) and proceeds to the code in \code{TRAIL-0} (lines 9--14), 
+respecting the deterministic execution order.
 %
@@
 %whole process is guaranteed to terminate in bounded time.
-Escaping a \code{loop} that contains parallel compositions also trigger the 
+Escaping a \code{loop} that contains parallel compositions also triggers the 
 same abortion process.

### 2015-09-29 paper.tex
Revisao "Related Work"

@@
 simultaneous lines of execution with an overhead of 2\% in comparison to the 
-original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
-programming)~\cite{ceu.sensys13}.
+original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \emph{C} for 
+event-driven programming)~\cite{ceu.sensys13}.
 
@@
     event void Boot.booted () {
-        ceu_go_init();
+        ceu_sys_init();
 #ifdef CEU_WCLOCKS
@@
     event void Timer.fired () {
-        ceu_go_wclock(10000);
+        int dt = 10000;
+        ceu_sys_go(EVT__WCLOCK, &dt);
     }
@@
     event void Photo.readDone (uint16_t val) {
-        ceu_go_event(EVT_PHOTO_READDONE, (void*)val);
+        ceu_sys_go(EVT_PHOTO_READDONE, &val);
     }
@@
     event void RadioSend.sendDone (message_t* msg) {
-        ceu_go_event(EVT_RADIO_SENDDONE, msg);
+        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
     }
@@
     event message_t* RadioReceive.receive (message_t* msg) {
-        ceu_go_event(EVT_RADIO_RECEIVE, msg);
+        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
         return msg;
@@
 
-TODO: refs/cite WSNs
-
-As an example, Figure~\ref{lst.impl.tinyos} shows our binding for \emph{TinyOS} 
-which maps \emph{nesC} callbacks to input events in \CEU.
+As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
+\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
 %
@@
 If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
-that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
-and advances the wall-clock time of \CEU (line 15)%
+that triggers callback \code{Timer.fired} (lines 13--18) every 10 milliseconds 
+and advances the wall-clock time of \CEU (line 16)%
 \footnote{We also offer a mechanism to start the underlying timer on demand to
@@
 The remaining lines map pre-defined TinyOS events that can be used in \CEU 
-programs, such as the light sensor (lines 19--23) and the radio transceiver 
-(lines 25--36).
+programs, such as the light sensor (lines 20--24) and the radio transceiver 
+(lines 26--37).
 
@@
 defines a time unit.
-The event-driven approach is popular in many 
-communities~\cite{TOS,games,GUI,callbacks,nodejs}, and is essentially 
-synchronous, even though it is not commonly/formally 
-noticed/acknowledged/recognized as such.
-(TODO: rewrite)
+%
+\CEU also distinguishes external events from stack-based internal events, which 
+provides a limited form of coroutines that supports reactive statements (e.g., 
+\code{await} and \code{par/or}).
+
+The event-driven approach~\cite{sync_async.whynotthreads} is well known and 
+popular in many software communities,
+such as web frameworks in the client side (e.g., \emph{jQuery}~\cite{js.jquery} 
+and \emph{Rx.js}~\cite{js.rx}) and server side (e.g., 
+\emph{Node.js}~\cite{js.node} and \emph{SEDA}~\cite{sync_async.seda}),
+GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java 
+Swing}~\cite{java.swing}),
+and Games~\cite{gamepatterns}.
+%
+Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
+through a queue and are dispatched sequentially and atomically to prevent race 
+conditions.
 %
@@
 the temporal analysis that enables safe shared-memory concurrency.
-%
-Finally, \CEU also distinguishes external events from stack-based internal 
-events, which provides a limited form of coroutines that supports \CEU's 
-reactive statements.
 

### 2015-09-29 paper.tex
Introducao

@@
 one aiming for resource efficiency and interoperability with C, and another for 
-code dissemination in sensor networks.
+dynamic reprogramming over the air.
 \end{abstract}
@@
 
+An established alternative to \emph{C} in the field of embedded systems is the 
+family of reactive synchronous languages \cite{rp.twelve}.
+%
+Two major styles of synchronous languages have evolved:
+in the \emph{control}--\emph{imperative} style, programs are structured with 
+control flow primitives, such as parallelism, repetition, and preemption;
+in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
+graphs of values, in which a change to a value is propagated through its 
+dependencies without explicit programming.
+%
+Among the control-based languages, Esterel~\cite{esterel.ieee91} influenced a 
+number of other embedded languages, such as \emph{Reactive-C}~\cite{rp.rc}, 
+\emph{OSM}~\cite{wsn.osm}, \emph{Sync-C}~\cite{rp.synchc}, and 
+\emph{PRET-C}~\cite{rp.pretc}.
+
+\CEU is another language targeting embedded systems based on Esterel with some 
+novel functionalities:
+%
 \begin{itemize}
-\item Relevancia de Esterel
-\item Ceu
-    \begin{itemize}
-        \item Foco inicial em "constrained WSNs"
-        \item Estendendo para outros dominios? (jogos, artigo Mod'15)
-        \item Usos: SenSys, Terra, sala de aula, GSoC
-        \item ideia de tamanhos, ROM,RAM,CPU
+%
+\item Stack-based internal events, which provide a limited form of coroutines.
+%
+\item A static temporal analysis and deterministic execution semantics that 
+allows programs to safely share memory.
+%
+\item A finalization mechanism for safe abortion of lines of execution.
+%
+\item First-class synchronized timers.
 \end{itemize}
-\item Limitacoes do modelo sincrono
-\item Semantica e Implementacao (nesse artigo, somente subset estatico)
+
+\CEU shares limitations with Esterel and synchronous languages in general:
+computations that run in unbounded time (e.g., cryptography, image processing) 
+do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
+elegantly implemented.
+
+In this work, we present a formal semantics of a small synchronous kernel that 
+represents a subset of \CEU covering these new functionalities.
+%
+We also present an implementation of \CEU with two backends:
+one aiming for efficiency and interoperability with \emph{C},
+and another based on a virtual machine that allows remote reprogramming.
+
+In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
+of constrained wireless sensor networks, developing a number of applications, 
+protocols, and drivers.
+%
+We evaluated the expressiveness of \CEU in comparison to event-driven code in 
+\emph{C}, and observed a reduction in source code size (around 25\%) with a 
+small increase in memory usage (around 5--10\% for ROM and 
+RAM)~\cite{ceu.sensys13}.
+%
+Considering the \emph{VM} backend, a simple application that blinks three LEDs 
+periodically occupies less than 100 bytes and can be completely transmitted in 
+4 radio messages.
+In contrast, without the preinstalled \emph{VM} with basic \emph{I/O} 
+functionality, a \emph{C}-compiled version occupies more than 2 
+kbytes~\cite{ceu.terra}.
+
+The rest of the paper is organized as follows:
+Section~\ref{sec.ceu} gives an overview of \CEU, exposing its fundamental 
+differences to Esterel.
+Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
+\CEU.
+Section~\ref{sec.impl} presents the \emph{C} and \emph{VM} implementation 
+backends.
+Section~\ref{sec.related} discusses other synchronous languages targeting 
+embedded systems.
+Section~\ref{sec.conclusion} concludes the paper.
+
+%\begin{itemize}
+        %\item Estendendo para outros dominios? (jogos, artigo Mod'15)
+        %\item Usos: sala de aula, GSoC
+%\item Semantica e Implementacao (nesse artigo, somente subset estatico)
 % Artigo de Rust, nao resolve os problemas
@@
 %state machines everywhere
-\end{itemize}
+%\end{itemize}
 
@@
 \section{Formal Semantics}
+\label{sec.sem}
 
@@
 For the sake of simplicity, we focus on the control aspects of the language, 
-leaving out side effects and $C$ calls (which behave like in conventional 
+leaving out side effects and \emph{C} calls (which behave like in conventional 
 imperative languages).
@@
 
-The $mem(id)$ primitive represents all accesses, assignments, and $C$ function 
-calls that affect a memory location identified by $id$.
+The $mem(id)$ primitive represents all accesses, assignments, and \emph{C} 
+function calls that affect a memory location identified by $id$.
 As the challenging parts of \CEU reside on its control structures, we are not 
@@
 \subsection{Operational Semantics}
-\label{sec.sem}
 
@@
 \section{Implementation}
+\label{sec.impl}
 
@@
 %
-This phase outputs code in $C$, given the way \CEU is tied to $C$ by design.
+This phase outputs code in \emph{C}, given the way \CEU is tied to \emph{C} by 
+design.
 %
@@
 \code{gcc}'s error handling.
-Therefore, we annotate the $C$ output file with \code{\#line} pragmas matching 
-the original file in \CEU.
+Therefore, we annotate the \emph{C} output file with \code{\#line} pragmas 
+matching the original file in \CEU.
 
@@
 
-The final phase packs the generated $C$ file with the \CEU runtime and 
+The final phase packs the generated \emph{C} file with the \CEU runtime and 
 platform-dependent functionality, compiling them with \emph{gcc} and generating 
@@
 The \CEU runtime comprehends the scheduler, timer management, and the external 
-$C$ API.
+\emph{C} API.
 %
@@
 %
-\item Nodes that represent expressions, assignments, $C$ calls, and 
+\item Nodes that represent expressions, assignments, \emph{C} calls, and 
 declarations simply reproduce $O=I$, as they do not await;
@@
 %
-Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
+Translating this idea to \emph{C} is straightforward~\cite{wsn.osm,wsn.ocram}: 
 memory for blocks in sequence are packed in a \code{struct}, while blocks in 
@@
 
-\subsection{The External $C$ API}
+\subsection{The External \emph{C} API}
 
@@
 From the implementation perspective, there are three external sources of input 
-into programs, which are all exposed as functions in a $C$ API:
+into programs, which are all exposed as functions in a \emph{C} API:
 
@@
 input event or the passage of time.
-In this case, the $C$ implementation makes a tail call to the corresponding 
-handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as synchronous 
-code has higher priority.
+In this case, the \emph{C} implementation makes a tail call to the 
+corresponding handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as 
+synchronous code has higher priority.
 
@@
 
-\section{Terra and the \CEU-VM}
+\subsection{The Terra Virtual Machine}
+\label{sec.terra}
 
 \section{Related Work}
+\label{sec.related}
 
@@
 A number of synchronous languages have been designed to interoperate with C, 
-such as \emph{Reactive C}~\cite{rp.rc}, Protothreads \cite{wsn.protothreads}, 
+such as \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
 \emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
@@
 \section{Final Remarks}
+\label{sec.conclusion}
 

### 2015-09-29 paper.tex
Texto do Adriano

@@
 
+Terra is a system that implements a specif programming model for wireless 
+sensor networks~\cite{Branco2015a}.
+Terra is based on virtual machine concepts that uses an implementation of Céu 
+language as its scripting language.
+
+Terra programming model combines the use of configurable component-based 
+virtual machines with a reactive scripting language which can be statically 
+analyzed to avoid unbounded execution and memory conflicts.
+This approach allows the flexibility of remotely uploading code on motes to be 
+combined with a set of guarantees for the programmer.
+The choice of the specific set of  components in a virtual machine 
+configuration defines the abstraction level seen by the application script.
+
+Terra uses \ceut as its scripting language and implements a component-based 
+virtual machine \vmt to be customized for different application domains.
+\ceut implements a variation of Céu programming language.
+This variation, on one hand, excludes some of Céu's characteristics and, on the 
+other hand, includes some language extensions.
+We use Céu as Terra's scripting language because of its reactive nature and its 
+high-level control primitives and compile-time safety guarantees.
+We want only the VM components to escape the safety analysis, so we took out 
+the facility to include arbitrary C code, but we did maintain all of Céu's 
+original control structures.
+
+The Terra Virtual machine \vmt, with its customizations, runs on WSN nodes with 
+limited resources like MicaZ\cite{Memsic2009b} and TelosB\cite{Memsic2009} 
+motes.
+We built \vmt using the nesC programming language~\cite{Gay2003} and the TinyOS 
+operating system~\cite{Levis2004}.
+We used TinyOS, basically, because it is the operating system for WSN  most 
+ported to different mote types.
+
+Figure \ref{fig:system} presents the three basic elements of Terra -- a 
+scripting language (\ceut), a set of customized pre-built components, and an 
+embedded virtual-machine engine.
+The \vmt runtime environment is the virtual machine combined with a set of 
+customized components.
+
+\begin{figure}[!htb]
+ \centering
+ \includegraphics[width=0.7\columnwidth]{fig_system}
+\caption{Terra programming system basic elements.}
+ \label{fig:system}
+\end{figure}
+
+A \ceut program is compiled to a bytecode file that can then be disseminated to 
+the network nodes, where it is interpreted by the \vmt, which implements the 
+bytecode interpreter, the execution model, the code dissemination service, and 
+some specific customized components.
+Figure \ref{fig.impl-t} illustrates the Terra compilation process.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=1.0\columnwidth]{impl-t}
+\caption{Terra Compilation process: from the source code in \ceut to the final 
+bytecode.
+\label{fig.impl-t}
+}
+\end{figure}
+
+Similar to the Céu compiler process, the Terra process also has the three main phases:
+Parsing, Temporal Analysis, Code Generation.
+The Terra parsing phase uses the original Céu parser adding some extensions to 
+parse specific \ceut elements as the data types, the data structures, and the 
+configuration block.
+A configuration block defines all events, functions, and data structures 
+available in a specific \vmt customization.
+The Terra Temporal Analysis phase  only differs in the code output. Instead of 
+generating code in C, Terra generates the \vmt assembly code using the original 
+Céu labels for trails entry points.
+In the final phase for code generation, Terra compiler generates the \vmt 
+bytecode file using the program address in place of Céu labels.
+The use of address allows the Céu scheduler to execute directly a trail without 
+a index table to replace the switch-case.
+Related to memory layout and trail allocation, Terra uses the same data 
+organization approach for all Céu controls and variables.
+All Céu data and control structures for a specific program are determined at 
+compile time.
+Terra arranges all these structures in the beginning of virtual machine memory 
+space.
+The original Céu run time control functions are maintained with a small 
+modification where accesses to control structures have an off-set relative to 
+the begin of the virtual machine memory space.
+
 \section{Related Work}

### 2015-09-29 paper.tex
Revisao Terra

--- 2015-09-29 paper.tex
@@
 \usepackage{xspace}
+\newcommand{\C}{\emph{C}\xspace}
+\newcommand{\VM}{\emph{VM}\xspace}
 \newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
@@
 
-An established alternative to \emph{C} in the field of embedded systems is the 
-family of reactive synchronous languages \cite{rp.twelve}.
+An established alternative to \C in the field of embedded systems is the family 
+of reactive synchronous languages \cite{rp.twelve}.
 %
@@
 We also present an implementation of \CEU with two backends:
-one aiming for efficiency and interoperability with \emph{C},
+one aiming for efficiency and interoperability with \C,
 and another based on a virtual machine that allows remote reprogramming.
@@
 We evaluated the expressiveness of \CEU in comparison to event-driven code in 
-\emph{C}, and observed a reduction in source code size (around 25\%) with a 
-small increase in memory usage (around 5--10\% for ROM and 
-RAM)~\cite{ceu.sensys13}.
-%
-Considering the \emph{VM} backend, a simple application that blinks three LEDs 
+\C, and observed a reduction in source code size (around 25\%) with a small 
+increase in memory usage (around 5--10\% for ROM and RAM)~\cite{ceu.sensys13}.
+%
+Considering the \VM backend, a simple application that blinks three LEDs 
 periodically occupies less than 100 bytes and can be completely transmitted in 
 4 radio messages.
-In contrast, without the preinstalled \emph{VM} with basic \emph{I/O} 
-functionality, a \emph{C}-compiled version occupies more than 2 
-kbytes~\cite{ceu.terra}.
+In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
+a \C-compiled version occupies more than 2 kbytes~\cite{ceu.terra}.
 
@@
 \CEU.
-Section~\ref{sec.impl} presents the \emph{C} and \emph{VM} implementation 
-backends.
+Section~\ref{sec.impl} presents the \C and \VM implementation backends.
 Section~\ref{sec.related} discusses other synchronous languages targeting 
@@
 The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
-with external \emph{C} libraries explicit and are required in \CEU.
+with external \C libraries explicit and are required in \CEU.
 }
@@
 \CEU tracks the interaction of \code{par/or} compositions with local variables 
-and stateful \emph{C} functions (e.g., device drivers) in order to preserve 
-safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
+and stateful \C functions (e.g., device drivers) in order to preserve safe 
+abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
 %
@@
 In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
-accompany the stateful \emph{C} call.
+accompany the stateful \C call.
 The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
@@
 %
-%Note also that \CEU environments rely on \emph{C} libraries that only provide 
+%Note also that \CEU environments rely on \C libraries that only provide 
 %asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.
@@
 For the sake of simplicity, we focus on the control aspects of the language, 
-leaving out side effects and \emph{C} calls (which behave like in conventional 
+leaving out side effects and \C calls (which behave like in conventional 
 imperative languages).
@@
 
-The $mem(id)$ primitive represents all accesses, assignments, and \emph{C} 
-function calls that affect a memory location identified by $id$.
+The $mem(id)$ primitive represents all accesses, assignments, and \C function 
+calls that affect a memory location identified by $id$.
 As the challenging parts of \CEU reside on its control structures, we are not 
@@
 %
-This phase outputs code in \emph{C}, given the way \CEU is tied to \emph{C} by 
-design.
+This phase outputs code in \C, given the way \CEU is tied to \C by design.
 %
@@
 \code{gcc}'s error handling.
-Therefore, we annotate the \emph{C} output file with \code{\#line} pragmas 
-matching the original file in \CEU.
+Therefore, we annotate the \C output file with \code{\#line} pragmas matching 
+the original file in \CEU.
 
@@
 
-The final phase packs the generated \emph{C} file with the \CEU runtime and 
+The final phase packs the generated \C file with the \CEU runtime and 
 platform-dependent functionality, compiling them with \emph{gcc} and generating 
@@
 The \CEU runtime comprehends the scheduler, timer management, and the external 
-\emph{C} API.
+\C API.
 %
@@
 %
-\item Nodes that represent expressions, assignments, \emph{C} calls, and 
-declarations simply reproduce $O=I$, as they do not await;
+\item Nodes that represent expressions, assignments, \C calls, and declarations 
+simply reproduce $O=I$, as they do not await;
 %
@@
 %
-Translating this idea to \emph{C} is straightforward~\cite{wsn.osm,wsn.ocram}: 
-memory for blocks in sequence are packed in a \code{struct}, while blocks in 
-parallel, in a \code{union}.
+Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
+for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
+in a \code{union}.
 %In the example, the code following the loop (identified as \code{...}) reuses 
@@
 simultaneous lines of execution with an overhead of 2\% in comparison to the 
-original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \emph{C} for 
+original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \C for 
 event-driven programming)~\cite{ceu.sensys13}.
@@
 
-In the final generated code in \emph{C}, each trail label representing an entry 
-point becomes a \emph{switch case} with the associated code to execute.
+In the final generated code in \C, each trail label representing an entry point 
+becomes a \emph{switch case} with the associated code to execute.
 %
@@
 
-\subsection{The External \emph{C} API}
+\subsection{The External \C API}
 
@@
 From the implementation perspective, there are three external sources of input 
-into programs, which are all exposed as functions in a \emph{C} API:
+into programs, which are all exposed as functions in a \C API:
 
@@
 input event or the passage of time.
-In this case, the \emph{C} implementation makes a tail call to the 
-corresponding handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as 
-synchronous code has higher priority.
+In this case, the \C implementation makes a tail call to the corresponding 
+handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as synchronous 
+code has higher priority.
 
@@
 
+Terra is a system for programming wireless sensor network applications that 
+uses \CEU as its scripting language~\cite{ceu.terra}.
+%
+Figure \ref{fig.terra} shows the three basic elements of Terra:
+\CEU as the scripting language,
+the set of customized pre-built components,
+and the embedded virtual-machine engine which can disseminate and install 
+bytecode images dynamically.
+%
+This approach aims to combine the flexibility of remotely uploading code with 
+the expressiveness and safety guarantees of \CEU.
+
+\begin{figure}[!htb]
+ \centering
+ \includegraphics[width=0.7\columnwidth]{terra}
+\caption{Terra programming system basic elements.}
+ \label{fig.terra}
+\end{figure}
+
+\CEU scripts in Terra cannot execute arbitrary \C code, instead, they rely on 
+pre-built components that can be customized for different application domains.
+%
+Considering the domain of WSNs, Terra already provides components organized in 
+four areas: radio communication, group management, data aggregation, and local 
+operations (e.g., access to sensors and actuators).
+%
+When creating an instance of the \VM, the programmer can choose whether or not 
+to include each component, setting different abstraction boundaries.
+%
+The generated \VM has to be preloaded into the embedded devices before they are 
+physically distributed.
+
+The communication between \CEU and the Terra components is mostly through 
+events:
+scripts make requests through \code{output} events and \code{await} answers 
+through \code{input} events.
+%
+Terra also provides system calls for initialization and configuration of 
+components.
+
+TODO: exemplo de implementacao/uso de um componente
+- Terra: binding em nesC
+- Ceu: header de configuracao
+- (analogo a Figura 19)
+
+Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
+the \C backend, the main difference resides in the \emph{Code Generation}, 
+which now outputs assembly instructions for the Terra \VM, instead of \C.
+
+TODO: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
+- (analogo a Figura 18)
+
+\begin{comment}
+Instead of generating code in C, Terra generates the vmt assembly code using 
+the original Céu labels for trails entry points.
+In the final phase for code generation, Terra compiler generates the vmt 
+bytecode file using the program address in place of Céu labels.
+The use of address allows the Céu scheduler to execute directly a trail without 
+a index table to replace the switch-case.
+Related to memory layout and trail allocation, Terra uses the same data 
+organization approach for all Céu controls and variables.
+All Céu data and control structures for a specific program are determined at 
+compile time.
+Terra arranges all these structures in the beginning of virtual machine memory 
+space.
+The original Céu run time control functions are maintained with a small 
+modification where accesses to control structures have an off-set relative to 
+the begin of the virtual machine memory space.
+\end{comment}
+
 \section{Related Work}
@@
 constructive semantics of Esterel.
-Then, it flattens the graph into sequential code in \emph{C} suitable for 
-static scheduling.
+Then, it flattens the graph into sequential code in \C suitable for static 
+scheduling.
 
@@
 \bibliographystyle{ACM-Reference-Format-Journals}
-\bibliography{my,other}
+\bibliography{my,other,terra}
                              % Sample .bib file with references that match those in

### 2015-09-29 paper.tex
Revisao Terra

@@
  \centering
- \includegraphics[width=0.7\columnwidth]{terra}
+ \includegraphics[width=0.5\columnwidth]{terra}
 \caption{Terra programming system basic elements.}
@@
 
-\CEU scripts in Terra cannot execute arbitrary \C code, instead, they rely on 
-pre-built components that can be customized for different application domains.
+\CEU scripts cannot execute arbitrary \C code, instead, they rely on pre-built 
+components that can be customized for different application domains.
 %
@@
 When creating an instance of the \VM, the programmer can choose whether or not 
-to include each component, setting different abstraction boundaries.
+to include each component, setting different abstraction boundaries for 
+scripts.
 %
@@
 
-The communication between \CEU and the Terra components is mostly through 
-events:
-scripts make requests through \code{output} events and \code{await} answers 
-through \code{input} events.
+The communication between scripts in \CEU and the components in the \VM is 
+mostly through events:
+scripts \code{emit} requests through \code{output} events and \code{await} 
+answers through \code{input} events.
 %
@@
 Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
-the \C backend, the main difference resides in the \emph{Code Generation}, 
-which now outputs assembly instructions for the Terra \VM, instead of \C.
+the \C backend, the main difference resides in the \emph{Code Generation} 
+phase, which here outputs assembly instructions for the Terra \VM, instead of 
+\C.
 

### 2015-09-30 paper.tex
Novo titulo: The Design, Semantics, and Implementation of Ceu: ... Revisao abstract, Intro

@@
 % Title portion
-\title{The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}
+\title{The Design, Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}
 
@@
 \CEU is a reactive language based on Esterel that targets constrained embedded 
-platforms.
-%
-Employing the synchronous programming model, it allows for a simple reasoning 
-about shared-memory concurrency.
-Furthermore, its restricted semantics enables deterministic and memory-safe 
-programs.
+platforms and ensures safe concurrency by handling threats at compile time, 
+rather than at runtime.
+%
+Based on the synchronous programming model, our design allows for a simple 
+reasoning about concurrency that enables compile-time analysis resulting in 
+deterministic and memory-safe programs.
 %
@@
 %
-In this work, we propose a formal semantics for \CEU focusing on its particular 
-control mechanisms, such as parallel compositions, finalization, and 
-stack-based internal events.
+In this work, we discuss the design of \CEU and propose a formal semantics 
+focusing on its particular control mechanisms, such as parallel compositions, 
+finalization, and stack-based internal events.
 %
 We also present an implementation with two backends:
-one aiming for resource efficiency and interoperability with C, and another for 
-dynamic reprogramming over the air.
+one aiming for resource efficiency and interoperability with \C,
+and another based on a virtual machine that allows remote reprogramming.
 \end{abstract}
@@
 %
-Among the control-based languages, Esterel~\cite{esterel.ieee91} influenced a 
-number of other embedded languages, such as \emph{Reactive-C}~\cite{rp.rc}, 
-\emph{OSM}~\cite{wsn.osm}, \emph{Sync-C}~\cite{rp.synchc}, and 
-\emph{PRET-C}~\cite{rp.pretc}.
-
-\CEU is another language targeting embedded systems based on Esterel with some 
-novel functionalities:
+Considering the control-based languages, Esterel~\cite{esterel.ieee91} was the 
+first to appear and influenced a number of embedded languages, such as 
+\emph{Reactive-C}~\cite{rp.rc}, \emph{OSM}~\cite{wsn.osm}, 
+\emph{Sync-C}~\cite{rp.synchc}, and \emph{PRET-C}~\cite{rp.pretc}.
+
+\CEU is another Esterel-based language targeting embedded systems with novel 
+functionalities:
 %
@@
 %
-\item Stack-based internal events, which provide a limited form of coroutines.
+\item Stack-based execution for internal events, which provide a limited form 
+of coroutines.
 %
@@
 
-\CEU shares limitations with Esterel and synchronous languages in general:
-computations that run in unbounded time (e.g., cryptography, image processing) 
-do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
-elegantly implemented.
-
-In this work, we present a formal semantics of a small synchronous kernel that 
-represents a subset of \CEU covering these new functionalities.
+In this work, we discuss the design of \CEU and propose a formal semantics of a 
+small synchronous kernel that represents a subset of the language covering 
+these new functionalities.
 %
 We also present an implementation of \CEU with two backends:
-one aiming for efficiency and interoperability with \C,
+one aiming for resource efficiency and interoperability with \C,
 and another based on a virtual machine that allows remote reprogramming.
+%
+Our implementations target resource-constrained devices, such as \emph{Arduino} 
+and \emph{MICAz} sensor nodes, which use the \emph{ATmega328} 8-bit 
+microcontroller with 32 Kbytes of FLASH and 2 Kbytes of SRAM.%
+\footnote{
+Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
+\\
+MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
+\\
+ATmega328: \url{http://www.atmel.com/devices/atmega328.aspx}}
 
 In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
-of constrained wireless sensor networks, developing a number of applications, 
-protocols, and drivers.
+of wireless sensor networks, developing a number of applications, protocols, 
+and drivers.
 %
 We evaluated the expressiveness of \CEU in comparison to event-driven code in 
-\C, and observed a reduction in source code size (around 25\%) with a small 
-increase in memory usage (around 5--10\% for ROM and RAM)~\cite{ceu.sensys13}.
+\C and attested a reduction in source code size (around 25\%) with a small 
+increase in memory usage (around 5--10\% for \emph{text} and 
+\emph{data})~\cite{ceu.sensys13}.
 %
@@
 In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
-a \C-compiled version occupies more than 2 kbytes~\cite{ceu.terra}.
+a \C-compiled version occupies more than 2 Kbytes~\cite{ceu.terra}.
 
 The rest of the paper is organized as follows:
-Section~\ref{sec.ceu} gives an overview of \CEU, exposing its fundamental 
+Section~\ref{sec.ceu} discusses the design of \CEU, exposing its fundamental 
 differences to Esterel.
@@
 
-\section{Overview of \CEU}
+\section{The Design of \CEU}
 \label{sec.ceu}
@@
 all possible paths~\cite{ceu.sensys13,esterel.primer}.
+%
+\CEU shares the same limitations with Esterel and synchronous languages in 
+general:
+computations that run in unbounded time (e.g., cryptography, image processing) 
+do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
+elegantly implemented.
 
@@
 
-TODO: exemplo de implementacao/uso de um componente
+TODO-VM: exemplo de implementacao/uso de um componente
 - Terra: binding em nesC
@@
 
-TODO: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
+TODO-VM: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
 - (analogo a Figura 18)
@@
 
+TODO-VM
+
 \begin{comment}

### 2015-09-30 paper.tex
Conclusao - backend => back end - schizophrenic statements

@@
 % Title portion
-\title{The Design, Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}
+\title{The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}
 
@@
 %
-We also present an implementation with two backends:
+We also present an implementation with two back ends:
 one aiming for resource efficiency and interoperability with \C,
@@
 
-\CEU is another Esterel-based language targeting embedded systems with novel 
-functionalities:
+\CEU is another Esterel-based language targeting embedded and (soft) real-time 
+systems with novel functionalities:
 %
@@
 %
-\item A finalization mechanism for safe abortion of lines of execution.
+\item A finalization mechanism for safe abortion of lines of execution holding 
+external resources.
 %
@@
 
-In this work, we discuss the design of \CEU and propose a formal semantics of a 
-small synchronous kernel that represents a subset of the language covering 
+In this work, we discuss the design of \CEU and propose a formal semantics for 
+a small synchronous kernel that represents a subset of the language covering 
 these new functionalities.
 %
-We also present an implementation of \CEU with two backends:
+We also present an implementation of \CEU with two back ends:
 one aiming for resource efficiency and interoperability with \C,
@@
 Our implementations target resource-constrained devices, such as \emph{Arduino} 
-and \emph{MICAz} sensor nodes, which use the \emph{ATmega328} 8-bit 
-microcontroller with 32 Kbytes of FLASH and 2 Kbytes of SRAM.%
+and \emph{MICAz} sensor nodes based on 8-bit microcontrollers.%
 \footnote{
@@
 \\
-ATmega328: \url{http://www.atmel.com/devices/atmega328.aspx}}
+Both use the \emph{ATmega328} with 32 Kbytes of FLASH and 2 Kbytes of SRAM:
+\url{http://www.atmel.com/devices/atmega328.aspx}}
 
@@
 %
-Considering the \VM backend, a simple application that blinks three LEDs 
+Considering the \VM back end, a simple application that blinks three LEDs 
 periodically occupies less than 100 bytes and can be completely transmitted in 
 4 radio messages.
-In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
-a \C-compiled version occupies more than 2 Kbytes~\cite{ceu.terra}.
+In contrast, a \C-compiled version occupies more than 2 Kbytes (as it includes 
+all \emph{I/O} functionality preloaded with the \VM)~\cite{ceu.terra}.
 
@@
 \CEU.
-Section~\ref{sec.impl} presents the \C and \VM implementation backends.
+Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
 Section~\ref{sec.related} discusses other synchronous languages targeting 
@@
 
-In the sections that follow, we review the main differences between \CEU and 
+In the sections that follow, we discuss the main differences between \CEU and 
 Esterel~\cite{ceu.sensys13}:
@@
 translated almost identically in the two languages (lines 4--9, in both 
-implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
-equivalent.
+implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
+constructs are equivalent.
 %
@@
 Internal computations within a reaction (e.g. expressions, assignments, and 
-native calls) are considered to take no time in accordance with the synchronous 
+system calls) are considered to take no time in accordance with the synchronous 
 hypothesis~\cite{rp.hypothesis}.
@@
 In contrast, programs can emit internal events but these follow a stack-based 
-execution policy, similar to subroutine calls in typical programming 
-languages~\cite{ceu.sensys13}.
+execution policy, similar to subroutine calls in typical programming languages.
 %
@@
 %
-In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
-and they both become active, resulting in the invocation of \code{f()} and 
-\code{g()} in no particular order.
+In the version in Esterel, when \code{A} occurs, \code{B} is emitted (lines 
+5--6) and both events become active, resulting in the invocation of \code{f()} 
+and \code{g()} in no particular order.
 %
@@
 \begin{minipage}[t]{0.43\linewidth}
-\begin{lstlisting}[mathescape=true]
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
 // ESTEREL
@@
 re-execution of statements in the same reaction inside loops (known as 
-\emph{schizophrenic statements})~\cite{esterel.schizo}.
+\emph{schizophrenic statements}~\cite{esterel.schizo,esterel.schizo2}).
 %
@@
     with
-        emit e;     // awakes 1s trail and restarts the loop
+        emit e;     // without the restriction, it awakes 1s trail and restarts the loop
         await f;
@@
 Even though both sides of the \code{par/or} have an \code{await} statement to 
-avoid instantaneous termination, the \code{loop} would restart instantaneously 
-if the \code{await} could awake in the same reaction that reaches it.
+avoid instantaneous termination, if the \code{await} could awake in the same 
+reaction that reaches it, the \code{loop} would restart instantaneously 
+resulting in infinite execution.
 
 In atypical scenarios requiring immediate awake, delayed awaits can be 
-circumvented by manually copying or making transformations in the code to 
-execute on awake.
+circumvented by manually copying or transforming the code to execute on awake.
 %
@@
 reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
-In this case, moving the \code{await} to the end of the loop makes the periodic 
-code to also execute immediately (and then in reactions to each \code{emit} 
-request).
-If the periodic \code{emit} depends on a condition, the code becomes more 
-intricate because we need to copy the condition test to the periodic code, as 
-illustrated in the right of Figure~\ref{lst.delay}.
+In this case, moving the \code{await} to the end of the loop (line 10) makes 
+the periodic code to also execute immediately (line 9), and then in reactions 
+to each \code{emit} request (line 5).
+If the periodic \code{emit} depends on a condition, as illustrated in the right 
+of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
+state variable (line 2) and to copy the condition test to the periodic code 
+(line 13).
+%
+On the one hand, we transfer the burden of dealing with these corner cases to 
+the programmer.
+On the other hand, we simplify the semantics of the language and eliminate the 
+need for analysis to deal with shizophrenic statements.
 
@@
 \begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
 event void e;
@@
 \CEU as the scripting language,
-the set of customized pre-built components,
+a set of customized pre-built components,
 and the embedded virtual-machine engine which can disseminate and install 
@@
 
-\CEU scripts cannot execute arbitrary \C code, instead, they rely on pre-built 
-components that can be customized for different application domains.
-%
-Considering the domain of WSNs, Terra already provides components organized in 
-four areas: radio communication, group management, data aggregation, and local 
-operations (e.g., access to sensors and actuators).
+In Terra, \CEU scripts cannot execute arbitrary \C code, instead, they rely on 
+pre-built components that can be customized for different application domains.
+%
+Considering the domain of sensor networks, Terra already provides components 
+organized in four areas: radio communication, group management, data 
+aggregation, and local operations (e.g., access to sensors and actuators).
 %
@@
 Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
-the \C backend, the main difference resides in the \emph{Code Generation} 
+the \C back end, the main difference resides in the \emph{Code Generation} 
 phase, which here outputs assembly instructions for the Terra \VM, instead of 
@@
 \CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
-disciplined synchronous-reactive model with support for lexical composition of 
+disciplined synchronous-reactive model and support for lexical composition of 
 lines of execution.
@@
 \CEU also distinguishes external events from stack-based internal events, which 
-provides a limited form of coroutines that supports reactive statements (e.g., 
+provide a limited form of coroutines supporting reactive statements (e.g., 
 \code{await} and \code{par/or}).
 
-The event-driven approach~\cite{sync_async.whynotthreads} is well known and 
-popular in many software communities,
-such as web frameworks in the client side (e.g., \emph{jQuery}~\cite{js.jquery} 
-and \emph{Rx.js}~\cite{js.rx}) and server side (e.g., 
-\emph{Node.js}~\cite{js.node} and \emph{SEDA}~\cite{sync_async.seda}),
-GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java 
-Swing}~\cite{java.swing}),
-and Games~\cite{gamepatterns}.
+The event-driven approach of \CEU is well known~\cite{sync_async.whynotthreads} 
+and popular in many software communities, such as client and server-side web 
+frameworks (e.g., \emph{jQuery}~\cite{js.jquery} and 
+\emph{Node.js}~\cite{js.node}), GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} 
+and \emph{Java Swing}~\cite{java.swing}), and Games~\cite{gamepatterns}.
 %
@@
 
-Esterel has different compilation backends that synthesizes to software and 
+Esterel has different compilation back ends that synthesizes to software and 
 also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
@@
 scheduling:
-The compiler slices programs into ``control points'' (analogous to our ``entry 
+the compiler slices programs into ``control points'' (analogous to our ``entry 
 points'') and rearranges them into a directed acyclic graph respecting the 
@@
 
-\subsection{Virtual Machines}
+TODO-VM
 
@@
 
-\section{Final Remarks}
+\section{Conclusion}
 \label{sec.conclusion}
+
+We presented the design, semantics, and implementation of \CEU, a synchronous 
+reactive language based on Esterel targeting constrained embedded systems.
+
+\CEU is a concurrency-safe language, employing a static analysis that encompass 
+all control constructs and ensures that the high degree of concurrency in 
+embedded systems does not pose safety threats to applications.
+%
+As a summary, the following safety properties hold for all programs that 
+successfully compile in \CEU:
+time and memory-bounded reactions to the environment (except for external 
+system calls),
+no race conditions in shared memory,
+reliable abortion for activities handling resources,
+and automatic synchronization for timers.
+%
+These properties are usually desirable in applications and are guaranteed as 
+preconditions in \CEU by design.
+
+\CEU is a resource-efficient language suitable for constrained embedded 
+systems.
+The reference implementation compiles to portable event-driven code in \C, with 
+no special requirements for OS threads or per-trail data stacks.
+The \VM implementation uses the same front end and imposes no extra 
+restrictions, being equally suitable for constrained systems.
+
+\CEU is a practical language with expressive control constructs, such as 
+lexically scoped parallel compositions, convenient first-class timers, and a 
+unique stack-based signaling mechanism.
+%
+Programs interoperate seamlessly with \C, and can take advantage of existing 
+libraries, lowering the entry barrier for adoption.
+%
+\CEU has an open source implementation and bindings for \emph{TinyOS}, 
+\emph{Arduino}, and the \emph{SDL} graphical library.%
+\footnote{Website of \CEU: \url{http://www.ceu-lang.org/}}
+
+For the past three years, we have been teaching \CEU for undergraduate and 
+graduate students in research projects and two hands-on courses on 
+\emph{distributed systems} and \emph{reactive programming}.
+%, in which the studnets designing protocols, embedded systems, and graphical 
+%applications.
+%
+Our experience shows that students take advantage of the sequential-imperative 
+style of \CEU and can implement non-trivial concurrent applications in a few of 
+weeks.
+
+% Idexx
+
+\begin{comment}
+We are also extending and applying \CEU in more dynamic domains, such as GUIs 
+and games~\cite{ceu.mod15}, by relaxing some restrictions in the language.
+In particular, programs in these domains typically require unbounded memory and 
+execution time, such as for dealing with dynamic collections.
+We provide unsafe annotations to circumvent the rigid semantics of \CEU, 
+requiring them to be explicit and trackable at the same time.
+\end{comment}
 

### 2015-10-04 paper.tex
Pequenos ajustes + Figuras do Adriano

@@
 A program with blocks in sequence and in parallel, with corresponding memory 
-layout.
+layout that the compiler generates.
 {\small %\textmd{
@@
 %{\small %\textmd{
-Static allocation of trails, entry-point labels, and dispatch loop.
+From left to right: static allocation of trails, entry-point labels, and 
+dispatch function.
+In the left, the comments identify the trail indexes inferred by the compiler.
+In the middle, each trail segment has an associated numeric identifier 
+generated by the compiler.
+In the right, the dispatcher uses a \code{switch} to associate each segment 
+identifier with the corresponding code to execute.
 %}%}
@@
     event void Timer.fired () {
-        int dt = 10000;
-        ceu_sys_go(EVT__WCLOCK, &dt);
+        ceu_sys_wclock(10000);
     }
@@
 
-#ifdef _EVT_PHOTO_READDONE
+#ifdef EVT_PHOTO_READDONE
     event void Photo.readDone (uint16_t val) {
@@
 
-#ifdef _EVT_RADIO_SENDDONE
+#ifdef EVT_RADIO_SENDDONE
     event void RadioSend.sendDone (message_t* msg) {
@@
 
-#ifdef _EVT_RADIO_RECEIVE
+#ifdef EVT_RADIO_RECEIVE
     event message_t* RadioReceive.receive (message_t* msg) {
@@
 If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
-that triggers callback \code{Timer.fired} (lines 13--18) every 10 milliseconds 
-and advances the wall-clock time of \CEU (line 16)%
+that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
+and advances the wall-clock time of \CEU (line 15)%
 \footnote{We also offer a mechanism to start the underlying timer on demand to
@@
 The remaining lines map pre-defined TinyOS events that can be used in \CEU 
-programs, such as the light sensor (lines 20--24) and the radio transceiver 
-(lines 26--37).
+programs, such as the light sensor (lines 19--23) and the radio transceiver 
+(lines 25--36).
 
@@
 components.
+%
+TODO: explicar Figura 21.
+\begin{comment}
+Figure~\ref{lst.terra.defs} illustrates a program interface declaration in \CEU
+(in the left), and associated binding code in \C
+
+ with respective binding customization scenario.
+This customization sample uses a small data structure, two output events, one input event, and two functions.
+The left listing is the definition file to be included into the \CEU script program and the right listing is the
+definition file to be included into the VM custom module implementation.
+\end{comment}
+
+\begin{figure}
+\begin{minipage}[t]{0.46\linewidth}
+\begin{lstlisting}[morekeywords={ubyte,ushort,output,function,regtype}]
+// Message definition
+regtype sensorMsg with
+    var ubyte  id;
+    var ushort value;
+end
+
+// Output events definition
+output void REQ_TEMP void          1;
+output void SEND_SENSOR sensorMsg  2;
+
+// Input events definition
+input void TEMP void               1;
+
+// functions definitions
+function ubyte getNodeId()         1;
+function ubyte queuePut(sensorMsg) 2;
+.
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.51\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
+morekeywords={nx_struct,nx_uint8_t,nx_uint16_t,uint8_t,uint32_t}]
+
+typedef nx_struct sensorMsg{
+  nx_uint8_t  id;
+  nx_uint16_t value;
+} sensorMsg_t;
+
+enum {
+  O_REQ_TEMP    = 1;
+  O_SEND_SENSOR = 2;
+
+  I_TEMP        = 1;
+
+  F_GETNODEID   = 1;
+  F_QUEUEPUT    = 2;
+};
+
+\end{lstlisting}
+\end{minipage}
+\caption{
+%{\small %\textmd{
+Terra control block sample and the equivalent definition for VM implementation.
+%}%}
+\label{lst.terra.defs}
+}
+\end{figure}
+
+TODO: dizer o que eh manual/automatico nos lsts de ceu
 
@@
 - (analogo a Figura 18)
+
+\begin{comment}
+%-------------------------------------------------
+Figure~\ref{lst.impl.custom} presents a partial implementation for the elements 
+introduced in the Figure~\ref{lst.impl.defs}.
+The two first functions are called from the VM decoder, the fist function \code{procOutevt()} dispatch any defined output events
+and the function \code{callFunction()} dispatch any defined customized functions.
+Lines 17--22 shows an example of a custom function that returns a value via stack.
+Lines 24--27 has a example of an external event call.
+Lines 29--34 presents an example how a input event is queued to the VM engine.
+\end{comment}
+
+%-----------------------------------------------------
+% Equivalent to the fig. 19
+
+\begin{figure}
+\begin{minipage}[t]{0.81\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
+morekeywords={uint8_t,uint16_t,uint32_t,error_t,command}]
+// Output event dispatcher
+command void VM.procOutEvt(uint8_t id, uint32_t value){
+  switch (id){
+    case O_REQ_TEMP    : proc_req_temp(id,value);    break;
+    case O_SEND_SENSOR : proc_send_sensor(id,value); break;
+  }
+}
+
+// Function dispatcher
+command void VM.callFunction(uint8_t id){
+  switch (id){
+    case F_GETNODEID : func_getNodeId(id); break;
+    case F_QUEUEPUT  : func_queuePut(id);  break;
+  }
+}
+
+// Pushing a value to the stack
+void  func_getNodeId(uint16_t id){
+  uint16_t stat;
+  stat = TOS_NODE_ID;
+  signal VM.push(stat);
+}
+
+// Calling a output event
+void  proc_req_temp(uint16_t id, uint32_t value){
+  call S_TEMP.read();
+}
+
+// Queueing an input event + value
+uint16_t lastTemp;
+event void S_TEMP.readDone(error_t result, uint16_t val)
+  lastTemp = val;
+  signal VM.queueEvt(I_TEMP, 0, &lastTemp);
+}
+
+\end{lstlisting}
+\end{minipage}
+
+\caption{
+%{\small %\textmd{
+VM Customization -- input/output events and functions.
+%}%}
+\label{lst.impl.custom}
+}
+\end{figure}
+
+%------------------------------------------------------
 

### 2015-10-05 paper.tex
Pequenos ajustes.

@@
 Based on the synchronous programming model, our design allows for a simple 
-reasoning about concurrency that enables compile-time analysis resulting in 
+reasoning about concurrency that enables compile-time analysis and results in 
 deterministic and memory-safe programs.
@@
 periodically occupies less than 100 bytes and can be completely transmitted in 
-4 radio messages.
-In contrast, a \C-compiled version occupies more than 2 Kbytes (as it includes 
-all \emph{I/O} functionality preloaded with the \VM)~\cite{ceu.terra}.
+4 radio messages, given that \emph{I/O} components are preloaded in the remote
+devices.
+In contrast, a \C-compiled version occupies more than 2 Kbytes, which includes 
+all \emph{I/O} functionality~\cite{ceu.terra}.
 
@@
 In the sections that follow, we discuss the main differences between \CEU and 
-Esterel~\cite{ceu.sensys13}:
+Esterel:
 queue-based external events and stack-based internal events 
@@
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
-restarts the \code{loop}.
+restarts the enclosing \code{loop}.
 
@@
 computations that run in unbounded time (e.g., cryptography, image processing) 
-do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
-elegantly implemented.
+do not fit the zero-delay hypothesis, and cannot be elegantly implemented.
 
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Internal signals (events) in Esterel and \CEU. \newline
+\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
+different semantics.
 \label{lst.prints}
@@
 %
-\CEU imposes a restriction that an \code{await} to an internal event is only 
-valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
-simultaneously in parallel trails, the \code{await} does not awake.
+\CEU imposes a restriction that an \code{await} is only valid for the next 
+reaction, i.e., if an \code{await} and \code{emit} occur simultaneously in 
+parallel trails, the \code{await} does not awake.
 %
@@
     with
-        emit e;     // without the restriction, it awakes 1s trail and restarts the loop
+        emit e; // w/o the restriction, awakes 1st trail and restarts the loop instantaneously
         await f;
@@
 of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
-state variable (line 2) and to copy the condition test to the periodic code 
-(line 13).
+state variable (line 2) and also to copy the condition test to the periodic 
+code (line 13).
 %
@@
 On the other hand, we simplify the semantics of the language and eliminate the 
-need for analysis to deal with shizophrenic statements.
+need for analysis to deal with schizophrenic statements.
 
@@
 In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
-\code{x} can never run concurrently, because each trail segment reacts to a 
-different input event and, according to the semantics of \CEU, cannot occur 
-simultaneously.
+\code{x} (lines 5,8) can never run concurrently, because each trail segment 
+reacts to a different input event (lines 4,7) and, according to the semantics 
+of \CEU, cannot occur simultaneously.
 %
 However, the program in the right is non-deterministic, because the two 
-assignments to \code{y} occur in the same reaction to input \code{A}.
+assignments to \code{y} (lines 5,8) occur in the same reaction to input 
+\code{A} (lines 4,7).
 
@@
 \begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
 input void A, B;
@@
 \begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
 input void A;
@@
 \CEU performs a temporal analysis at compile time and detects concurrent 
-accesses to shared variables, as follows~\cite{ceu.sensys13}:
+accesses to shared variables, as follows:
 \emph{if a variable is written in a trail segment, then a concurrent trail 
@@
 shared-memory concurrency is beneficial, given the extensive use of memory 
-mapped ports for I/O and the lack of hardware support for real parallelism.
+mapped ports for \emph{I/O} and the lack of hardware support for real 
+parallelism.
 Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
-\emph{PRET-C}\cite{rp.pretc}, made a similar design choice.
+\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
 %For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
@@
 
-To illustrate threats related to abortion, consider the unsafe example, which 
-does not compile, in the left of Figure~\ref{lst.fin}, describing the state 
-machine of a data collection protocol for sensor 
+To illustrate threats related to abortion, consider the unsafe example in the 
+left of Figure~\ref{lst.fin}, which does not compile.
+It describes the state machine of a data collection protocol for sensor 
 networks~\cite{wsn.ctp,ceu.sensys13}.
@@
 one handles the immediate retransmission request (line 7);
-the other transmits the packet%
+the other transmits the packet and waits for a confirmation (lines 9--16).
+%
+%
+The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
+takes at least one minute before looping, in accordance with the specification.
+Note that the call to \code{\_send\_enqueue}%
 \footnote{
@@
 }
-and waits for a confirmation (lines 9--16).
-%
-%
-The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
-takes at least one minute before looping, in accordance with the specification.
-Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
-the radio driver a pointer to the lexically-scoped packet.
+(line 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
+lexically-scoped packet (line 12).
 The driver makes the transmission in the background, holding the packet until 
-it signals the application with the \code{SENDACK} to acknowledge completion.
+it signals the application with the \code{SENDACK} to acknowledge completion 
+(line 15).
 %
@@
 and stateful \C functions (e.g., device drivers) in order to preserve safe 
-abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
+abortion of trails.
 %
@@
 The \code{await} statement of \CEU supports wall-clock time and handles deltas 
-automatically~\cite{ceu.sensys13}, resulting in more robust applications.
+automatically, resulting in more robust applications.
 For the example in the left of Figure~\ref{lst.timers}, suppose that after the 

### 2015-10-05 paper.tex
lines => ln.

@@
 The first phrase of the specification, awaiting and emitting the events, is 
-translated almost identically in the two languages (lines 4--9, in both 
+translated almost identically in the two languages (ln. 4--9, in both 
 implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
@@
 For the second phrase, the reset behavior, the Esterel version uses a 
-\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
-\code{par/or} (lines 3--12):
+\code{abort-when} (ln. 3--10), which serves the same purpose of \CEU's 
+\code{par/or} (ln. 3--12):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
@@
 %
-In the version in Esterel, when \code{A} occurs, \code{B} is emitted (lines 
+In the version in Esterel, when \code{A} occurs, \code{B} is emitted (ln.  
 5--6) and both events become active, resulting in the invocation of \code{f()} 
@@
 \setlength{\itemsep}{0pt}
-\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
+\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.\\
     \emph{stack: [1st-trail]}
-\item 2nd trail awakes (line 9), calls \code{\_g()}, and terminates.\\
+\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.\\
     \emph{stack: [1st-trail]}
@@
 Figure~\ref{lst.sub}.
-The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
-awaits its identifying event (line 4), incrementing the value passed as 
-reference (line 5).
-A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
-\code{A} through an \code{emit} (line 10).
+The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
+awaits its identifying event (ln. 4), incrementing the value passed as 
+reference (ln. 5).
+A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
+\code{A} through an \code{emit} (ln. 10).
 Given the stacked execution for internal events, the calling trail pauses, the 
-subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
-awaits the next ``call'' (line 4, again).
+subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
+awaits the next ``call'' (ln. 4, again).
 Only after this sequence that the calling trail resumes and passes the 
-assertion test (line 11).
+assertion test (ln. 11).
 
@@
 \end{lstlisting}
-\caption{ Subroutine \code{inc} is defined in a loop (lines 3--6), in parallel 
-with the caller (lines 8--11).
+\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
+with the caller (ln. 8--11).
 \label{lst.sub}
@@
 reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
-In this case, moving the \code{await} to the end of the loop (line 10) makes 
-the periodic code to also execute immediately (line 9), and then in reactions 
-to each \code{emit} request (line 5).
+In this case, moving the \code{await} to the end of the loop (ln. 10) makes the 
+periodic code to also execute immediately (ln. 9), and then in reactions to 
+each \code{emit} request (ln. 5).
 If the periodic \code{emit} depends on a condition, as illustrated in the right 
 of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
-state variable (line 2) and also to copy the condition test to the periodic 
-code (line 13).
+state variable (ln. 2) and also to copy the condition test to the periodic code 
+(ln. 13).
 %
@@
 In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
-\code{x} (lines 5,8) can never run concurrently, because each trail segment 
-reacts to a different input event (lines 4,7) and, according to the semantics 
-of \CEU, cannot occur simultaneously.
+\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
+reacts to a different input event (ln. 4,7) and, according to the semantics of 
+\CEU, cannot occur simultaneously.
 %
 However, the program in the right is non-deterministic, because the two 
-assignments to \code{y} (lines 5,8) occur in the same reaction to input 
-\code{A} (lines 4,7).
+assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
+(ln. 4,7).
 
@@
 %
-The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
+The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (ln. 1) 
 represent the external interface of the protocol.
@@
 The protocol is composed of two main trails:
-one simply monitors the stopping event (line 3);
-the other periodically transmits the packet (lines 5--18).
-%
-The periodic transmission is a loop that starts two other trails (lines 6--17):
-one handles the immediate retransmission request (line 7);
-the other transmits the packet and waits for a confirmation (lines 9--16).
-%
-%
-The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
+one simply monitors the stopping event (ln. 3);
+the other periodically transmits the packet (ln. 5--18).
+%
+The periodic transmission is a loop that starts two other trails (ln. 6--17):
+one handles the immediate retransmission request (ln. 7);
+the other transmits the packet and waits for a confirmation (ln. 9--16).
+%
+%
+The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
 takes at least one minute before looping, in accordance with the specification.
@@
 }
-(line 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
-lexically-scoped packet (line 12).
+(ln. 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
+lexically-scoped packet (ln. 12).
 The driver makes the transmission in the background, holding the packet until 
 it signals the application with the \code{SENDACK} to acknowledge completion 
-(line 15).
-%
-At any time, the client may request a retransmission (line 7), which terminates 
-the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
-idle), and restarts the loop (line 5).
-%
-The client may also request to stop the whole protocol at any time (line 3).
+(ln. 15).
+%
+At any time, the client may request a retransmission (ln. 7), which terminates 
+the \code{par/or} (ln. 6), aborts the ongoing transmission (ln. 14, if not 
+idle), and restarts the loop (ln. 5).
+%
+The client may also request to stop the whole protocol at any time (ln. 3).
 %
@@
 The event $.$ (dot) represents the ``boot'' reaction.
-The assignments to \code{y} in parallel (lines 5,8 in the code) have an empty 
-intersection of $I$ (lines 6,9 in the AST), hence, they do not conflict.
-Note that although the accesses in lines 5,11 in the code (lines 6,11 in the 
-AST) do have an intersection, they are not in parallel and are also safe.
+The assignments to \code{y} in parallel (ln. 5,8 in the code) have an empty 
+intersection of $I$ (ln. 6,9 in the AST), hence, they do not conflict.
+Note that although the accesses in ln. 5,11 in the code (ln. 6,11 in the AST) 
+do have an intersection, they are not in parallel and are also safe.
 
@@
 points and associated trails, e.g., the label \code{Awake\_e} will execute on 
-\code{TRAIL-0} (line 7).
+\code{TRAIL-0} (ln. 7).
 %
@@
 %
-The entry points translate to an \code{enum} in the generated code (lines 
-1--10, in the middle of the figure).
+The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
+in the middle of the figure).
 The state of trails translate to a vector of type \code{trail\_t} with the 
-maximum number of simultaneous trails (lines 12--15).
+maximum number of simultaneous trails (ln. 12--15).
 On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
-(line 13), while all others are set to \code{INACTIVE} (in the example, only 
-one, in line 14).
+(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
+one, in ln. 14).
 
@@
 pass, given that the \code{Main} label is the only one that can be active at 
-the stack level 0 (line 13, in the middle of Figure~\ref{lst.impl.trails}).
+the stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).
 
@@
 When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
-(lines 4--7) and proceeds to the code in \code{TRAIL-0} (lines 9--14), 
-respecting the deterministic execution order.
+(ln. 4--7) and proceeds to the code in \code{TRAIL-0} (ln. 9--14), respecting 
+the deterministic execution order.
 %
@@
 The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
-registered label \code{And\_sub\_2} (lines 16--21), which sets \code{TRAIL-1} 
-to await \code{EVT\_A} and also halts.
+registered label \code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to 
+await \code{EVT\_A} and also halts.
 
@@
 %
-The file \code{ceu.h} (included in line 3) contains all definitions for the 
-compiled \CEU program, which are further queried through \code{\#ifdef}'s.
-The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
-pointing to the labels defined in the program.
-The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
-so we initialize \CEU inside it (line 7).
-If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
-that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
-and advances the wall-clock time of \CEU (line 15)%
+The file \code{ceu.h} (ln. 3) contains all definitions for the compiled \CEU 
+program, which are further queried through \code{\#ifdef}'s.
+The file \code{ceu.c} (ln. 4) contains the main loop of \CEU pointing to the 
+labels defined in the program.
+The callback \code{Boot.booted} (ln. 6--11) is called by TinyOS on startup, so 
+we initialize \CEU inside it (ln. 7).
+If the \CEU program uses timers, we also start a periodic timer (ln. 8--10) 
+that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds 
+and advances the wall-clock time of \CEU (ln. 15)%
 \footnote{We also offer a mechanism to start the underlying timer on demand to
@@
 The remaining lines map pre-defined TinyOS events that can be used in \CEU 
-programs, such as the light sensor (lines 19--23) and the radio transceiver 
-(lines 25--36).
+programs, such as the light sensor (ln. 19--23) and the radio transceiver (ln.  
+25--36).
 

### 2015-10-07 paper.tex
Reorganizacao das figuras.
Revisao ate 3.3

@@
 \\
-Both use the \emph{ATmega328} with 32 Kbytes of FLASH and 2 Kbytes of SRAM:
-\url{http://www.atmel.com/devices/atmega328.aspx}}
+(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
+Kbytes of SRAM.)}
+%:\url{http://www.atmel.com/devices/atmega328.aspx}}
 
@@
 %
-Considering the \VM back end, a simple application that blinks three LEDs 
-periodically occupies less than 100 bytes and can be completely transmitted in 
-4 radio messages, given that \emph{I/O} components are preloaded in the remote
-devices.
-In contrast, a \C-compiled version occupies more than 2 Kbytes, which includes 
-all \emph{I/O} functionality~\cite{ceu.terra}.
+For the \VM back end, a simple application that blinks three LEDs periodically 
+occupies less than 100 bytes and can be completely transmitted in 4 radio 
+messages.
+In contrast, the full image of a \C-compiled version occupies more than 2 
+Kbytes~\cite{ceu.terra}.%
+\footnote{
+The \VM is preloaded in the devices with some \emph{I/O} functionality, whereas 
+\C-based binaries are full images.
+}
 
 The rest of the paper is organized as follows:
-Section~\ref{sec.ceu} discusses the design of \CEU, exposing its fundamental 
+Section~\ref{sec.ceu} discusses the design of \CEU, focusing on the fundamental 
 differences to Esterel.
@@
 
-\CEU is a synchronous reactive language based on Esterel~\cite{esterel.ieee91} 
-with support for multiple concurrent lines of execution known as \emph{trails}.
+\CEU is a synchronous reactive language based on Esterel with support for 
+multiple concurrent lines of execution known as \emph{trails}.
 By reactive, we mean that programs are stimulated by the environment through 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \begin{minipage}[t]{0.48\linewidth}
@@
 
-\CEU, following the Esterel mindset, has a strong imperative flavor, with 
-explicit control flow through sequences, loops, parallels, and also 
-assignments.
+\CEU has a strong imperative flavor, with explicit control flow through 
+sequences, loops, parallels, and also assignments.
 Being designed for control-intensive applications, it provides support for 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \begin{minipage}[t]{0.43\linewidth}
@@
 awaits the next ``call'' (ln. 4, again).
-Only after this sequence that the calling trail resumes and passes the 
-assertion test (ln. 11).
+Only after this sequence the calling trail resumes and passes the assertion 
+test (ln. 11).
 
@@
 These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
-re-execution of statements in the same reaction inside loops (known as 
-\emph{schizophrenic statements}~\cite{esterel.schizo,esterel.schizo2}).
+re-execution of statements in the same reaction (known as \emph{schizophrenic 
+statements}~\cite{esterel.schizo,esterel.schizo2}).
 %
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=2em]
 event void e,f;
@@
     with
-        emit e; // w/o the restriction, awakes 1st trail and restarts the loop instantaneously
-        await f;
-    end
-end
-\end{lstlisting}
-
-Even though both sides of the \code{par/or} have an \code{await} statement to 
-avoid instantaneous termination, if the \code{await} could awake in the same 
-reaction that reaches it, the \code{loop} would restart instantaneously 
-resulting in infinite execution.
+        emit e;     // w/o the restriction, the emit awakes 1st trail
+        await f;    // and restarts the loop instantaneously
+    end
+end
+\end{lstlisting}
+
+Both sides of the \code{par/or} have an \code{await} statement to avoid 
+instantaneous termination (ln. 4,7).
+However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
+same reaction that reaches them, the \code{par/or} would terminate and restart 
+the \code{loop} instantaneously, resulting in infinite execution.
 
@@
     loop do
-        <...>
+        <...>     // code that awaits
         emit e;   // periodic request
@@
     loop do
-        <...>
-        if <...> then
+        <...>         // code that awaits
+        if <...> then // some conditon
             should_execute = true;
@@
 
-Embedded applications make extensive use of shared memory, such as for sharing 
-resources through memory-mapped registers.
+Embedded applications make extensive use of global memory, such as for 
+accessing resources through memory-mapped registers.
 Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
@@
 Concurrency in \CEU is characterized when two or more trail segments in 
-parallel execute during the same reaction chain.
+parallel execute during the same reaction.
 A trail segment is a sequence of statements followed by an \code{await} (or 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \begin{minipage}[t]{0.50\linewidth}
@@
 To illustrate threats related to abortion, consider the unsafe example in the 
-left of Figure~\ref{lst.fin}, which does not compile.
+left of Figure~\ref{lst.fin}, which does not compile in \CEU.
 It describes the state machine of a data collection protocol for sensor 
@@
 The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
-takes at least one minute before looping, in accordance with the specification.
+takes at least one minute before looping (ln. 10), in accordance with the 
+specification.
 Note that the call to \code{\_send\_enqueue}%
@@
 the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
-it can at least ensure that the program always terminates with \code{v=1}:
+it can at least ensure that the program always terminates with \code{v=1}.
 %
@@
 We describe a small synchronous kernel with broadcast communication
-highlighting the main differences to Esterel, in particular the stack-based
+highlighting the peculiarities of \CEU, in particular the stack-based
 execution for internal events.
@@
 $emitting$, and $p~@~loop; p$), which are discussed further, all other 
-expressions map to their counterparts in the concrete language.
+expressions are equivalent to their counterparts in the concrete language.
 
@@
 An $await$ is simply transformed into an $awaiting$ (rule \textbf{await}) as an 
-artifact to remember the external sequence number $n+1$ it can awake:
+artifice to remember the external sequence number $n+1$ it can awake:
 an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
 event $id$ matches the top of the stack and it was reached in a previous 
-reaction (i.e., sequence number $m<n$).
+reaction (i.e., sequence number $m<=n$).
 %
@@
     {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
-    & \textbf{(or-adv2)}   \\
+    & \textbf{(or-adv2)}   %\\
 \end{eqnarray*}
@@
 \begin{align*}
-  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m = n)   \\
+  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m > n)   \\
   isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
@@
 
+%%%
+
 For a parallel $and$, if one of the sides terminates, the composition is simply 
-substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}).
-The last two rules \textbf{and-brk1} and \textbf{and-brk2} deal with a $break$ 
-in each of the sides in parallel.
-To escape the innermost loop, a $break$ aborts the other side by applying the 
-$clear$ function (to be described further):
+substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
+as follows).
+%
+For a parallel $or$, if one of the sides terminates, the whole composition 
+terminates, also applying the $clear$ function to properly finalize the aborted 
+side (rules \textbf{or-nop1} and \textbf{or-nop2}):
 %
@@
 %%%
-& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
-    & \textbf{(and-brk1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
-    & \textbf{(and-brk2)}   \\
-\end{eqnarray*}
-}
-%
-For a parallel $or$, if one of the sides terminates, the whole composition 
-terminates, also applying the $clear$ function to the aborted side (rules 
-\textbf{or-nop1} and \textbf{or-nop2}).
-Finally, a $break$ (rules \textbf{or-brk1} and \textbf{or-brk2}) behaves in the 
-same way as in a parallel $and$:
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
 & \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
@@
     {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
-    & \textbf{(or-nop2)}   \\
-%%%
-& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
-    & \textbf{(or-brk1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
-    & \textbf{(or-brk2)}   %\\
+    & \textbf{(or-nop2)}   %\\
 \end{eqnarray*}
@@
 
-\begin{figure}
+\begin{figure}[b]
 {\small
@@
 
+Finally, a $break$ in one of the sides in parallel escapes the closest 
+enclosing $loop$, properly aborting the other side by applying the $clear$ 
+function:
+%
+{ \setlength{\jot}{20pt}
+\begin{eqnarray*}
+& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
+    & \textbf{(and-brk1)}   \\
+%%%
+& \frac
+    {\DS isBlocked(n,S,p) }
+%   -----------------------------------------------------------
+    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
+    & \textbf{(and-brk2)}   \\
+%%%
+& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
+    & \textbf{(or-brk1)}   \\
+%%%
+& \frac
+    {\DS isBlocked(n,S,p) }
+%   -----------------------------------------------------------
+    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
+    & \textbf{(or-brk2)}   %\\
+\end{eqnarray*}
+}
+%
 A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
@@
 However, there are some significant mismatches between the concrete and 
-abstract \CEU, and we discuss appropriate mappings in this section.
+abstract \CEU, and we (informally) propose appropriate mappings in this 
+section.
 %
@@
 expressions.
+The concrete code in the left communicates the value \code{1} from the 
+\code{emit} to the \code{await}.
+The abstract code uses a shared variable to hold the value.
 \label{lst.map.emit.await}
@@
 
-To encompass first-class timers, we introduce a special external \code{TICK} 
+To encompass first-class timers, we introduce a special external \code{DT} 
 event that is intercalated with each other event occurrence in an application 
@@
 \begin{align*}
-\LL [TICK], p \RR
+\LL [DT], p \RR
     & \xRightarrow[~~1~~]{*}
@@
 \\
-\LL [TICK], p'' \RR
+\LL [DT], p'' \RR
     & \xRightarrow[~~3~~]{*}
@@
 
-The \code{TICK} event has an associated variable \code{TICK\_} carrying the 
+The \code{DT} event has an associated variable \code{DT\_} carrying the 
 wall-clock time elapsed between the two occurrences, as depicted by the 
@@
 \code{loop} that decrements the elapsed number of microseconds for each 
-occurrence of \code{TICK}.
+occurrence of \code{DT}.
 When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \begin{minipage}[t]{0.30\linewidth}
@@
 loop do
-    await TICK;
-    tot = tot - TICK_;
+    await DT;
+    tot = tot - DT_;
     if tot <= 0 then
@@
 loop(
-    await(TICK);
+    await(DT);
     mem;
@@
 
-\begin{figure}
+\begin{figure}[b]
 \centering
@@
 
-\begin{figure}
+In the final generated code in \C, each trail label representing an entry point 
+becomes a \emph{switch case} with the associated code to execute.
+%
+Figure~\ref{lst.impl.trails} illustrates the generation process.
+For the program in the left of the figure, the compiler extracts the entry 
+points and associated trails, e.g., the label \code{Awake\_e} will execute on 
+\code{TRAIL-0} (ln. 7).
+%
+For each statement that pauses (\code{emit} and \code{await}), resumes 
+(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
+and \code{break}), the compiler splits the trail into segments with associated 
+entry points.
+%
+The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
+in the middle of the figure).
+The state of trails translate to a vector of type \code{trail\_t} with the 
+maximum number of simultaneous trails (ln. 12--15).
+On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
+(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
+one, in ln. 14).
+
+\begin{figure}[t]
 \begin{minipage}[t]{0.31\linewidth}
@@
 
-In the final generated code in \C, each trail label representing an entry point 
-becomes a \emph{switch case} with the associated code to execute.
-%
-Figure~\ref{lst.impl.trails} illustrates the generation process.
-For the program in the left of the figure, the compiler extracts the entry 
-points and associated trails, e.g., the label \code{Awake\_e} will execute on 
-\code{TRAIL-0} (ln. 7).
-%
-For each statement that pauses (\code{emit} and \code{await}), resumes 
-(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
-and \code{break}), the compiler splits the trail into segments with associated 
-entry points.
-%
-The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
-in the middle of the figure).
-The state of trails translate to a vector of type \code{trail\_t} with the 
-maximum number of simultaneous trails (ln. 12--15).
-On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
-(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
-one, in ln. 14).
-
 The scheduler executes in two passes:
@@
 
-\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
-disciplined synchronous-reactive model and support for lexical composition of 
-lines of execution.
+\CEU has a strong influence from Esterel and embraces the disciplined 
+synchronous-reactive model and support for lexical composition of lines of 
+execution.
 %

### 2015-10-07 paper.tex
Revisao ate 4.4

@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{minipage}[t]{0.32\linewidth}
@@
 
-To encompass first-class timers, we introduce a special external \code{DT} 
-event that is intercalated with each other event occurrence in an application 
-(e.g.  \emph{e1, e2}):
+To encompass first-class timers, we introduce a special external event 
+\code{DT} that is intercalated with each other event occurrence in an 
+application (e.g.  \emph{e1, e2}):
 
@@
 
-The \code{DT} event has an associated variable \code{DT\_} carrying the 
-wall-clock time elapsed between the two occurrences, as depicted by the 
+The event \code{DT} has an associated variable \code{DT\_} carrying the 
+wall-clock time elapsed between two occurrences in sequence, as depicted by the 
 two-step translation of Figure~\ref{lst.map.timers.2}.
-In the concrete program in the left, the variable \code{dt} will hold the 
-residual delta time (as described in Section~\ref{sec.ceu.timers}) after 
-awaking from the timer.
+In the concrete program in the left, the variable \code{dt} holds the residual 
+delta time (as described in Section~\ref{sec.ceu.timers}) after awaking from 
+the timer.
 In the first step of the translation, we expand the \code{await 10ms} to a 
@@
     if tot <= 0 then
-        dt = tot;
+        dt = -tot;
         break;
@@
 
-The biggest mismatch between concrete and abstract \CEU is regarding 
-finalization blocks, which require more complex modifications in the program
-for a proper mapping using the $fin$ expressions.
+The biggest mismatch between concrete and abstract \CEU is regarding the 
+\code{finalize} blocks, which require more complex modifications in the program
+for a proper mapping using $fin$ expressions.
 In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
@@
 \code{\_hold}.
-In the translation, we first need to catch the \code{do-end} termination to run 
-the finalization code.
+In the translation, we first need to catch the outermost \code{do-end} 
+termination to run the finalization code.
 For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
-original body in parallel with a $fin$ to run the finalization code.
-This way, the $fin$ body executes whenever the \code{par/or} terminates, either 
-normally (after the \code{await B}) or aborted from an outer composition.
+original body in parallel with a $fin$ expression to run the finalization code.
+Note that the $fin$ has no transition rules in the semantics, keeping the 
+\code{par/or} alive.
+This way, the $fin$ body only executes when the \code{par/or} terminates either 
+normally (after the \code{await B}), or aborted from an outer composition.
 %
@@
 This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
-suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
-such as building a symbol table for checking variable declarations.
-%
-This phase outputs code in \C, given the way \CEU is tied to \C by design.
+suspicious accesses to shared memory, and also performs ``classical'' semantic 
+analysis, such as building a symbol table for checking variable declarations.
+%
+For the (standard) \C back end, this phase outputs code in \C.
 %
@@
 
-\item[Code Generation]
+\item[Final Generation]
 
@@
 
-\CEU favors a fine-grained use of trails, being common the use of trails that 
-await a single event.
+\CEU favors a fine-grained use of trails, being common to use trails that await 
+a single event.
 For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
@@
 At any given moment, a trail can be awaiting in one of the following states: 
-\code{INACTIVE}, \code{STACKED}, \code{FIN}, or in any of the events defined in 
-the program:
+\code{INACTIVE}, \code{STACKED}, \code{FINALIZE}, or in any of the events 
+defined in the program:
 
@@
     STACKED,
-    FIN,
+    FINALIZE,
     EVT_A,      // input void A;
@@
 %
-A \code{FIN} trail represents a hanged finalization block which is only 
+A \code{FINALIZE} trail represents a hanged finalization block which is only 
 scheduled when its corresponding block goes out of scope.
@@
 awaiting); the field \code{lbl} holds the entry point in the code to execute 
-when the trail is scheduled; the third field depends on the \code{evt} field 
-and may hold the \code{seqno} for an event, or the stack level \code{stk} for a
+when the trail segment is scheduled; the third field depends on the \code{evt} 
+field and may hold the \code{seqno} for an event, or the stack level \code{stk} 
+for a
 \code{STACKED} state.
@@
 
-In the final generated code in \C, each trail label representing an entry point 
-becomes a \emph{switch case} with the associated code to execute.
+In the final generated code in \C, each trail segment label representing an 
+entry point becomes a \emph{switch case} with the associated code to execute.
 %
@@
 Regarding abortion and finalization, when a \code{par/or} terminates, the 
-scheduler makes a \emph{broadcast} pass for the \code{FIN} event, but limited 
-to the range of trails covered by the terminating \code{par/or}.
-Trails that do not match the \code{FIN} are set to \code{INACTIVE}, as they 
-have to be aborted.
+scheduler makes a \emph{broadcast} pass for the \code{FINALIZE} event, but 
+limited to the range of trails covered by the terminating \code{par/or}.
+Trails that do not match the \code{FINALIZE} are set to \code{INACTIVE}, as 
+they have to be aborted.
 Given that trails in parallel are allocated in subsequent slots in the static 

### 2015-10-08 paper.tex
Revisao ate of final Mais texto do Adriano

@@
 The core of our semantics is a relation that, given a sequence number $n$ 
-identifying the current reaction chain, maps a program $p$ and a stack of 
-events $S$ in a single step to a modified program and stack:
+identifying the current reaction, maps a program $p$ and a stack of events $S$ 
+in a single step to a modified program and stack:
 %
@@
 n     &\in \mathds{N}
-    && (unique~identifier~for~the~reaction~chain)
+    && (unique~identifier~for~the~reaction)
 \end{align*}
 %
-At the beginning of a reaction chain, the stack is initialized with the 
-occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
-events on top of it (we discuss how they are popped further).
+At the beginning of a reaction, the stack is initialized with the occurring 
+external event $ext$ ($S=[ext]$), but $emit$ expressions can push new events on 
+top of it (we discuss how they are popped further).
 The ever-increasing sequence number $n$ prevents that awaiting expressions 
@@
 %
-Although the value here is arbitrary, it is unique in a reaction chain, because 
-a given expression can execute only once within it (remember that $loops$ must 
+Although the value here is arbitrary, it is unique in a reaction, because a 
+given expression can execute only once within it (remember that $loops$ must 
 contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
@@
 $mem$ expressions, i.e., they are guaranteed to execute entirely within a 
-reaction chain.
+reaction.
 
@@
 %
-A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
+A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
 parallel trails.
@@
 If all trails hangs only in $awaiting$ expressions, it means that the program 
-cannot advance in the current reaction chain.
+cannot advance in the current reaction.
 %
@@
 %
-To describe a \emph{reaction chain} in \CEU, i.e., how a program behaves in 
-reaction to a single external event, we use the reflexive transitive closure of 
+To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
+to a single external event, we use the reflexive transitive closure of 
 \textbf{relation-outer}:
@@
 corresponding variable that is explicitly set by the environment before each 
-reaction chain.
+reaction.
 
@@
 in the middle of the figure).
-The state of trails translate to a vector of type \code{trail\_t} with the 
+The state of trails translates to a vector of type \code{trail\_t} with the 
 maximum number of simultaneous trails (ln. 12--15).
@@
 %
-In the \emph{broadcast} pass, the scheduler sets all trails that are waiting 
-for the current event to \code{STACKED} in the current stack level.
-%
-In the \emph{dispatch} pass, the scheduler executes each trail that is 
-\code{STACKED} to run in the current level, setting it immediately to 
-\code{INACTIVE}.
+in the \emph{broadcast} pass, it sets all trails that are waiting for the 
+current event to \code{STACKED} in the current stack level;
+%
+in the \emph{dispatch} pass, it executes each trail that is \code{STACKED} to 
+run in the current level, setting it immediately to \code{INACTIVE} (the trail 
+segment may reset it in sequence if it doesn't terminate).
 
@@
 level and resumes the previous execution.
-For the first reaction chain, the scheduler starts from the \emph{dispatch} 
-pass, given that the \code{Main} label is the only one that can be active at 
-the stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).
-
-The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
-according to the current label to execute.
+For the first reaction, the scheduler starts from the \emph{dispatch} pass, 
+given that the \code{Main} label is the only one that can be active at the 
+stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).
+
+The code in the right of the Figure~\ref{lst.impl.trails} dispatches a trail 
+segment according to the current label to execute.
 %
@@
 \begin{description}
-\item[{\textbf\code{ceu\_go\_init()}}:] initializes the program (e.g. trails) 
-and executes the ``boot'' reaction (i.e., the \code{Main} label).
+\item[{\textbf\code{ceu\_go\_init()}}:] initializes the program (e.g., the 
+initial state of trails) and executes the ``boot'' reaction (i.e., the 
+\code{Main} label).
 
 \item[{\textbf\code{ceu\_go\_event(id,param)}}:] executes the reaction for the 
-received event id and associated parameter.
+received external event id and associated parameter.
 
@@
 As an example, it is possible to set different priorities for events that occur 
-concurrently (i.e. while a reaction chain is running).
-However, a binding must never interleave or run multiple functions in parallel.
+concurrently (i.e. while a reaction is running).
+However, a binding must never interleave or run multiple API calls in parallel.
 This would break the \CEU sequential/discrete semantics of time.
@@
 {
-    #include "ceu.h"
-    #include "ceu.c"
+    #include "ceu_app.h"
+    #include "ceu_app.c"
 
@@
 %
-The file \code{ceu.h} (ln. 3) contains all definitions for the compiled \CEU 
-program, which are further queried through \code{\#ifdef}'s.
-The file \code{ceu.c} (ln. 4) contains the main loop of \CEU pointing to the 
-labels defined in the program.
+The file \code{ceu\_app.h} (ln. 3) contains all definitions for the compiled 
+\CEU program, which are further queried through \code{\#ifdef}'s.
+The file \code{ceu\_app.c} (ln. 4) contains the runtime of \CEU with the 
+scheduler and dispatcher pointing to the labels defined in the program.
 The callback \code{Boot.booted} (ln. 6--11) is called by TinyOS on startup, so 
@@
 25--36).
+%
+The scheduler of the TinyOS is already synchronous by default and always 
+execute \code{event} handlers atomically, hence, the API calls to \CEU are 
+properly serialized.
 
@@
 - (analogo a Figura 18)
+
+During compilation process the script is broken in trails in the same way it is 
+done in the \CEU compiler explained in Section 4.
+The Terra virtual machine implements a similar \CEU scheduler to execute the 
+script trails, but the VM
+uses the real trails address instead of labels. In this way is possible to jump directly to each script
+partitions without using a dispatcher.
+The Terra VM instruction set includes operations to support the \CEU execution 
+model.
+Basically, it is composed by instructions to define a pending trail (\code{tkins}),
+a pending event(using \code{set\_c} to set a value in the memory),
+a check operation to verify the \code{par/and} termination(\code{chkret}),
+a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
+All pending entries are managed by the scheduler, that also responds for timer and input events.
+All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.
+
+Figure~\ref{lst.impl.assembly} presents the bytecode generated for the initial 
+part for the example of Figure~\ref{lst.impl.trails}.
+After initializing the trail and event slots in the memory, the VM engine starts the execution from the \code{main} label.
+In this example it registers in the trails slots the label addresses for the two block of the \code{par/and} control (lines 3--6).
+The scheduler, first, executes the code in the label \code{and\_sub\_1} that registers the label \code{awk\_e} into the slot of event \code{e} (lines 8--10).
+Then, executes the code in the label \code{and\_sub\_2} that registers the label \code{awk\_temp} into the slot of the event \code{I\_TEMP} (lines 17--18).
+In this moment the VM engine suspends its execution waiting for some new input or timer event.
+In the case of this example, the engine waits for the \code{I\_TEMP} event to continue the script execution.
+
+\begin{figure}
+\begin{minipage}[t]{0.91\linewidth}
+\begin{lstlisting}[language=C,numbers=left,xleftmargin=2.5em,keywords={tkins,end,set_c,ushort,ubyte,exec}]
+  <...> // Initializes the trails and events slots
+
+:main -- Activate TRAILs 0 & 1
+  tkins and_sub_1           // Activate TRAIL 0 - lbl and_sub_1
+  tkins and_sub_2           // Activate TRAIL 1 - lbl and_sub_2
+  end                       // Suspend execution
+
+:and_sub_1
+  set_c ushort slot_e awk_e // await e
+  end                       // Suspend execution
+
+:awk_e
+  set_c ubyte slot_1 1      // Close TRAIL 0 - lbl and_sub_1
+  exec chkret               // goto check par/and termination
+
+:and_sub_2
+  set_c ushort I_TEMP awk_temp // await TEMP
+  end                       // Suspend execution
+
+  <....> // Other entries: chkret, awk_TEMP, etc..
+\end{lstlisting}
+\end{minipage}
+\caption{
+%{\small %\textmd{
+Assembly code for a \code{par/and} control structure.
+%}%}
+\label{lst.impl.assembly}
+}
+\end{figure}
 
@@
 Instead of generating code in C, Terra generates the vmt assembly code using 
-the original Céu labels for trails entry points.
+the original \CEU labels for trails entry points.
 In the final phase for code generation, Terra compiler generates the vmt 
-bytecode file using the program address in place of Céu labels.
-The use of address allows the Céu scheduler to execute directly a trail without 
-a index table to replace the switch-case.
+bytecode file using the program address in place of \CEU labels.
+The use of address allows the \CEU scheduler to execute directly a trail 
+without a index table to replace the switch-case.
 Related to memory layout and trail allocation, Terra uses the same data 
-organization approach for all Céu controls and variables.
-All Céu data and control structures for a specific program are determined at 
+organization approach for all \CEU controls and variables.
+All \CEU data and control structures for a specific program are determined at 
 compile time.
@@
 space.
-The original Céu run time control functions are maintained with a small 
+The original \CEU run time control functions are maintained with a small 
 modification where accesses to control structures have an off-set relative to 
@@
 \CEU has a strong influence from Esterel and embraces the disciplined 
-synchronous-reactive model and support for lexical composition of lines of 
+synchronous-reactive model with support for lexical composition of lines of 
 execution.
@@
 However, there are fundamental semantic differences that prevents the design of 
-\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
-%
-In particular, Esterel has a notion of time similar to that of digital circuits 
-in which multiple signals can be active at a clock tick.
+\CEU as pure extensions to Esterel.
+%
+In particular, Esterel has a notion of time similar to digital circuits in 
+which multiple signals can be active at a clock tick.
 In fact, Esterel is also used in hardware design.
@@
 order for threads, based on either explicit priorities, or source code lexical 
-order (like \CEU).
+order.
 %
@@
 TODO-VM
+
+Plummer et al.~\cite{esterel.vm} present a reactive virtual machine for 
+embedded systems.
+This VM uses the Esterel programming language as scripting language and implements
+a specific instruction set to the Esterel concurrent and multi-thread semantic.
+They focus in code reduction instead of on efficiency.
+Experimentally, they found the bytecode was about half the size of optimized
+native assembly code generated from C.
+The Terra virtual machine also implements a customized instruction set for a reactive
+programming model, but it uses the \CEU model for concurrence and event 
+control.
+The Terra component based approach also contributes to have a very small bytecode size.
+In our experiments, depending of the abstraction level of the pre-built customized components,
+the bytecode size reduction may reach about 99\%.
 
@@
 %
-These properties are usually desirable in applications and are guaranteed as 
-preconditions in \CEU by design.
+These properties are usually desirable in embedded applications and are 
+guaranteed as preconditions in \CEU by design.
 

### 2015-10-09 paper.tex
Secao VM

@@
           interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
-          bool, data, tag, escape, new, traverse},
+          bool, data, tag, escape, new, traverse,implementation,output},
     emphstyle={\bfseries},
@@
 A program with blocks in sequence and in parallel, with corresponding memory 
-layout that the compiler generates.
+layout generated by the compiler.
 {\small %\textmd{
@@
 #ifdef EVT_PHOTO_READDONE
-    event void Photo.readDone (uint16_t val) {
+    event void Photo.readDone (int val) {
         ceu_sys_go(EVT_PHOTO_READDONE, &val);
@@
 The \emph{TinyOS} binding for \CEU.
+This platform-dependent template includes the \C files generated from the 
+original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
+\emph{Final generation} phase of Figure~\ref{fig.impl}.
 %}%}
@@
 Terra also provides system calls for initialization and configuration of 
+components (e.g., \emph{getters} and \emph{setters}).
+%
+The left of Figure~\ref{lst.terra.defs} shows a \CEU interface with the 
+available functionality for a customized \VM with temperature and radio 
 components.
-%
-TODO: explicar Figura 21.
-\begin{comment}
-Figure~\ref{lst.terra.defs} illustrates a program interface declaration in \CEU
-(in the left), and associated binding code in \C
-
- with respective binding customization scenario.
-This customization sample uses a small data structure, two output events, one input event, and two functions.
-The left listing is the definition file to be included into the \CEU script program and the right listing is the
-definition file to be included into the VM custom module implementation.
-\end{comment}
+The right of the figure shows the associated binding for output events (ln.  
+1--8), input events (ln. 10--14), and system calls (ln. 16--22).
+%
+Note that all applications for the customized \VM must comply with the same 
+interface.
+In contrast, the template-based \C back end allows applications to choose all 
+possible combinations of functionalities from the underlying platform at 
+compile time (as illustrated in Figure~\ref{lst.impl.tinyos}).
 
 \begin{figure}
-\begin{minipage}[t]{0.46\linewidth}
-\begin{lstlisting}[morekeywords={ubyte,ushort,output,function,regtype}]
-// Message definition
-regtype sensorMsg with
-    var ubyte  id;
-    var ushort value;
-end
-
-// Output events definition
-output void REQ_TEMP void          1;
-output void SEND_SENSOR sensorMsg  2;
-
-// Input events definition
-input void TEMP void               1;
-
-// functions definitions
-function ubyte getNodeId()         1;
-function ubyte queuePut(sensorMsg) 2;
-.
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.51\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
-morekeywords={nx_struct,nx_uint8_t,nx_uint16_t,uint8_t,uint32_t}]
-
-typedef nx_struct sensorMsg{
-  nx_uint8_t  id;
-  nx_uint16_t value;
-} sensorMsg_t;
-
-enum {
-  O_REQ_TEMP    = 1;
-  O_SEND_SENSOR = 2;
-
-  I_TEMP        = 1;
-
-  F_GETNODEID   = 1;
-  F_QUEUEPUT    = 2;
-};
-
-\end{lstlisting}
-\end{minipage}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[morekeywords={function}]
+// Output events
+output void REQUEST_TEMPERATURE;
+output int  REQUEST_SEND;    // sends int
+
+// Input events
+input int  TEMPERATURE_DONE; // receives int
+input void SEND_DONE;
+
+// System calls
+function int getRadioID (void);
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
+// Output events
+void VM.out(int evt_id, void* args) {
+    switch (id){
+        case O_REQUEST_TEMPERATURE:
+            call TINYOS_TEMP.read();
+        <...>; // O_REQUEST_SEND
+    }
+}
+
+// Input events
+event TINYOS_TEMP.done (int val) {
+    VM.enqueue(I_TEMPERATURE_DONE, &val);
+}
+<...> // TINYOS_SEND.done
+
+// System calls
+void VM.function(int id, void* params) {
+    switch (id) {
+        case F_GET_RADIO_ID:
+            VM.push(TINYOS_NODE_ID);
+    }
+}
+\end{lstlisting}
+\end{minipage}
+%
 \caption{
 %{\small %\textmd{
-Terra control block sample and the equivalent definition for VM implementation.
-%}%}
+\CEU interface (in the left) with associated customized \VM (in the right).
+%
+\code{VM.out} redirects all output events to the corresponding OS call (ln.  
+1--8).
+%
+Each \emph{TinyOS} event callback calls \code{VM.enqueue} for the corresponding 
+input event (ln 10--14).
+%
+System calls use \code{VM.push} for immediate return values (ln. 16--22).
+%
 \label{lst.terra.defs}
@@
 \end{figure}
-
-TODO: dizer o que eh manual/automatico nos lsts de ceu
-
-TODO-VM: exemplo de implementacao/uso de um componente
-- Terra: binding em nesC
-- Ceu: header de configuracao
-- (analogo a Figura 19)
 
@@
 \C.
-
-TODO-VM: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
-- (analogo a Figura 18)
+%
+The \VM includes special instructions to support complex and recurrent 
+operations from the runtime of \CEU, such as dispatching events, clearing 
+aborted trails
+Basically, they map
+
+Basically, it is composed by instructions to define a pending trail (\code{tkins}),
+a pending event(using \code{set\_c} to set a value in the memory),
+a check operation to verify the \code{par/and} termination(\code{chkret}),
+a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
+All pending entries are managed by the scheduler, that also responds for timer and input events.
+All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.
 
@@
 partitions without using a dispatcher.
-The Terra VM instruction set includes operations to support the \CEU execution 
-model.
-Basically, it is composed by instructions to define a pending trail (\code{tkins}),
-a pending event(using \code{set\_c} to set a value in the memory),
-a check operation to verify the \code{par/and} termination(\code{chkret}),
-a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
-All pending entries are managed by the scheduler, that also responds for timer and input events.
-All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.
 
@@
 %-------------------------------------------------
-Figure~\ref{lst.impl.custom} presents a partial implementation for the elements 
-introduced in the Figure~\ref{lst.impl.defs}.
-The two first functions are called from the VM decoder, the fist function \code{procOutevt()} dispatch any defined output events
-and the function \code{callFunction()} dispatch any defined customized functions.
-Lines 17--22 shows an example of a custom function that returns a value via stack.
-Lines 24--27 has a example of an external event call.
-Lines 29--34 presents an example how a input event is queued to the VM engine.
 \end{comment}
@@
 % Equivalent to the fig. 19
-
-\begin{figure}
-\begin{minipage}[t]{0.81\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
-morekeywords={uint8_t,uint16_t,uint32_t,error_t,command}]
-// Output event dispatcher
-command void VM.procOutEvt(uint8_t id, uint32_t value){
-  switch (id){
-    case O_REQ_TEMP    : proc_req_temp(id,value);    break;
-    case O_SEND_SENSOR : proc_send_sensor(id,value); break;
-  }
-}
-
-// Function dispatcher
-command void VM.callFunction(uint8_t id){
-  switch (id){
-    case F_GETNODEID : func_getNodeId(id); break;
-    case F_QUEUEPUT  : func_queuePut(id);  break;
-  }
-}
-
-// Pushing a value to the stack
-void  func_getNodeId(uint16_t id){
-  uint16_t stat;
-  stat = TOS_NODE_ID;
-  signal VM.push(stat);
-}
-
-// Calling a output event
-void  proc_req_temp(uint16_t id, uint32_t value){
-  call S_TEMP.read();
-}
-
-// Queueing an input event + value
-uint16_t lastTemp;
-event void S_TEMP.readDone(error_t result, uint16_t val)
-  lastTemp = val;
-  signal VM.queueEvt(I_TEMP, 0, &lastTemp);
-}
-
-\end{lstlisting}
-\end{minipage}
-
-\caption{
-%{\small %\textmd{
-VM Customization -- input/output events and functions.
-%}%}
-\label{lst.impl.custom}
-}
-\end{figure}
 
@@
 
-\emph{(As subsecoes serao retiradas do texto ao final...)}
-
-\subsection{Semantics}
+%\emph{(As subsecoes serao retiradas do texto ao final...)}
+
+%\subsection{Semantics}
 
@@
 
-A number of synchronous languages have been designed to interoperate with C, 
-such as \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
+Many synchronous languages have been designed to interoperate with C, such as 
+\emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
 \emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
@@
 
-TODO-VM
-
-Plummer et al.~\cite{esterel.vm} present a reactive virtual machine for 
-embedded systems.
-This VM uses the Esterel programming language as scripting language and implements
-a specific instruction set to the Esterel concurrent and multi-thread semantic.
-They focus in code reduction instead of on efficiency.
-Experimentally, they found the bytecode was about half the size of optimized
-native assembly code generated from C.
-The Terra virtual machine also implements a customized instruction set for a reactive
-programming model, but it uses the \CEU model for concurrence and event 
-control.
-The Terra component based approach also contributes to have a very small bytecode size.
-In our experiments, depending of the abstraction level of the pre-built customized components,
-the bytecode size reduction may reach about 99\%.
-
-\begin{comment}
-More recently, Wireless Sensor Networks (WSNs) emerged as an active research 
-area for highly constrained embedded concurrency, resulting in the development 
-of many synchronous languages~\cite{wsn.protothreads,wsn.sol,wsn.osm}.
-%
-offer lightweight cooperative multi-threading for embedded systems.
-Its stackless implementation reduces memory consumption but precludes support 
-for local variables.
-\CEU also avoids the use of stacks for trails, but preserves support for locals 
-by calculating the required memory at compile time.
-%
-SOL~\cite{wsn.sol} and OSM~\cite{wsn.osm} provide parallel state machines for 
-WSNs, offering a formal and mature model for programming embedded systems.
-However, the main contributions of \CEU, stacked execution for internal 
-events and safe support for shared-memory concurrency, do not directly adapt to 
-the state-machine formalism.
-
-In common among the referred works is the agreement in providing low-level 
-access (e.g., systems calls and shared-memory) and lock-free concurrency that 
-precludes race conditions on programs.
-However, they do not specify an execution order for tasks reacting to the same 
-external stimulus~\cite{esterel.primer,wsn.protothreads,wsn.osm}.
-This way, if two tasks access the same resource concurrently, even if the 
-accesses are race free, the final outcome is nondeterministic.
-As discussed in Section~\ref{sec.safety.det}, \CEU refuses programs with such 
-behavior.
-
-On the opposite side of the spectrum of concurrency models, asynchronous 
-languages for embedded systems~\cite{wsn.mantisos,arduino.occam}
-assume time independence among processes and are more appropriate for 
-applications with a low synchronization rate or for those involving
-algorithmic-intensive problems.
-\end{comment}
-
-\subsection{Implementation}
+%\subsection{Implementation}
 
@@
 
-TODO-VM
+%\subsection{VM}
+
+A number of virtual machines have been proposed for embedded systems.
+%
+The \emph{Sun SPOT} Java-based platform with the \emph{Squawk 
+JVM}~\cite{wsn.sunspot} brings Java for the embedded domain, but requires a 
+much powerful hardware.%
+\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
+KBytes of SRAM.}
+%
+\emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
+are complete \emph{JVMs} targeting constrained embedded systems with support 
+for multithreading and garbage collection.
+%
+Java has antagonistic design choices in comparison to \CEU:
+it does not impose static bounds on memory usage and execution time, and 
+provides preemptive multithreading with synchronization primitives for 
+accessing shared memory.
+%
+Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
+design choices to our work.
+To reduce code size, the \VM has a instruction set to deal with event handling 
+and concurrency constructs that are particular to Esterel.
+However, the proposed \VM is only a proof of concept, with no support for 
+arithmetic operations, external system calls, or remote reprogramming.
 

### 2015-10-09 paper.tex
Exemplo finalize: output em vez de funcao

@@
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
-input void STOP, RETRANSMIT, SENDACK;
+input void STOP, RETRANSMIT, SEND_ACK;
+output _pkt_t* SEND_ENQUEUE, SEND_CANCEL;
 par/or do
@@
                 <fill-buffer-info>
-                _send_enqueue(&buffer);
-                await SENDACK;
-            end
-        end
-    end
-end
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.41\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
+                emit SEND_ENQUEUE => &buffer;
+                await SEND_ACK;
+            end
+        end
+    end
+end
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=12]
 <...>
@@
     finalize
-        _send_enqueue(&buffer);
+        emit SEND_ENQUEUE => &buffer;
     with
-        _send_cancel(&buffer);
-    end
-    await SENDACK;
+        emit SEND_CANCEL => &buffer;
+    end
+    await SEND_ACK;
 <...>
@@
 %
-The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (ln. 1) 
-represent the external interface of the protocol.
-%
-The protocol has to transmit a packet every minute, unless it receives a 
-\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
-request to terminate.
+The input and output events represent the external interface of the protocol 
+(ln. 1--2).
+%
+The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
+unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
+or a \code{STOP} request to terminate.
 %
 The protocol is composed of two main trails:
-one simply monitors the stopping event (ln. 3);
-the other periodically transmits the packet (ln. 5--18).
-%
-The periodic transmission is a loop that starts two other trails (ln. 6--17):
-one handles the immediate retransmission request (ln. 7);
-the other transmits the packet and waits for a confirmation (ln. 9--16).
-%
-%
-The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
-takes at least one minute before looping (ln. 10), in accordance with the 
+one simply monitors the stopping event (ln. 4);
+the other periodically transmits the packet (ln. 6--19).
+%
+The periodic transmission is a loop that starts two other trails (ln. 7--18):
+one handles the immediate retransmission request (ln. 8);
+the other transmits the packet and waits for a confirmation (ln. 10--17).
+%
+%
+The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
+takes at least one minute before looping (ln. 11), in accordance with the 
 specification.
-Note that the call to \code{\_send\_enqueue}%
-\footnote{
-The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
-with external \C libraries explicit and are required in \CEU.
-}
-(ln. 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
-lexically-scoped packet (ln. 12).
+Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
+handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
 The driver makes the transmission in the background, holding the packet until 
-it signals the application with the \code{SENDACK} to acknowledge completion 
-(ln. 15).
-%
-At any time, the client may request a retransmission (ln. 7), which terminates 
-the \code{par/or} (ln. 6), aborts the ongoing transmission (ln. 14, if not 
-idle), and restarts the loop (ln. 5).
-%
-The client may also request to stop the whole protocol at any time (ln. 3).
+it signals the application with the \code{SEND\_ACK} to acknowledge completion 
+(ln. 16).
+%
+At any time, the client may request a retransmission (ln. 8), which terminates 
+the \code{par/or} (ln. 7), aborts the ongoing transmission (ln. 15, if not 
+idle), and restarts the loop (ln. 6).
+%
+The client may also request to stop the whole protocol at any time (ln. 4).
 %
@@
 \CEU tracks the interaction of \code{par/or} compositions with local variables 
-and stateful \C functions (e.g., device drivers) in order to preserve safe 
-abortion of trails.
+and stateful output events and \C calls in order to preserve safe abortion of 
+trails.
 %
@@
 In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
-accompany the stateful \C call.
+accompany the output request.
 The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
@@
 
-The $mem(id)$ primitive represents all accesses, assignments, and \C function 
-calls that affect a memory location identified by $id$.
+The $mem(id)$ primitive represents all accesses, assignments, \C function 
+calls, and output events that affect a memory location identified by $id$.
 As the challenging parts of \CEU reside on its control structures, we are not 

### 2015-10-09 paper.tex
Pequenos ajustes

@@
 \affil{Departamento de Inform\'atica, PUC--Rio}
-Adriano Branco
-\affil{Departamento de Inform\'atica, PUC--Rio}
 Roberto Ierusalimschy
@@
 \affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
+Adriano Branco
+\affil{Departamento de Inform\'atica, PUC--Rio}
 }
@@
 \CEU is a reactive language based on Esterel that targets constrained embedded 
-platforms and ensures safe concurrency by handling threats at compile time, 
-rather than at runtime.
+platforms and ensures safe concurrency by handling threats at compile time.
 %
@@
 %
-In this work, we discuss the design of \CEU and propose a formal semantics 
-focusing on its particular control mechanisms, such as parallel compositions, 
-finalization, and stack-based internal events.
-%
-We also present an implementation with two back ends:
+We discuss the design of \CEU and propose a formal semantics for its particular 
+control mechanisms, such as parallel compositions, finalization, and internal 
+events.
+%
+We also present two implementation back ends:
 one aiming for resource efficiency and interoperability with \C,
@@
 Therefore, if the sending trail is aborted by the \code{STOP} or 
-\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
-a \emph{dangling pointer} in the radio driver, which will possibly transmit 
-corrupted data.
+\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
+leaving behind a \emph{dangling pointer} in the radio driver, which will 
+possibly transmit corrupted data.
 %
@@
 \CEU tracks the interaction of \code{par/or} compositions with local variables 
-and stateful output events and \C calls in order to preserve safe abortion of 
-trails.
+and stateful output events calls in order to preserve safe abortion of trails.
 %
@@
 transmission when the block of \code{buffer} goes out of scope, i.e., the 
-finalization clause (after the \code{with}) executes automatically on external 
-abortion.%
+finalization clause (after the \code{with}, ln. 18) executes automatically on 
+external abortion.%
 \footnote{
@@
 The left of Figure~\ref{lst.terra.defs} shows a \CEU interface with the 
-available functionality for a customized \VM with temperature and radio 
-components.
-The right of the figure shows the associated binding for output events (ln.  
+available functionality for a customized \VM (with temperature and radio 
+components).
+The right of the figure shows the associated bindings for output events (ln.  
 1--8), input events (ln. 10--14), and system calls (ln. 16--22).
@@
 interface.
-In contrast, the template-based \C back end allows applications to choose all 
-possible combinations of functionalities from the underlying platform at 
-compile time (as illustrated in Figure~\ref{lst.impl.tinyos}).
+In contrast, the template-based \C back end (illustrated in 
+Figure~\ref{lst.impl.tinyos}) allows applications to choose all possible 
+combinations of functionalities from the underlying platform at compile time.
 
@@
 phase, which here outputs assembly instructions for the Terra \VM, instead of 
-\C.
+statements in \C.
 %
@@
 operations from the runtime of \CEU, such as dispatching events, clearing 
-aborted trails
+aborted trails,
+TODO
+\begin{comment}
 Basically, they map
@@
 
-\begin{comment}
-%-------------------------------------------------
-\end{comment}
-
-%-----------------------------------------------------
-% Equivalent to the fig. 19
-
-%------------------------------------------------------
-
-\begin{comment}
 Instead of generating code in C, Terra generates the vmt assembly code using 
@@
 %
-The \emph{Sun SPOT} Java-based platform with the \emph{Squawk 
-JVM}~\cite{wsn.sunspot} brings Java for the embedded domain, but requires a 
-much powerful hardware.%
+The \emph{Sun SPOT} Java-based platform with the \emph{Squawk JVM} brings Java 
+for the embedded domain~\cite{wsn.sunspot}, but requires a much powerful 
+hardware.%
 \footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
@@
 design choices to our work.
-To reduce code size, the \VM has a instruction set to deal with event handling 
-and concurrency constructs that are particular to Esterel.
+To reduce code size, the \VM has a specialized instruction set to deal with 
+events and concurrency constructs that are particular to Esterel.
 However, the proposed \VM is only a proof of concept, with no support for 

### 2015-10-09 paper.tex
Revisao Terra

@@
 
-Terra is a system for programming wireless sensor network applications that 
+Terra is a system for programming wireless sensor network applications which 
 uses \CEU as its scripting language~\cite{ceu.terra}.
@@
 
+The main difference between the standard \C back end and the Terra \VM is the 
+\emph{Code Generation} phase (as illustrated in Figure~\ref{fig.impl}), which 
+here outputs assembly instructions for the \VM, instead of statements in \C.
+%
+To reduce the memory footprint of applications, the \VM includes special 
+instructions for complex and recurrent operations from the runtime of \CEU, 
+such as handling events and trails.
+
 In Terra, \CEU scripts cannot execute arbitrary \C code, instead, they rely on 
@@
 \begin{figure}
-\begin{minipage}[t]{0.50\linewidth}
+\begin{minipage}[t]{0.51\linewidth}
 \begin{lstlisting}[morekeywords={function}]
@@
 output void REQUEST_TEMPERATURE;
-output int  REQUEST_SEND;    // sends int
+output int  REQUEST_SEND;    // sends int value
 
 // Input events
-input int  TEMPERATURE_DONE; // receives int
+input int  TEMPERATURE_DONE; // recvs int value
 input void SEND_DONE;
@@
 %
-\begin{minipage}[t]{0.50\linewidth}
+\begin{minipage}[t]{0.49\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=2.5em]
@@
 %{\small %\textmd{
-\CEU interface (in the left) with associated customized \VM (in the right).
-%
-\code{VM.out} redirects all output events to the corresponding OS call (ln.  
-1--8).
+\CEU interface (in the left) with customized \VM (in the right).
+%
+The routine \code{VM.out} redirects all output events to the corresponding OS 
+calls (ln.  1--8).
 %
@@
 
-Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
-the \C back end, the main difference resides in the \emph{Code Generation} 
-phase, which here outputs assembly instructions for the Terra \VM, instead of 
-statements in \C.
-%
-The \VM includes special instructions to support complex and recurrent 
-operations from the runtime of \CEU, such as dispatching events, clearing 
-aborted trails,
-TODO
 \begin{comment}
@@
 %
-The \emph{Sun SPOT} Java-based platform with the \emph{Squawk JVM} brings Java 
-for the embedded domain~\cite{wsn.sunspot}, but requires a much powerful 
-hardware.%
+The \emph{Sun SPOT} platform with the \emph{Squawk JVM} brings Java for the 
+embedded domain~\cite{wsn.sunspot}, but requires a much powerful hardware.%
 \footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 

### 2015-10-10 paper.tex
based on Esterel => inspired by Esterel

@@
 % Title portion
-\title{The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}
+\title{The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language}
 
@@
 \begin{abstract}
-\CEU is a reactive language based on Esterel that targets constrained embedded 
-platforms and ensures safe concurrency by handling threats at compile time.
+\CEU is a reactive language inspired by Esterel that targets constrained 
+embedded platforms and ensures safe concurrency by handling threats at compile 
+time.
 %
@@
 
-\CEU is a synchronous reactive language based on Esterel with support for 
+\CEU is a synchronous reactive language inspired by Esterel with support for 
 multiple concurrent lines of execution known as \emph{trails}.
@@
 We presented the design, semantics, and implementation of \CEU, a synchronous 
-reactive language based on Esterel targeting constrained embedded systems.
+reactive language inspired by Esterel targeting constrained embedded systems.
 

### 2015-10-10 paper.tex
Novo titulo Nova intro   - complexidades de Esterel Revisoes Roberto

@@
 % Title portion
-\title{The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language}
+\title{The Design, Semantics, and Implementation of the Synchronous Reactive 
+Language \CEU}
 
@@
 Considering the control-based languages, Esterel~\cite{esterel.ieee91} was the 
-first to appear and influenced a number of embedded languages, such as 
-\emph{Reactive-C}~\cite{rp.rc}, \emph{OSM}~\cite{wsn.osm}, 
+first to appear and succeed, influencing a number of embedded languages, such 
+as \emph{Reactive-C}~\cite{rp.rc}, \emph{OSM}~\cite{wsn.osm}, 
 \emph{Sync-C}~\cite{rp.synchc}, and \emph{PRET-C}~\cite{rp.pretc}.
 
-\CEU is another Esterel-based language targeting embedded and (soft) real-time 
-systems with novel functionalities:
+Despite its success and influence, Esterel has a overly complex semantics that 
+requires ingenious static analysis to detect and refuse programs with 
+\emph{causality} and \emph{schizophrenia} 
+problems~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
+%
+The complex semantics not only challenges the analysis and compilation of 
+programs, but also results in incompatible or non-compliant implementations.
+%
+More importantly, it also affects the programmer's understanding capacity, 
+which, after all, have to solve the errors when facing corner cases in 
+programs.
+%
+Another drawback of the Esterel semantics consists of non-deterministic 
+execution for intra-reaction statements, which prevents threads to share memory 
+and interact with side-effect and stateful system calls.
+
+In this work, we present \CEU, a new programming language that inherits the 
+synchronous and imperative mindset of Esterel but diverges in fundamental 
+semantic aspects.
+%
+Overall, \CEU has a simple semantics with fine-grained execution control, and a 
+straightforward implementation targeting resource-constrained systems.
+%
+The list that follows summarizes the semantic peculiarities of \CEU:
 %
@@
 \end{itemize}
-
-In this work, we discuss the design of \CEU and propose a formal semantics for 
-a small synchronous kernel that represents a subset of the language covering 
-these new functionalities.
-%
-We also present an implementation of \CEU with two back ends:
+%
+We discuss the design of \CEU and present a formal semantics for a small 
+synchronous kernel that represents a subset of the language covering these new 
+functionalities.
+%
+We also present a lightweight implementation of \CEU with two back ends:
 one aiming for resource efficiency and interoperability with \C,
@@
 Our implementations target resource-constrained devices, such as \emph{Arduino} 
-and \emph{MICAz} sensor nodes based on 8-bit microcontrollers.%
+and \emph{MICAz} sensor nodes based on 8-bit microcontrollers, showing that the 
+peculiarities in the semantics of \CEU does not pose practical obstacles.%
 \footnote{
@@
 messages.
-In contrast, the full image of a \C-compiled version occupies more than 2 
-Kbytes~\cite{ceu.terra}.%
-\footnote{
-The \VM is preloaded in the devices with some \emph{I/O} functionality, whereas 
-\C-based binaries are full images.
-}
 
@@
 By reactive, we mean that programs are stimulated by the environment through 
-input events that are broadcast to all awaiting trails.
-By synchronous, we mean that all trails at any given moment are either reacting 
-to the current event or are awaiting another event;
-in other words, trails are never reacting to different events.
+input events, which are broadcast to all awaiting trails.
+By synchronous, we mean that trails at any given moment are either reacting to 
+the current event or are awaiting another event;
+in other words, trails never react to different events simultaneously.
 
@@
 For the second phrase, the reset behavior, the Esterel version uses a 
-\code{abort-when} (ln. 3--10), which serves the same purpose of \CEU's 
-\code{par/or} (ln. 3--12):
+\code{abort-when} (ln. 3--10), which, in this case, serves the same purpose of 
+\CEU's \code{par/or} (ln. 3--12):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
@@
 
-\begin{figure}[b]
+\begin{figure}
 \begin{minipage}[t]{0.48\linewidth}
@@
 
-\CEU has a strong imperative flavor, with explicit control flow through 
-sequences, loops, parallels, and also assignments.
-Being designed for control-intensive applications, it provides support for 
+\CEU employs the synchronous model, in which programs advance in a sequence of 
+discrete reactions to external events.
+%
+It also has a strong imperative flavor, with explicit control flow through 
+sequences, loops, and parallels, and also assignments.
+%
+Being designed for control-intensive applications, \CEU provides support for 
 concurrent lines of execution and broadcast communication through events.
 %
-In addition, \CEU also employs the synchronous model, in which programs advance 
-in a sequence of discrete reactions to external events.
 Internal computations within a reaction (e.g. expressions, assignments, and 
@@
 hypothesis~\cite{rp.hypothesis}.
-The \code{await} statements are the only ones that halt a running reaction and 
-allow a program to advance in this notion of time.
+An \code{await} is the only statement that halts a running reaction and allows 
+a program to advance in this notion of time.
 %
@@
 computations that run in unbounded time (e.g., cryptography, image processing) 
-do not fit the zero-delay hypothesis, and cannot be elegantly implemented.
+do not fit the zero-delay hypothesis, and cannot be directly implemented.
 
@@
 
-Esterel makes no semantic distinctions between internal and external signals, 
-both having only the notion of either presence or absence during an entire 
-reaction~\cite{esterel.preemption}.
-%
-In \CEU, external input events are unique within reactions and programs cannot 
-emit them, resulting in an intrinsic queue-based handling.
+Esterel makes no semantic distinctions between internal and external signals.
+In particular, programs can emit multiple external input signals, which may 
+coexist during a reaction.
+%
+In \CEU, a reaction starts from the occurrence of an external input event and 
+programs cannot emit inputs.
+Therefore, the occurring input event is unique during the entire reaction, 
+resulting in intrinsic queue-based handling.
 %
@@
 %
-In the version in Esterel, when \code{A} occurs, \code{B} is emitted (ln.  
-5--6) and both events become active, resulting in the invocation of \code{f()} 
-and \code{g()} in no particular order.
-%
-In the version in \CEU, the occurrence of \code{A} makes the program behave as 
-follows (with the stack contents in italics):
+In Esterel, when \code{A} occurs, \code{B} is emitted (ln.  5--6) and both 
+events become active, resulting in the invocation of \code{f()} and \code{g()} 
+in no particular order.
+%
+In \CEU, the occurrence of \code{A} makes the program behave as follows:
 %
@@
 \setlength{\itemsep}{0pt}
-\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.\\
-    \emph{stack: [1st-trail]}
-\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.\\
-    \emph{stack: [1st-trail]}
+\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.
+\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.
 \item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
-    terminates.\\
-    \emph{stack: []}
+    terminates.
 \item Both trails have terminated, so the \code{par/and} rejoins, and the 
-program also terminates;
+program also terminates.
 \end{enumerate}
@@
 %
-\CEU imposes a restriction that an \code{await} is only valid for the next 
-reaction, i.e., if an \code{await} and \code{emit} occur simultaneously in 
-parallel trails, the \code{await} does not awake.
+\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
+for the next reaction, i.e., if an \code{await} and \code{emit} occur 
+simultaneously in parallel trails, the \code{await} does not awake.
 %
@@
 %
-The example that follows relies on this restriction to avoid infinite 
-execution:
-
+The example in Figure~\ref{lst.await} illustrates this distinction, which 
+prevents infinite execution by design.
+Both sides of the \code{par/or} have an \code{await} statement to avoid 
+instantaneous termination (ln. 4,7).
+However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
+same reaction that reaches them, the \code{par/or} would terminate and restart 
+the \code{loop} instantaneously, resulting in infinite execution.
+
+\begin{figure}
 \begin{lstlisting}[numbers=left,xleftmargin=2em]
@@
     with
-        emit e;     // w/o the restriction, the emit awakes 1st trail
+        emit e;     // w/o delayed awaits, the emit awakes 1st trail
         await f;    // and restarts the loop instantaneously
@@
 \end{lstlisting}
-
-Both sides of the \code{par/or} have an \code{await} statement to avoid 
-instantaneous termination (ln. 4,7).
-However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
-same reaction that reaches them, the \code{par/or} would terminate and restart 
-the \code{loop} instantaneously, resulting in infinite execution.
+\caption{ Delayed awaits prevents re-execution of statements by design.
+\label{lst.await}
+}
+\end{figure}
 
@@
 %
-From our experience, in some cases we need to execute a block of code 
-periodically from internal event requests, \emph{including the current 
-reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
-In this case, moving the \code{await} to the end of the loop (ln. 10) makes the 
+For instance, sometimes we need to execute a block of code immediately, and 
+then periodically, from internal event requests, as illustrated in 
+Figure~\ref{lst.delay}.
+In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
 periodic code to also execute immediately (ln. 9), and then in reactions to 
 each \code{emit} request (ln. 5).
-If the periodic \code{emit} depends on a condition, as illustrated in the right 
-of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
-state variable (ln. 2) and also to copy the condition test to the periodic code 
-(ln. 13).
+If the periodic \code{emit} depends on a condition, then the code 
+transformation becomes more intricate, requiring an extra condition test around 
+the periodic code to prevent its immediate execution.
 %
@@
 
-\begin{figure}
-\begin{minipage}[t]{0.50\linewidth}
+\begin{figure}[t]
+%\begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
@@
     loop do
-        <...>     // code that awaits
+        <...>     // code that awaits some period
         emit e;   // periodic request
@@
     loop do
-        <...>     // code to execute
+        <...>     // code to execute immediately and then periodically
         await e;  // await after
@@
 \end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-event void e;
-var bool should_execute = false;
-par do
-    loop do
-        <...>         // code that awaits
-        if <...> then // some conditon
-            should_execute = true;
-            emit e;
-        end
-    end
-with
-    loop do
-        if should_execute then
-            <...>     // code to execute
-        end
-        await e;
-    end
-end
-\end{lstlisting}
-\end{minipage}
-%
-\caption{ Examples that circumvent the \emph{delayed await} restriction by 
-post-fixing the \code{await} inside the \code{loop} (in the left), and by 
-copying the condition test (in the right).
+%\end{minipage}
+%
+%\begin{minipage}[t]{0.50\linewidth}
+%\begin{lstlisting}[numbers=left,xleftmargin=3em]
+%event void e;
+%var bool should_execute = false;
+%par do
+%    loop do
+%        <...>         // code that awaits
+%        if <...> then // some conditon
+%            should_execute = true;
+%            emit e;
+%        end
+%    end
+%with
+%    loop do
+%        if should_execute then
+%            <...>     // code to execute
+%        end
+%        await e;
+%    end
+%end
+%\end{lstlisting}
+%\end{minipage}
+%
+\caption{ An example that circumvents the \emph{delayed await} restriction by 
+post-fixing the \code{await} inside the \code{loop}.
+% (in the left), and by copying the condition test (in the right).
 \label{lst.delay}
@@
 
-Embedded applications make extensive use of global memory, such as for 
-accessing resources through memory-mapped registers.
+Embedded applications make extensive use of global memory and shared resources, 
+such as through memory-mapped registers and system calls to drivers.
 Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
-with concurrent lines of execution sharing memory.
-%
+with concurrent lines of execution sharing memory and interacting with the 
+environment through system calls.
+
 Esterel is only deterministic with respect to reactive control: ``the same 
@@
 %
-However, the execution order for operations with side-effects within a reaction 
-is non-deterministic: ``if there is no control dependency, as in \code{<<call 
-f1() || call f2()>>}, the order is unspecified and it would be an error to rely 
-on it''~\cite{esterel.primer}.
-%
-Therefore, Esterel forbids sharing memory between lines of execution: 
-\emph{``if a variable is written by some thread, then it can neither be read 
-nor be written by concurrent threads''}~\cite{esterel.primer}.
+However, the execution order for operations within a reaction is 
+non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
+|| call f2()>>}, the order is unspecified and it would be an error to rely on 
+it''~\cite{esterel.primer}.
+%
+For this reason, Esterel assumes that external calls to the host language do 
+not perform side effects.
+%
+Therefore, Esterel syntactically forbids sharing memory between lines of 
+execution to preserve determinism: \emph{``if a variable is written by some 
+thread, then it can neither be read nor be written by concurrent 
+threads''}~\cite{esterel.primer}.
 
@@
 \code{x} (ln. 5,8) can never run concurrently, because each trail segment 
-reacts to a different input event (ln. 4,7) and, according to the semantics of 
-\CEU, cannot occur simultaneously.
+reacts to different input events (ln. 4,7), which cannot occur simultaneously 
+(according to Section~\ref{sec.ceu.ints}).
 %
@@
 
-\begin{figure}[b]
+\begin{figure}
 \begin{minipage}[t]{0.50\linewidth}
@@
 accesses to shared variables, as follows:
-\emph{if a variable is written in a trail segment, then a concurrent trail 
-segment cannot read or write to that variable, nor dereference a pointer of 
-that variable type.}
+if a variable is written in a trail segment, then a concurrent trail segment 
+cannot read or write to that variable, nor dereference a pointer of that 
+variable type.
 An analogous policy is applied for pointers \emph{vs} variables and pointers 
 \emph{vs} pointers, as well as for system calls with side effects (e.g., 
-\code{printf}).
+\code{printf}).%
+\footnote{\CEU assumes all system calls perform side effects, unless they are 
+annotated as ``\texttt{@pure}''.}
 %The algorithm for the analysis holds the set of all events in preceding 
@@
 
-To illustrate threats related to abortion, consider the unsafe example in the 
+To illustrate the risks related to abortion, consider the unsafe example in the 
 left of Figure~\ref{lst.fin}, which does not compile in \CEU.
-It describes the state machine of a data collection protocol for sensor 
-networks~\cite{wsn.ctp,ceu.sensys13}.
+It describes the state machine of the data collection protocol \emph{CTP} for 
+sensor networks~\cite{wsn.ctp}.
 %
@@
 %
-The protocol is composed of two main trails:
+The protocol is implemented with two main trails:
 one simply monitors the stopping event (ln. 4);
@@
 The driver makes the transmission in the background, holding the packet until 
-it signals the application with the \code{SEND\_ACK} to acknowledge completion 
-(ln. 16).
+it signals the application with a \code{SEND\_ACK} event to acknowledge 
+completion (ln. 16).
 %
 At any time, the client may request a retransmission (ln. 8), which terminates 
-the \code{par/or} (ln. 7), aborts the ongoing transmission (ln. 15, if not 
-idle), and restarts the loop (ln. 6).
+the \code{par/or} (ln. 7) and restarts the loop (ln. 6), but does not abort the 
+ongoing transmission initiated with the \code{emit SEND\_ENQUEUE} (ln. 15).
 %
@@
 %
-In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
-accompany the output request.
+\CEU enforces a \emph{finalization} clause to accompany the output request.
 The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
@@
       $|$ p @ loop p                 (unwinded loop)
+      $|$ nop                        (terminated expression)
 \end{lstlisting}
@@
 
-Figure~\ref{lst.formal.syntax} shows the BNF-like syntax for a subset of \CEU 
-that is sufficient to describe all semantic peculiarities of the language.
-%
-Except for $fin$ and all semantic-derived expressions (i.e., $awating$, 
-$emitting$, and $p~@~loop; p$), which are discussed further, all other 
-expressions are equivalent to their counterparts in the concrete language.
+Figure~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
+sufficient to describe all semantic peculiarities of the language.
+%
+Except for $fin$ and the expressions used internally by the semantics (i.e., 
+$awating$, $emitting$, $p~@~loop; p$, and $nop$), all other expressions are 
+equivalent to their counterparts in the concrete language.
 
@@
 calls, and output events that affect a memory location identified by $id$.
+%
+According to the synchronous hypothesis of \CEU, $mem$ expressions are 
+considered to be atomic and instantaneous.
+%
 As the challenging parts of \CEU reside on its control structures, we are not 
@@
 occurrences in programs.
-%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
-%considered to be atomic and instantaneous
-%
-The special notation $nop$ is used to represent an innocuous $mem$ expression 
-(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
-unused identifier).
+%
+%The special notation $nop$ is used to represent an innocuous $mem$ expression 
+%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
+%unused identifier).
 %
@@
 
-The core of our semantics is a relation that, given a sequence number $n$ 
-identifying the current reaction, maps a program $p$ and a stack of events $S$ 
-in a single step to a modified program and stack:
+The core of our semantics describes how a program reacts to a single external 
+input event, i.e., starting from the input event, how the program behaves and 
+becomes idle again for the subsequent reaction.
+%
+We use a set of small-step operational rules, which are built in such a way 
+that at most one transition is possible at any time, resulting in deterministic 
+reactions.
+%
+Each reaction is identified by a ever-increasing $n$ that remains constant 
+during the entire reaction.
+The transition rules map a program $p$ and a stack of events $S$ in a single 
+step to a modified program and stack:
 %
@@
 \LL S', p' \RR
-    & \textbf{(relation-inner)}
+    & \textbf{(rule-inner)}
 \end{align*}
@@
 S, S' &\in id^*
-    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
+    &&(stack~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
 p, p' &\in P
@@
 n     &\in \mathds{N}
-    && (unique~identifier~for~the~reaction)
+    && (unique~identifier~for~the~entire~reaction)
 \end{align*}
@@
 top of it (we discuss how they are popped further).
-The ever-increasing sequence number $n$ prevents that awaiting expressions 
-awake in the same reaction they are reached (the \emph{delayed awaits} as 
-explained in Section~\ref{sec.ceu.ints}).
-
-We describe this relation with a set of small-step structural semantics rules, 
-which are built in such a way that at most one transition is possible at any 
-time, resulting in deterministic reaction chains.
-%
+The sequence number $n$ prevents that awaiting expressions awake in the same 
+reaction they are reached (the \emph{delayed awaits} as explained in 
+Section~\ref{sec.ceu.ints}).
+
 The transition rules for the primary expressions are as follows:
@@
 \begin{align*}
+\LL S,\1mem(id) \RR &\ST
+\LL S,\1nop \RR
+    & \textbf{(mem)}        \\
+%%%
 \LL S,\1await(id) \RR &\ST
@@
 \LL id:S,\1awaiting(id,m) \RR &\ST
-\LL id:S,\1nop \RR, \2m<=n
+\LL id:S,\1nop \RR, \2if~m\leq n
     & \textbf{(awake)}   \\
@@
 %%%
-\LL S,\1emitting(|S|) \RR &\ST
-\LL S,\1nop \RR
-    & \textbf{(pop)}
+\LL S,\1emitting(k) \RR &\ST
+\LL S,\1nop \RR, \2 if~k=|S|
+    & \textbf{(pop)}        %\\
 \end{align*}
@@
 %
-An $await$ is simply transformed into an $awaiting$ (rule \textbf{await}) as an 
+A $mem$ operation executes immediately and becomes a $nop$ to indicate 
+termination (rule \textbf{mem}).
+An $await$ is transformed into an $awaiting$ (rule \textbf{await}) as an 
 artifice to remember the external sequence number $n+1$ it can awake:
@@
 %%%
-& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
+& \LL S, (nop~;~q) \RR \ST  \LL S, q \RR
     & \textbf{(seq-nop)}      \\
@@
 %%%
-& \LL S, (mem(id)~@~loop~p) \RR \ST \LL S, loop~p \RR
+& \LL S, (nop~@~loop~p) \RR \ST \LL S, loop~p \RR
     & \textbf{(loop-nop)}    \\
@@
 \textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
-$mem(id)$.
+$nop$.
 However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
@@
                                     isBlocked(n,S,q)             \\
-  isBlocked(n,S, \_)             &= false \2  (mem,await,      \\
+  isBlocked(n,S, \_)             &= false \2  (nop,await,      \\
                                   &    \5\5\5\2 emit,break,if,loop)   %\\
@@
 \begin{eqnarray*}
-& \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
+& \LL S, (nop~and~q) \RR \ST \LL S, q \RR
     & \textbf{(and-nop1)}   \\
 %%%
-& \LL S, (p~and~mem(id)) \RR \ST \LL S, p \RR
+& \LL S, (p~and~nop) \RR \ST \LL S, p \RR
     & \textbf{(and-nop2)}   \\
 %%%
-& \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
+& \LL S, (nop~or~q) \RR \ST \LL S, clear(q) \RR
     & \textbf{(or-nop1)}   \\
@@
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
+    {\DS \LL S, (p~or~nop) \RR \ST \LL S, clear(p) \RR }
     & \textbf{(or-nop2)}   %\\
@@
 %
-Therefore, we define another relation that behaves as \textbf{relation-inner} 
+Therefore, we define another rule that behaves as \textbf{rule-inner} 
 (presented above) if the program can advance, and, otherwise, pops the stack:
@@
     { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
-\5\5\5\textbf{(relation-outer)}
+\5\5\5\textbf{(rule-outer)}
 $$
@@
 to a single external event, we use the reflexive transitive closure of 
-\textbf{relation-outer}:
+\textbf{rule-outer}:
 %
@@
 
-\begin{figure}[b]
+\begin{figure}
 \begin{minipage}[t]{0.30\linewidth}
@@
 \item The program terminates, i.e., the small-step rules transform the whole 
-program into a $mem(id)$.
+program into a $nop$.
 \end{itemize}
@@
 
-\begin{comment}
-As a static language, much of the complexity in the implementation of \CEU{} 
-resides in the compile phase.
-Nonetheless, some complexity is left to the runtime phase, which has to manage 
-first-class timers, finalization blocks, and all bookkeeping related to trails.
-\end{comment}
-
-The compilation process of a program in \CEU is composed of three main phases, 
-as illustrated in Figure~\ref{fig.impl}:
-
-\begin{figure}[b]
-\centering
-\includegraphics[scale=0.20]{impl}
-\caption{ Compilation process: from the source code in \CEU to the final 
-binary.
-\label{fig.impl}
-}
-\end{figure}
-
-\begin{comment}
-The program below is used as our guiding example for this chapter:
-
-\begin{lstlisting}[numbers=left,xleftmargin=2em]
-input int A, B, C;
-var int ret;
-loop do
-   par/or do
-      do                            // 1st trail
-          var int a = await A;
-          ret = ret + a;
-      end
-      do
-          var int b = await B;
-          ret = ret + b;
-      end
-      break;
-   with
-      par/and do                    // 2nd trail
-         finalize with
-            ret = ret + 10;
-         end
-         await C;
-      with
-         await A;
-      end
-   end
-end
-...                                 // code after the loop
-\end{lstlisting}
-\end{comment}
-
-\begin{description}
-
-\item[Parsing]
-
-The parser of \CEU is written in \emph{LPeg}~\cite{lua.lpeg}, a pattern 
-matching library that also recognize grammars, making it possible to write the 
-tokenizer and grammar with the same tool.
-%
-The source code is then converted to an \emph{abstract syntax tree (AST)} to be 
-used in further phases.
-%
-This phase may be aborted due to syntax errors in the \CEU source file.
-
-\item[Temporal Analysis]
-
-This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
-suspicious accesses to shared memory, and also performs ``classical'' semantic 
-analysis, such as building a symbol table for checking variable declarations.
-%
-For the (standard) \C back end, this phase outputs code in \C.
-%
-Some type checking is delayed to the last phase to take advantage of 
-\code{gcc}'s error handling.
-Therefore, we annotate the \C output file with \code{\#line} pragmas matching 
-the original file in \CEU.
-
-\item[Final Generation]
-
-The final phase packs the generated \C file with the \CEU runtime and 
-platform-dependent functionality, compiling them with \emph{gcc} and generating 
-the final binary.
-%
-The \CEU runtime comprehends the scheduler, timer management, and the external 
-\C API.
-%
-The platform files include libraries for I/O and bindings to invoke the \CEU 
-scheduler on external events.
-
-\end{description}
-
-%In the sections that follow, we discuss the most sensible parts of the 
-%\emph{code generation} phase.
-%compiler considering our design, such as the temporal analysis, runtime 
-%scheduler, and the external API.
+The compilation process of a program in \CEU for the original \C back end is 
+composed of three main phases:
+the \emph{parsing phase} converts the source code in \CEU to an \emph{abstract 
+syntax tree (AST)};
+the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
+unbounded loops and suspicious accesses to shared memory;
+the \emph{code generation phase} converts the \emph{AST} to standard \C code 
+and packs it with platform-dependent functionality and the runtime of \CEU, 
+compiling everything with \emph{gcc} to generate the final binary.
 
@@
 
-\subsection{The External \C API}
-
-As a reactive language, the execution of a program in \CEU is guided entirely 
-by the occurrence of external events.
-From the implementation perspective, there are three external sources of input 
-into programs, which are all exposed as functions in a \C API:
-
-\begin{description}
-\item[{\textbf\code{ceu\_go\_init()}}:] initializes the program (e.g., the 
-initial state of trails) and executes the ``boot'' reaction (i.e., the 
-\code{Main} label).
-
-\item[{\textbf\code{ceu\_go\_event(id,param)}}:] executes the reaction for the 
-received external event id and associated parameter.
-
-\item[{\textbf\code{ceu\_go\_wclock(us)}}:] increments the current time in 
-microseconds and runs a reaction if any timer expires.
-
-%\item[{\textbf\code{ceu\_go\_async}}:] executes a single loop iteration for 
-%the next \code{async}, switching among them in a \emph{round robin} policy.
-\end{description}
-
-Given the semantics of \CEU, the functions are guaranteed to take a bounded 
-time to execute.
-They also return a status code that says if the \CEU{} program has terminated 
-after the reactions.
-Further calls to the API have no effect on terminated programs.
-
-\begin{comment}
-Note that \CEU{} code running from a call to \code{ceu\_go\_async} may emit an 
-input event or the passage of time.
-In this case, the \C implementation makes a tail call to the corresponding 
-handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as synchronous 
-code has higher priority.
-
-The API reflects the \emph{global asynchronous} part of \CEU{}, as discussed in 
-Section~\ref{sec:ceu:gals}.
-A simple and opaque API hides local state from the environment, suggesting that 
-the execution varies entirely according to the sequence (and parameters) of API 
-calls.
-\end{comment}
-
-The bindings for the specific platforms are responsible for calling the 
-functions in the API in the order that better suit their requirements.
-As an example, it is possible to set different priorities for events that occur 
-concurrently (i.e. while a reaction is running).
-However, a binding must never interleave or run multiple API calls in parallel.
+\subsection{Interaction with the Environment}
+
+As a reactive language, the execution of programs in \CEU is guided entirely by 
+the occurrence of external input events.
+%
+The binding for a specific platform (environment) is responsible for calling 
+hook functions in the API of the runtime of \CEU whenever an external event 
+occurs.
+%
+However, the binding must never interleave or run multiple API calls in 
+parallel.
 This would break the \CEU sequential/discrete semantics of time.
-%, as discussed in Section~\ref{sec:ceu}.
-
-\begin{figure}
-\begin{lstlisting}[numbers=left,xleftmargin=2em]
-implementation
-{
-    #include "ceu_app.h"
-    #include "ceu_app.c"
-
-    event void Boot.booted () {
-        ceu_sys_init();
-#ifdef CEU_WCLOCKS
-        call Timer.startPeriodic(10);
-#endif
-    }
-
-#ifdef CEU_WCLOCKS
-    event void Timer.fired () {
-        ceu_sys_wclock(10000);
-    }
-#endif
-
-#ifdef EVT_PHOTO_READDONE
-    event void Photo.readDone (int val) {
-        ceu_sys_go(EVT_PHOTO_READDONE, &val);
-    }
-#endif
-
-#ifdef EVT_RADIO_SENDDONE
-    event void RadioSend.sendDone (message_t* msg) {
-        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
-    }
-#endif
-
-#ifdef EVT_RADIO_RECEIVE
-    event message_t* RadioReceive.receive (message_t* msg) {
-        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
-        return msg;
-    }
-#endif
-
-    <...>   // other events
-}
-\end{lstlisting}
-%\rule{14cm}{0.37pt}
-\caption{
-%{\small %\textmd{
-The \emph{TinyOS} binding for \CEU.
-This platform-dependent template includes the \C files generated from the 
-original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
-\emph{Final generation} phase of Figure~\ref{fig.impl}.
-%}%}
-\label{lst.impl.tinyos}
-}
-\end{figure}
 
 As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
-\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
+\emph{TinyOS}~\cite{wsn.tos}, which maps driver callbacks to input events in 
+\CEU.
 %
@@
 
+\begin{figure}
+\begin{lstlisting}[numbers=left,xleftmargin=2em]
+implementation
+{
+    #include "ceu_app.h"
+    #include "ceu_app.c"
+
+    event void Boot.booted () {
+        ceu_sys_init();
+#ifdef CEU_WCLOCKS
+        call Timer.startPeriodic(10);
+#endif
+    }
+
+#ifdef CEU_WCLOCKS
+    event void Timer.fired () {
+        ceu_sys_wclock(10000);
+    }
+#endif
+
+#ifdef EVT_PHOTO_READDONE
+    event void Photo.readDone (int val) {
+        ceu_sys_go(EVT_PHOTO_READDONE, &val);
+    }
+#endif
+
+#ifdef EVT_RADIO_SENDDONE
+    event void RadioSend.sendDone (message_t* msg) {
+        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
+    }
+#endif
+
+#ifdef EVT_RADIO_RECEIVE
+    event message_t* RadioReceive.receive (message_t* msg) {
+        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
+        return msg;
+    }
+#endif
+
+    <...>   // other events
+}
+\end{lstlisting}
+%\rule{14cm}{0.37pt}
+\caption{
+%{\small %\textmd{
+The \emph{TinyOS} binding for \CEU.
+This platform-dependent template includes the \C files generated from the 
+original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
+\emph{Final generation} phase of Figure~\ref{fig.impl}.
+%}%}
+\label{lst.impl.tinyos}
+}
+\end{figure}
+
 \subsection{The Terra Virtual Machine}
@@
 The main difference between the standard \C back end and the Terra \VM is the 
-\emph{Code Generation} phase (as illustrated in Figure~\ref{fig.impl}), which 
-here outputs assembly instructions for the \VM, instead of statements in \C.
+\emph{code generation phase}, which here outputs assembly instructions for the 
+\VM, instead of statements in \C.
 %

### 2015-10-10 paper.tex
Subfiguras (vs left/right)

@@
 Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
-implementations in Esterel and \CEU for the following control 
+implementations in Esterel (a) and \CEU (b) for the following control 
 specification~\cite{esterel.primer}:
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
-// ESTEREL
 loop
@@
 end
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(a) Esterel}
 \end{minipage}
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3.5em]
-// CEU
 loop do
@@
 \end{lstlisting}
+\centering\small{(b) \CEU}
 \end{minipage}
@@
 Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
-Esterel and \CEU.
+Esterel (a) and \CEU (b).
 %
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
-// ESTEREL
 input A;    // external
@@
 \end{lstlisting}
+\centering\small{(a) Esterel}
 \end{minipage}
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3.1em]
-// CEU
 input void A;  // external (in uppercase)
@@
 %
-In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
-\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
-reacts to different input events (ln. 4,7), which cannot occur simultaneously 
-(according to Section~\ref{sec.ceu.ints}).
-%
-However, the program in the right is non-deterministic, because the two 
-assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
-(ln. 4,7).
+In the example of Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
+(ln.  5,8) can never run concurrently, because each trail segment reacts to 
+different input events (ln. 4,7), which cannot occur simultaneously (according 
+to Section~\ref{sec.ceu.ints}).
+%
+However, the example of Figure~\ref{lst.shared}.b is non-deterministic, because 
+the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
+\code{A} (ln. 4,7).
 
@@
 \end{lstlisting}
+\centering\small{(a)}
 \end{minipage}
@@
 \end{lstlisting}
+\centering\small{(b)}
 \end{minipage}
@@
 \caption{ Shared-memory concurrency in \CEU:
-the code in the left is safe because the trails access \code{x} atomically in 
-different reactions;
-the code in the right is unsafe because both trails access \code{y} in the same 
-reaction.
+Example (a) is safe because the trails access \code{x} atomically in different 
+reactions;
+Example (b) is unsafe because both trails access \code{y} in the same reaction.
 \label{lst.shared}
@@
 %
-Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
-suspicious, the assignments to \code{y} are both atomic and deterministic, 
-i.e., after the reaction to \code{A} terminates, the value of \code{y} is $4$ 
-(\code{(1+1)*2}).
+Therefore, even though the example of Figure~\ref{lst.shared}.b is suspicious, 
+the assignments to \code{y} are both atomic and deterministic, i.e., after the 
+reaction to \code{A} terminates, the value of \code{y} is $4$ (\code{(1+1)*2}).
 %
@@
 Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
-in Esterel and \CEU to summarize the semantic difference regarding 
+in Esterel (a) and \CEU (b) to summarize the semantic difference regarding 
 (non-)determinism.
@@
 \begin{lstlisting}[mathescape=true]
-// ESTEREL
 input A;
@@
 \begin{lstlisting}
-// CEU
 input void A;
@@
 <...>
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(b)}
 \end{minipage}
 %\rule{8.5cm}{0.37pt}
-\caption{The unsafe network protocol in the left, which does not compile, is 
-extended with a finalization clause in the right to handle abortion properly.
+\caption{
+(a) The unsafe network protocol does not compile.
+(b) The finalization clause extends the protocol to handle abortion properly.
 %\caption{ Parallel compositions can describe complex state machines.
@@
 
-To illustrate the risks related to abortion, consider the unsafe example in the 
-left of Figure~\ref{lst.fin}, which does not compile in \CEU.
+To illustrate the risks related to abortion, consider the unsafe example of 
+Figure~\ref{lst.fin}.a, which does not compile in \CEU.
 It describes the state machine of the data collection protocol \emph{CTP} for 
@@
 
-The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
-\CEU tracks the interaction of \code{par/or} compositions with local variables 
-and stateful output events calls in order to preserve safe abortion of trails.
+The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
+tracks the interaction of \code{par/or} compositions with local variables and 
+stateful output events calls in order to preserve safe abortion of trails.
 %
@@
 \CEU enforces a \emph{finalization} clause to accompany the output request.
-The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
-transmission when the block of \code{buffer} goes out of scope, i.e., the 
-finalization clause (after the \code{with}, ln. 18) executes automatically on 
-external abortion.%
+The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
+when the block of \code{buffer} goes out of scope, i.e., the finalization 
+clause (after the \code{with}, ln. 18) executes automatically on external 
+abortion.%
 \footnote{
@@
 \begin{lstlisting}
-// ESTEREL
 abort
@@
 \begin{lstlisting}
-// CEU
 par/or do
@@
 automatically, resulting in more robust applications.
-For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
-first \code{await} request, the underlying system gets busy and takes 15ms to 
-check for expiring awaits.
+For the example of Figure~\ref{lst.timers}.a, suppose that after the first 
+\code{await} request, the underlying system gets busy and takes 15ms to check 
+for expiring awaits.
 The \CEU scheduler will notice that the \code{await 10ms} has not only already 
@@
 v = 2;
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(a)}
 \end{minipage}
@@
 be added and compared.
-For instance, for the program in the right of Figure~\ref{lst.timers}, although 
-the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
-it can at least ensure that the program always terminates with \code{v=1}.
+For instance, for the example of Figure~\ref{lst.timers}.b, although the 
+scheduler cannot guarantee that the first trail terminates exactly in 11ms, it 
+can at least ensure that the program always terminates with \code{v=1}.
 %
@@
 the concrete program in \CEU, which communicates the value $1$ between the 
-\code{emit} and \code{await} in parallel (left-most code).
-In the intermediate translation, we include the shared variable \code{e\_} to 
-hold the value being communicated between the two trails in order to simplify 
-the \code{emit}.
-Finally, we convert the program into the equivalent in the abstract syntax, 
+\code{emit} and \code{await} in parallel (a).
+In the intermediate translation (b), we include the shared variable \code{e\_} 
+to hold the value being communicated between the two trails in order to 
+simplify the \code{emit}.
+Finally, we convert the program into the equivalent in the abstract syntax (c), 
 translating side-effect statements into $mem$ expressions.
@@
 end
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(a)}
 \end{minipage}
@@
 await(e) ; mem ; mem
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(c)}
 \end{minipage}
@@
 expressions.
-The concrete code in the left communicates the value \code{1} from the 
-\code{emit} to the \code{await}.
-The abstract code uses a shared variable to hold the value.
+The concrete code in (a) communicates the value \code{1} from the \code{emit} 
+to the \code{await}.
+The abstract code in (c) uses a shared variable to hold the value.
 \label{lst.map.emit.await}
@@
 two-step translation of Figure~\ref{lst.map.timers.2}.
-In the concrete program in the left, the variable \code{dt} holds the residual 
-delta time (as described in Section~\ref{sec.ceu.timers}) after awaking from 
-the timer.
-In the first step of the translation, we expand the \code{await 10ms} to a 
+In the concrete program (a), the variable \code{dt} holds the residual delta 
+time (as described in Section~\ref{sec.ceu.timers}) after awaking from the 
+timer.
+In the first step of the translation (b), we expand the \code{await 10ms} to a 
 \code{loop} that decrements the elapsed number of microseconds for each 
@@
 the variable \code{dt} to contain the appropriate delta.
-In the last step, we convert the program to the abstract syntax.
+In the last step (c), we convert the program to the abstract syntax.
 
@@
 dt = await 10ms;
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(a)}
 \end{minipage}
@@
 end
-\end{lstlisting}
+.
+\end{lstlisting}
+\centering\small{(b)}
 \end{minipage}
@@
 \end{lstlisting}
+\centering\small{(c)}
 \end{minipage}
@@
 In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
-concrete program (\code{CODE-1}) that uses a \code{finalize} to safely 
-\code{\_release} the reference to \code{ptr} kept after the call to 
-\code{\_hold}.
+concrete program (a) that uses a \code{finalize} to safely \code{\_release} the 
+reference to \code{ptr} kept after the call to \code{\_hold}.
 In the translation, we first need to catch the outermost \code{do-end} 
 termination to run the finalization code.
-For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
-original body in parallel with a $fin$ expression to run the finalization code.
+For this, we translate the block into a \code{par/or} (b) with the original 
+body in parallel with a $fin$ expression to run the finalization code.
 Note that the $fin$ has no transition rules in the semantics, keeping the 
@@
 To deal with this issue, for each $fin$ we need a corresponding flag to keep 
-track of code that needs to be finalized (\code{CODE-3}).
+track of code that needs to be finalized (c).
 %
@@
 The complete translation substitutes the side-effect operations with $mem$ 
-expressions (\code{CODE-4}).
+expressions (d).
 
@@
 
-// CODE-1
-\end{lstlisting}
+.
+\end{lstlisting}
+\centering\small{(a)}
 \end{minipage}
@@
 
-// CODE-2
-\end{lstlisting}
+.
+\end{lstlisting}
+\centering\small{(b)}
 \end{minipage}
@@
 end
-
-// CODE-3
-\end{lstlisting}
+.
+\end{lstlisting}
+\centering\small{(c)}
 \end{minipage}
@@
 )
-// CODE-4
-\end{lstlisting}
+\end{lstlisting}
+\centering\small{(d)}
 \end{minipage}
@@
 
-The example in the left of Figure~\ref{lst.impl.ast} has a corresponding $AST$, 
-in the right of the figure, with the sets $I$ and $O$ for each node.
+The example of Figure~\ref{lst.impl.ast}.a has a corresponding $AST$, in 
+Figure~\ref{lst.impl.ast}.b, with the sets $I$ and $O$ for each node.
 The event $.$ (dot) represents the ``boot'' reaction.
@@
 Figure~\ref{lst.impl.trails} illustrates the generation process.
-For the program in the left of the figure, the compiler extracts the entry 
-points and associated trails, e.g., the label \code{Awake\_e} will execute on 
-\code{TRAIL-0} (ln. 7).
+For the program in (a), the compiler extracts the entry points and associated 
+trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln. 7).
 %
@@
 The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
-in the middle of the figure).
+in (b)).
 The state of trails translates to a vector of type \code{trail\_t} with the 
@@
 %{\small %\textmd{
-From left to right: static allocation of trails, entry-point labels, and 
-dispatch function.
-In the left, the comments identify the trail indexes inferred by the compiler.
-In the middle, each trail segment has an associated numeric identifier 
+(a) Static allocation of trails: the comments identify the trail indexes 
+inferred by the compiler;
+(b) Entry-point labels: each trail segment has an associated numeric identifier 
 generated by the compiler.
-In the right, the dispatcher uses a \code{switch} to associate each segment 
+(c) Dispatch function: uses a \code{switch} to associate each segment 
 identifier with the corresponding code to execute.
@@
 
-The code in the right of the Figure~\ref{lst.impl.trails} dispatches a trail 
-segment according to the current label to execute.
+The code in Figure~\ref{lst.impl.trails}.c dispatches a trail segment according 
+to the current label to execute.
 %
@@
 %
-The left of Figure~\ref{lst.terra.defs} shows a \CEU interface with the 
-available functionality for a customized \VM (with temperature and radio 
-components).
-The right of the figure shows the associated bindings for output events (ln.  
-1--8), input events (ln. 10--14), and system calls (ln. 16--22).
+Figure~\ref{lst.terra.defs}.a shows a \CEU interface with the available 
+functionality for a customized \VM (with temperature and radio components).
+Figure~\ref{lst.terra.defs}.b shows the associated bindings for output events 
+(ln.  1--8), input events (ln. 10--14), and system calls (ln.  16--22).
 %
@@
 function int getRadioID (void);
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{(a)}
 \end{minipage}
@@
 %{\small %\textmd{
-\CEU interface (in the left) with customized \VM (in the right).
-%
-The routine \code{VM.out} redirects all output events to the corresponding OS 
-calls (ln.  1--8).
+(a) \CEU interface with customized \VM.
+%
+(b) The routine \code{VM.out} redirects all output events to the corresponding 
+OS calls (ln.  1--8).
 %

### 2015-10-10 paper.tex
Typos
Revisao geral

@@
 % Title portion
-\title{The Design, Semantics, and Implementation of the Synchronous Reactive 
-Language \CEU}
+\title{The Design, Semantics, and Implementation of the Synchronous Language 
+\CEU}
 
@@
 
-Despite its success and influence, Esterel has a overly complex semantics that 
+Despite its success and influence, Esterel has an overly complex semantics that 
 requires ingenious static analysis to detect and refuse programs with 
-\emph{causality} and \emph{schizophrenia} 
-problems~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
+\emph{causality} and \emph{schizophrenia} problems, with extensive coverage in 
+the 
+literature~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
 %
 The complex semantics not only challenges the analysis and compilation of 
-programs, but also results in incompatible or non-compliant implementations.
-%
-More importantly, it also affects the programmer's understanding capacity, 
-which, after all, have to solve the errors when facing corner cases in 
-programs.
-%
-Another drawback of the Esterel semantics consists of non-deterministic 
-execution for intra-reaction statements, which prevents threads to share memory 
-and interact with side-effect and stateful system calls.
+programs, but also results in incompatible and non-compliant implementations.
+%
+Furthermore, it also affects the capacity of the programmer's understanding, 
+which, after all, has to solve the errors when facing corner cases.
+%
+Another drawback of the Esterel semantics consists of loose and 
+non-deterministic execution for intra-reaction statements, which prevents 
+threads to share memory and interact with side-effect and stateful system calls 
+appropriately.
+%
+For instance, although Esterel supports orthogonal abortion of 
+threads~\cite{esterel.preemption}, it doesn't offer a finalization mechanism to 
+release resources in use.
 
@@
 %
-The list that follows summarizes the semantic peculiarities of \CEU:
+The list that follows summarizes the semantic peculiarities of our design:
 %
@@
 Our implementations target resource-constrained devices, such as \emph{Arduino} 
-and \emph{MICAz} sensor nodes based on 8-bit microcontrollers, showing that the 
-peculiarities in the semantics of \CEU does not pose practical obstacles.%
+and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
 \footnote{
@@
 (Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
-Kbytes of SRAM.)}
-%:\url{http://www.atmel.com/devices/atmega328.aspx}}
+Kbytes of SRAM.)
+}, showing that the peculiarities in the semantics of \CEU does not pose 
+practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}
 
@@
 For the \VM back end, a simple application that blinks three LEDs periodically 
-occupies less than 100 bytes and can be completely transmitted in 4 radio 
-messages.
+occupies less than 100 bytes and can be transmitted over the air in a couple of 
+packets.
 
@@
 implementations in Esterel (a) and \CEU (b) for the following control 
-specification~\cite{esterel.primer}:
+specification:
 %
 \emph{``Emit an output O as soon as two inputs A and B have occurred.
-Reset this behavior each time the input R occurs''.}
+Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
 %
 The first phrase of the specification, awaiting and emitting the events, is 
-translated almost identically in the two languages (ln. 4--9, in both 
+translated almost identically in the two languages (ln. 3--8, in both 
 implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
@@
 For the second phrase, the reset behavior, the Esterel version uses a 
-\code{abort-when} (ln. 3--10), which, in this case, serves the same purpose of 
-\CEU's \code{par/or} (ln. 3--12):
+\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
+\CEU's \code{par/or} (ln. 2--11):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
@@
 Esterel makes no semantic distinctions between internal and external signals.
-In particular, programs can emit multiple external input signals, which may 
-coexist during a reaction.
-%
-In \CEU, a reaction starts from the occurrence of an external input event and 
-programs cannot emit inputs.
+In particular, programs can emit different external input signals 
+simultaneously, with all coexisting during a reaction.
+%
+In \CEU, a reaction starts from an external input event and programs cannot 
+emit inputs at all.
 Therefore, the occurring input event is unique during the entire reaction, 
@@
 %
-In Esterel, when \code{A} occurs, \code{B} is emitted (ln.  5--6) and both 
+In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
 events become active, resulting in the invocation of \code{f()} and \code{g()} 
@@
 \setlength{\itemsep}{0pt}
-\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.
-\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.
+\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
+\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
 \item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
@@
 
-\begin{figure}[b]
+\begin{figure}
 \begin{minipage}[t]{0.43\linewidth}
@@
 
-Internal events bring support for a limited form of subroutines, as depicted in 
+Internal events provide a fine-grained intra-reaction control mechanism.
+For instance, it brings a limited form of subroutines, as depicted in 
 Figure~\ref{lst.sub}.
@@
 
-\begin{figure}
+\begin{figure}[b]
 \begin{lstlisting}[numbers=left,xleftmargin=2em]
@@
 statements}~\cite{esterel.schizo,esterel.schizo2}).
-%
-The example in Figure~\ref{lst.await} illustrates this distinction, which 
+
+The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
 prevents infinite execution by design.
@@
 For instance, sometimes we need to execute a block of code immediately, and 
-then periodically, from internal event requests, as illustrated in 
+then, periodically from internal event requests, as illustrated in 
 Figure~\ref{lst.delay}.
@@
 with concurrent lines of execution sharing memory and interacting with the 
-environment through system calls.
+environment.
 
@@
 %
-Therefore, Esterel syntactically forbids sharing memory between lines of 
-execution to preserve determinism: \emph{``if a variable is written by some 
-thread, then it can neither be read nor be written by concurrent 
+Also, Esterel syntactically forbids sharing memory between lines of execution 
+to preserve determinism: \emph{``if a variable is written by some thread, then 
+it can neither be read nor be written by concurrent 
 threads''}~\cite{esterel.primer}.
@@
 %
-In the example of Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
+In the example in Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
 (ln.  5,8) can never run concurrently, because each trail segment reacts to 
@@
 %
-However, the example of Figure~\ref{lst.shared}.b is non-deterministic, because 
+However, the example in Figure~\ref{lst.shared}.b is non-deterministic, because 
 the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
@@
 \code{printf}).%
-\footnote{\CEU assumes all system calls perform side effects, unless they are 
-annotated as ``\texttt{@pure}''.}
+\footnote{\CEU assumes that all system calls perform side effects, unless they 
+are annotated as ``\texttt{@pure}''.}
 %The algorithm for the analysis holds the set of all events in preceding 
@@
 %
-Therefore, even though the example of Figure~\ref{lst.shared}.b is suspicious, 
+Therefore, even though the example in Figure~\ref{lst.shared}.b is suspicious, 
 the assignments to \code{y} are both atomic and deterministic, i.e., after the 
@@
 
-Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
-in Esterel (a) and \CEU (b) to summarize the semantic difference regarding 
+Figure~\ref{lst.det} compares two syntactically equivalent code fragments in 
+Esterel (a) and \CEU (b) to summarize the semantic differences regarding 
 (non-)determinism.
@@
 
-The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
+The introductory example in Figure~\ref{lst.abro} illustrates how synchronous 
 languages can abort awaiting lines of execution without tweaking them with 
@@
 when the block of \code{buffer} goes out of scope, i.e., the finalization 
-clause (after the \code{with}, ln. 18) executes automatically on external 
-abortion.%
+clause (ln. 18) executes automatically on external abortion.%
 \footnote{
@@
 \begin{comment}
-The code fragments of Figure~\ref{lst.abortion} shows a corner case regarding
+The code fragments in Figure~\ref{lst.abortion} shows a corner case regarding
 abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
@@
 automatically, resulting in more robust applications.
-For the example of Figure~\ref{lst.timers}.a, suppose that after the first 
+For the example in Figure~\ref{lst.timers}.a, suppose that after the first 
 \code{await} request, the underlying system gets busy and takes 15ms to check 
@@
 be added and compared.
-For instance, for the example of Figure~\ref{lst.timers}.b, although the 
+For instance, for the example in Figure~\ref{lst.timers}.b, although the 
 scheduler cannot guarantee that the first trail terminates exactly in 11ms, it 
@@
 For the sake of simplicity, we focus on the control aspects of the language, 
-leaving out side effects and \C calls (which behave like in conventional 
+leaving out side effects and system calls (which behave like in conventional 
 imperative languages).
@@
 
-The $mem(id)$ primitive represents all accesses, assignments, \C function 
-calls, and output events that affect a memory location identified by $id$.
+The $mem(id)$ primitive represents all accesses, assignments, system calls, and 
+output events that affect a memory location identified by $id$.
 %
@@
 input event, i.e., starting from the input event, how the program behaves and 
-becomes idle again for the subsequent reaction.
+becomes idle again to proceed to the subsequent reaction.
 %
@@
 top of it (we discuss how they are popped further).
-The sequence number $n$ prevents that awaiting expressions awake in the same 
-reaction they are reached (the \emph{delayed awaits} as explained in 
-Section~\ref{sec.ceu.ints}).
+The sequence number $n$, which is incremented each reaction, prevents that 
+awaiting expressions awake in the same reaction they are reached (the 
+\emph{delayed awaits} as explained in Section~\ref{sec.ceu.ints}).
 
@@
 event $id$ matches the top of the stack and it was reached in a previous 
-reaction (i.e., sequence number $m<=n$).
+reaction (i.e., sequence number $m \leq n$).
 %
@@
 With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
-rule \textbf{pop} expects its parameter to match the current stack level.
+rule \textbf{pop} expects its parameter $k$ to match the current stack level.
 This trick provides the desired stack-based semantics for internal events.
@@
 %
-Note that there is a syntactic restriction that a $fin$ body can only contain 
-$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
-reaction.
+Note also that there is a syntactic restriction that a $fin$ body can only 
+contain $mem$ expressions in sequence, i.e., they are guaranteed to execute 
+entirely within a reaction.
 
@@
 However, $emitting$ expressions are pending in lower stack indexes and should 
-eventualy resume in the ongoing reaction (see rule \textbf{pop}).
+eventually resume in the ongoing reaction (see rule \textbf{pop}).
 %
 Therefore, we define another rule that behaves as \textbf{rule-inner} 
-(presented above) if the program can advance, and, otherwise, pops the stack:
+(presented above) if the program can advance, and, otherwise, pops the stack to 
+resume the lower level:
 %
@@
 Most statements from \CEU (``concrete \CEU'') map directly to those presented 
-in the reduced syntax of Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
+in the reduced syntax in Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
 %, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
@@
 However, there are some significant mismatches between the concrete and 
-abstract \CEU, and we (informally) propose appropriate mappings in this 
+abstract \CEU, and we (informally) present appropriate mappings in this 
 section.
@@
 values between them.
-In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
-the concrete program in \CEU, which communicates the value $1$ between the 
-\code{emit} and \code{await} in parallel (a).
+In the two-step translation in Figure~\ref{lst.map.emit.await}, we start with 
+the concrete program in \CEU (a), which communicates the value $1$ between the 
+\code{emit} and \code{await} in parallel.
 In the intermediate translation (b), we include the shared variable \code{e\_} 
@@
 translating side-effect statements into $mem$ expressions.
-External events require a similar translation, i.e., each external event has a 
-corresponding variable that is explicitly set by the environment before each 
+External events have a similar translation, i.e., each external event requires 
+a corresponding variable that is explicitly set by the environment before each 
 reaction.
@@
 wall-clock time elapsed between two occurrences in sequence, as depicted by the 
-two-step translation of Figure~\ref{lst.map.timers.2}.
+two-step translation in Figure~\ref{lst.map.timers.2}.
 In the concrete program (a), the variable \code{dt} holds the residual delta 
@@
 for a proper mapping using $fin$ expressions.
-In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
+In the three-step translation in Figure~\ref{lst.map.fin}, we start with a 
 concrete program (a) that uses a \code{finalize} to safely \code{\_release} the 
@@
 the program until it blocks and no rules can be applied.
-The \emph{isBlocked} predicate of Figure~\ref{fig.isBlocked} identifies a 
+The \emph{isBlocked} predicate in Figure~\ref{fig.isBlocked} identifies a 
 blocked program based on its structure and event at the top of the stack.
@@
 the \emph{code generation phase} converts the \emph{AST} to standard \C code 
-and packs it with platform-dependent functionality and the runtime of \CEU, 
-compiling everything with \emph{gcc} to generate the final binary.
+and packs it with platform-dependent functionality (e.g., system calls) and the 
+runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
+binary.
+
+In the sections that follow, we discuss the implementation insights related to 
+the peculiarities of \CEU:
+the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
+static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
+and~\ref{sec.impl.trails}),
+static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
+interaction with the environment (Section~\ref{sec.impl.environment}),
+and the \VM back end (Section~\ref{sec.terra}),
 
 \subsection{Temporal Analysis for Shared-Memory Concurrency}
+\label{sec.impl.analysis}
 
@@
 
-The example of Figure~\ref{lst.impl.ast}.a has a corresponding $AST$, in 
+The example in Figure~\ref{lst.impl.ast}.a has a corresponding $AST$, in 
 Figure~\ref{lst.impl.ast}.b, with the sets $I$ and $O$ for each node.
@@
 
-\subsection{Memory Layout}
-\label{sec:impl:memory}
+\subsection{Static Memory Layout}
+\label{sec.impl.data}
 
@@
 
-\subsection{Trail Allocation}
-\label{sec:impl:gates}
+\subsection{Static and Lightweight Trail Allocation}
+\label{sec.impl.trails}
 
@@
 simultaneous lines of execution with an overhead of 2\% in comparison to the 
-original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \C for 
-event-driven programming)~\cite{ceu.sensys13}.
-
-\subsection{Code Generation and Scheduling}
+original version in \C~\cite{ceu.sensys13}.
+
+\subsection{Static Scheduling and Trail Finalization}
+\label{sec.impl.scheduler}
 
@@
 given that the \code{Main} label is the only one that can be active at the 
-stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).
+stack level 0 (ln. 13, in Figure~\ref{lst.impl.trails}.b).
 
@@
 \subsection{Interaction with the Environment}
+\label{sec.impl.environment}
 
@@
 As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
-\emph{TinyOS}~\cite{wsn.tos}, which maps driver callbacks to input events in 
+\emph{TinyOS}~\cite{wsn.tos}, which maps system callbacks to input events in 
 \CEU.
@@
     event void Boot.booted () {
-        ceu_sys_init();
+        ceu_init();
 #ifdef CEU_WCLOCKS
@@
     event void Timer.fired () {
-        ceu_sys_wclock(10000);
+        ceu_wclock(10000);
     }
@@
     event void Photo.readDone (int val) {
-        ceu_sys_go(EVT_PHOTO_READDONE, &val);
+        ceu_go(EVT_PHOTO_READDONE, &val);
     }
@@
     event void RadioSend.sendDone (message_t* msg) {
-        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
+        ceu_go(EVT_RADIO_SENDDONE, &msg);
     }
@@
     event message_t* RadioReceive.receive (message_t* msg) {
-        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
+        ceu_go(EVT_RADIO_RECEIVE, &msg);
         return msg;
@@
 \CEU also distinguishes external events from stack-based internal events, which 
-provide a limited form of coroutines supporting reactive statements (e.g., 
-\code{await} and \code{par/or}).
-
-The event-driven approach of \CEU is well known~\cite{sync_async.whynotthreads} 
-and popular in many software communities, such as client and server-side web 
-frameworks (e.g., \emph{jQuery}~\cite{js.jquery} and 
-\emph{Node.js}~\cite{js.node}), GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} 
-and \emph{Java Swing}~\cite{java.swing}), and Games~\cite{gamepatterns}.
+provide a limited form of coroutines supporting reactive statements.
+
+The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
+and popular in many software communities, such as web frameworks (e.g., 
+\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
+(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
+and Games~\cite{gamepatterns}.
 %
@@
 \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
-\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
+\emph{PRET-C}~\cite{rp.pretc} and \emph{Sync-C}~\cite{rp.synchc}.
 %
@@
 order for threads, based on either explicit priorities, or source code lexical 
-order.
+order (similar to \CEU).
 %
@@
 Our experience shows that students take advantage of the sequential-imperative 
-style of \CEU and can implement non-trivial concurrent applications in a few of 
+style of \CEU and can implement non-trivial concurrent applications in a few 
 weeks.

### 2015-10-13 paper.tex
(*) submitted

@@
 Despite its success and influence, Esterel has an overly complex semantics that 
-requires ingenious static analysis to detect and refuse programs with 
-\emph{causality} and \emph{schizophrenia} problems, with extensive coverage in 
-the 
-literature~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
+requires careful static analysis to detect and refuse programs with 
+\emph{causality} and \emph{schizophrenia} problems.
+Both problems have an extensive coverage and debate in the literature 
+~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
 %
@@
 %
-Furthermore, it also affects the capacity of the programmer's understanding, 
+Furthermore, it also affects the programmer's understanding about the code, 
 which, after all, has to solve the errors when facing corner cases.
 %
-Another drawback of the Esterel semantics consists of loose and 
+Another drawback of the Esterel semantics consists of the loose and 
 non-deterministic execution for intra-reaction statements, which prevents 
@@
 For instance, although Esterel supports orthogonal abortion of 
-threads~\cite{esterel.preemption}, it doesn't offer a finalization mechanism to 
+threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
 release resources in use.
@@
 %
-\item Stack-based execution for internal events, which provide a limited form 
+\item Stack-based execution for internal events, which provides a limited form 
 of coroutines.
 %
-\item A static temporal analysis and deterministic execution semantics that 
+\item Static temporal analysis and deterministic execution semantics that 
 allows programs to safely share memory.
 %
-\item A finalization mechanism for safe abortion of lines of execution holding 
+\item Finalization mechanism for safe abortion of lines of execution holding 
 external resources.
@@
 \footnote{
-Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
-\\
-MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
-\\
-(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
-Kbytes of SRAM.)
+Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
+microcontroller with 32K of FLASH and 2K of SRAM.
 }, showing that the peculiarities in the semantics of \CEU does not pose 
@@
 %
-For the \VM back end, a simple application that blinks three LEDs periodically 
-occupies less than 100 bytes and can be transmitted over the air in a couple of 
-packets.
+For the \VM back end, applications have a bytecode footprint in the order of 
+hundreds of bytes can be transmitted over the air in a few 
+packets~\cite{ceu.terra}.
 
@@
 
-\begin{figure}[b]
+\begin{figure}
 \begin{lstlisting}[numbers=left,xleftmargin=2em]
@@
 re-execution of statements in the same reaction (known as \emph{schizophrenic 
-statements}~\cite{esterel.schizo,esterel.schizo2}).
+statements}~\cite{esterel.constructive}).
 
@@
 %
-On the one hand, we transfer the burden of dealing with these corner cases to 
-the programmer.
+On the one hand, we transfer the burden of dealing with these specific corner 
+cases to the programmer.
 On the other hand, we simplify the semantics of the language and eliminate the 
-need for analysis to deal with schizophrenic statements.
+need for complex analysis to deal with schizophrenic statements.
 
@@
 %
-\caption{ An example that circumvents the \emph{delayed await} restriction by 
-post-fixing the \code{await} inside the \code{loop}.
+\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
+the \code{await} inside the \code{loop}.
 % (in the left), and by copying the condition test (in the right).
@@
 The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
-tracks the interaction of \code{par/or} compositions with local variables and 
-stateful output events calls in order to preserve safe abortion of trails.
+tracks the interaction of \code{par/or} compositions with pointers to local 
+variables and stateful output events calls in order to preserve safe abortion 
+of trails.
 %
@@
 operational semantics to formally describe the language.
-We describe a small synchronous kernel with broadcast communication
-highlighting the peculiarities of \CEU, in particular the stack-based
-execution for internal events.
+We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
+in particular the stack-based execution for internal events.
 %
@@
 Finally, to describe the complete execution of a program, we trigger multiple 
-``invocations'' of reaction chains in sequence:
+``invocations'' of reactions in sequence:
 %
@@
 %
-Each invocation starts with an external event at the top of the stack and 
-finishes with a modified program and an empty stack.
+Each invocation starts with the occurring external event at the top of the 
+stack and finishes with a modified program and an empty stack.
 After each invocation, we increment the sequence number.
@@
 
-\begin{comment}
-
-There are also some restriction on valid programs in \CEU, which the semantic 
-allows:
-- loops
-- awaits in fins
-- local scopes
-
-handled in separate, in a parsing phase that is not related to the execution
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The semantic rules are continuously applied (note the `$*$' klenee operator) 
-until
-consider only the top of the stack and continuously apply transformations to 
-the program until it blocks and no rules can be applied.
-The \emph{isBlocked} predicate in Figure~\ref{fig.isBlocked} identifies a 
-blocked program based on its structure and event at the top of the stack.
-xxx
-A program becomes blocked when all parallel branches are hanged in $awaiting$, 
-$stacked$, and/or $emitting$ primitives, as defined in 
-Figure~\ref{fig:isBlocked}.
-an \code{awaiting} is unblocked only if its event matches the top of the stack
-the \code{emitting} primitive only proceeds once its stack level is restored.
-
-\begin{itemize}
-\item The program is awaiting in all trails, i.e., function $pop$ returns 
-$(0,\{\})$.
-\item The program terminates, i.e., the small-step rules transform the whole 
-program into a $nop$.
-\end{itemize}
-%
-In Section~\ref{XXX} we show that by imposing syntactic restrictions to 
-programs, reaction chains always reach one of these conditions in a finite 
-number of steps, meaning that reactions to the environment always execute in 
-bounded time.
-
-To be compliant with the reactive nature of \CEU, we assume that all programs 
-start awaiting the main event ``$\$$'', which is emitted once by the 
-environment on startup, i.e., $(i,E)=(1,\{\$\})$ for the very first big step.
-
-As briefly introduced, small-step rules continuously apply transformations to 
-unblocked trails.
-A program becomes blocked when all parallel branches are hanged in $awaiting$, 
-$stacked$, and/or $emitting$ primitives, as defined in 
-Figure~\ref{fig:isBlocked}.
-
-All small-step rules are associated with the current (deepest) stack depth 
-level $i$ acquired from the previous big step.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{comment}
-
 \section{Implementation}
@@
 
-In the sections that follow, we discuss the implementation insights related to 
-the peculiarities of \CEU:
+In the sections that follow, we discuss the implementation highlights related 
+to the peculiarities of \CEU:
 the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
@@
 original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
-\emph{Final generation} phase of Figure~\ref{fig.impl}.
+\emph{code generation phase}.
 %}%}
@@
 %
-The \emph{Sun SPOT} platform with the \emph{Squawk JVM} brings Java for the 
-embedded domain~\cite{wsn.sunspot}, but requires a much powerful hardware.%
-\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
-KBytes of SRAM.}
+%The \emph{Sun SPOT} platform with the \emph{Squawk JVM} brings Java for the 
+%embedded domain~\cite{wsn.sunspot}, but requires a much powerful hardware.%
+%\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
+%KBytes of SRAM.}
 %
 \emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
-are complete \emph{JVMs} targeting constrained embedded systems with support 
-for multithreading and garbage collection.
+are complete \emph{Java VMs} targeting constrained embedded systems with 
+support for multithreading and garbage collection.
 %
@@
 it does not impose static bounds on memory usage and execution time, and 
-provides preemptive multithreading with synchronization primitives for 
-accessing shared memory.
+provides preemptive multithreading which requires synchronization primitives 
+for accessing shared memory.
 %

### 2016-02-24 formsem.tex

\documentclass[11pt,a4paper,oneside]{amsart}
\title{Notes on the semantics of Céu}
\renewcommand{\datename}{}
\date{Guilherme F.~Lima, \today}
%---------------------------------------------------------------------------
% TODO:
% 1. Adicionar as instruções @\bot (erro) e @\top (terminação).
% 2. Tornar a instrução 'fin p' bloqueante.
% 3. Tentar tirar o número n da relação \step.
% 4. Adicionar sequência de saída (emit p. eventos externos) à configuração.
% 5. Completar as provas: determinismo, terminação, ordenação das
%    finalizações.
% 6. Definir relação de equivalência entre programas.
%
% FUTURO:
% - Comparar as semânticas construtivas de Esterel.
% - Colocar comportamentos (behaviors, instrução spawn) na semântica.
%---------------------------------------------------------------------------
\usepackage[protrusion=true]{microtype}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{aliascnt}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{hyperref}
\hypersetup{colorlinks=true,allcolors=black}
\usepackage{mathtools}
\usepackage[varg]{txfonts}
\usepackage{xcolor}
\usepackage[nameinlink]{cleveref}

% To-do.
\def\FIXME#1{{\color{red}{\textbf{#1}}}}
\newenvironment{TODO}
  {\bigskip\bgroup\color{red}\textbf{TODO}.~}
  {\egroup\bigskip}

% Theorems.
\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\Crefname{thm}{Theorem}{Theorems}

\theoremstyle{definition}
\newaliascnt{defn}{thm}
\newtheorem{defn}[defn]{Definition}
\aliascntresetthe{defn}
\Crefname{defn}{Definition}{Definitions}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\Crefname{remark}{Remark}{Remarks}

% Bussproofs.
\usepackage{bussproofs}
\def\labelSpacing{0em}
\def\ScoreOverhang{0em}
\def\defaultHypSeparation{\quad}
\EnableBpAbbreviations

% Auxiliary.
\def\R#1{\hyperref[Rule:#1]{$R_\text{#1}$}}
\def\Rtag#1{\tag{$R_\text{#1}$}}

% Symbols.
\def\Ceu{C\'eu}
\let\nil=\varepsilon
\def\<#1>{\langle#1\rangle}
\def\|#1|{\left|#1\right|}
\def\val{\mathop{\mathit{val}}}
\let\clear=\kappa
\makeatletter
\def\@raise#1#2#3{
  \setbox0=\hbox{#1}%
  \mathbin{%
    \hbox to\wd0{%
      \rlap{\box0}\hfill\raise#2\hbox{$\scriptstyle{#3}$}\hfill
    }%
  }%
}
\def\step#1{\@raise{$\to$}{1.1ex}{#1}}
\def\stepx#1{\@raise{$\step{#1}$}{-.75ex}{\ast}}
\def\eval#1{\@raise{$\Rightarrow$}{1.1ex}{#1}}
\def\quadtext#1{\quad\text{#1}\quad}
\def\Case#1.{[Case~#1]\enskip}
\makeatother

% Abstract syntax.
\makeatletter
\def\@ceuop#1{\mathop{\texttt{#1}}}%
\def\@ceubin#1{\mathbin{\texttt{#1}}}%
\def\ceu#1{%
  \bgroup
  \def\Skip{\@ceuop{skip}}%
  \def\Mem{\@ceuop{mem}}%
  \def\Await{\@ceuop{await}}%
  \def\Emit{\@ceuop{emit}}%
  \def\Break{\@ceuop{break}}%
  \def\Ifelse##1##2##3{%
    \@ceuop{if}%
    \@ceuop{mem}(##1)%
    \@ceuop{then}{##2}%
    \@ceuop{else}{##3}}%
  \def\Loop{\@ceuop{loop}}%
  \def\And{\@ceubin{and}}%
  \def\Or{\@ceubin{or}}%
  \def\Fin{\@ceuop{fin}}%
  \def\Awaiting{\@ceuop{@awaiting}}%
  \def\Emitting{\@ceuop{@emitting}}%
  \def\Atloop{\@ceuop{@loop}}%
  \ensuremath{#1}\ignorespaces
  \egroup
}
\makeatother

\begin{document}
\maketitle

\section{Abstract syntax}
\label{Section:syntax}

The \emph{abstract syntax} of \Ceu\ programs is given by the following
grammar:
%%
\bgroup
\vskip\abovedisplayskip
\noindent
\hfil\hbox{%
  \vtop{%
    \tabskip0pt
    \offinterlineskip
    \halign{\strut\hfil$#$&$\;#$\hfil&\qquad#\hfil\cr
      p\in{P}\Coloneqq
          & \ceu{\Mem(v)}         & read or write variable~$v$\cr
      \mid& \ceu{\Await(e)}       & await event~$e$\cr
      \mid& \ceu{\Emit(e)}        & emit event~$e$\cr
      \mid& \ceu{\Break}          & break innermost loop\cr
      \mid& \ceu{\Ifelse{v}{p_1}{p_2}}  & conditional\cr
      \mid& \ceu{p_1;p_2}         & sequence\cr
      \mid& \ceu{\Loop p_1}       & repetition\cr
      \mid& \ceu{p_1\And p_2}     & par/and\cr
      \mid& \ceu{p_1\Or p_2}      & par/or\cr
      \mid& \ceu{\Fin p_1}        & finalization\cr
      \mid& \ceu{\Awaiting(e,n)}  & awaiting~$e$ since reaction~$n$\cr
      \mid& \ceu{\Emitting(n)}    & emitting~$e$ on stack level~$n$\cr
      \mid& \ceu{p_1\Atloop p_2}  & unwinded loop,\cr
    }%
  }%
}\hfil%
\vskip\belowdisplayskip
\egroup
%%
\noindent where~$n\in{N}$ is an integer, $v\in{V}$~is a memory location
(variable) identifier, $e\in{E}$~is an event identifier, and~$p$, $p_1$,
$p_2\in{P}$ are programs.  We use the abbreviation~``\ceu\Skip'' (do
nothing) for the vacuous memory access~``\ceu{\Mem(\nil)}'', where~$\nil$ is
the empty string.  We thus assume that~$\nil\in{V}$.

\section{Reaction step}
\label{Section:step}

For our purposes, the \emph{state} of a \Ceu\ program within a reaction is
represented by a stack of event identifiers~$e_1e_2\dots{}e_n\in{}E^*$,
where the leftmost element is the top-of-stack.  A \emph{configuration} is a
triple~$\<p,\alpha,n>\in{P\times{E^*}\times{N}}=\Delta$ that represents the
situation of program~$p$ waiting to be evaluated in state~$\alpha$ and
reaction~$n$.  Given an initial configuration, each program reaction is
computed by successive applications of the reaction-step
relation~$\mathord{\to}\subseteq{\Delta\times\Delta}$ such
that~$\<p,\alpha,n>\step{}\<p',\alpha',n>$ iff: an execution step of
program~$p$ in state~$\alpha$ and reaction~$n$ evaluates to a modified
program~$p'$ and a modified state~$\alpha'$ in the same reaction.  Since
relation~$\step{}$ is defined in a way such that it can only relate
configurations with the same~$n$, we often
write~$\<p,\alpha>\step{n}\<p',\alpha'>$
for~$\<p,\alpha,n>\step{}\<p',\alpha',n>$.\footnote{The reaction number~$n$
  is used only to prevent~\ceu\Await\ instructions from awaking in the same
  reaction they are reached.}

The reaction-step relation~$\step{\null}$ is defined inductively by the
rules presented in
%%
\Cref{%
Section:step:await-and-emit,%
Section:step:conditionals,%
Section:step:sequences-and-loops,%
Section:step:parand-and-paror,%
}.

\subsection{Await and emit}
\label{Section:step:await-and-emit}

For all~$n$, $n'\in{N}$, $e\in{E}$, and~$\alpha\in{E^*}$:
\begin{alignat*}{2}
\Rtag{await}
\<\ceu{\Await(e)},\alpha>
&\step{n}\<\ceu{\Awaiting(e,n')},\alpha>
&&\qquad\text{with~$n'=n+1$}\\
%%
\Rtag{awake}
\<\ceu{\Awaiting(e,n')},e\alpha>
&\step{n}\<\ceu\Skip,e\alpha>
&&\qquad\text{if~$n'\leq{n}$}\\
%%
\Rtag{emit-push}
\<\ceu{\Emit(e)},\alpha>
&\step{n}\<\ceu{\Emitting(n')},e\alpha>
&&\qquad\text{with~$n'=\|\alpha|$}\\
%%
\Rtag{emit-pop}
\<\ceu{\Emitting(n')},\alpha>
&\step{n}\<\ceu\Skip,\alpha>
&&\qquad\text{if~$n'=\|\alpha|$\,.}
\end{alignat*}

\subsection{Conditionals}
\label{Section:step:conditionals}

For all~$n\in{N}$, $v\in{V}$, $\alpha\in{E^*}$, and~$p_1$, $p_2\in{P}$:
\begin{alignat*}{3}
\Rtag{if-true}
\<\ceu{\Ifelse{v}{p_1}{p_2}},\alpha>
&\step{n}\<p_1,\alpha>
&&\qquad\text{if~$\val(v,n)\ne0$}\\
%%
\Rtag{if-false}
\<\ceu{\Ifelse{v}{p_1}{p_2}},\alpha>
&\step{n}\<p_2,\alpha>
&&\qquad\text{if~$\val(v,n)=0$\,,}
\end{alignat*}
%%
where~$\val(v,n)$ denotes the value of variable~$v$ in reaction~$n$.  For
simplicity, and without loss of generality, we deal only with integer
variables, i.e., $\val(v,n)\in{N}$, for all~$v$ and~$n$.  Moreover, we
assume that~$\val(\nil,n)=0$, for all~$n$.

\subsection{Sequences and loops}
\label{Section:step:sequences-and-loops}

For all~$n\in{N}$, $v\in{V}$, $\alpha$, $\alpha'\in{E^*}$, and~$p$, $p_1$,
$p_1'$, $p_2\in{P}$:
\begin{alignat*}{2}
\Rtag{seq-mem}
\<\ceu{\Mem(v);p},\alpha>&\step{n}\<p,\alpha>&&\\
%%
\Rtag{seq-brk}
\<\ceu{\Break};p,\alpha>&\step{n}\<\ceu{\Break},\alpha>&&\\
%%
\Rtag{seq-adv}&\hskip-4.15em
\AXC{$\<p_1,\alpha>\step{n}\<p_1',\alpha'>$}
\UIC{$\<p_1;p_2,\alpha>\step{n}\<p_1';p_2,\alpha'>$}
\DP&&\\
%%
\Rtag{loop-expd}
\<\ceu{\Loop p},\alpha>&\step{n}\<\ceu{p\Atloop p},\alpha>&&\\
%%
\Rtag{loop-mem}
\<\ceu{\Mem(v)\Atloop p},\alpha>&\step{n}\<\ceu{\Loop p},\alpha>&&\\
%%
\Rtag{loop-brk}
\<\ceu{\Break\Atloop p},\alpha>&\step{n}\<\ceu\Skip,\alpha>&&\\
%%
\Rtag{loop-adv}&\hskip-6.7em
\AXC{$\<p_1,\alpha>\step{n}\<p_1',\alpha'>$}
\UIC{$\<\ceu{p_1\Atloop p_2},\alpha>
  \step{n}\<\ceu{p_1'\Atloop p_2},\alpha'>$}
\DP&&
\end{alignat*}

\subsection{Par/and and par/or}
\label{Section:step:parand-and-paror}

For all~$n\in{N}$, $v\in{V}$, $\alpha$, $\alpha'\in{E^*}$, and~$p_1$,
$p_1'$, $p_2$, $p_2'\in{P}$:
\begin{alignat*}{2}
\Rtag{and-mem1}
\<\ceu{\Mem(v)\And p_2},\alpha>&\step{n}\<p_2,\alpha>&&\\
%%
\Rtag{and-mem2}
\<\ceu{p_1\And\Mem(v)},\alpha>&\step{n}\<p_1,\alpha>&&\\
%%
\Rtag{and-brk1}
\<\ceu{\Break{}\And p_2},\alpha>&\step{n}\<\ceu{p_2';\Break},\alpha>
&&\quad\text{with~$p_2'=\clear(p_2)$}\\
%%
\Rtag{and-brk2}
\<\ceu{p_1 \And\Break{}},\alpha>&\step{n}\<\ceu{p_1';\Break},\alpha>
&&\quad\text{with~$p_1'=\clear(p_1)$}\\
%%
\Rtag{and-adv1}&\hskip-5.75em
\AXC{$\<p_1,\alpha>\step{n}\<p_1',\alpha'>$}
\UIC{$\<\ceu{p_1\And p_2},\alpha>\step{n}\<\ceu{p_1'\And p_2},\alpha'>$}
\DP&&\\
%%
\Rtag{and-adv2}&\hskip-5.75em
\AXC{$\<p_2,\alpha>\step{n}\<p_2',\alpha'>$}
\UIC{$\<\ceu{p_1\And p_2},\alpha>\step{n}\<\ceu{p_1\And p_2'},\alpha'>$}
\DP&&\quad\text{if~$\#(p_1,\alpha,n)$\,,}
\end{alignat*}
where the call~$\clear(p)$ (read ``clear~$p$'') concatenates in a sequence
the body~$p'$ of all active finalization instructions~\ceu{\Fin{p'}} in~$p$,
and the predicate~$\#(p,\alpha,n)$ (read ``$p$ is blocked'') is true iff all
execution trails of~$p$ are hanged in instructions that cannot be consumed
in the current reaction step, viz., \ceu\Awaiting\ or \ceu\Emitting.  See
\Cref{Appendix:extra:clear,Appendix:extra:is-blocked} for the precise
definition of these functions.

The evaluation rules for par/or instructions (below) are slightly different.
While in the case par/and both sides must terminate (evaluate
to~\ceu{\Mem(v)} or~\ceu\Break) for the composition to terminate, in the
case of par/or the termination of \emph{any} side causes the whole
composition to terminate.  Note, however, that in both cases, par/and and
par/or, the right-hand side of the instruction can only be evaluated if the
its left-hand side is blocked.

For all~$n\in{N}$, $v\in{V}$, $\alpha$, $\alpha'\in{E^*}$, and~$p_1$,
$p_1'$, $p_2$, $p_2'\in{P}$:
\begin{alignat*}{2}
\Rtag{or-mem1}
\<\ceu{\Mem(v)\Or p_2},\alpha>&\step{n}\<\ceu{p_2'},\alpha>
&&\enspace\text{with~$p_2'=\clear(p_2)$}\\
%%
\Rtag{or-mem2}
\<\ceu{p_1\Or\Mem(v)},\alpha>&\step{n}\<\ceu{p_1'},\alpha>
&&\enspace\text{if~$\#(p_1,\alpha,n)$, with~$p_1'=\clear(p_1)$}\\
%%
\Rtag{or-brk1}
\<\ceu{\Break{}\Or p_2},\alpha>&\step{n}\<\ceu{p_2';\Break},\alpha>
&&\enspace\text{with~$p_2'=\clear(p_2)$}\\
%%
\Rtag{or-brk2}
\<\ceu{p_1\Or\Break},\alpha>&\step{n}\<\ceu{p_1';\Break},\alpha>
&&\enspace\text{if~$\#(p_1,\alpha,n)$, with~$p_1'=\clear(p_1)$}\\
%%
\Rtag{or-adv1}&\hskip-5.2em
\AXC{$\<\ceu{p_1},\alpha>\step{n}\<p_1',\alpha'>$}
\UIC{$\<\ceu{p_1\Or p_2},\alpha>\step{n}\<\ceu{p_1'\Or p_2},\alpha'>$}
\DP&&\\
%%
\Rtag{or-adv2}&\hskip-5.2em
\AXC{$\<\ceu{p_2},\alpha>\step{n}\<p_2',\alpha'>$}
\UIC{$\<\ceu{p_1\Or p_2},\alpha>\step{n}\<\ceu{p_1\Or p_2'},\alpha'>$}
\DP&&\enspace\text{if~$\#(n,\alpha,p_1)$}
\end{alignat*}

\section{Properties of the reaction-step relation}
\label{Section:step-properties}

\begin{thm}[Determinism]
For all~$p$, $p_1$, $p_2\in{P}$, $\alpha$, $\alpha_1$, $\alpha_2\in{E^*}$,
and~$n\in{N}$:
\begin{equation*}
(\<p,\alpha>\step{n}\<p_1,\alpha_1>
\quadtext{and}
\<p,\alpha>\step{n}\<p_2,\alpha_2>)
\quadtext{implies}
\<p_1,\alpha_1>=\<p_2,\alpha_2>\,.
\end{equation*}
\end{thm}
\begin{proof}
Suppose~$\<p,\alpha>\step{n}\<p_1,\alpha_1>$
and~$\<p,\alpha>\step{n}\<p_2,\alpha_2>$, for arbitrary~$p$, $p_1$,
$p_2\in{P}$, $\alpha$, $\alpha_1$, $\alpha_2\in{E^*}$, and~$n\in{N}$.  We
proceed by structural induction on~$p$.  The following ten cases are
possible.  (The cases where~$p=\ceu{\Mem(v)}$, $p=\ceu{\Break}$, and
$p=\ceu{\Fin{p'}}$ are not considered since there are no rules to evaluate
them, i.e., by the theorem hypothesis, $p\ne\ceu{\Mem(v)}$,
$p\ne\ceu\Break$, and~$p\ne\ceu{\Fin{p'}}$.)

\Case{1}. $p=\ceu{\Await(e)}$, for some~$e\in{E}$.  By rule~\R{await},
$p_1=p_2=\ceu{\Awaiting(e,n')}$, with~$n'=n+1$,
and~$\alpha_1=\alpha_2=\alpha$.

\Case{2}. $p=\ceu{\Awaiting(e,n')}$, for some~$e\in{E}$.  By rule~\R{awake},
$p_1=p_2=\ceu\Skip$, with~$n'\leq{n}$, and~$\alpha_1=\alpha_2=\alpha$.

\Case{3}. $p=\ceu{\Emit(e)}$, for some~$e\in{E}$.
\FIXME{\dots}

\Case{4}. $p=\ceu{\Emitting(n')}$, for some~$n'\in{N}$.
\FIXME{\dots}

\Case{5}. $p=\ceu{\Ifelse{v}{p'}{p''}}$, for some~$v\in{V}$ and~$p'$,
$p''\in{P}$.
\FIXME{\dots}

\Case{6}. $p=\ceu{p';p''}$, for some~$p'$, $p''\in{P}$.
\FIXME{\dots}

\Case{7}. $p=\ceu{\Loop p'}$, for some~$p'\in{P}$.
\FIXME{\dots}

\Case{8}. $p=\ceu{p'\Atloop p''}$, for some~$p'$, $p''\in{P}$.
\FIXME{\dots}

\Case{9}. $p=\ceu{p'\And p''}$, for some~$p'$, $p''\in{P}$.
\FIXME{\dots}

\Case{10}. $p=\ceu{p'\Or p''}$, for some~$p'$, $p''\in{P}$.
\FIXME{\dots}

\begin{TODO}
Aqui eu esbarro no primeiro problema.  A indução em~$P$ fica
  complicada porque há programas que não avaliam para nada.  Por exemplo,
  se~$p=\ceu{\Ifelse{v}{p'}{p''}}$ com $p'=\ceu\Break$, então a hipótese
  indutiva não vale~$p'$, mesmo com~$p'\prec{p}$.  Ou seja, eu teria que
  identificar todos os casos em que o~$p'$ da hipótese indutiva não vale.
  Eu vejo duas soluções simples para esse problema.

  A primeira solução é fazer com que todo programa avalie para alguma coisa,
  e.g., definir regras do tipo:
  \begin{align*}
    \<\ceu{\Mem(v)},\alpha>&\step{n}\<\ceu\Skip,\alpha>\\
    \<\ceu\Break,\alpha>&\step{n}\<\ceu\Break,\alpha>\\
    \<\ceu{\Fin{p_1}},\alpha>&\step{n}\<\ceu{\Fin{p_1}},\alpha>\,.\\
  \end{align*}
  Nesse caso, teríamos que ver se as regras acima são suficientes e se elas
  não causam conflitam com as existentes.

  A segunda solução, é deixar em~$P$ apenas os programas que avaliam para
  alguma coisa.  Ou seja, \ceu{\Mem(v)} e \ceu\Break\ não fariam parte
  de~$P$ mas sim de um conjunto auxiliar.  Dessa forma, quando eu fizesse
  uma indução em~$P$ não teria que lidar com esses casos.  Voltado ao
  exemplo, \ceu{\Ifelse{v}{\Break}{p''}} seria tratado como um caso base, já
  que na gramática eu teria algo do tipo:
  \begin{align*}
    p\Coloneqq\ceu{\Ifelse{v}{\Break}{p_2}}\mid\cdots{}
  \end{align*}
  A desvantagem dessa segunda opção é que o número de regras pode aumentar
  significativamente.

  Qualquer que seja a solução, se eu tenho~$\<p,\alpha>$ sempre avaliando
  para alguma coisa, na hora de fazer o fecho (i.e., na definição da
  relação~$\eval{}$) eu não preciso do fecho reflexivo---o transitivo basta.
\end{TODO}
\end{proof}

\begin{thm}[Termination]
For all~$\<p,\alpha>\in\Delta$,
\begin{equation*}
(\exists!\<p',\alpha'>\in\Delta)(\<p,\alpha>\step{n}\<p',\alpha'>\,).
\end{equation*}
\end{thm}
\begin{proof}
By structural induction on~$P$.

\begin{TODO}
Aqui eu caio no memso problema anterior.  Há programas em~$P$ que não
avaliam para nada.
\end{TODO}
\end{proof}

\section{Reaction}
\label{Section:reaction}

\begin{TODO}
Eu tinha definido a seguinte regra:
%%
\begin{quote}
  For all~$n\in{N}$, $e\in{E}$, $\alpha\in{E^*}$, and~$p\in{P}$:
  \label{Rule:pop}
  \begin{equation*}
    \Rtag{pop}\hskip1.7em
    \<\ceu{p},e\alpha>\step{n}\<\ceu{p},\alpha>
    \qquad\qquad\text{if~$\#(p,e\alpha,n)$\,.}
  \end{equation*}
  %%
  This rule is necessary to ensure that pending~\ceu\Emitting\ instructions
  (in lower stack indexes) are eventually resumed by subsequent reaction
  steps.
\end{quote}
Dessa forma, o ``pop'' faria parte da própria definição de passo.  Mas
depois eu percebi que essa regra complica (inviabiliza?) qualquer indução
simples em~$P$ (acho que eu teria que fazer uma indução em
pares~$\<p,\alpha>$).  Talvez ela tenha que ficar fora mesmo---i.e., na
definição da relação~$\eval{}$, igual o original---ou talvez haja alguma
forma de implementá-la em~$\step{}$ sem complicar as provas.
\end{TODO}

The reaction-step relation ($\step{}$), defined in \Cref{Section:step}
computes a single step within a reaction.  To compute a full reaction we now
define the reaction relation~$\eval{n}\subseteq\Delta\times{P}$ such
that~$\<p,\alpha>\eval{n}{p'}$ iff program~$p$ in initial state~$\alpha$
evaluates (reacts) to a modified program~$p'$ in the empty state in a finite
number of steps.  More formally, for all~$\alpha\in{E}$ and~$p$, $p'\in{P}$:
%%
\begin{equation*}
  \<p,\alpha>\eval{n}{p'}\quadtext{iff}\<p,\alpha>\stepx{n'}\<p,\nil>\,,
\end{equation*}
%%
where~$\stepx{}$ is the reflexive-transitive closure of relation~$\step{}$.
We use~$\<p,\alpha>\eval{n}{p'}$ as an abbreviation
for~$\<p,\alpha,n>\eval{}p'$.

Finally, we describe the complete execution of a \Ceu\ program, given some
sequence of external events, by applying the reaction relation repeatedly to
each event in the sequence, with the reaction number incremented after each
application.

\section{Properties of the reaction relation}
\label{Section:reaction-properties}

\begin{thm}[Determinism]
For all~$p$, $p_1$, $p_2\in{P}$, $\alpha\in{E^*}$, and~$n\in{N}$:
\begin{equation*}
(\<p,\alpha>\eval{n}p_1\quadtext{and}\<p,\alpha>\eval{n}p_2)
\quadtext{implies}p_1=p_2\,.
\end{equation*}
\end{thm}
\begin{proof}
By induction on the number of reaction steps.  \FIXME{\dots}
\end{proof}

\begin{thm}[Termination]
For all~$p$, $p'\in{P}$, $\alpha\in{E^*}$, and~$n\in{N}$:
\begin{equation*}
(\exists!p'\in{P})(\<p,\alpha>\eval{n}p')\,.
\end{equation*}
\end{thm}
\begin{proof}
By induction on the number of reaction steps.  \FIXME{\dots}

\begin{TODO}
  Talvez eu tenha que partir para uma indução em pares
  ordenados~$\<p,\alpha>$, já que tanto~$p$ pode diminur quanto~$\alpha$.
\end{TODO}
\end{proof}

\section{Alternative formulation (Pure \Ceu)}
\label{Section:pure-ceu}

\begin{TODO}
Uma abordagem alternativa seria simplificar a formalização.  Minha proposta
é, num primeiro momento, definir uma versão reduzida (pura) de
\Ceu\ retirando da linguagem as construções \ceu{\Mem(v)} e
\ceu{\Ifelse{v}{p_1}{p_2}} e talvez outras coisas, e.g., loop e finalização.
E dado essa versão pura, tentar definir uma semântica big-step que se
comporte da maneira desejada, i.e., avaliação da esquerda para direita com
instruções \emph{await} atrasadas.  (Eu acho mais fácil trabalhar com a
big-step, mas se não tiver jeito a gente volta para a small-step.)

Num segundo momento, minha idéia é pegar as semânticas construtivas de
Esterel e ver como eles resolvem o problema do loop causal no \emph{await}
que acorda no mesmo ciclo.  Talvez seja interessante não atrasar os
\emph{awaits} em \Ceu.

Feito isso, i.e., definida uma base semântica e provados o determinismo e
convergência, eu posso partir para as provas de equivalência entre programas
e, quem sabe, encontrar possíveis formas normais.

(A propósito, do jeito que está definido hoje as reações de \Ceu\ não
retornam nada para o ambiente, i.e., não há evento externo de saída.  É isso
mesmo?)
\end{TODO}

\appendix
\section{Auxiliary definitions}
\label{Appendix:extra}

\subsection{The clear function}
\label{Appendix:extra:clear}

The clear function~$\clear\colon{P}\to{P'}$ concatenates in a sequence the
body of all active finalization instructions~\ceu{\Fin p'} within some
program~$p$.  Function~$\clear$ is defined by recursion on~$P$ as follows:
\begin{align*}
\clear(\ceu\Skip)&=\ceu\Skip\\
\clear(\ceu{\Mem(v)})&=\ceu\Skip\\
\clear(\ceu{\Await(e)})&=\ceu\Skip\\
\clear(\ceu{\Emit(e)})&=\ceu\Skip\\
\clear(\ceu{\Break})&=\ceu\Skip\\
\clear(\ceu{\Ifelse{v}{p_1}{p_2}})&=\ceu\Skip\\
\clear(\ceu{p_1;p_2})&=\clear(p_1)\\
\clear(\ceu{\Loop p_1})&=\ceu\Skip\\
\clear(\ceu{p_1\And p_2})&=\ceu{\clear(p_1);\clear(p_2)}\\
\clear(\ceu{p_2\Or p_2})&=\ceu{\clear(p_1);\clear(p_2)}\\
\clear(\ceu{\Fin p_1})&=\ceu{p_1}\\
\clear(\ceu{\Awaiting(e,n)})&=\ceu\Skip\\
\clear(\ceu{\Emitting(n)})&=\ceu\Skip\\
\clear(\ceu{p_1\Atloop p_2})&=\ceu\Skip\\
\end{align*}

The form of program~$p'\in{P'}\subseteq{P}$ returned by~$\clear$ is given by
the grammar:
\begin{equation*}
p'\in{P'}\Coloneqq\ceu{\Skip}\mid\ceu{\Mem(v)}\mid\ceu{p_1';p_2'}\,.
\end{equation*}
(Note that the definition of~$\clear$ assumes that the body~$p_1$ of all
instructions~$\ceu{\Fin p_1}$ is necessarily a member of~$P'$, i.e., $p_1$
is either~\ceu\Skip\ or~\ceu{\Mem(v)} or some finite sequence of such
instructions.)

\subsection{The is-blocked predicate}
\label{Appendix:extra:is-blocked}

The is-blocked predicate~$\#\colon{P}\times{E^*}\times{N}\to\{\top,\bot\}$
checks if all execution trails of a given program~$p$ in state~$\alpha$ and
reaction~$n$ are blocked, i.e., if all its trails are hanged in instructions
\ceu\Awaiting\ or \ceu\Emitting, which cannot be consumed in the current
reaction step.  Function~$\#$ is defined by recursion on~$P$ as follows:

\begin{align*}
\#(\ceu\Skip,e\alpha,n)&=\bot\\
\#(\ceu{\Mem(v)},e\alpha,n)&=\bot\\
\#(\ceu{\Await(e')},e\alpha,n)&=\bot\\
\#(\ceu{\Emit(e')},e\alpha,n)&=\bot\\
\#(\ceu{\Break},e\alpha,n)&=\bot\\
\#(\ceu{\Ifelse{v}{p_1}{p_2}},e\alpha,n)&=\bot\\
\#(\ceu{p_1;p_2},e\alpha,n)&=\#(p_1,e\alpha,n)\\
\#(\ceu{\Loop p_1},e\alpha,n)&=\bot\\
\#(\ceu{p_1\And p_2},e\alpha,n)&=\#(p_1,e\alpha,n)\land\#(p_2,e\alpha,n)\\
\#(\ceu{p_1\Or p_2},e\alpha,n)&=\#(p_1,e\alpha,n)\land\#(p_2,e\alpha,n)\\
\#(\ceu{\Fin p_1},e\alpha,n)&=\bot\FIXME{??}\\
\#(\ceu{\Awaiting(e',n')},e\alpha,n)&=e\ne{e'}\lor{n=n'}\\
\#(\ceu{\Emitting(n')},e\alpha,n)&=\|e\alpha|\ne{n'}\\
\#(\ceu{p_1\Atloop p_2},e\alpha,n)&=\#(p_1,e\alpha,n)\FIXME{??}\\
\end{align*}

\end{document}

### 2016-03-26 paper.tex
(+) some TODOs
...

@@
           interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
-          bool, data, tag, escape, new, traverse,implementation,output},
+          bool, data, tag, escape, new, traverse,implementation,output,
+          native,@const,@pure,@safe,define},
     emphstyle={\bfseries},
@@
 
+% TODO: SCCharts?, SynchCharts?
+
 Despite its success and influence, Esterel has an overly complex semantics that 
@@
 %
-Furthermore, it also affects the programmer's understanding about the code, 
-which, after all, has to solve the errors when facing corner cases.
+Above all, it also affects the programmer's understanding about the code, 
+which, ultimately, has to solve the errors when facing corner cases.
 %
@@
 non-deterministic execution for intra-reaction statements, which prevents 
-threads to share memory and interact with side-effect and stateful system calls 
-appropriately.
-%
-For instance, although Esterel supports orthogonal abortion of 
-threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
-release resources in use.
+threads to share memory and interact with stateful system calls safely.
+%
+%For instance, although Esterel supports orthogonal abortion of 
+%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
+%release resources in use.
 
@@
 %
-\item Stack-based execution for internal events, which provides a limited form 
-of coroutines.
-%
-\item Static temporal analysis and deterministic execution semantics that 
-allows programs to safely share memory.
-%
-\item Finalization mechanism for safe abortion of lines of execution holding 
-external resources.
-%
-\item First-class synchronized timers.
+\item Unique and queue-based external events, which define the notion of time 
+in \CEU (Section~\ref{sec.ceu.exts}).
+%
+\item Stack-based internal events for intra-reaction communication, which also
+provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
+%
+\item Static temporal analysis to detect suspicious concurrent statements 
+(Section~\ref{sec.ceu.analysis}).
+%
+\item Safe integration with \C via function annotations and enforced 
+finalization for external resources (Section~\ref{sec.ceu.c}).
+%
+\item First-class synchronized timers with a dedicated syntax 
+(Section~\ref{sec.ceu.timers}).
 \end{itemize}
@@
 functionalities.
+% TODO: why? stack, fin, principalmente
 %
@@
 microcontroller with 32K of FLASH and 2K of SRAM.
-}, showing that the peculiarities in the semantics of \CEU does not pose 
+}, showing that the peculiarities in the semantics of \CEU do not pose 
 practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}
@@
 of wireless sensor networks, developing a number of applications, protocols, 
-and drivers.
+and device drivers.
 %
@@
 For the \VM back end, applications have a bytecode footprint in the order of 
-hundreds of bytes can be transmitted over the air in a few 
+hundreds of bytes which can be transmitted over the air in a few 
 packets~\cite{ceu.terra}.
@@
 Esterel:
-queue-based external events and stack-based internal events 
-(Section~\ref{sec.ceu.ints}),
-shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
-safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
-and first-class timers (Section~\ref{sec.ceu.timers}).
+unique and queue-based external events (Section~\ref{sec.ceu.exts}),
+stack-based internal events (Section~\ref{sec.ceu.ints}),
+static temporal analysis for concurrent statements (Section~\ref{sec.ceu.analysis}),
+safe integration with \C (Section~\ref{sec.ceu.c}), and
+first-class synchronized timers (Section~\ref{sec.ceu.timers}).
 
@@
 restarts the enclosing \code{loop}.
+
+% TODO: input/output declarations
 
@@
 
-\subsection{Queue-Based External Events and Stack-Based Internal Events}
+\subsection{ Unique and Queue-Based External Events }
+\label{sec.ceu.exts}
+
+Esterel defines a logical unit of time as a discrete sequence of instants or 
+``ticks''.
+At each tick, depending on external stimuli the environment provides, an 
+arbitrary number of input events can be active simultaneously (including zero).
+The set of input events awakes the program, which must react within the current 
+tick, in bounded time.
+
+In \CEU, each unique external stimulus itself defines a unit of time.
+The execution model for \CEU programs is as follows~\cite{ceu.sensys13}:
+%
+\begin{enumerate}
+\item The program initiates the ``boot reaction'' in a single trail (but 
+      parallel constructs may create new trails).
+\item Active trails execute until they await or terminate.
+      This step is named a \emph{reaction chain}, and always runs in bounded 
+      time.
+\item The program goes idle and the environment takes control.
+\item On the occurrence of a new external input event, the environment awakes 
+      \emph{all} trails awaiting that event.
+      It then goes to step 2.
+\end{enumerate}
+%
+A program must react completely to an occurring event completely before 
+handling the next.
+%
+Based on the synchronous hypothesis, a program takes no time on step 2 and is 
+always idle on step 3.
+In practice, if a new external input event occurs while a reaction chain is 
+running (step 2), it is enqueued to run in the next reaction.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\columnwidth]{tick}
+\caption{The discrete notions of time in Esterel and \CEU.
+\label{fig.ticks}
+}
+\end{figure}
+
+Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
+\CEU.
+The box \code{Real World} shows event occurrences over a continuous timeline 
+divided in units of 10 milliseconds.
+Since Esterel and \CEU define discrete logical units of time, we describe how 
+the events (which are the same in all boxes) fit into their timelines.
+
+\code{Box-1} uses fixed-length ticks in Esterel, within which reactions to 
+occurring inputs have to fit~\cite{esterel.fixed} .
+%
+Here, we assume a \code{R(boot)} reaction which happens before any input at 
+\code{tick-0}.
+%
+The input \code{A} ``physically'' occurs during the boot reaction but, because 
+time is logical and discrete, the event is delayed to the next tick.
+%
+Note that the reaction \code{R(A)} takes more time than \code{tick-1}, causing 
+a \emph{timing violation} in the program execution (after the end of 
+\code{tick-1} and before the end of \code{R(A)}).
+%
+The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
+happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
+%
+Since no new events occurred during \code{tick-2}, the CPU stays idle during 
+the whole \code{tick-3}.
+%
+Finally, one instance of event \code{D} and two instances of event \code{E} 
+occur ``simultaneously'' during the idle \code{tick-3}.
+However, because the program cannot detect both occurrences of \code{E}, only 
+one is considered in \code{R(D+E)}.
+
+\code{Box-2} is an alternative that allows variable-length ticks in Esterel 
+that adjust to the time the corresponding reaction takes to 
+complete~\cite{esterel.variable}.
+This approach avoids the time violation for \code{R(A)} and also yields
+smaller idle periods.
+For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
+start reaction \code{R(D)} on \code{tick-4}.
+Similarly to the fixed-tick approach, only one of the two simultaneous 
+occurrences of \code{E} is considered on \code{tick-5}.
+
+\code{Box-3} illustrates the unique and queue-based policy for input events in 
+\CEU.
+We also assume a \code{R(boot)} reaction before any input.
+%
+Because the event \code{A} is unique during \code{tick-0}, the behavior in \CEU 
+is similar to \code{Box-2} for the first two units of time.
+%
+However, the events \code{B} and \code{C} are not simultaneous in \CEU and are 
+handled in subsequent reactions \code{R(B)} and \code{R(C)}.
+%
+We assume that the CPU time for \code{R(B+C)} in Esterel is roughly the same as 
+\code{R(B)+R(C)} in \CEU.
+This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
+%
+Finally, \CEU recognizes and reacts to the two instances of \code{E} 
+independently, which are handled in sequence.
+
+We decided for the unique and queue-based semantics for the reasons that 
+follow:
+%
+\begin{itemize}
+\item \textbf{A ``tick'' is too abstract and imprecise:}
+Outside the domain of hardware specification, a tick has no natural counterpart 
+in the real world.
+Also, since no time regularity is required for ticks~\cite{esterel.multiclock}, 
+the two approaches for Esterel in Figure~\ref{fig.ticks} lead to different 
+behaviors for the same sequence of inputs.
+%
+\item \textbf{Events are never simultaneous:}
+From a rigorous point of view, event occurrences are infinitesimal, having zero 
+probability of being simultaneous.
+This way, we believe that the notion of simultaneity should not be imposed by 
+the language, but explicitly defined for each use case.
+In the end of the section, we present a specification to detect 
+``simultaneous'' button clicks.
+%
+Note that in the case of Esterel, simultaneity depends on the imprecise 
+definition of ticks.
+For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
+simultaneous, even though \code{A} and \code{B} actually happen much closer to 
+one another.
+\item \textbf{Unique input events imply mutual exclusion:}
+Because they are atomic, reactions to different events never overlap.
+The possibility to reason about each input individually is a prerequisite for 
+the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
+\end{itemize}
+
+\begin{figure}
+\begin{minipage}[t]{0.57\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em,mathescape=true]
+input void LEFT_CLICK;
+input void RIGHT_CLICK;
+event void middle_click;
+loop do
+    par/or do
+        AWAIT_AND(LEFT_CLICK, RIGHT_CLICK);
+        emit middle_click;
+    with
+        AWAIT_OR(LEFT_CLICK, RIGHT_CLICK);
+        await 200 ms;
+    end
+end
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.43\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=13]
+#define AWAIT_AND(e1, e2)   \
+    par/and do              \
+        await e1;           \
+    with                    \
+        await e2;           \
+    end
+#define AWAIT_OR(e1, e2)    \
+    par/or do               \
+        await e1;           \
+    with                    \
+        await e2;           \
+    end
+\end{lstlisting}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Application defines that a \code{middle\_click} event occurs whenever 
+both \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} occur within 200 milliseconds.
+\label{lst.simultaneous}
+}
+\end{figure}
+
+Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
+terms of \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} (ln. 1--2).
+If both events occur, we emit the internal event \code{middle\_click} (ln.  
+6--7).
+However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
+we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
+with the enclosing \code{loop} (ln. 4).
+The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
+simple expansions to a \code{par/and} and \code{par/or} for better readability.
+%
+Here, ``simultaneous'' means ``within 200 milliseconds'', which is a huge 
+amount of time for a language-defined tick, and which would break the 
+synchronous hypothesis.
+%
+A similar implementation for Esterel would not rely on the tick notion of 
+simultaneity either.
+%
+We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
+Section~\ref{sec.ceu.timers}.
+
+The synchronous hypothesis for \CEU holds if the reactions run faster than the 
+rate of incoming input events.
+Otherwise, the system will accumulate delays between the real occurrence and 
+actual reaction to events.
+%
+This is also the case for the variable-length-tick approach of Esterel, since 
+the more inputs to handle, the longer the reaction takes, and the more inputs 
+will accumulate for subsequent ticks.
+%
+For the fixed-length-tick approach of Esterel, a breach in the synchronous 
+hypothesis causes timing violations, which can be avoided with \emph{worst case 
+reaction time} analysis to infer an appropriate value for the tick 
+length~\cite{esterel.fixed}.
+
+\begin{comment}
+> Multiclock Esterel
+The Esterel synchronous reactive language [5,4,6], which we call Classic Esterel
+in this paper, is based on a single-clock instantaneous interaction principle. The
+behavior of a program is defined by a sequence of reactions to input sequences.
+The execution environment decides when the program is provided an input, and
+a reaction is viewed as the simultaneous production of an output response to
+an input event. We call ticks the instants in which reactions occur, and master
+clock the sequence of these instants (it is a logical clock and no time regularity
+is required).
+
+An Esterel Processor with Full Preemption Support and its Worst Case Reaction 
+Time Analysis
+3. ESTEREL AND THE KIEL ESTEREL PROCESSOR
+The execution of an Esterel program is divided into (logical) instants, or 
+(logical) ticks, which are conceptually executed infinitely fast.
+An Esterel program interacts with its environment through signals, which are 
+either present throughout a logical instant or absent.
+Input signals are sampled at each tick, and each tick may generate output 
+signals.
+It would be possible to have the system just compute one logical instant after 
+the other, to start with the next reaction as soon as the previous one has 
+finished; however, typically the ticks are executed at some fixed frequency, 
+resulting in an interval T between each tick, where T is determined by the 
+real-time requirements of the system.
+The conceptually infinitely fast computation of the reaction within a logical 
+tick, plus in the case of Esterel the unique signal presence/absence status 
+throughout a logical tick, together constitute the synchrony hypothesis. In 
+practice, the results of a logical instant of course cannot be computed 
+infinitely fast; however, to maintain the abstraction of the synchrony 
+hypothesis, a logical instant must be computed within some desired reaction 
+time T.
+
+4. WORST CASE REACTION TIME ANALYSIS
+TODO: requires this, what about this in the context of MThreads??
+Towards Direct Execution of Esterel Programs on Reactive Processors
+3.1 Preserving the Synchrony Hypothesis
+The synchronous model implies that reactions to input signals are instantaneous 
+and occur at discrete logical instants called ticks.
+As a result, a sequence of actions may need to be performed within the duration 
+of a single tick.
+In order to model the logical tick of Esterel in RePIC, a tick with variable 
+length in terms of absolute time is implemented.
+The duration of one tick is determined by the time required to execute all the 
+instantaneous instructions that are placed between any two consecutive await 
+(i.e. TAWAIT, SAWAIT or CAWAIT) or halt instructions which are the tick 
+delimiting instructions.
+(TODO: tem figuras)
+\end{comment}
+
+\subsection{ Stack-Based Internal Events }
 \label{sec.ceu.ints}
@@
 \end{figure}
+
+% TODO: .. will probably expect that duplicating the  emit inc => &v; line 
+% would result in v==3.
 
@@
 %\newpage %TTT
-\subsection{Shared-Memory Concurrency and Determinism}
-\label{sec.ceu.det}
+
+\subsection{ Static Temporal Analysis for Concurrent Statements }
+\label{sec.ceu.analysis}
 
 Embedded applications make extensive use of global memory and shared resources, 
-such as through memory-mapped registers and system calls to drivers.
+such as through memory-mapped registers and system calls to device drivers.
 Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
@@
 %
-For this reason, Esterel assumes that external calls to the host language do 
-not perform side effects.
-%
-Also, Esterel syntactically forbids sharing memory between lines of execution 
-to preserve determinism: \emph{``if a variable is written by some thread, then 
-it can neither be read nor be written by concurrent 
-threads''}~\cite{esterel.primer}.
-
-Concurrency in \CEU is characterized when two or more trail segments in 
-parallel execute during the same reaction.
-A trail segment is a sequence of statements followed by an \code{await} (or 
-termination).
-%
-In the example in Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
-(ln.  5,8) can never run concurrently, because each trail segment reacts to 
-different input events (ln. 4,7), which cannot occur simultaneously (according 
-to Section~\ref{sec.ceu.ints}).
-%
-However, the example in Figure~\ref{lst.shared}.b is non-deterministic, because 
-the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
-\code{A} (ln. 4,7).
+A number of Esterel-based synchronous languages, such as 
+\emph{SOL}~\cite{wsn.sol},
+\emph{SC}~\cite{rp.sychc}, and
+\emph{PRET-C}~\cite{rp.pretc},
+enforce intra-reaction determinism with an arbitrary execution order for 
+statements in multiple lines of execution.
+%, e.g., based on lexical order or manual priority assignment.
+%
+\CEU also takes the deterministic approach and when multiple trails are active 
+during the same reaction, they are scheduled in the order they appear in the 
+program source code.
 
@@
 
-\CEU performs a temporal analysis at compile time and detects concurrent 
-accesses to shared variables, as follows:
+Even so, we consider that enforcing an arbitrary execution order can be 
+misleading in some cases.
+%
+For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
+which define a shared variable (ln. 2), and assign to it parallel trails (ln.  
+5, 8).
+%
+In the example \code{(a)}, the two assignments to \code{x} execute from 
+reactions to different events \code{A} and \code{B} which, by definition, 
+cannot occur simultaneously (Section~\ref{sec.ceu.exts}).
+Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
+for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
+%
+In the example \code{(b)}, the two assignments to \code{y} execute from a 
+reaction to the same event \code{A}, and are simultaneous from an external 
+point of view.
+Since \CEU employs lexical order for intra-reaction statements, the execution 
+is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
+%
+However, any (apparently innocuous) change in the order of trails can change 
+the semantics of a program, which we consider unsafe.
+
+To mitigate this threat, \CEU performs a temporal analysis at compile time and 
+detects concurrent accesses to shared variables, as follows:
 if a variable is written in a trail segment, then a concurrent trail segment 
@@
 variable type.
-An analogous policy is applied for pointers \emph{vs} variables and pointers 
-\emph{vs} pointers, as well as for system calls with side effects (e.g., 
-\code{printf}).%
-\footnote{\CEU assumes that all system calls perform side effects, unless they 
-are annotated as ``\texttt{@pure}''.}
-%The algorithm for the analysis holds the set of all events in preceding 
-%\code{await} statements for each variable access.
-%Then, the sets for all accesses in parallel trails are compared to assert that 
-%no events are shared among them.
-%Otherwise the compiler warns about the suspicious accesses.
-
-Regardless of the temporal analysis of \CEU, when multiple trails are active 
-during the same reaction, they are scheduled in the order they appear in the 
-program source code.
-%
-Therefore, even though the example in Figure~\ref{lst.shared}.b is suspicious, 
-the assignments to \code{y} are both atomic and deterministic, i.e., after the 
-reaction to \code{A} terminates, the value of \code{y} is $4$ (\code{(1+1)*2}).
-%
-On the one hand, enforcing an execution order for concurrent operations may 
-seen arbitrary and also precludes true parallelism.
-On the other hand, it provides a priority scheme for trails, and makes 
-shared-memory concurrency more tractable.
-%
-For constrained embedded development, we believe that deterministic 
-shared-memory concurrency is beneficial, given the extensive use of memory 
-mapped ports for \emph{I/O} and the lack of hardware support for real 
-parallelism.
-Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
-\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
-%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
-%parallelism is inherent.
-
-Figure~\ref{lst.det} compares two syntactically equivalent code fragments in 
-Esterel (a) and \CEU (b) to summarize the semantic differences regarding 
-(non-)determinism.
-%
-Even though the program in \CEU executes deterministically, the compiler still 
-issues a warning, because an apparently innocuous reordering of trails modifies 
-the semantics of the program.
-%
-Note that in Esterel multiple external events can coexist in the same reaction, 
-which disallows a similar temporal analysis.
+%An analogous policy is applied for pointers \emph{vs} variables and pointers 
+%\emph{vs} pointers.
+%
+A trail segment is a sequence of statements followed by an \code{await} (or 
+termination).
+Concurrency in \CEU is characterized when two or more trail segments in 
+parallel react to the same input event.
+%
+Considering the examples in Figure~\ref{lst.shared}:
+%
+\begin{itemize}
+\item The assignments to \code{x} in lines 2 and 5 in \code{(a)}
+      \textbf{cannot} be concurrent because they are not in parallel trails.
+\item The assignments to \code{x} in lines 5 and 8 in \code{(a)}
+      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
+      the same reaction.
+\item The assignments to \code{y} in lines 5 and 8 in \code{(b)}
+      \textbf{can} be concurrent because they are in parallel trails and 
+      \textbf{can} execute during the same reaction.
+\end{itemize}
+%
+The algorithm for the analysis, which is depicted in 
+Section~\ref{sec.impl.analsyis}, inspects all possible \code{await} statements 
+that precede a variable access and hold the associated awaking events.
+Then, it checks all accesses in parallel trails to see if they share an awaking 
+event.
+If it is the case, the compiler warns about the suspicious accesses.
+
+Note that this analysis is only possible due to the uniqueness of input events 
+within reactions.
+Otherwise, any two trail segments in parallel could be concurrent, even if they 
+react to different input events.
+
+\subsection{ Safe Integration with \C }
+\label{sec.ceu.c}
+
+In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
+the \C compiler that generates the final binary.
+Therefore, access to \C is seamless and, more importantly, easily trackable.
+%
+Similarly to Esterel, which supports the \code{call} primitive for external 
+code, calls are assumed to be instantaneous~\cite{esterel.primer}.
+%
+Evidently, programs should only resort to \C for asynchronous functionality, 
+such as non-blocking I/O, or simple \code{struct} accessors, but never for 
+control purposes.
 
 \begin{figure}[b]
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[mathescape=true]
-input A;
-[
-    await A;
-    call f1();
-$\|$
-    await A;
-    call f2();
-];
-\end{lstlisting}
-\centering\small{(a) Esterel}
-\end{minipage}
-%
-\begin{minipage}[t]{0.48\linewidth}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+native do
+    #define NUM 10
+    void f  (void)  { <...> }
+    void g  (int v) { <...> }
+    int  id (int v) { <...> }
+end
+par/and do
+    _f();
+with
+    _g(_id(_NUM));
+end
+\end{lstlisting}
+\centering\small{(a)}
+\end{minipage}
+%
+\begin{minipage}[t]{0.44\linewidth}
 \begin{lstlisting}
-input void A;
-par/and do
-    await A;
-    _f1();
-with
-    await A;
-    _f2();
-end
-\end{lstlisting}
-\centering\small{(b) \CEU}
+native @const _NUM;
+native @pure  _id;
+native @safe  _f with _g;
+
+.
+\end{lstlisting}
+\centering\small{(b)}
 \end{minipage}
 %\rule{8.4cm}{0.37pt}
-\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
-unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
-deterministic scheduling based on lexical order.
-\label{lst.det}
+\caption{ TODO
+\label{lst.c}
 }
@@
 
-\subsection{Safe Abortion with Finalization}
-\label{sec.ceu.abrt}
-
-The introductory example in Figure~\ref{lst.abro} illustrates how synchronous 
-languages can abort awaiting lines of execution without tweaking them with 
-synchronization primitives.
+In addition, \CEU also employs the temporal analysis for calls and accesses to 
+external symbols in \C.
+%
+As an example, the program in Figure~\ref{lst.c}.a defines four external 
+symbols inside a \code{native} block for standard declarations in \C (ln.  
+1--6).
+%
+The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
+during the boot reaction (ln. 8,10):
+the first trail calls the symbol \code{\_f},
+while the second calls \code{\_g} and \code{\_id}, and also accesses 
+\code{\_NUM}.
+%
+Since \CEU does not inspect any code defined in \C, it has no idea about the 
+meaning of the symbols and complains about suspicious concurrent accesses 
+between \code{\_f} against all symbols in the second trail.
+
+The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
+compiler about the semantics of the \C symbols in use in \code{(a)}:
+%
+\begin{itemize}
+\item \code{\_NUM} is a constant symbol, meaning that it is safe to use it 
+      concurrently with any other symbol in the program.
+\item \code{\_id} is a pure function, also meaning that it is safe to call it
+      concurrently with any other symbol in the program.
+\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
+      conflict and they can be safely called concurrently.
+\end{itemize}
+
+Esterel's \code{abort} and \CEU's \code{par/or} statements illustrate how 
+synchronous languages can abort awaiting lines of execution without tweaking 
+them with synchronization primitives.
 In contrast, traditional (asynchronous) multi-threaded languages cannot express 
 thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
-Still, handling abortion when dealing with external resources is challenging 
-because they are not subject to the same synchronous execution discipline.
+%
+Still, aborting lines of execution that deal with external resources is 
+challenging because they may end up in an inconsistent state.
+%
+For this reason, Esterel and \CEU provide a \code{finalize} construct with a 
+clause that executes even if the enclosing block aborts and does not terminate 
+normally.
+
+Figure~\ref{lst.fin} uses \code{lock} and \code{unlock} calls to represent an 
+external resource.
+The normal behavior is to \code{lock} the resource, perform some operations in 
+subsequent reactions to input \code{A}, and then \code{unlock} the resource.
+However, if the aborting input \code{B} occurs after the \code{lock} but before 
+the reactions to \code{A}, we still want to call \code{unlock} to safely 
+release the resource.
+%
+In Esterel and \CEU, the finalize clauses (ln. a:10 and b:5) are automatically 
+called if the enclosing blocks (ln. a:3--1 and b:3--10) are externally aborted 
+(ln. a:12 and b:12).
+
+We completely missed the "finalize" construct of Esterel and recognize the 
+similarity with our proposal.
+We now emphasize that Céu tracks pointers entering and leaving the program to 
+enforce appropriate use of "finalize" (as part of the section about interfacing 
+with C).
 
 \begin{figure}
-\begin{minipage}[t]{0.55\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-input void STOP, RETRANSMIT, SEND_ACK;
-output _pkt_t* SEND_ENQUEUE, SEND_CANCEL;
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
+input A, B;
+abort
+    finalize
+        call lock();
+        await A;
+        <...>;      // do something
+        await A;
+        <...>;      // do something
+    with
+        call unlock();
+    end
+when B
+.
+\end{lstlisting}
+\centering\small{(a) Esterel}
+\end{minipage}
+%
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+input void A, B;
 par/or do
-    await STOP;
+    _lock();
+    finalize with
+        _unlock();
+    end
+    await A;
+    <...>;      // do something
+    await A;
+    <...>;      // do something
 with
-    loop do
-        par/or do
-            await RETRANSMIT;
-        with
-            par/and do
-                await 1min;
-            with
-                var _pkt_t buffer;
-                <fill-buffer-info>
-                emit SEND_ENQUEUE => &buffer;
-                await SEND_ACK;
-            end
-        end
-    end
-end
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=12]
-<...>
-    var _pkt_t buffer;
-    <fill-buffer-info>
-    finalize
-        emit SEND_ENQUEUE => &buffer;
-    with
-        emit SEND_CANCEL => &buffer;
-    end
-    await SEND_ACK;
-<...>
-
-.
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%\rule{8.5cm}{0.37pt}
+    await B;
+end
+\end{lstlisting}
+\centering\small{(b) \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
 \caption{
-(a) The unsafe network protocol does not compile.
-(b) The finalization clause extends the protocol to handle abortion properly.
-%\caption{ Parallel compositions can describe complex state machines.
+Finalization in Esterel and \CEU: after the call to \code{lock}, both languages 
+guarantee to call \code{unlock} if the enclosing block aborts when \code{B} 
+occurs.
 \label{lst.fin}
@@
 
-To illustrate the risks related to abortion, consider the unsafe example of 
-Figure~\ref{lst.fin}.a, which does not compile in \CEU.
-It describes the state machine of the data collection protocol \emph{CTP} for 
-sensor networks~\cite{wsn.ctp}.
-%
-The input and output events represent the external interface of the protocol 
-(ln. 1--2).
-%
-The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
-unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
-or a \code{STOP} request to terminate.
-%
-The protocol is implemented with two main trails:
-one simply monitors the stopping event (ln. 4);
-the other periodically transmits the packet (ln. 6--19).
-%
-The periodic transmission is a loop that starts two other trails (ln. 7--18):
-one handles the immediate retransmission request (ln. 8);
-the other transmits the packet and waits for a confirmation (ln. 10--17).
-%
-%
-The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
-takes at least one minute before looping (ln. 11), in accordance with the 
-specification.
-Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
-handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
-The driver makes the transmission in the background, holding the packet until 
-it signals the application with a \code{SEND\_ACK} event to acknowledge 
-completion (ln. 16).
-%
-At any time, the client may request a retransmission (ln. 8), which terminates 
-the \code{par/or} (ln. 7) and restarts the loop (ln. 6), but does not abort the 
-ongoing transmission initiated with the \code{emit SEND\_ENQUEUE} (ln. 15).
-%
-The client may also request to stop the whole protocol at any time (ln. 4).
-%
-Therefore, if the sending trail is aborted by the \code{STOP} or 
-\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
-leaving behind a \emph{dangling pointer} in the radio driver, which will 
-possibly transmit corrupted data.
-%
-%\subsubsection{Finalization}
-%\label{sec.ceu.fin}
-
-The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
-tracks the interaction of \code{par/or} compositions with pointers to local 
-variables and stateful output events calls in order to preserve safe abortion 
-of trails.
-%
-%Finalization clauses are fundamental to preserve the orthogonality of 
-%\code{par/or} compositions in SSRP.
-%
-\CEU enforces a \emph{finalization} clause to accompany the output request.
-The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
-when the block of \code{buffer} goes out of scope, i.e., the finalization 
-clause (ln. 18) executes automatically on external abortion.%
-\footnote{
-Note that the compiler only enforces the programmer to write the finalization 
-clause, but cannot check if it actually handles the resource properly.
-}
-%
+\CEU goes one step further and enforces the use of \code{finalize} for system 
+calls that deal with pointers representing resources:
+%
+\begin{itemize}
+\item If a system call \textbf{receives} a pointer from \CEU, the pointer 
+represents a \textbf{local} resource that requires finalization.
+\item If a system call \textbf{returns} a pointer to \CEU, the pointer 
+represents an \textbf{external} resource that requires finalization.
+\end{itemize}
+%
+\CEU tracks the interaction of system calls with pointers and requires 
+finalization clauses to accompany them.
+%
+In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
+(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
+(ln. 5), which is an asynchronous call that transmits the buffer in the 
+background.
+If the block aborts (ln. 11) before receiving an acknowledge from the 
+environment (ln. 9), the local \code{msg} goes out of scope and the external 
+transmission now holds a \emph{dangling pointer}.
+The finalization ensures that the transmission also aborts (ln. 7).
+%
+In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} 
+returns an external file resource as a pointer.
+If the block aborts (ln. 12) during the \code{await 1s} (ln. 9), the file 
+remains open as a \emph{memory leak}.
+The finalization ensures that the file closes properly (ln. 6).
+%
+In both cases, the code does not compile if the programmer forgets to use the
+\code{finalize} construct.
+
+Note that the illustrative example of Figure~\ref{lst.fin} does not manipulate 
+pointers (i.e., the resource is a \emph{singleton}).
+That case is an example of a bad and unsafe API to expose to \CEU because the 
+compiler will not enforce the use of finalization.
+%
+%TODO: @nohold
 %Note also that \CEU environments rely on \C libraries that only provide 
@@
 
-\begin{comment}
-The code fragments in Figure~\ref{lst.abortion} shows a corner case regarding
-abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
-%
-For instance, it is not clear in the code in Esterel if the call to \code{f} 
-should execute or not after \code{A}, given that the body and abortion events 
-are the same.
-%
-For this reason, Esterel provides \emph{weak} and \emph{strong} variations for 
-the \code{abort} statement.
-With \emph{strong} abortion (the default), the body is aborted immediately and 
-the call does not execute.
-%
-In \CEU, given the deterministic scheduling rules, strong and weak abortions 
-can be chosen by reordering trails inside a \code{par/or}, e.g., in 
-Figure~\ref{lst.abortion}, the second trail is strongly aborted by the first 
-trail and the call to \code{\_f} never executes.
-
 \begin{figure}
 \begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}
-abort
-    await A;
-    call f();
-when A;
-
-\end{lstlisting}
-\centering\small{(a) Esterel}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
+par/or do
+   var _buffer_t msg;
+   <...> // prepare msg
+   finalize
+      _send_request(&msg);
+   with
+      _send_cancel(&msg);
+   end
+   await SEND_ACK;
+with
+   <...>
+end
+.
+\end{lstlisting}
+\centering\small{(a) Local resource finalization}
 \end{minipage}
@@
 \begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
 par/or do
-    await A;
+   var _FILE* f;
+   finalize
+      f = _fopen(...);
    with
-    await A;
-    _f();
-end
-\end{lstlisting}
-\centering\small{(b) \CEU}
-\end{minipage}
-%
+      _fclose(f);
+   end
+   _fwrite(..., f);
+   await 1s;
+   _fwrite(..., f);
+with
+   <...>
+end
+\end{lstlisting}
+\centering\small{(b) External resource finalization}
+\end{minipage}
 %\rule{8.4cm}{0.37pt}
-\caption{ Strong abortion in Esterel and \CEU. %\newline
-\label{lst.abortion}
+\caption{
+\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
+local resources and \emph{memory leaks} for external resources.
+\label{lst.fin.ceu}
 }
 \end{figure}
-\end{comment}
-
-\subsection{First-Class Timers}
+
+\subsection{ First-Class Timers }
 \label{sec.ceu.timers}
@@
 
-\section{Formal Semantics}
-\label{sec.sem}
-
-%\begin{document}
-
-In this section, we introduce a reduced syntax of \CEU and propose an 
-operational semantics to formally describe the language.
-We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
-in particular the stack-based execution for internal events.
-%
-For the sake of simplicity, we focus on the control aspects of the language, 
-leaving out side effects and system calls (which behave like in conventional 
-imperative languages).
-
-\subsection{Abstract Syntax}
-\label{sec.sem.syntax}
-
-\begin{figure}
-\begin{lstlisting}[mathescape=true]
-                                   // primary expressions
-  p ::= mem(id)                   (any memory access to `id')
-      $|$ await(id)                  (await event `id')
-      $|$ emit(id)                   (emit event `id')
-      $|$ break                      (loop escape)
-                                   // compound expressions
-      $|$ if mem(id) then p else p   (conditional)
-      $|$ p ; p                      (sequence)
-      $|$ loop p                     (repetition)
-      $|$ p and p                    (par/and)
-      $|$ p or p                     (par/or)
-      $|$ fin p                      (finalization)
-                                   // derived by semantic rules
-      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
-      $|$ emitting(n)                (emitting on stack level `n')
-      $|$ p @ loop p                 (unwinded loop)
-      $|$ nop                        (terminated expression)
-\end{lstlisting}
-\caption{
-    Reduced syntax of \CEU.
-\label{lst.formal.syntax}
-}
-\end{figure}
-
-Figure~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
-sufficient to describe all semantic peculiarities of the language.
-%
-Except for $fin$ and the expressions used internally by the semantics (i.e., 
-$awating$, $emitting$, $p~@~loop; p$, and $nop$), all other expressions are 
-equivalent to their counterparts in the concrete language.
-
-The $mem(id)$ primitive represents all accesses, assignments, system calls, and 
-output events that affect a memory location identified by $id$.
-%
-According to the synchronous hypothesis of \CEU, $mem$ expressions are 
-considered to be atomic and instantaneous.
-%
-As the challenging parts of \CEU reside on its control structures, we are not 
-concerned here with a precise semantics for side effects, but only with their 
-occurrences in programs.
-%
-%The special notation $nop$ is used to represent an innocuous $mem$ expression 
-%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
-%unused identifier).
-%
-Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
-an identifier is either a variable or an event.
-
-\subsection{Operational Semantics}
-
-The core of our semantics describes how a program reacts to a single external 
-input event, i.e., starting from the input event, how the program behaves and 
-becomes idle again to proceed to the subsequent reaction.
-%
-We use a set of small-step operational rules, which are built in such a way 
-that at most one transition is possible at any time, resulting in deterministic 
-reactions.
-%
-Each reaction is identified by a ever-increasing $n$ that remains constant 
-during the entire reaction.
-The transition rules map a program $p$ and a stack of events $S$ in a single 
-step to a modified program and stack:
-%
-\begin{align*}
-\LL S, p \RR &\ST
-\LL S', p' \RR
-    & \textbf{(rule-inner)}
-\end{align*}
-%
-where
-%
-\begin{align*}
-S, S' &\in id^*
-    &&(stack~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
-p, p' &\in P
-    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
-n     &\in \mathds{N}
-    && (unique~identifier~for~the~entire~reaction)
-\end{align*}
-%
-At the beginning of a reaction, the stack is initialized with the occurring 
-external event $ext$ ($S=[ext]$), but $emit$ expressions can push new events on 
-top of it (we discuss how they are popped further).
-The sequence number $n$, which is incremented each reaction, prevents that 
-awaiting expressions awake in the same reaction they are reached (the 
-\emph{delayed awaits} as explained in Section~\ref{sec.ceu.ints}).
-
-The transition rules for the primary expressions are as follows:
-%
-{ \setlength{\jot}{20pt}
-\begin{align*}
-\LL S,\1mem(id) \RR &\ST
-\LL S,\1nop \RR
-    & \textbf{(mem)}        \\
-%%%
-\LL S,\1await(id) \RR &\ST
-\LL S,\1awaiting(id,n+1) \RR
-    & \textbf{(await)}      \\
-%%%
-\LL id:S,\1awaiting(id,m) \RR &\ST
-\LL id:S,\1nop \RR, \2if~m\leq n
-    & \textbf{(awake)}   \\
-%%%
-\LL S,\1emit(id) \RR &\ST
-\LL id:S,\1emitting(|S|) \RR
-    & \textbf{(emit)}       \\
-%%%
-\LL S,\1emitting(k) \RR &\ST
-\LL S,\1nop \RR, \2 if~k=|S|
-    & \textbf{(pop)}        %\\
-\end{align*}
-}
-%
-A $mem$ operation executes immediately and becomes a $nop$ to indicate 
-termination (rule \textbf{mem}).
-An $await$ is transformed into an $awaiting$ (rule \textbf{await}) as an 
-artifice to remember the external sequence number $n+1$ it can awake:
-an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
-event $id$ matches the top of the stack and it was reached in a previous 
-reaction (i.e., sequence number $m \leq n$).
-%
-%Remember that in \CEU, the \code{await} statement returns the value associated 
-%with the corresponding event: the yielded $mem$ represents the operation to 
-%query that value.
-%
-An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
-stands for the stack length), and pushing the referred event on the stack (rule 
-\textbf{emit}).
-With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
-rule \textbf{pop} expects its parameter $k$ to match the current stack level.
-This trick provides the desired stack-based semantics for internal events.
-
-Proceeding to compound expressions, the rules for conditionals and sequences 
-are straightforward:
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \frac
-    { \DS val(id,n) \neq 0 }
-%   -----------------------------------------------------------
-    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
-          \LL S, p \RR }
-    & \textbf{(if-true)}       \\
-%%%
-& \frac
-    { \DS val(id,n) = 0 }
-%   -----------------------------------------------------------
-    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
-          \LL S, q \RR }
-    & \textbf{(if-false)}       \\
-%%%
-& \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
-    & \textbf{(seq-adv)}      \\
-%%%
-& \LL S, (nop~;~q) \RR \ST  \LL S, q \RR
-    & \textbf{(seq-nop)}      \\
-%%%
-& \LL S, (break~;~q) \RR \ST \LL S, break \RR
-    & \textbf{(seq-brk)}
-\end{eqnarray*}
-}
-%
-Given that our semantics focuses on control, rules \textbf{if-true} and 
-\textbf{if-false} are the only to query $mem$ expressions.
-%
-The ``magical'' function $val$ receives a memory identifier and the current 
-reaction sequence number, returning the current memory value.
-%
-Although the value here is arbitrary, it is unique in a reaction, because a 
-given expression can execute only once within it (remember that $loops$ must 
-contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
-reaction they are reached).
-%For all other rules, we omit these values (e.g., \textbf{seq-nop}).
-
-The rules for loops are analogous to sequences, but use \code{`@'} as 
-separators to properly bind breaks to their enclosing loops:
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL S, (loop~p) \RR \ST \LL S, (p~@~loop~p) \RR
-    & \textbf{(loop-expd)}       \\
-%%%
-& \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
-% -----------------------------------------------------------
-    {\DS \LL S, (p~@~loop~q) \RR \ST \LL S', (p'~@~loop~q) \RR }
-    & \textbf{(loop-adv)}    \\
-%%%
-& \LL S, (nop~@~loop~p) \RR \ST \LL S, loop~p \RR
-    & \textbf{(loop-nop)}    \\
-%%%
-& \LL S, (break~@~loop~p) \RR \ST \LL S, nop \RR
-    & \textbf{(loop-brk)}
-\end{eqnarray*}
-}
-%
-When a program encounters a $loop$, it first expands its body in sequence with 
-itself (rule \textbf{loop-expd}).
-Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
-\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
-$nop$.
-However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
-Note that if we used \code{`;'} as a separator in loops, rules 
-\textbf{loop-brk} and \textbf{seq-brk} would conflict.
-%
-Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
-a $nop$.
-%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
-%into a $clear(p)$.
-%We cannot simply transform the loop into a $nop$ because its body may be a 
-%parallel composition containing finalization blocks.
-
-Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
-always force transitions on their left branches $p$ to occur before their right 
-branches $q$:
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
-    & \textbf{(and-adv1)}      \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
-    & \textbf{(and-adv2)}      \\
-%%%
-& \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
-    & \textbf{(or-adv1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
-%   -----------------------------------------------------------
-    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
-    & \textbf{(or-adv2)}   %\\
-\end{eqnarray*}
-}
-%
-The deterministic behavior of the semantics relies on the \emph{isBlocked} 
-predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
-\textbf{and-adv2} and \textbf{or-adv2}.
-These rules require the left branch $p$ to be blocked in order to allow the 
-right transition from $q$ to $q'$.
-%
-Basically, the \emph{isBlocked} predicate determines that an expression becomes 
-blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
-expressions that cannot advance.
-
-\begin{figure}
-{\small
-\begin{align*}
-  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m > n)   \\
-  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
-  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
-  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
-  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
-                                    isBlocked(n,S,q)             \\
-  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
-                                    isBlocked(n,S,q)             \\
-  isBlocked(n,S, \_)             &= false \2  (nop,await,      \\
-                                  &    \5\5\5\2 emit,break,if,loop)   %\\
-\end{align*}
-}%
-%\rule{14cm}{0.37pt}
-\caption{
-The recursive predicate $isBlocked$ is true only if all branches in parallel 
-are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
-\label{fig.isBlocked}
-}
-\end{figure}
-
-%%%
-
-For a parallel $and$, if one of the sides terminates, the composition is simply 
-substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
-as follows).
-%
-For a parallel $or$, if one of the sides terminates, the whole composition 
-terminates, also applying the $clear$ function to properly finalize the aborted 
-side (rules \textbf{or-nop1} and \textbf{or-nop2}):
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL S, (nop~and~q) \RR \ST \LL S, q \RR
-    & \textbf{(and-nop1)}   \\
-%%%
-& \LL S, (p~and~nop) \RR \ST \LL S, p \RR
-    & \textbf{(and-nop2)}   \\
-%%%
-& \LL S, (nop~or~q) \RR \ST \LL S, clear(q) \RR
-    & \textbf{(or-nop1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~or~nop) \RR \ST \LL S, clear(p) \RR }
-    & \textbf{(or-nop2)}   %\\
-\end{eqnarray*}
-}
-%
-The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
-all active $fin$ bodies of the side being aborted, so that they execute before 
-the composition rejoins.
-Note that there are no transition rules for $fin$ expressions.
-This is because once reached, a $fin$ expression halts and will only execute 
-when it is aborted by a trail in parallel and is expanded by the $clear$ 
-function.
-In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
-the concrete language to a $fin$ in the formal semantics.
-%
-Note also that there is a syntactic restriction that a $fin$ body can only 
-contain $mem$ expressions in sequence, i.e., they are guaranteed to execute 
-entirely within a reaction.
-
-\begin{figure}[b]
-{\small
-\begin{align*}
-  clear( fin~p )       &= p                   \\
-  clear( p~;~q )       &= clear(p)            \\
-  clear( p~@~loop~q) ) &= clear(p)            \\
-  clear( p~and~q )     &= clear(p)~;~clear(q) \\
-  clear( p~or~q )      &= clear(p)~;~clear(q) \\
-  clear( \_ )          &= nop
-\end{align*}
-}%
-%\rule{14cm}{0.37pt}
-\caption{
-The function $clear$ extracts $fin$ expressions in parallel and put their 
-bodies in sequence.
-\label{fig.formal.clear}
-}
-\end{figure}
-
-Finally, a $break$ in one of the sides in parallel escapes the closest 
-enclosing $loop$, properly aborting the other side by applying the $clear$ 
-function:
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
-    & \textbf{(and-brk1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
-    & \textbf{(and-brk2)}   \\
-%%%
-& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
-    & \textbf{(or-brk1)}   \\
-%%%
-& \frac
-    {\DS isBlocked(n,S,p) }
-%   -----------------------------------------------------------
-    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
-    & \textbf{(or-brk2)}   %\\
-\end{eqnarray*}
-}
-%
-A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
-parallel trails.
-%
-If all trails hangs only in $awaiting$ expressions, it means that the program 
-cannot advance in the current reaction.
-%
-However, $emitting$ expressions are pending in lower stack indexes and should 
-eventually resume in the ongoing reaction (see rule \textbf{pop}).
-%
-Therefore, we define another rule that behaves as \textbf{rule-inner} 
-(presented above) if the program can advance, and, otherwise, pops the stack to 
-resume the lower level:
-%
-$$
-\frac
-    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
-%   -----------------------------------------------------------
-    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
-%
-\5\5\5
-%
-\frac
-    { \DS isBlocked(n,\1s:S,\1p) }
-%   -----------------------------------------------------------
-    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
-\5\5\5\textbf{(rule-outer)}
-$$
-%
-To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
-to a single external event, we use the reflexive transitive closure of 
-\textbf{rule-outer}:
-%
-$$
-    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
-$$
-%
-Finally, to describe the complete execution of a program, we trigger multiple 
-``invocations'' of reactions in sequence:
-%
-\begin{align*}
-\LL [e1], p \RR
-    & \xRightarrow[~~1~~]{*}
-\LL [  ], p' \RR
-\\
-\LL [e2], p' \RR
-    & \xRightarrow[~~2~~]{*}
-\LL [  ], p'' \RR
-\\
-\LL [e3], p'' \RR
-    & \xRightarrow[~~3~~]{*}
-\LL [  ], p''' \RR
-\\
-& ...
-\end{align*}
-%
-Each invocation starts with the occurring external event at the top of the 
-stack and finishes with a modified program and an empty stack.
-After each invocation, we increment the sequence number.
-
-\subsection{Concrete Language Mapping}
-
-Most statements from \CEU (``concrete \CEU'') map directly to those presented 
-in the reduced syntax in Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
-%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
-%\code{par/or}.
-For instance, the \code{if} in the concrete language behaves exactly like the 
-$if$ in the formal semantics.
-However, there are some significant mismatches between the concrete and 
-abstract \CEU, and we (informally) present appropriate mappings in this 
-section.
-%
-Again, we are not considering side-effects, which are all mapped to the $mem$ 
-semantic construct.
-
-\begin{comment}
-$
-    \code{if}      \mapsto if   ,\2
-    \code{';'}     \mapsto ~';' ,\2
-    \code{loop}    \mapsto loop ,\2
-    \code{par/and} \mapsto and  ,\2
-    \code{par/or}  \mapsto or
-$.
-\end{comment}
-
-\subsubsection{await and emit}
-
-The concrete \code{await} and \code{emit} primitives support communication of 
-values between them.
-In the two-step translation in Figure~\ref{lst.map.emit.await}, we start with 
-the concrete program in \CEU (a), which communicates the value $1$ between the 
-\code{emit} and \code{await} in parallel.
-In the intermediate translation (b), we include the shared variable \code{e\_} 
-to hold the value being communicated between the two trails in order to 
-simplify the \code{emit}.
-Finally, we convert the program into the equivalent in the abstract syntax (c), 
-translating side-effect statements into $mem$ expressions.
-External events have a similar translation, i.e., each external event requires 
-a corresponding variable that is explicitly set by the environment before each 
-reaction.
-
-\begin{figure}[t]
-\begin{minipage}[t]{0.32\linewidth}
-\begin{lstlisting}
-par/or do
-  <...>
-  emit e => 1;
-with
-  v = await e;
-  _printf("%d\n",v);
-end
-
-.
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.32\linewidth}
-\begin{lstlisting}
-par/or do
-  <...>
-  e_ = 1;
-  emit e;
-with
-  await e;
-  v = e_;
-  _printf("%d\n",v);
-end
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.32\linewidth}
-\begin{lstlisting}
-<...> ; mem ; emit(e)
-        or
-await(e) ; mem ; mem
-
-.
-\end{lstlisting}
-\centering\small{(c)}
-\end{minipage}
-\caption{
-Two-step translation from concrete to abstract \code{emit} and \code{await} 
-expressions.
-The concrete code in (a) communicates the value \code{1} from the \code{emit} 
-to the \code{await}.
-The abstract code in (c) uses a shared variable to hold the value.
-\label{lst.map.emit.await}
-}
-\end{figure}
-
-\subsubsection{First-class Timers}
-
-To encompass first-class timers, we introduce a special external event 
-\code{DT} that is intercalated with each other event occurrence in an 
-application (e.g.  \emph{e1, e2}):
-
-\begin{align*}
-\LL [DT], p \RR
-    & \xRightarrow[~~1~~]{*}
-\LL [    ], p' \RR
-\\
-\LL [e1], p' \RR
-    & \xRightarrow[~~2~~]{*}
-\LL [  ], p'' \RR
-\\
-\LL [DT], p'' \RR
-    & \xRightarrow[~~3~~]{*}
-\LL [    ], p''' \RR
-\\
-\LL [e2], p''' \RR
-    & \xRightarrow[~~4~~]{*}
-\LL [  ], p'''' \RR
-\\
-& ...
-\end{align*}
-
-The event \code{DT} has an associated variable \code{DT\_} carrying the 
-wall-clock time elapsed between two occurrences in sequence, as depicted by the 
-two-step translation in Figure~\ref{lst.map.timers.2}.
-In the concrete program (a), the variable \code{dt} holds the residual delta 
-time (as described in Section~\ref{sec.ceu.timers}) after awaking from the 
-timer.
-In the first step of the translation (b), we expand the \code{await 10ms} to a 
-\code{loop} that decrements the elapsed number of microseconds for each 
-occurrence of \code{DT}.
-When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
-the variable \code{dt} to contain the appropriate delta.
-In the last step (c), we convert the program to the abstract syntax.
-
-\begin{figure}
-\begin{minipage}[t]{0.30\linewidth}
-\begin{lstlisting}
-dt = await 10ms;
-
-.
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.37\linewidth}
-\begin{lstlisting}
-var int tot = 10000; // 10ms
-loop do
-    await DT;
-    tot = tot - DT_;
-    if tot <= 0 then
-        dt = -tot;
-        break;
-    end
-end
-.
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.30\linewidth}
-\begin{lstlisting}
-mem;
-loop(
-    await(DT);
-    mem;
-    if mem then
-        mem;
-        break
-    else
-        nop
-)
-\end{lstlisting}
-\centering\small{(c)}
-\end{minipage}
-\caption{
-Two-step translation from concrete to abstract timer.
-\label{lst.map.timers.2}
-}
-\end{figure}
-
-\subsubsection{Finalization Blocks}
-\label{sec.formal.fins}
-
-The biggest mismatch between concrete and abstract \CEU is regarding the 
-\code{finalize} blocks, which require more complex modifications in the program
-for a proper mapping using $fin$ expressions.
-In the three-step translation in Figure~\ref{lst.map.fin}, we start with a 
-concrete program (a) that uses a \code{finalize} to safely \code{\_release} the 
-reference to \code{ptr} kept after the call to \code{\_hold}.
-In the translation, we first need to catch the outermost \code{do-end} 
-termination to run the finalization code.
-For this, we translate the block into a \code{par/or} (b) with the original 
-body in parallel with a $fin$ expression to run the finalization code.
-Note that the $fin$ has no transition rules in the semantics, keeping the 
-\code{par/or} alive.
-This way, the $fin$ body only executes when the \code{par/or} terminates either 
-normally (after the \code{await B}), or aborted from an outer composition.
-%
-However, the $fin$ still (incorrectly) executes even if the call to 
-\code{\_hold} is not reached in the body due to an abort before awaking from 
-the \code{await A}.
-%
-To deal with this issue, for each $fin$ we need a corresponding flag to keep 
-track of code that needs to be finalized (c).
-%
-The flag is initially set to false, avoiding the finalization code to execute.
-Only after the call to \code{\_hold} that we set the flag to true and enable 
-the $fin$ body to execute.
-%
-The complete translation substitutes the side-effect operations with $mem$ 
-expressions (d).
-
-\begin{figure}
-\begin{minipage}[t]{0.26\linewidth}
-\begin{lstlisting}
-do
-  var int* ptr = <...>;
-  await A;
-  finalize
-    _hold(ptr);
-  with
-    _release(ptr);
-  end
-  await B;
-end
-
-.
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.26\linewidth}
-\begin{lstlisting}
-par/or do
-  var int* ptr = <...>;
-  await A;
-  _hold(ptr);
-  await B;
-with
-  { fin
-      _release(ptr); }
-end
-
-.
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.26\linewidth}
-\begin{lstlisting}
-f_ = 0;
-par/or do
-  var int* ptr = <...>;
-  await A;
-  _hold(ptr);
-  f_ = 1;
-  await B;
-with
-  { fin
-      if f_ then
-        _release(ptr);
-      end }
-end
-.
-\end{lstlisting}
-\centering\small{(c)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.18\linewidth}
-\begin{lstlisting}
-mem;
-(
-   mem;
-   await(A);
-   mem;
-   mem;
-   await(B);
-or
-   fin
-     if mem then
-       mem
-     else
-       nop
-)
-\end{lstlisting}
-\centering\small{(d)}
-\end{minipage}
-%
-\caption{
-Three-step translation from concrete to abstract finalization.
-\label{lst.map.fin}
-}
-\end{figure}
+\subsection{ Discussion }
+\label{sec.ceu.dicsussion}
+
+TODO Link the sections, connect the dots, draw conclusions, eliminate remaining xxx.
+
+Timers allows for an internal, defined by the programmer in charge of the 
+specification, definition of simultaneity.
+
+On the one hand, enforcing an execution order for concurrent operations may 
+seen arbitrary and also precludes true parallelism.
+On the other hand, it provides a priority scheme for trails, and makes 
+shared-memory concurrency more tractable.
+%
+For constrained embedded development, we believe that deterministic 
+shared-memory concurrency is beneficial, given the extensive use of memory 
+mapped ports for \emph{I/O} and the lack of hardware support for real 
+parallelism.
+TODO: we could detect that trails do not share memory and execute in multiple 
+threads
+
+On the one hand, enforcing an execution order for concurrent operations may 
+seen arbitrary and also precludes true parallelism.
+On the other hand, it provides a priority scheme for trails, and makes 
+shared-memory concurrency more tractable.
+%
+For constrained embedded development, we believe that deterministic 
+shared-memory concurrency is beneficial, given the extensive use of memory 
+mapped ports for \emph{I/O} and the lack of hardware support for real 
+parallelism.
+%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
+%parallelism is inherent.
 
@@
 
-\subsection{Temporal Analysis for Shared-Memory Concurrency}
+\subsection{Temporal Analysis for Concurrent Statements}
 \label{sec.impl.analysis}
+
+TODO: internal events, emits are propagated from preceding awaits to input
 
@@
 Here, we focus on the algorithm that detects suspicious access to shared 
-variables, as discussed in Section~\ref{sec.ceu.det}.
+variables, as discussed in Section~\ref{sec.ceu.analysis}.
 
@@
 
-Instead of generating code in C, Terra generates the vmt assembly code using 
+Instead of generating code in \C, Terra generates the vmt assembly code using 
 the original \CEU labels for trails entry points.
@@
 
-Many synchronous languages have been designed to interoperate with C, such as 
+Many synchronous languages have been designed to interoperate with \C, such as 
 \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
-\emph{PRET-C}~\cite{rp.pretc} and \emph{Sync-C}~\cite{rp.synchc}.
+\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
 %

### 2016-03-26 paper.tex
...

@@
 Overall, \CEU has a simple semantics with fine-grained execution control, and a 
-straightforward implementation targeting resource-constrained systems.
+straightforward single-threaded implementation targeting resource-constrained 
+systems.
 %
@@
 returns an external file resource as a pointer.
-If the block aborts (ln. 12) during the \code{await 1s} (ln. 9), the file 
+If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
 remains open as a \emph{memory leak}.
@@
    _fwrite(..., f);
-   await 1s;
+   await A;
    _fwrite(..., f);
@@
 
+TODO add a remark about tasks/async/thread in "Discussion"
+
 TODO Link the sections, connect the dots, draw conclusions, eliminate remaining xxx.
@@
 parallelism.
+
 TODO: we could detect that trails do not share memory and execute in multiple 
@@
 
-On the one hand, enforcing an execution order for concurrent operations may 
-seen arbitrary and also precludes true parallelism.
-On the other hand, it provides a priority scheme for trails, and makes 
-shared-memory concurrency more tractable.
-%
-For constrained embedded development, we believe that deterministic 
-shared-memory concurrency is beneficial, given the extensive use of memory 
-mapped ports for \emph{I/O} and the lack of hardware support for real 
-parallelism.
 %For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
@@
 a single event.
-For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
+For this reason, \CEU does not allocate per-trail stacks; instead, all data 
 resides in fixed memory slots---this is true for the program variables as well 
@@
 
+\subsection{Single-Threaded Dispatching}
+
+The implementation of \CEU dispatches active trails sequentially in a single 
+thread, taking no advantage of multi-core CPUs.
+%
+This decision comes not only from the fact that \CEU targets constrained 
+embedded systems with a single CPU, but also because \CEU imposes deterministic 
+execution for intra-reaction statements.
+
+Nonetheless, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
+analysis of \CEU infers precisely trails that are concurrent yet do not share 
+resources.
+Hence, these non-conflicting trails could potentially execute with real 
+parallelism in multiple cores.
+%
+However, our experiments with a multi-threaded implementation in multi-core 
+CPUs execute slower than the single-threaded implementation in the same CPUs.
+Considering that we use \CEU primarily in control-dominated applications, this 
+result is not surprising~\cite{esterel.multicore.1,esterel.multicore.2}.
+We believe that
+    the continuous fork-and-rejoin overhead due to small reactions
+as well as
+    the excessive locality of data due to stackless threads sharing contiguous static memory
+seem to prevent any speed-up gains.
+
+If we consider data-dominated applications, multi-core implementations can 
+offer considerable speed-up gains.
+However, data-intensive computations do not typically require a disciplined 
+step-wise execution and can actually execute in asynchronous calls.
+Esterel provides a \code{task} primitive for this 
+purpose~\cite{esterel.primer}, and \CEU provides an equivalent 
+\code{async/thread} primitive (which are out of the scope of this paper).
+
 \subsection{Interaction with the Environment}
@@
 
-\CEU has a strong influence from Esterel and embraces the disciplined 
+\CEU has a strong influence from Esterel, embracing its disciplined 
 synchronous-reactive model with support for lexical composition of lines of 
@@
 which multiple signals can be active at a clock tick.
-In fact, Esterel is also used in hardware design.
+%In fact, Esterel is also used in hardware design.
 %

### 2016-03-26 paper.tex
...

@@
 
-We completely missed the "finalize" construct of Esterel and recognize the 
-similarity with our proposal.
-We now emphasize that Céu tracks pointers entering and leaving the program to 
-enforce appropriate use of "finalize" (as part of the section about interfacing 
-with C).
-
 \begin{figure}
@@
 
+Since loops in \CEU must contain \code{await} statements, reactions run in 
+bounded time, which guarantees that successive calls to \code{dispatch} never 
+block the scheduler for long.
+However, the code generation phase does not inspect C calls and also has no 
+extra analysis such as for worst-case reaction 
+times~\cite{esterel.wcrt,esterel.fixed}.
+
 \subsection{Interaction with the Environment}

### 2016-03-26 paper.tex
...

@@
 
+Single-thread dispatching may not be suitable for hard real-time activities.
+In a previous work~\cite{ceu.sensys13}, we measure how synchronous lengthy 
+computations in C (e.g., hashing and compression) can block the scheduler and 
+affect higher-priority activities such as a radio driver.
+In such cases, the system requires careful testing to avoid undersized hardware 
+deployment.
+
 \subsection{Interaction with the Environment}
@@
 
-\begin{comment}
-Basically, they map
-
-Basically, it is composed by instructions to define a pending trail (\code{tkins}),
-a pending event(using \code{set\_c} to set a value in the memory),
-a check operation to verify the \code{par/and} termination(\code{chkret}),
-a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
-All pending entries are managed by the scheduler, that also responds for timer and input events.
-All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.
-
-During compilation process the script is broken in trails in the same way it is 
-done in the \CEU compiler explained in Section 4.
-The Terra virtual machine implements a similar \CEU scheduler to execute the 
-script trails, but the VM
-uses the real trails address instead of labels. In this way is possible to jump directly to each script
-partitions without using a dispatcher.
-
-Figure~\ref{lst.impl.assembly} presents the bytecode generated for the initial 
-part for the example of Figure~\ref{lst.impl.trails}.
-After initializing the trail and event slots in the memory, the VM engine starts the execution from the \code{main} label.
-In this example it registers in the trails slots the label addresses for the two block of the \code{par/and} control (lines 3--6).
-The scheduler, first, executes the code in the label \code{and\_sub\_1} that registers the label \code{awk\_e} into the slot of event \code{e} (lines 8--10).
-Then, executes the code in the label \code{and\_sub\_2} that registers the label \code{awk\_temp} into the slot of the event \code{I\_TEMP} (lines 17--18).
-In this moment the VM engine suspends its execution waiting for some new input or timer event.
-In the case of this example, the engine waits for the \code{I\_TEMP} event to continue the script execution.
-
-\begin{figure}
-\begin{minipage}[t]{0.91\linewidth}
-\begin{lstlisting}[language=C,numbers=left,xleftmargin=2.5em,keywords={tkins,end,set_c,ushort,ubyte,exec}]
-  <...> // Initializes the trails and events slots
-
-:main -- Activate TRAILs 0 & 1
-  tkins and_sub_1           // Activate TRAIL 0 - lbl and_sub_1
-  tkins and_sub_2           // Activate TRAIL 1 - lbl and_sub_2
-  end                       // Suspend execution
-
-:and_sub_1
-  set_c ushort slot_e awk_e // await e
-  end                       // Suspend execution
-
-:awk_e
-  set_c ubyte slot_1 1      // Close TRAIL 0 - lbl and_sub_1
-  exec chkret               // goto check par/and termination
-
-:and_sub_2
-  set_c ushort I_TEMP awk_temp // await TEMP
-  end                       // Suspend execution
-
-  <....> // Other entries: chkret, awk_TEMP, etc..
-\end{lstlisting}
-\end{minipage}
-\caption{
-%{\small %\textmd{
-Assembly code for a \code{par/and} control structure.
-%}%}
-\label{lst.impl.assembly}
-}
-\end{figure}
-
-Instead of generating code in \C, Terra generates the vmt assembly code using 
-the original \CEU labels for trails entry points.
-In the final phase for code generation, Terra compiler generates the vmt 
-bytecode file using the program address in place of \CEU labels.
-The use of address allows the \CEU scheduler to execute directly a trail 
-without a index table to replace the switch-case.
-Related to memory layout and trail allocation, Terra uses the same data 
-organization approach for all \CEU controls and variables.
-All \CEU data and control structures for a specific program are determined at 
-compile time.
-Terra arranges all these structures in the beginning of virtual machine memory 
-space.
-The original \CEU run time control functions are maintained with a small 
-modification where accesses to control structures have an off-set relative to 
-the begin of the virtual machine memory space.
-\end{comment}
-
 \section{Related Work}
@@
 
-\begin{comment}
-- Kiel Esterel compiler
-
-- We are not concerned with WCET:
-Predictable system design is concerned with the challenge o
-f building sys-
-tems in such a way that requirements can be guaranteed from th
-e design. This
-means that an off-line analysis should demonstrate satisfa
-ction of timing require-
-ments, subject to assumptions made on operating conditions
-foreseen for the sys-
-tem [Stankovic and Ramamritham 1990]
-
-http://embedded.cs.uni-saarland.de/publications/BuildingTimingPredictableEmbeddedSystems.pdf
-survey of timeing-predictable embedded systems,
-synchronous languages do some at the lang semantic level
-\cite{pret.building}
-\end{comment}
-
 \section{Conclusion}

### 2016-03-27 paper.tex
...

@@
 
+TODO: recover the logical notion of simultaneaty substituting 200ms with TICK
+
 The synchronous hypothesis for \CEU holds if the reactions run faster than the 
@@
 
+TODO: move POE to here, simpler more natural for software dev
+
+- static analsysis also simpler
+    - highlight that is not only important to have a comptutally-sound 
+      algorithm, but that something that programmers can understand
+- same with 1st-class timers, also, too much common to be ignored
+
 TODO add a remark about tasks/async/thread in "Discussion"
@@
 shared-memory concurrency more tractable.
+    - again, simple
 %

### 2016-03-28 paper.tex
...

@@
 % Title portion
-\title{The Design, Semantics, and Implementation of the Synchronous Language 
-\CEU}
+\title{The Design and Implementation of the Synchronous Language \CEU}
 
@@
 %
-The list that follows summarizes the semantic peculiarities of our design:
+The list that follows summarizes the contributions behind the design of \CEU:
 %
@@
 \end{itemize}
-%
-We discuss the design of \CEU and present a formal semantics for a small 
-synchronous kernel that represents a subset of the language covering these new 
-functionalities.
-% TODO: why? stack, fin, principalmente
 %
@@
 microcontroller with 32K of FLASH and 2K of SRAM.
-}, showing that the peculiarities in the semantics of \CEU do not pose 
-practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}
+}, showing the practical aspect of our clean semantics.
+%:\url{http://www.atmel.com/devices/atmega328.aspx}}
 
@@
 differences to Esterel.
-Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
-\CEU.
+%Section~\ref{sec.sem} presents a formal semantics for the control primitives 
+%of \CEU.
 Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
@@
 in the real world.
-Also, since no time regularity is required for ticks~\cite{esterel.multiclock}, 
-the two approaches for Esterel in Figure~\ref{fig.ticks} lead to different 
-behaviors for the same sequence of inputs.
+Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
+approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
+different behaviors for the same sequence of inputs.
 %
@@
 with the enclosing \code{loop} (ln. 4).
+%
+We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
+Section~\ref{sec.ceu.timers}.
+%
 The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
@@
 %
-Here, ``simultaneous'' means ``within 200 milliseconds'', which is a huge 
-amount of time for a language-defined tick, and which would break the 
-synchronous hypothesis.
-%
-A similar implementation for Esterel would not rely on the tick notion of 
-simultaneity either.
-%
-We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
-Section~\ref{sec.ceu.timers}.
-
-TODO: recover the logical notion of simultaneaty substituting 200ms with TICK
+In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
+which is a huge amount of time for a language-defined tick, and which would 
+break the synchronous hypothesis.
+%
+Note that a similar implementation for Esterel would not rely on the tick 
+notion of simultaneity either.
+%
+Note also that the application in \CEU could define an external input 
+\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
+10) to recover the abstract notion of simultaneity.
 
@@
 
+lexical order pragamatics
+    - again, familiar to programmers
+        - register order
+    - paralax, logging
+        - simple, understandeable, imperative
+        - multiple calls to compose, printf, are not commutative
+        - define with intimidating name
+    - implementation aspects in Section XX
+
 TODO: move POE to here, simpler more natural for software dev
@@
 - same with 1st-class timers, also, too much common to be ignored
+
+- understandable semantics
+    - implementable
 
@@
 
-We presented the design, semantics, and implementation of \CEU, a synchronous 
-reactive language inspired by Esterel targeting constrained embedded systems.
+We presented the design and implementation of \CEU, a synchronous reactive 
+language inspired by Esterel targeting constrained embedded systems.
 
@@
 weeks.
-
-% Idexx
 

### 2016-03-28 paper.tex
...

@@
 the language, but explicitly defined for each use case.
-In the end of the section, we present a specification to detect 
-``simultaneous'' button clicks.
-%
 Note that in the case of Esterel, simultaneity depends on the imprecise 
@@
 one another.
+After introducing internal events and first-class timers, we show an example 
+that detects simultaneous button clicks in Section~\ref{sec.ceu.discussion}.
+%
 \item \textbf{Unique input events imply mutual exclusion:}
@@
 
-\begin{figure}
-\begin{minipage}[t]{0.57\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em,mathescape=true]
-input void LEFT_CLICK;
-input void RIGHT_CLICK;
-event void middle_click;
-loop do
-    par/or do
-        AWAIT_AND(LEFT_CLICK, RIGHT_CLICK);
-        emit middle_click;
-    with
-        AWAIT_OR(LEFT_CLICK, RIGHT_CLICK);
-        await 200 ms;
-    end
-end
-\end{lstlisting}
-\end{minipage}
-%
-\begin{minipage}[t]{0.43\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=13]
-#define AWAIT_AND(e1, e2)   \
-    par/and do              \
-        await e1;           \
-    with                    \
-        await e2;           \
-    end
-#define AWAIT_OR(e1, e2)    \
-    par/or do               \
-        await e1;           \
-    with                    \
-        await e2;           \
-    end
-\end{lstlisting}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Application defines that a \code{middle\_click} event occurs whenever 
-both \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} occur within 200 milliseconds.
-\label{lst.simultaneous}
-}
-\end{figure}
-
-Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
-terms of \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} (ln. 1--2).
-If both events occur, we emit the internal event \code{middle\_click} (ln.  
-6--7).
-However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
-we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
-with the enclosing \code{loop} (ln. 4).
-%
-We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
-Section~\ref{sec.ceu.timers}.
-%
-The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
-simple expansions to a \code{par/and} and \code{par/or} for better readability.
-%
-In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
-which is a huge amount of time for a language-defined tick, and which would 
-break the synchronous hypothesis.
-%
-Note that a similar implementation for Esterel would not rely on the tick 
-notion of simultaneity either.
-%
-Note also that the application in \CEU could define an external input 
-\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
-10) to recover the abstract notion of simultaneity.
-
 The synchronous hypothesis for \CEU holds if the reactions run faster than the 
@@
 length~\cite{esterel.fixed}.
+% TODO: Such analsysis is orthogonal
 
@@
 Esterel makes no semantic distinctions between internal and external signals.
-In particular, programs can emit different external input signals 
-simultaneously, with all coexisting during a reaction.
-%
-In \CEU, a reaction starts from an external input event and programs cannot 
-emit inputs at all.
-Therefore, the occurring input event is unique during the entire reaction, 
-resulting in intrinsic queue-based handling.
-%
-In contrast, programs can emit internal events but these follow a stack-based 
-execution policy, similar to subroutine calls in typical programming languages.
+In particular, programs can emit external and internal signals simultaneously, 
+with all coexisting during the entire reaction.
+%
+%In \CEU, a reaction starts from an external input event and programs cannot 
+%emit inputs at all.
+%Therefore, the occurring input event is unique during the entire reaction, 
+%resulting in intrinsic queue-based handling.
+%
+In \CEU, subsequent external events coming from the environment define a 
+timeline, which programs cannot interfere by emitting new external events.
+%
+Howerver, \CEU supports internal events which programs can \code{emit} in 
+addition to \code{await}.
+In contrast with external events, internal events follow a stack-based 
+execution policy similar to subroutine calls in typical programming languages.
 %
@@
 
-The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
-prevents infinite execution by design.
-Both sides of the \code{par/or} have an \code{await} statement to avoid 
-instantaneous termination (ln. 4,7).
-However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
-same reaction that reaches them, the \code{par/or} would terminate and restart 
-the \code{loop} instantaneously, resulting in infinite execution.
-
-\begin{figure}
+\begin{figure}[b]
 \begin{lstlisting}[numbers=left,xleftmargin=2em]
@@
 
-In atypical scenarios requiring immediate awake, delayed awaits can be 
-circumvented by manually copying or transforming the code to execute on awake.
-%
-For instance, sometimes we need to execute a block of code immediately, and 
-then, periodically from internal event requests, as illustrated in 
-Figure~\ref{lst.delay}.
-In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
-periodic code to also execute immediately (ln. 9), and then in reactions to 
-each \code{emit} request (ln. 5).
-If the periodic \code{emit} depends on a condition, then the code 
-transformation becomes more intricate, requiring an extra condition test around 
-the periodic code to prevent its immediate execution.
-%
-On the one hand, we transfer the burden of dealing with these specific corner 
-cases to the programmer.
-On the other hand, we simplify the semantics of the language and eliminate the 
-need for complex analysis to deal with schizophrenic statements.
-
-\begin{figure}[t]
+The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
+prevents infinite execution by design.
+Both sides of the \code{par/or} have an \code{await} statement to avoid 
+instantaneous termination (ln. 4,7).
+However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
+same reaction that reaches them, the \code{par/or} would terminate and restart 
+the \code{loop} instantaneously, resulting in infinite execution.
+
+\begin{figure}
 %\begin{minipage}[t]{0.50\linewidth}
@@
 
+In atypical scenarios requiring immediate awake, delayed awaits can be 
+circumvented by manually copying or transforming the code to execute on awake.
+%
+For instance, sometimes we need to execute a block of code immediately, and 
+then, periodically from internal event requests, as illustrated in 
+Figure~\ref{lst.delay}.
+In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
+periodic code to also execute immediately (ln. 9), and then in reactions to 
+each \code{emit} request (ln. 5).
+If the periodic \code{emit} depends on a condition, then the code 
+transformation becomes more intricate, requiring an extra condition test around 
+the periodic code to prevent its immediate execution.
+%
+On the one hand, we transfer the burden of dealing with these specific corner 
+cases to the programmer.
+On the other hand, we simplify the semantics of the language and eliminate the 
+need for complex analysis to deal with schizophrenic statements.
+
 %\newpage %TTT
@@
 (ln. a:12 and b:12).
+
+\CEU goes one step further and enforces the use of \code{finalize} for system 
+calls that deal with pointers representing resources:
+%
+\begin{itemize}
+\item If a system call \textbf{receives} a pointer from \CEU, the pointer 
+represents a \textbf{local} resource that requires finalization.
+\item If a system call \textbf{returns} a pointer to \CEU, the pointer 
+represents an \textbf{external} resource that requires finalization.
+\end{itemize}
+%
+\CEU tracks the interaction of system calls with pointers and requires 
+finalization clauses to accompany them.
+%
+In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
+(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
+(ln. 5), which is an asynchronous call that transmits the buffer in the 
+background.
+If the block aborts (ln. 11) before receiving an acknowledge from the 
+environment (ln. 9), the local \code{msg} goes out of scope and the external 
+transmission now holds a \emph{dangling pointer}.
+The finalization ensures that the transmission also aborts (ln. 7).
+%
+In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} 
+returns an external file resource as a pointer.
+If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
+remains open as a \emph{memory leak}.
+The finalization ensures that the file closes properly (ln. 6).
+%
+In both cases, the code does not compile if the programmer forgets to use the
+\code{finalize} construct.
 
@@
 
-\CEU goes one step further and enforces the use of \code{finalize} for system 
-calls that deal with pointers representing resources:
-%
-\begin{itemize}
-\item If a system call \textbf{receives} a pointer from \CEU, the pointer 
-represents a \textbf{local} resource that requires finalization.
-\item If a system call \textbf{returns} a pointer to \CEU, the pointer 
-represents an \textbf{external} resource that requires finalization.
-\end{itemize}
-%
-\CEU tracks the interaction of system calls with pointers and requires 
-finalization clauses to accompany them.
-%
-In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
-(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
-(ln. 5), which is an asynchronous call that transmits the buffer in the 
-background.
-If the block aborts (ln. 11) before receiving an acknowledge from the 
-environment (ln. 9), the local \code{msg} goes out of scope and the external 
-transmission now holds a \emph{dangling pointer}.
-The finalization ensures that the transmission also aborts (ln. 7).
-%
-In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} 
-returns an external file resource as a pointer.
-If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
-remains open as a \emph{memory leak}.
-The finalization ensures that the file closes properly (ln. 6).
-%
-In both cases, the code does not compile if the programmer forgets to use the
-\code{finalize} construct.
-
 Note that the illustrative example of Figure~\ref{lst.fin} does not manipulate 
@@
 \label{sec.ceu.dicsussion}
+
+\CEU aims to offer a simpler semantics than Esterel while keeping the power of 
+synchronous languages, with a tractable concurrency model, reactive and 
+deterministic execution, and safe support for orthogonal abortion of lines of 
+execution.
+
+The notion of time based on events instead of ticks 
+(Section~\ref{sec.ceu.exts}) is widespread~\cite{sync_async.whynotthreads} and 
+popular in many software communities, such as web frameworks (e.g., 
+\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
+(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
+and Games~\cite{gamepatterns}.
+%
+Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
+through a queue and are dispatched sequentially and atomically to prevent race 
+conditions.
+%
+We believe that for software design, this approach is more familiar to 
+programmers and simplifies the reasoning about concurrency.
+For instance, the uniqueness of external events in \CEU is a prerequisite for 
+the temporal analysis for concurrent statements.
+
+With the event-based semantics, \CEU does not impose a notion of simultaneity 
+for events, requiring each application to define it only when necessary.
+%
+Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
+terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
+\code{RIGHT\_CLICK} (ln. 1--2).
+If both events occur, we emit the internal event \code{middle\_click} (ln.  
+6--7).
+However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
+we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
+with the enclosing \code{loop} (ln. 4).
+%
+In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
+which is a huge amount of time for a language-defined tick, and which would 
+break the synchronous hypothesis.
+%
+Note that a similar implementation for Esterel would not rely on the tick 
+notion of simultaneity either.
+%
+Note also that the application in \CEU could define an external input 
+\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
+10) to recover the abstract notion of simultaneity.
+
+\begin{figure}
+\begin{minipage}[t]{0.57\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em,mathescape=true]
+input void LEFT_CLICK;
+input void RIGHT_CLICK;
+event void middle_click;
+loop do
+    par/or do
+        AWAIT_AND(LEFT_CLICK, RIGHT_CLICK);
+        emit middle_click;
+    with
+        AWAIT_OR(LEFT_CLICK, RIGHT_CLICK);
+        await 200 ms;
+    end
+end
+\end{lstlisting}
+\end{minipage}
+%
+\begin{minipage}[t]{0.43\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=13]
+#define AWAIT_AND(e1, e2)   \
+    par/and do              \
+        await e1;           \
+    with                    \
+        await e2;           \
+    end
+#define AWAIT_OR(e1, e2)    \
+    par/or do               \
+        await e1;           \
+    with                    \
+        await e2;           \
+    end
+\end{lstlisting}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Application defines that a \code{middle\_click} event occurs whenever 
+both \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} occur within 200 milliseconds.
+%
+The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
+simple expansions to a \code{par/and} and \code{par/or} for better readability.
+\label{lst.simultaneous}
+}
+\end{figure}
+
+internal events
+    - flexible, coroutines, resumable exceptions
+    - yet simple, programmers understand stacks
+    - most of the time doesn't matter
+
+Intra-reaction determinism is another
+The temporal analsysi (Section~\ref{sec.ceu.analysis}).
+
+The temporal analsysi (Section~\ref{sec.ceu.analysis}).
+simple, programmers understand
+
+targets control-intensive applications and provides abstractions that can 
+express program flow specifications concisely.
 
@@
 
-\begin{figure}[t]
+\begin{figure}[b]
 \begin{minipage}[t]{0.31\linewidth}
@@
 
-The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
-and popular in many software communities, such as web frameworks (e.g., 
-\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
-(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
-and Games~\cite{gamepatterns}.
-%
-Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
-through a queue and are dispatched sequentially and atomically to prevent race 
-conditions.
-%
-We believe that for software design, this approach is more familiar to 
-programmers and simplifies the reasoning about concurrency.
-For instance, the uniqueness of external events in \CEU is a prerequisite for 
-the temporal analysis that enables safe shared-memory concurrency.
-
 Many synchronous languages have been designed to interoperate with \C, such as 

### 2016-03-28 paper.tex
...

@@
 Despite its success and influence, Esterel has an overly complex semantics that 
-requires careful static analysis to detect and refuse programs with 
+requires careful static analysis such as to detect and refuse programs with 
 \emph{causality} and \emph{schizophrenia} problems.
@@
 non-deterministic execution for intra-reaction statements, which prevents 
-threads to share memory and interact with stateful system calls safely.
+threads to interact with stateful system calls safely and makes shared-memory 
+concurrency not as straightforward as reading and writing to shared variables.
 %
@@
 %
-Overall, \CEU has a simple semantics with fine-grained execution control, and a 
-straightforward single-threaded implementation targeting resource-constrained 
-systems.
+Overall, \CEU has a simple semantics with fine-grained control for 
+intra-reaction execution which is amenable to a simple temporal analysis to 
+improve safety:
 %
@@
 %
-\item Safe integration with \C via function annotations and enforced 
-finalization for external resources (Section~\ref{sec.ceu.c}).
+\item Safe integration with \C which enforces finalization for external 
+resources (Section~\ref{sec.ceu.c}).
 %
@@
 %
-We also present a lightweight implementation of \CEU with two back ends:
+We also present a lightweight single-threaded implementation of \CEU with two 
+back ends:
 one aiming for resource efficiency and interoperability with \C,
@@
 
-\CEU is a synchronous reactive language inspired by Esterel with support for 
-multiple concurrent lines of execution known as \emph{trails}.
-By reactive, we mean that programs are stimulated by the environment through 
-input events, which are broadcast to all awaiting trails.
-By synchronous, we mean that trails at any given moment are either reacting to 
-the current event or are awaiting another event;
-in other words, trails never react to different events simultaneously.
-
-In the sections that follow, we discuss the main differences between \CEU and 
-Esterel:
-unique and queue-based external events (Section~\ref{sec.ceu.exts}),
-stack-based internal events (Section~\ref{sec.ceu.ints}),
-static temporal analysis for concurrent statements (Section~\ref{sec.ceu.analysis}),
-safe integration with \C (Section~\ref{sec.ceu.c}), and
-first-class synchronized timers (Section~\ref{sec.ceu.timers}).
-
-Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
-implementations in Esterel (a) and \CEU (b) for the following control 
-specification:
+\CEU is a synchronous reactive language inspired by Esterel in which programs 
+advance in a sequence of discrete reactions to external events.
+%
+%Regarding the similarities with Esterel,
+%\CEU has a strong imperative flavor, with explicit control flow through 
+%sequences, loops, and parallel compositions.
+%
+Like Esterel, \CEU is designed for control-intensive applications, supporting 
+concurrent lines of execution, known as \emph{trails}, and broadcast 
+communication through events.
+%
+Internal computations within a reaction (e.g. expressions, assignments, and 
+system calls) are considered to take no time in accordance with the synchronous 
+hypothesis~\cite{rp.hypothesis}.
+An \code{await} is the only statement that halts a running reaction and allows 
+a program to advance in this notion of time.
+%
+To ensure that reactions run in bounded time and programs always progress, 
+loops are statically required to contain at least one \code{await} statement in 
+all possible paths~\cite{ceu.sensys13,esterel.primer}.
+%
+\CEU shares the same limitations with Esterel and synchronous languages in 
+general:
+computations that run in unbounded time (e.g., cryptography, image processing) 
+do not fit the zero-delay hypothesis, and cannot be directly implemented.
+
+Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel (a) and 
+\CEU (b) for the following control specification:
 %
@@
 
-\CEU employs the synchronous model, in which programs advance in a sequence of 
-discrete reactions to external events.
-%
-It also has a strong imperative flavor, with explicit control flow through 
-sequences, loops, and parallels, and also assignments.
-%
-Being designed for control-intensive applications, \CEU provides support for 
-concurrent lines of execution and broadcast communication through events.
-%
-Internal computations within a reaction (e.g. expressions, assignments, and 
-system calls) are considered to take no time in accordance with the synchronous 
-hypothesis~\cite{rp.hypothesis}.
-An \code{await} is the only statement that halts a running reaction and allows 
-a program to advance in this notion of time.
-%
-To ensure that reactions run in bounded time and programs always progress, 
-loops are statically required to contain at least one \code{await} statement in 
-all possible paths~\cite{ceu.sensys13,esterel.primer}.
-%
-\CEU shares the same limitations with Esterel and synchronous languages in 
-general:
-computations that run in unbounded time (e.g., cryptography, image processing) 
-do not fit the zero-delay hypothesis, and cannot be directly implemented.
-
 % TODO: par/or vs abort, strong/weak abortion
+
+In the sections that follow, we discuss the main differences between \CEU and 
+Esterel:
+Unique and queue-based external events
+    (\ref{sec.ceu.exts});
+Stack-based internal events
+    (\ref{sec.ceu.ints});
+Static temporal analysis for concurrent statements
+    (\ref{sec.ceu.analysis});
+Safe integration with \C
+    (\ref{sec.ceu.c});
+and
+First-class synchronized timers
+    (\ref{sec.ceu.timers}).
 
@@
 
-Esterel defines a logical unit of time as a discrete sequence of instants or 
+Esterel defines time as a discrete sequence of logical unit instants or 
 ``ticks''.
-At each tick, depending on external stimuli the environment provides, an 
-arbitrary number of input events can be active simultaneously (including zero).
-The set of input events awakes the program, which must react within the current 
-tick, in bounded time.
-
-In \CEU, each unique external stimulus itself defines a unit of time.
-The execution model for \CEU programs is as follows~\cite{ceu.sensys13}:
+At each tick, the program reacts to an arbitrary number of simultaneous input 
+events (including zero), depending on external stimuli the environment 
+provides.
+In contrast, \CEU defines time as a discrete sequence of reactions to
+external stimuli.
+At each input event, which constitutes a logical unit of time, the program 
+reacts exclusively to it.
+The execution of a program in \CEU is as follows~\cite{ceu.sensys13}:
 %
@@
 %
-A program must react completely to an occurring event completely before 
-handling the next.
+A program must react to an occurring event completely before handling the next.
 %
@@
 divided in units of 10 milliseconds.
-Since Esterel and \CEU define discrete logical units of time, we describe how 
-the events (which are the same in all boxes) fit into their timelines.
-
-\code{Box-1} uses fixed-length ticks in Esterel, within which reactions to 
-occurring inputs have to fit~\cite{esterel.fixed} .
-%
-Here, we assume a \code{R(boot)} reaction which happens before any input at 
-\code{tick-0}.
+Esterel and \CEU use discrete logical units of time in which the occurring 
+events (which are the same in all boxes) have to fit somehow:
+%
+\begin{itemize}
+\item \code{Box-1}:
+Esterel with fixed-length ticks,
+within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
+%
+We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
+input.
 %
 The input \code{A} ``physically'' occurs during the boot reaction but, because 
-time is logical and discrete, the event is delayed to the next tick.
-%
-Note that the reaction \code{R(A)} takes more time than \code{tick-1}, causing 
-a \emph{timing violation} in the program execution (after the end of 
-\code{tick-1} and before the end of \code{R(A)}).
+time is discrete, its corresponding reaction only executes inthe next tick.
+%
+Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
+\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
 %
@@
 %
-Since no new events occurred during \code{tick-2}, the CPU stays idle during 
-the whole \code{tick-3}.
+Since no new events occur during \code{tick-2}, the CPU stays idle during the 
+whole \code{tick-3}.
 %
@@
 occur ``simultaneously'' during the idle \code{tick-3}.
-However, because the program cannot detect both occurrences of \code{E}, only 
-one is considered in \code{R(D+E)}.
-
-\code{Box-2} is an alternative that allows variable-length ticks in Esterel 
-that adjust to the time the corresponding reaction takes to 
-complete~\cite{esterel.variable}.
-This approach avoids the time violation for \code{R(A)} and also yields
+However, the program can only detect one occurrence of \code{E}, which is 
+considered in \code{R(D+E)}.
+%
+\item \code{Box-2}:
+Esterel with variable-length ticks that adjust to the CPU time the 
+corresponding reaction takes to complete~\cite{esterel.variable}.
+This approach avoids the time violation for \code{R(A)} and also results in
 smaller idle periods.
@@
 Similarly to the fixed-tick approach, only one of the two simultaneous 
-occurrences of \code{E} is considered on \code{tick-5}.
-
-\code{Box-3} illustrates the unique and queue-based policy for input events in 
-\CEU.
+occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
+takes too long.
+%
+\item \code{Box-3}:
+\CEU with unique and queue-based policy for input events.
 We also assume a \code{R(boot)} reaction before any input.
 %
-Because the event \code{A} is unique during \code{tick-0}, the behavior in \CEU 
-is similar to \code{Box-2} for the first two units of time.
-%
-However, the events \code{B} and \code{C} are not simultaneous in \CEU and are 
-handled in subsequent reactions \code{R(B)} and \code{R(C)}.
+Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
+behavior in \CEU is similar to \code{Box-2} for the first two units of time 
+(\code{boot-0} and \code{A-1}).
+%
+However, \CEU does not consider the events \code{B} and \code{C} simultaneous, 
+which are handled in subsequent reactions \code{R(B)} and \code{R(C)}.
 %
@@
 independently, which are handled in sequence.
-
-We decided for the unique and queue-based semantics for the reasons that 
-follow:
+\end{itemize}
+
+We decided for the unique and queue-based semantics in \CEU for the reasons 
+that follow:
 %
@@
 %
-\item \textbf{Events are never simultaneous:}
+\item \textbf{Events are never absolutely simultaneous:}
 From a rigorous point of view, event occurrences are infinitesimal, having zero 
@@
 This way, we believe that the notion of simultaneity should not be imposed by 
-the language, but explicitly defined for each use case.
-Note that in the case of Esterel, simultaneity depends on the imprecise 
-definition of ticks.
-For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
-simultaneous, even though \code{A} and \code{B} actually happen much closer to 
-one another.
-After introducing internal events and first-class timers, we show an example 
-that detects simultaneous button clicks in Section~\ref{sec.ceu.discussion}.
+the language, but defined explicitly for each use case.
+In the case of Esterel, simultaneity depends on the imprecise length of 
+discrete ticks.
+For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
+events \code{B} and \code{C} are simultaneous, even though \code{A} and 
+\code{B} actually happen much closer to one another.
+In Section~\ref{sec.ceu.discussion}, after introducing internal events and 
+first-class timers, we show an example that detects simultaneous button clicks.
 %
 \item \textbf{Unique input events imply mutual exclusion:}
-Because they are atomic, reactions to different events never overlap.
-The possibility to reason about each input individually is a prerequisite for 
-the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
+Reactions to multiple events never overlap because they are atomic.
+Automated mutual exclusion simplifies the reasoning about the program and is a 
+prerequisite for the temporal analysis to be discussed in 
+Section~\ref{sec.ceu.analysis}.
 \end{itemize}
@@
 rate of incoming input events.
-Otherwise, the system will accumulate delays between the real occurrence and 
-actual reaction to events.
+Otherwise, the application continuously accumulates delays between the real 
+occurrence and actual reaction to events.
 %
@@
 the more inputs to handle, the longer the reaction takes, and the more inputs 
-will accumulate for subsequent ticks.
+accumulates for subsequent ticks.
 %
@@
 
+TODO: queue-based/unique in the semantics of the language and not in the 
+library
+
+permeates all aspects
+
 \CEU aims to offer a simpler semantics than Esterel while keeping the power of 

### 2016-03-29 paper.tex
...

@@
 Overall, \CEU has a simple semantics with fine-grained control for 
-intra-reaction execution which is amenable to a simple temporal analysis to 
-improve safety:
+intra-reaction execution which is amenable to a simple temporal analysis that
+improves safety.
 %
@@
 \item Unique and queue-based external events, which define the notion of time 
-in \CEU (Section~\ref{sec.ceu.exts}).
+in \CEU.
+% (Section~\ref{sec.ceu.exts}).
 %
 \item Stack-based internal events for intra-reaction communication, which also
-provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
-%
-\item Static temporal analysis to detect suspicious concurrent statements 
-(Section~\ref{sec.ceu.analysis}).
+provides a limited form of coroutines.
+% (Section~\ref{sec.ceu.ints}).
+%
+\item Static temporal analysis to detect suspicious concurrent statements.
+%(Section~\ref{sec.ceu.analysis}).
 %
 \item Safe integration with \C which enforces finalization for external 
-resources (Section~\ref{sec.ceu.c}).
-%
-\item First-class synchronized timers with a dedicated syntax 
-(Section~\ref{sec.ceu.timers}).
+resources.
+% (Section~\ref{sec.ceu.c}).
+%
+\item First-class synchronized timers with a dedicated syntax.
+%(Section~\ref{sec.ceu.timers}).
 \end{itemize}
@@
 one aiming for resource efficiency and interoperability with \C,
-and another based on a virtual machine that allows remote reprogramming.
+and another as a virtual machine that allows remote reprogramming.
 %
@@
 %
-We evaluated the expressiveness of \CEU in comparison to event-driven code in 
-\C and attested a reduction in source code size (around 25\%) with a small 
-increase in memory usage (around 5--10\% for \emph{text} and 
+We evaluated the expressiveness of \CEU in comparison to hand-crafted 
+event-driven code in \C and attested a reduction in source code size (around 
+25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
 \emph{data})~\cite{ceu.sensys13}.
@@
 translated almost identically in the two languages (ln. 3--8, in both 
-implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
-constructs are equivalent.
+implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
+are equivalent.
 %
@@
 
-In the sections that follow, we discuss the main differences between \CEU and 
-Esterel:
+In the subsections that follow, we discuss the main differences between \CEU 
+and Esterel:
 Unique and queue-based external events
@@
     (\ref{sec.ceu.timers}).
+We finish with a summary of our design
+    (\ref{sec.ceu.discussion}).
 
@@
 At each tick, the program reacts to an arbitrary number of simultaneous input 
-events (including zero), depending on external stimuli the environment 
-provides.
+events, depending on external stimuli the environment provides.
 In contrast, \CEU defines time as a discrete sequence of reactions to
-external stimuli.
+unique input events.
 At each input event, which constitutes a logical unit of time, the program 
@@
 In practice, if a new external input event occurs while a reaction chain is 
-running (step 2), it is enqueued to run in the next reaction.
+running (step 2), it is enqueued to occur in the next reaction.
 
@@
 \begin{itemize}
-\item \code{Box-1}:
-Esterel with fixed-length ticks,
+\item \code{[Box-1]}:
+``Esterel with fixed-length ticks'',
 within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
@@
 The input \code{A} ``physically'' occurs during the boot reaction but, because 
-time is discrete, its corresponding reaction only executes inthe next tick.
+time is discrete, its corresponding reaction only executes in the next tick.
 %
@@
 %
-\item \code{Box-2}:
-Esterel with variable-length ticks that adjust to the CPU time the 
+\item \code{[Box-2]}:
+``Esterel with variable-length ticks'', which are adjusted to the CPU time the 
 corresponding reaction takes to complete~\cite{esterel.variable}.
@@
 For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
-start reaction \code{R(D)} on \code{tick-4}.
+react alone as \code{R(D)} on \code{tick-4}.
 Similarly to the fixed-tick approach, only one of the two simultaneous 
@@
 %
-\item \code{Box-3}:
-\CEU with unique and queue-based policy for input events.
+\item \code{[Box-3]}:
+``\CEU with unique and queue-based input events''.
 We also assume a \code{R(boot)} reaction before any input.
@@
 %
-However, \CEU does not consider the events \code{B} and \code{C} simultaneous, 
-which are handled in subsequent reactions \code{R(B)} and \code{R(C)}.
-%
-We assume that the CPU time for \code{R(B+C)} in Esterel is roughly the same as 
-\code{R(B)+R(C)} in \CEU.
+However, \CEU does not consider the events \code{B} and \code{C} as 
+simultaneous, and handles each in subsequent reactions \code{R(B)} and 
+\code{R(C)}.
+%
+We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
+\CEU to be roughly the same.
 This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.

### 2016-03-29 paper.tex
...

@@
 Otherwise, the application continuously accumulates delays between the real 
-occurrence and actual reaction to events.
+occurrence and actual reaction of a given event.
 %
@@
 %
-Howerver, \CEU supports internal events which programs can \code{emit} in 
-addition to \code{await}.
+To cope with intra-reaction communication, \CEU supports internal events which 
+programs can \code{emit} and \code{await}.
 In contrast with external events, internal events follow a stack-based 
@@
 For instance, sometimes we need to execute a block of code immediately, and 
-then, periodically from internal event requests, as illustrated in 
+then periodically from internal event requests, as illustrated in 
 Figure~\ref{lst.delay}.
@@
 \emph{SOL}~\cite{wsn.sol},
-\emph{SC}~\cite{rp.sychc}, and
+\emph{SC}~\cite{rp.synchc}, and
 \emph{PRET-C}~\cite{rp.pretc},
@@
 For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
-which define a shared variable (ln. 2), and assign to it parallel trails (ln.  
-5, 8).
+which define a shared variable (ln. 2), and assign to it in parallel trails 
+(ln.  5, 8).
 %
@@
 %
-In the example \code{(b)}, the two assignments to \code{y} execute from a 
-reaction to the same event \code{A}, and are simultaneous from an external 
-point of view.
+In the example \code{(b)}, the two assignments to \code{y} are simultaneous 
+because they execute in reaction to the same event \code{A}.
 Since \CEU employs lexical order for intra-reaction statements, the execution 
@@
 %
-However, any (apparently innocuous) change in the order of trails can change 
-the semantics of a program, which we consider unsafe.
+However, an (apparently innocuous) change in the order of trails changes the 
+semantics of the program, which we consider unsafe.
 
@@
 The algorithm for the analysis, which is depicted in 
-Section~\ref{sec.impl.analsyis}, inspects all possible \code{await} statements 
-that precede a variable access and hold the associated awaking events.
+Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
+that precede a variable access and keep a list with all corresponding awaking 
+events.
 Then, it checks all accesses in parallel trails to see if they share an awaking 
@@
 within reactions.
-Otherwise, any two trail segments in parallel could be concurrent, even if they 
+Otherwise, any two trail segments in parallel can be concurrent, even if they 
 react to different input events.
@@
 %
-Evidently, programs should only resort to \C for asynchronous functionality, 
+This way, programs should only resort to \C for asynchronous functionality, 
 such as non-blocking I/O, or simple \code{struct} accessors, but never for 
-control purposes.
+control purposes (the compiler can restrict the available symbols in \C).
 
@@
 \end{lstlisting}
-\centering\small{(a)}
+\centering\small{(a) Definition and use of external symbols}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{(b)}
+\centering\small{(b) Annotations for the external symbols in (a).}
 \end{minipage}
 %\rule{8.4cm}{0.37pt}
-\caption{ TODO
+\caption{ The unsafe program in (a) only compiles with the annotations in (b).
 \label{lst.c}
@@
 
-In addition, \CEU also employs the temporal analysis for calls and accesses to 
-external symbols in \C.
+We extend the temporal analysis of \CEU for calls and accesses to external 
+symbols in \C.
 %
 As an example, the program in Figure~\ref{lst.c}.a defines four external 
-symbols inside a \code{native} block for standard declarations in \C (ln.  
+symbols inside a \code{native} block with standard declarations in \C (ln.  
 1--6).
@@
 %
-Since \CEU does not inspect any code defined in \C, it has no idea about the 
-meaning of the symbols and complains about suspicious concurrent accesses 
-between \code{\_f} against all symbols in the second trail.
+Since \CEU does not inspect any code defined in \C, it complains about 
+suspicious concurrent accesses between \code{\_f} and all symbols in the second 
+trail.
 
@@
 
-Esterel's \code{abort} and \CEU's \code{par/or} statements illustrate how 
+With the annotations in \code{(b)}, the program in \code{(a)} compiles 
+successfully.
+
+Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish how 
 synchronous languages can abort awaiting lines of execution without tweaking 
@@
 %
-For this reason, Esterel and \CEU provide a \code{finalize} construct with a 
-clause that executes even if the enclosing block aborts and does not terminate 
-normally.
-
-Figure~\ref{lst.fin} uses \code{lock} and \code{unlock} calls to represent an 
-external resource.
-The normal behavior is to \code{lock} the resource, perform some operations in 
-subsequent reactions to input \code{A}, and then \code{unlock} the resource.
-However, if the aborting input \code{B} occurs after the \code{lock} but before 
-the reactions to \code{A}, we still want to call \code{unlock} to safely 
-release the resource.
-%
-In Esterel and \CEU, the finalize clauses (ln. a:10 and b:5) are automatically 
-called if the enclosing blocks (ln. a:3--1 and b:3--10) are externally aborted 
-(ln. a:12 and b:12).
+For this reason, Esterel and \CEU provide a \code{finalize} construct to 
+unconditionally execute a series of statements even if the enclosing block is 
+aborted and does not terminate normally.
+
+The example in Figure~\ref{lst.fin} in Esterel (\code{(a)}) and \CEU 
+(\code{(b)}) uses the \code{lock} and \code{unlock} calls which represent 
+accesses to an external resource.
+The normal behavior is to \code{lock} the resource (ln. \code{(a)}:4 and 
+\code{(b)}:3), perform some operations in subsequent reactions to input 
+\code{A} (ln. \code{(a)}:5--8 and \code{(b)}:7--10), and then \code{unlock} the 
+resource (ln. \code{(a)}:10 and \code{(b)}:5).
+However, if the aborting input \code{B} (ln. \code{(a)}:12 and \code{(b)}:12) 
+occurs after the \code{lock} but before the reactions to \code{A}, we still 
+want to call \code{unlock} to safely release the resource.
+%
+In Esterel and \CEU, the finalize clauses (ln. \code{(a)}:10 and \code{(b)}:5) 
+are automatically called if the enclosing blocks (ln. \code{(a)}:3--1 and 
+\code{(b)}:3--10) are externally aborted (ln. \code{(a)}:12 and \code{(b)}:12).
 
@@
 \subsection{ Discussion }
-\label{sec.ceu.dicsussion}
+\label{sec.ceu.discussion}
 

### 2016-03-29 paper.tex
...

@@
 \newcommand{\code}[1] {{\small{\texttt{#1}}}}
+\newcommand{\ax}{\code{[a]}\xspace}
+\newcommand{\bx}{\code{[b]}\xspace}
 \newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}
@@
 
-Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel (a) and 
-\CEU (b) for the following control specification:
+Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
+\CEU \bx for the following control specification:
 %
@@
 \end{lstlisting}
-\centering\small{(a) Esterel}
+\centering\small{\ax Esterel}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{(b) \CEU}
+\centering\small{\bx \CEU}
 \end{minipage}
@@
 Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
-Esterel (a) and \CEU (b).
+Esterel \ax and \CEU \bx.
 %
@@
 \end{lstlisting}
-\centering\small{(a)}
+\centering\small{\ax}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{(b)}
+\centering\small{\bx}
 \end{minipage}
@@
 \caption{ Shared-memory concurrency in \CEU:
-Example (a) is safe because the trails access \code{x} atomically in different 
+Example \ax is safe because the trails access \code{x} atomically in different 
 reactions;
-Example (b) is unsafe because both trails access \code{y} in the same reaction.
+Example \bx is unsafe because both trails access \code{y} in the same reaction.
 \label{lst.shared}
@@
 %
-In the example \code{(a)}, the two assignments to \code{x} execute from 
-reactions to different events \code{A} and \code{B} which, by definition, 
-cannot occur simultaneously (Section~\ref{sec.ceu.exts}).
+In the example \ax, the two assignments to \code{x} execute from reactions to 
+different events \code{A} and \code{B} which, by definition, cannot occur 
+simultaneously (Section~\ref{sec.ceu.exts}).
 Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
@@
 %
-In the example \code{(b)}, the two assignments to \code{y} are simultaneous 
-because they execute in reaction to the same event \code{A}.
+In the example \bx, the two assignments to \code{y} are simultaneous because 
+they execute in reaction to the same event \code{A}.
 Since \CEU employs lexical order for intra-reaction statements, the execution 
@@
 \begin{itemize}
-\item The assignments to \code{x} in lines 2 and 5 in \code{(a)}
+\item The assignments to \code{x} in lines 2 and 5 in \ax
       \textbf{cannot} be concurrent because they are not in parallel trails.
-\item The assignments to \code{x} in lines 5 and 8 in \code{(a)}
+\item The assignments to \code{x} in lines 5 and 8 in \ax
       \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
       the same reaction.
-\item The assignments to \code{y} in lines 5 and 8 in \code{(b)}
+\item The assignments to \code{y} in lines 5 and 8 in \bx
       \textbf{can} be concurrent because they are in parallel trails and 
@@
 \end{lstlisting}
-\centering\small{(a) Definition and use of external symbols}
+\centering\small{\ax Definition and use of external symbols}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{(b) Annotations for the external symbols in (a).}
+\centering\small{\bx Annotations for the external symbols in \ax.}
 \end{minipage}
 %\rule{8.4cm}{0.37pt}
-\caption{ The unsafe program in (a) only compiles with the annotations in (b).
+\caption{ The unsafe program in \ax only compiles with the annotations in \bx.
 \label{lst.c}
@@
 The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
-compiler about the semantics of the \C symbols in use in \code{(a)}:
+compiler about the semantics of the \C symbols in use in \ax:
 %
@@
 
-With the annotations in \code{(b)}, the program in \code{(a)} compiles 
-successfully.
+With the annotations in \bx, the program in \ax compiles successfully.
 
@@
 
-The example in Figure~\ref{lst.fin} in Esterel (\code{(a)}) and \CEU 
-(\code{(b)}) uses the \code{lock} and \code{unlock} calls which represent 
-accesses to an external resource.
-The normal behavior is to \code{lock} the resource (ln. \code{(a)}:4 and 
-\code{(b)}:3), perform some operations in subsequent reactions to input 
-\code{A} (ln. \code{(a)}:5--8 and \code{(b)}:7--10), and then \code{unlock} the 
-resource (ln. \code{(a)}:10 and \code{(b)}:5).
-However, if the aborting input \code{B} (ln. \code{(a)}:12 and \code{(b)}:12) 
-occurs after the \code{lock} but before the reactions to \code{A}, we still 
-want to call \code{unlock} to safely release the resource.
-%
-In Esterel and \CEU, the finalize clauses (ln. \code{(a)}:10 and \code{(b)}:5) 
-are automatically called if the enclosing blocks (ln. \code{(a)}:3--1 and 
-\code{(b)}:3--10) are externally aborted (ln. \code{(a)}:12 and \code{(b)}:12).
+The example in Figure~\ref{lst.fin} in Esterel \ax and \CEU \bx uses the 
+\code{lock} and \code{unlock} calls which represent accesses to an external 
+resource.
+The normal behavior is to \code{lock} the resource (ln. \ax:4 and \bx:3), 
+perform some operations in subsequent reactions to input \code{A} (ln.  
+\ax:5--8 and \bx:7--10), and then \code{unlock} the resource (ln.  \ax:10 and 
+\bx:5).
+However, if the aborting input \code{B} (ln. \ax:12 and \bx:12) occurs after 
+the \code{lock} but before the reactions to \code{A}, we still want to call 
+\code{unlock} to safely release the resource.
+%
+In Esterel and \CEU, the finalize clauses (ln. \ax:10 and \bx:5) are 
+automatically called if the enclosing blocks (ln. \ax:3--1 and \bx:3--10) are 
+externally aborted (ln. \ax:12 and \bx:12).
 
@@
 \end{lstlisting}
-\centering\small{(a) Local resource finalization}
+\centering\small{\ax Local resource finalization}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{(b) External resource finalization}
+\centering\small{\bx External resource finalization}
 \end{minipage}
@@
 Figure~\ref{lst.impl.trails} illustrates the generation process.
-For the program in (a), the compiler extracts the entry points and associated 
+For the program in \ax, the compiler extracts the entry points and associated 
 trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln. 7).
@@
 The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
-in (b)).
+in \bx).
 The state of trails translates to a vector of type \code{trail\_t} with the 
@@
 %{\small %\textmd{
-(a) Static allocation of trails: the comments identify the trail indexes 
+\ax Static allocation of trails: the comments identify the trail indexes 
 inferred by the compiler;
-(b) Entry-point labels: each trail segment has an associated numeric identifier 
+\bx Entry-point labels: each trail segment has an associated numeric identifier 
 generated by the compiler.
@@
 %{\small %\textmd{
-(a) \CEU interface with customized \VM.
-%
-(b) The routine \code{VM.out} redirects all output events to the corresponding 
+\ax \CEU interface with customized \VM.
+%
+\bx The routine \code{VM.out} redirects all output events to the corresponding 
 OS calls (ln.  1--8).

### 2016-03-30 paper.tex
...

@@
     (\ref{sec.ceu.timers}).
-We finish with a summary of our design
-    (\ref{sec.ceu.discussion}).
+%We finish with a summary of our design
+    %(\ref{sec.ceu.discussion}).
 
@@
 \code{B} actually happen much closer to one another.
-In Section~\ref{sec.ceu.discussion}, after introducing internal events and 
+In Section~\ref{sec.ceu.timers}, after introducing internal events and 
 first-class timers, we show an example that detects simultaneous button clicks.
@@
 \setlength{\itemsep}{0pt}
-\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
+\item 1st trail awakes (ln. 4), broadcasts \code{b}, and pauses.
 \item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
+\item No other trails awake to \code{b}.
 \item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
@@
 \end{lstlisting}
-\centering\small{\ax}
+\centering\small{\ax Accesses to \code{x} are safe}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{\bx}
+\centering\small{\bx Accesses to \code{y} are unsafe}
 \end{minipage}
@@
 %
-This way, programs should only resort to \C for asynchronous functionality, 
-such as non-blocking I/O, or simple \code{struct} accessors, but never for 
-control purposes (the compiler can restrict the available symbols in \C).
+This way, programs should only resort to \C for asynchronous functionality%
+\footnote{
+In \CEU, it is possible to restrict the available \C symbols as a compile-time 
+option.
+}, such as non-blocking I/O, or simple \code{struct} accessors, but never for 
+control purposes.
 
@@
 \end{lstlisting}
-\centering\small{\ax Definition and use of external symbols}
+\centering\small{\ax Definitions and uses of symbols}
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{\bx Annotations for the external symbols in \ax.}
+\centering\small{\bx Annotations for the symbols in \ax}
 \end{minipage}
@@
 
-We extend the temporal analysis of \CEU for calls and accesses to external 
-symbols in \C.
+The temporal analysis of Section~\ref{sec.ceu.analysis} also considers calls 
+and accesses to external symbols in \C.
 %
@@
 during the boot reaction (ln. 8,10):
-the first trail calls the symbol \code{\_f},
-while the second calls \code{\_g} and \code{\_id}, and also accesses 
-\code{\_NUM}.
-%
-Since \CEU does not inspect any code defined in \C, it complains about 
-suspicious concurrent accesses between \code{\_f} and all symbols in the second 
-trail.
+the first trail calls symbol \code{\_f},
+while the second calls \code{\_g} and \code{\_id}, and also reads \code{\_NUM}.
+%
+Since \CEU does not inspect any code in \C, it complains about suspicious 
+concurrent accesses between \code{\_f} and all symbols in the second trail.
 
 The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
-compiler about the semantics of the \C symbols in use in \ax:
+compiler about the semantics of the \C symbols in use in \ax.
+With the annotations in \bx, the program in \ax compiles successfully:
 %
@@
 
-With the annotations in \bx, the program in \ax compiles successfully.
-
-Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish how 
-synchronous languages can abort awaiting lines of execution without tweaking 
-them with synchronization primitives.
+Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish 
+synchronous from asynchronous languages with respect to orthogonal abortion of
+lines of execution (i.e., without tweaking them with synchronization 
+primitives)~\cite{esterel.preemption}.
 In contrast, traditional (asynchronous) multi-threaded languages cannot express 
-thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
+thread termination safely~\cite{sync_async.threadsstop}.
 %
@@
 
-\subsection{ Discussion }
-\label{sec.ceu.discussion}
-
-TODO: queue-based/unique in the semantics of the language and not in the 
-library
-
-permeates all aspects
-
-\CEU aims to offer a simpler semantics than Esterel while keeping the power of 
-synchronous languages, with a tractable concurrency model, reactive and 
-deterministic execution, and safe support for orthogonal abortion of lines of 
-execution.
-
-The notion of time based on events instead of ticks 
-(Section~\ref{sec.ceu.exts}) is widespread~\cite{sync_async.whynotthreads} and 
-popular in many software communities, such as web frameworks (e.g., 
-\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
-(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
-and Games~\cite{gamepatterns}.
-%
-Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
-through a queue and are dispatched sequentially and atomically to prevent race 
-conditions.
-%
-We believe that for software design, this approach is more familiar to 
-programmers and simplifies the reasoning about concurrency.
-For instance, the uniqueness of external events in \CEU is a prerequisite for 
-the temporal analysis for concurrent statements.
-
-With the event-based semantics, \CEU does not impose a notion of simultaneity 
-for events, requiring each application to define it only when necessary.
+In Section~\ref{sec.ceu.exts}, we argue that events occurrences are 
+infinitesimal and can never be absolutely simultaneous.
+However, the ``sensation of simultaneity'' is not infinitesimal, but increases 
+with the inaccuracy of the observer (e.g., a human being).
+Therefore, simultaneity should be defined with respect to the observer, which 
+varies from case to case.
+
+First-class timers simplify the implementation of application-defined 
+simultaneity.
 %
@@
 \end{figure}
-
-internal events
-    - flexible, coroutines, resumable exceptions
-    - yet simple, programmers understand stacks
-    - most of the time doesn't matter
-
-Intra-reaction determinism is another
-The temporal analsysi (Section~\ref{sec.ceu.analysis}).
-
-The temporal analsysi (Section~\ref{sec.ceu.analysis}).
-simple, programmers understand
-
-targets control-intensive applications and provides abstractions that can 
-express program flow specifications concisely.
-
-lexical order pragamatics
-    - again, familiar to programmers
-        - register order
-    - paralax, logging
-        - simple, understandeable, imperative
-        - multiple calls to compose, printf, are not commutative
-        - define with intimidating name
-    - implementation aspects in Section XX
-
-TODO: move POE to here, simpler more natural for software dev
-
-- static analsysis also simpler
-    - highlight that is not only important to have a comptutally-sound 
-      algorithm, but that something that programmers can understand
-- same with 1st-class timers, also, too much common to be ignored
-
-- understandable semantics
-    - implementable
-
-TODO add a remark about tasks/async/thread in "Discussion"
-
-TODO Link the sections, connect the dots, draw conclusions, eliminate remaining xxx.
-
-Timers allows for an internal, defined by the programmer in charge of the 
-specification, definition of simultaneity.
-
-On the one hand, enforcing an execution order for concurrent operations may 
-seen arbitrary and also precludes true parallelism.
-On the other hand, it provides a priority scheme for trails, and makes 
-shared-memory concurrency more tractable.
-    - again, simple
-%
-For constrained embedded development, we believe that deterministic 
-shared-memory concurrency is beneficial, given the extensive use of memory 
-mapped ports for \emph{I/O} and the lack of hardware support for real 
-parallelism.
-
-TODO: we could detect that trails do not share memory and execute in multiple 
-threads
-
-%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
-%parallelism is inherent.
 
@@
 
+The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
+and popular in many software communities, such as web frameworks (e.g., 
+\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
+(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
+and Games~\cite{gamepatterns}.
+%
+Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
+through a queue and are dispatched sequentially and atomically to prevent race 
+conditions.
+%
+We believe that for software design, this approach is more familiar to 
+programmers and simplifies the reasoning about concurrency.
+For instance, the uniqueness of external events in \CEU is a prerequisite for 
+the temporal analysis that enables safe shared-memory concurrency.
+
 Many synchronous languages have been designed to interoperate with \C, such as 
@@
 
+\begin{comment}
+%\subsection{ Discussion }
+\label{sec.ceu.discussion}
+
+TODO: related
+- simplicity
+    - intra-reaction det
+    - internal events?
+        - more fine grained
+
+% SIMPLE
+
+\CEU aims to offer a simpler semantics than Esterel without giving up its 
+tractable synchronous concurrency model:
+
+% unique external events
+
+The notion of time based on events instead of ticks 
+(Section~\ref{sec.ceu.exts}) is widespread~\cite{sync_async.whynotthreads} and 
+popular in many software communities, such as web frameworks (e.g., 
+\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
+(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
+and Games~\cite{gamepatterns}.
+%
+Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
+through a queue and are dispatched sequentially and atomically to prevent race 
+conditions.
+%
+We believe that for software design, this approach is more familiar to 
+programmers and simplifies the reasoning about concurrency.
+For instance, the uniqueness of external events in \CEU is a prerequisite for 
+the temporal analysis for concurrent statements.
+
+% internal events
+
+Internal events
+
+% lexical order
+
+    - flexible, coroutines, resumable exceptions
+    - yet simple, programmers understand stacks
+    - most of the time doesn't matter
+
+Intra-reaction determinism is another
+The temporal analsysi (Section~\ref{sec.ceu.analysis}).
+
+The temporal analsysi (Section~\ref{sec.ceu.analysis}).
+simple, programmers understand
+
+targets control-intensive applications and provides abstractions that can 
+express program flow specifications concisely.
+
+lexical order pragamatics
+    - again, familiar to programmers
+        - register order
+    - paralax, logging
+        - simple, understandeable, imperative
+        - multiple calls to compose, printf, are not commutative
+        - define with intimidating name
+    - implementation aspects in Section XX
+
+TODO: move POE to here, simpler more natural for software dev
+
+- static analsysis also simpler
+    - highlight that is not only important to have a comptutally-sound 
+      algorithm, but that something that programmers can understand
+- same with 1st-class timers, also, too much common to be ignored
+
+- understandable semantics
+    - implementable
+
+TODO add a remark about tasks/async/thread in "Discussion"
+
+TODO Link the sections, connect the dots, draw conclusions, eliminate remaining xxx.
+
+Timers allows for an internal, defined by the programmer in charge of the 
+specification, definition of simultaneity.
+
+On the one hand, enforcing an execution order for concurrent operations may 
+seen arbitrary and also precludes true parallelism.
+On the other hand, it provides a priority scheme for trails, and makes 
+shared-memory concurrency more tractable.
+    - again, simple
+%
+For constrained embedded development, we believe that deterministic 
+shared-memory concurrency is beneficial, given the extensive use of memory 
+mapped ports for \emph{I/O} and the lack of hardware support for real 
+parallelism.
+
+TODO: we could detect that trails do not share memory and execute in multiple 
+threads
+
+%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
+%parallelism is inherent.
+
+Furthermore, our broad safety analysis, encompassing all proposed concurrency 
+mechanisms, ensures that the high degree of concurrency in WSNs does not pose 
+safety threats to applications.
+%
+%Even with the restrictions that were needed to enable this analysis, \CEU was 
+%shown to be expressive enough to imply in a decrease in code complexity for a 
+%range of typical WSN applications.
+%
+%By restricting the expressiveness of the language, we could enable a broad 
+%safety analysis in programs at compile time.
+%The analysis encompasses all proposed concurrency mechanisms, such as parallel 
+%compositions, first-class timers, and communication via events.
+%
+
+% SAFE
+
+Esterel XXX
+thanks to the uniqueness of events, pointer tracking, and first-class timers, 
+\CEU also provides additional safety protections:
+%
+No concurrency in accesses to shared variables and system calls sharing a 
+resource
+Finalization for blocks going out of scope
+Auto-adjustment for timers in sequence and synchronization for timers in parallel
+
+These properties are desirable in any application and are guaranteed as 
+preconditions in \CEU by design.
+Ensuring or even extracting these properties from less restricted languages 
+requires significant manual analysis.
+
+% LIMITATIONS
+
+Even though the achieved expressiveness and overhead of \CEU meet the 
+requirements of WSNs, its design imposes two inherent limitations:
+the lack of dynamic loading which would forbid the static analysis,
+and the lack of hard real-time guarantees.
+
+Regarding the first limitation, dynamic features are already discouraged due to 
+resource constraints.
+For instance, even object-oriented languages targeting WSNs forbid dynamic 
+allocation~\cite{wsn.flowtalk,wsn.virgil}.
+%
+Given that we focus on system-level development which does not require rich 
+dynamic functionality, we leave for a future work an in-depth discussion about 
+this issue.
+Note that queues, stacks, and other simple dynamic data structures for handling 
+message packets can be made available to \CEU through its safe integration with 
+$C$.
+
+To deal with the second limitation, which can be critical in the presence of 
+lengthy computations, we can consider the following approaches:
+(1) manually placing \code{pause} statements in unbounded loops;
+(2) integrating \CEU with a preemptive system.
+%
+The first option requires the lengthy operations to be rewritten in \CEU using
+\code{pause} statements so that other trails can be interleaved with them.
+This option is the one recommended in many related work that provide a similar 
+cooperative primitive (e.g. \code{pause}~\cite{esterel.primer}, 
+\code{PT\_YIELD}~\cite{wsn.protothreads}, \code{yield}~\cite{wsn.sol}, 
+\code{post}~\cite{wsn.nesc}).
+%
+Considering the second option, \CEU and preemptive threads are not mutually 
+exclusive.
+% and the second option is appealing to be explored in a future work.
+For instance, TOSThreads~\cite{wsn.tosthreads} proposes a message-based 
+integration with \emph{nesC} that is safe and matches the semantics of \CEU 
+external events.
+\end{comment}
+
 \section{Conclusion}
 \label{sec.conclusion}
+
+% CONTRIBS
+    - queue in the semantics
+    - permeates all aspects
+    - As far as we know, \CEU is the first Esterel-based language to HAVE?
+      queue-based/unique in the semantics of the language and not in the
+      library
 

### 2016-03-31 paper.tex
...

@@
 \begin{abstract}
-\CEU is a reactive language inspired by Esterel that targets constrained 
-embedded platforms and ensures safe concurrency by handling threats at compile 
-time.
-%
-Based on the synchronous programming model, our design allows for a simple 
-reasoning about concurrency that enables compile-time analysis and results in 
-deterministic and memory-safe programs.
-%
-% TODO: some uses?
-%
-We discuss the design of \CEU and propose a formal semantics for its particular 
-control mechanisms, such as parallel compositions, finalization, and internal 
-events.
+\CEU is a synchronous language inspired by Esterel which aims to offer a 
+simpler semantics and more fine-grained control over the program execution.
+%
+\CEU employs an event-triggered notion of time that allows for a simple
+reasoning about reactions in parallel and enables compile-time analysis to 
+produce deterministic and concurrency-safe programs.
+%
+We discuss the particularities of our design in comparison to Esterel, such as 
+stack-based internal events, temporal analysis for concurrent statements, safe 
+integration with \C, and first-class timers.
 %
@@
 The first phrase of the specification, awaiting and emitting the events, is 
-translated almost identically in the two languages (ln. 3--8, in both 
+translated almost identically in the two languages (ln. 5--10, in both 
 implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
@@
 For the second phrase, the reset behavior, the Esterel version uses a 
-\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
-\CEU's \code{par/or} (ln. 2--11):
+\code{abort-when} (ln. 4--11), which, in this case, serves the same purpose of 
+\CEU's \code{par/or} (ln. 4--13):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
 restarts the enclosing \code{loop}.
-
-% TODO: input/output declarations
 
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
+input  A, B;
+output O;
 loop
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+input  void A, B;
+output void O;
 loop do
@@
 %
-To cope with intra-reaction communication, \CEU supports internal events which 
-programs can \code{emit} and \code{await}.
+To cope with deterministic intra-reaction communication, \CEU supports internal 
+events which programs can \code{emit} and \code{await}.
 In contrast with external events, internal events follow a stack-based 
@@
 \item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
-\item No other trails awake to \code{b}.
+      \emph{(No other trails awake to \code{b}.)}
 \item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
@@
       library
+    - stack
+        - deterministic intra-reaction execution
+    - imposed intra-reaction execution
+        - for inter-reaction commm
+        - for intra-reaction commm
+        - for stmts
 

### 2016-03-31 paper.tex
...

@@
 
-To mitigate this threat, \CEU performs a temporal analysis at compile time and 
-detects concurrent accesses to shared variables, as follows:
+To mitigate this threat, \CEU performs a temporal analysis at compile time to 
+detect concurrent accesses to shared variables:
 if a variable is written in a trail segment, then a concurrent trail segment 
@@
 
-TODO: internal events, emits are propagated from preceding awaits to input
-
 The compile-time \emph{temporal analysis} phase detects inconsistencies in \CEU 
 programs.
-Here, we focus on the algorithm that detects suspicious access to shared 
-variables, as discussed in Section~\ref{sec.ceu.analysis}.
+Here, we focus on the algorithm that detects suspicious concurrent statements, 
+such as accesses to shared variables, as discussed in 
+Section~\ref{sec.ceu.analysis}.
 
 For each node representing a statement in the program AST, we keep the set of 
-events $I$ (for \emph{incoming}) that can lead to the execution of the node, 
-and also the set of events $O$ (for \emph{outgoing}) that can terminate the 
-node.
+input events $I$ (for \emph{incoming}) that can lead to the execution of the 
+node, and also the set of input events $O$ (for \emph{outgoing}) that can 
+terminate the node.
 
@@
 %
-\item An \code{await e} statement has $O=\{e\}$.
+\item An \code{await E} statement has $O=\{E\}$, where \code{E} is an input 
+event (see internal events below).
 %
@@
 $I_n=O_{n-1}$.
-The first child inherits $I$ from its parent node, and the set $O$ for the 
-sequence node is copied from its last child, i.e., $O=O_n$.
+The first child inherits $I$ from its parent node, while the set $O$ for the 
+whole node is copied from its last child, i.e., $O=O_n$.
 %
@@
 O_1 \cup ... \cup O_n$.
+\item For internal events, an \code{await} awakes from any input that leads to 
+    any matching \code{emit} in a trail in parallel:
+    \begin{itemize}
+    \item An \code{emit e} terminates in the same reaction, having $O=I$.
+    \item An \code{await e} has $O=I_{e1} \cup ... \cup I_{eN}$,
+        where $e1...eN$ are \code{emit e} statements in trails in parallel.
+    \end{itemize}
 \end{itemize}

### 2016-03-31 paper.tex
...

@@
 
-% TODO: SCCharts?, SynchCharts?
+% TODO-no: SCCharts?, SynchCharts?
 
@@
 
-% TODO: par/or vs abort, strong/weak abortion
+% TODO-no: par/or vs abort, strong/weak abortion
 
@@
 and Esterel:
+%
 Unique and queue-based external events
@@
     %(\ref{sec.ceu.discussion}).
+%
+We provide the formal specification of the semantics of \CEU in a separate 
+work~\cite{ceu.phd}.
 
@@
 length~\cite{esterel.fixed}.
-% TODO: Such analsysis is orthogonal
+% TODO-no: Such analsysis is orthogonal
 
@@
 4. WORST CASE REACTION TIME ANALYSIS
-TODO: requires this, what about this in the context of MThreads??
+TODO-no: requires this, what about this in the context of MThreads??
 Towards Direct Execution of Esterel Programs on Reactive Processors
@@
 delimiting instructions.
-(TODO: tem figuras)
+(TODO-no: tem figuras)
 \end{comment}
@@
 
-% TODO: .. will probably expect that duplicating the  emit inc => &v; line 
+% TODO-no: .. will probably expect that duplicating the  emit inc => &v; line 
 % would result in v==3.
@@
 %
-%TODO: @nohold
+%TODO-no: @nohold
 %Note also that \CEU environments rely on \C libraries that only provide 
@@
 %
-Note also that the application in \CEU could define an external input 
-\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
-10) to recover the abstract notion of simultaneity.
+Conversely, an environment for \CEU could provide \code{TICK} input and 
+broadcast it periodically to allow applications to recover the abstract notion 
+of simultaneity (e.g., changing ln. 10 to \code{await TICK}).
+%the expiring event from \code{200ms} to \code{TICK} in ln.  10).
 
@@
 binary.
+
+In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
+in comparison to hand-crafted event-driven code in \emph{nesC}~\cite{wsn.nesc} 
+(a \C variant) available in a stable codebase%
+\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}.
+%
+Figure~\ref{fig.eval} compares source size (\emph{tokens}), binary size 
+(\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
+network protocols and a radio driver.
+The small increase in resource usage shows that \CEU fits the context of 
+constrained embedded systems which is typically reserved to \C only.
+
+\begin{figure*}[t]
+\includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
+\caption{ Resource usage for \CEU and \emph{nesC} in the domain of sensor 
+networks.
+\label{fig.eval}
+}
+\end{figure*}
 
@@
 
+- TODO: no available Esterel compiler for the constrained target platform
+    - we compare with C, which is a benchmark, technically unbeatable for 
+      resource efficiency
+
 % SIMPLE

### 2016-04-01 paper.tex
...

@@
 reacts exclusively to it.
-The execution of a program in \CEU is as follows~\cite{ceu.sensys13}:
+The event-triggered execution of a program in \CEU is as 
+follows~\cite{ceu.sensys13}:
 %
@@
 % TODO-no: Such analsysis is orthogonal
+
+A limitation of event-triggered execution is that all program behavior must be 
+purely reactive, given that no code can execute in the absence of inputs.
+%
+Tick-triggered execution allows for active behavior, since code can execute 
+regulary on every tick.
+%
+Although it supports asynchronous active execution~\cite{ceu.tr}, we still 
+consider \CEU as purely reactive since its synchronous core cannot express 
+active behavior.
 
@@
 
+\subsection{Summary}
+
+Contributions:
+%
+\begin{itemize}
+\item \textbf{Event-triggered notion of time bound to the semantics of the 
+language}, which allows ....
+\item \textbf{Deterministic and fine-grained intra-reaction execution and 
+communication}, lexical, stack-based internal events
+\item \textbf{Static temporal analysis for concurrent statements}, which
+      prevent logical race conditions on shared variables and system calls.
+\item \textbf{Safe integration with \C}, with enforced resource clean up.
+\item \textbf{First-class timers} with dedicated syntax and automatic 
+synchronization.
+\end{itemize}
+
+Limitations:
+%
+\begin{itemize}
+\item purely reactive execution,
+\item bounded-only reactions,
+\item sequential
+\end{itemize}
+%
+We believe that a simple-tractable synchronous core
+and extend it with memory-isolated parallelizabe asynchronous execution.
+
+    - permeates all aspects
+    - As far as we know, \CEU is the first Esterel-based language to HAVE?
+      queue-based/unique in the semantics of the language and not in the
+      library
+    - stack
+        - deterministic intra-reaction execution
+    - imposed intra-reaction execution
+        - for inter-reaction commm
+        - for intra-reaction commm
+        - for stmts
+
 \section{Implementation}
@@
 
+TODO: comparison with Esterel
+
 %\emph{(As subsecoes serao retiradas do texto ao final...)}
@@
 external events.
+
+TODO: search "limitation" in the paper and condense all
+
+A limitation of event-triggered execution is that all program behavior must be 
+purely reactive, given that no code can execute in the absence of inputs.
+%
+Tick-triggered execution allows for active behavior, since code can execute 
+regulary on every tick.
+%
+Although it supports asynchronous active execution~\cite{ceu.tr}, we still 
+consider \CEU as purely reactive since its synchronous core cannot express 
+active behavior.
+
 \end{comment}
@@
 
-% CONTRIBS
-    - queue in the semantics
-    - permeates all aspects
-    - As far as we know, \CEU is the first Esterel-based language to HAVE?
-      queue-based/unique in the semantics of the language and not in the
-      library
-    - stack
-        - deterministic intra-reaction execution
-    - imposed intra-reaction execution
-        - for inter-reaction commm
-        - for intra-reaction commm
-        - for stmts
-
-We presented the design and implementation of \CEU, a synchronous reactive 
-language inspired by Esterel targeting constrained embedded systems.
+We presented the design, semantics, and implementation of \CEU, a synchronous 
+reactive language inspired by Esterel targeting constrained embedded systems.
 
@@
 
-\begin{comment}
-We are also extending and applying \CEU in more dynamic domains, such as GUIs 
-and games~\cite{ceu.mod15}, by relaxing some restrictions in the language.
-In particular, programs in these domains typically require unbounded memory and 
-execution time, such as for dealing with dynamic collections.
-We provide unsafe annotations to circumvent the rigid semantics of \CEU, 
-requiring them to be explicit and trackable at the same time.
-\end{comment}
-
 % Bibliography

### 2016-04-01 paper.tex
...

@@
 
+From our experience, however, we find that it is not uncommon to require 
+stateful and non-commutative calls to execute concurrently.
+%
+This is the case for logging (e.g., calls to \code{\_printf} in parallel) or 
+redrawing objects in the screen.
+%
+Figure~\ref{fig.redraw}.a shows an abstract code to animate and redraw objects 
+\code{back} and \code{fore} in the screen in trails in parallel.
+In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
+conflicting pixels.
+This way, \code{\_redraw} is non commutative and the code does not compile 
+correctly.
+%
+However, since the \code{back} object should be redrawn before the \code{fore} 
+object, we want to rely on \CEU's intra-reaction deterministic execution based 
+on lexical order.
+Therefore, in Figure~\ref{fig.redraw}, we redefine the redrawing function to an 
+``intimidating'' name \code{\_redraw\_non\_commutative} to be explicit about 
+its effect (ln. 2), and annotate it as safe (ln. 4--5) to compile successfully.
+
+\begin{figure}
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}
+par do
+    <...> // animate and redraw "back"
+        _redraw(back);
+with
+    <...> // animate and redraw "fore"
+        _redraw(fore);
+end
+
+.
+\end{lstlisting}
+\centering\small{\ax \code{\_redraw} \textbf{cannot} be concurrent}
+\end{minipage}
+%
+\begin{minipage}[t]{0.55\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+native do
+    #define redraw_non_commutative redraw
+end
+@safe _redraw_non_commutative with
+      _redraw_non_commutative;
+par do
+    <...> // animate and redraw "back"
+        _redraw_non_commutative(back);
+with
+    <...> // animate and redraw "fore"
+        _redraw_non_commutative(fore);
+end
+\end{lstlisting}
+\centering\small{\bx \code{\_redraw\_non\_commutative} \textbf{can} be concurrent}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
+\label{lst.c}
+}
+\end{figure}
+
 Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish 
@@
 
+\CEU aims to offer a simpler semantics than Esterel without giving up its 
+tractable synchronous concurrency model:
+
+\CEU is a synchronous language inspired by Esterel which aims to offer a 
+simpler semantics and more fine-grained control over the program execution.
+%
+\CEU employs an event-triggered notion of time that allows for a simple
+reasoning about reactions in parallel and enables compile-time analysis to 
+produce deterministic and concurrency-safe programs.
+%
+We discuss the particularities of our design in comparison to Esterel, such as 
+stack-based internal events, temporal analysis for concurrent statements, safe 
+integration with \C, and first-class timers.
+%
+We also present two implementation back ends:
+one aiming for resource efficiency and interoperability with \C,
+and another based on a virtual machine that allows remote reprogramming.
+
+Overall, \CEU has a simple semantics with fine-grained control for 
+intra-reaction execution which is amenable to a simple temporal analysis that
+improves safety.
+
 Contributions:
@@
 \end{figure*}
+
+- TODO: no available Esterel compiler for the constrained target platform
+    - we compare with C, which is a benchmark, technically unbeatable for 
+      resource efficiency
 
@@
 
-- TODO: no available Esterel compiler for the constrained target platform
-    - we compare with C, which is a benchmark, technically unbeatable for 
-      resource efficiency
-
 % SIMPLE
-
-\CEU aims to offer a simpler semantics than Esterel without giving up its 
-tractable synchronous concurrency model:
-
-% unique external events
-
-The notion of time based on events instead of ticks 
-(Section~\ref{sec.ceu.exts}) is widespread~\cite{sync_async.whynotthreads} and 
-popular in many software communities, such as web frameworks (e.g., 
-\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
-(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
-and Games~\cite{gamepatterns}.
-%
-Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
-through a queue and are dispatched sequentially and atomically to prevent race 
-conditions.
-%
-We believe that for software design, this approach is more familiar to 
-programmers and simplifies the reasoning about concurrency.
-For instance, the uniqueness of external events in \CEU is a prerequisite for 
-the temporal analysis for concurrent statements.
-
-% internal events
-
-Internal events
-
-% lexical order
 
@@
         - register order
-    - paralax, logging
-        - simple, understandeable, imperative
-        - multiple calls to compose, printf, are not commutative
-        - define with intimidating name
     - implementation aspects in Section XX

### 2016-04-01 paper.tex
...

@@
 %
-\item First-class synchronized timers with a dedicated syntax.
+\item First-class timers with dedicated syntax and automatic synchronization.
 %(Section~\ref{sec.ceu.timers}).
@@
 
-\CEU aims to offer a simpler semantics than Esterel without giving up its 
-tractable synchronous concurrency model:
-
-\CEU is a synchronous language inspired by Esterel which aims to offer a 
-simpler semantics and more fine-grained control over the program execution.
-%
-\CEU employs an event-triggered notion of time that allows for a simple
-reasoning about reactions in parallel and enables compile-time analysis to 
-produce deterministic and concurrency-safe programs.
-%
-We discuss the particularities of our design in comparison to Esterel, such as 
-stack-based internal events, temporal analysis for concurrent statements, safe 
-integration with \C, and first-class timers.
-%
-We also present two implementation back ends:
-one aiming for resource efficiency and interoperability with \C,
-and another based on a virtual machine that allows remote reprogramming.
-
-Overall, \CEU has a simple semantics with fine-grained control for 
-intra-reaction execution which is amenable to a simple temporal analysis that
-improves safety.
-
-Contributions:
+\CEU aims to offer a deterministic and simpler semantics than Esterel with a 
+more fine-grained control over the program execution.
+%
+The list that follows enumerate the contributions of our design in this 
+direction:
 %
 \begin{itemize}
-\item \textbf{Event-triggered notion of time bound to the semantics of the 
-language}, which allows ....
-\item \textbf{Deterministic and fine-grained intra-reaction execution and 
-communication}, lexical, stack-based internal events
-\item \textbf{Static temporal analysis for concurrent statements}, which
-      prevent logical race conditions on shared variables and system calls.
-\item \textbf{Safe integration with \C}, with enforced resource clean up.
-\item \textbf{First-class timers} with dedicated syntax and automatic 
-synchronization.
+%
+\item Event-triggered notion of time bound to the semantics of the language.
+Event-driven programming is popular in many domains, such as server and GUI 
+development.
+We believe that programmers are more familiar with this notion of time, which 
+simplifies the reasoning about concurrency.
+For instance, the uniqueness of external events is a prerequisite for the 
+temporal analysis of \CEU.
+%
+\item Deterministic intra-reaction execution and communication.
+Determinism in \CEU is all-inclusive and does not depend on additional levels 
+of static analysis.
+It encompasses the whole language, including memory accesses, system calls, and 
+the expressive stack-based internal events.
+For instance, programmers can always figure out which statement executes next, 
+making runtime analysis and debugging easier.
+%
+\item Static temporal analysis for concurrent statements.
+Although execution is deterministic, the compiler of \CEU still advises about 
+suspicious statements reacting concurrently to the same event.
+This way, the complier forces the programmer to serialize the statements 
+accordingly.
+%
+\item Safe integration with \C.
+\CEU forces the programmer to deal with pointers representing resources through 
+finalization clauses.
+When dealing with concurrent system calls, programmers can provide the compiler 
+with annotations to remove false positives from the temporal analysis or to 
+allow non commutative concurrent behavior.
+%
+\item First-class timers with dedicated syntax and automatic synchronization.
+Given the omnipresence of timers in embedded systems, a dedicated syntax can 
+simplify the development and readability of programs.
+Furthermore, automatic synchronization releases the programmer from the burden 
+of accumulating timers in sequence and comparing timers in parallel.
+%
 \end{itemize}
 
-Limitations:
+The design decisions XXX, raise some important limitations as follows:
 %
 \begin{itemize}
-\item purely reactive execution,
+\item Purely reactive execution.
+    hardware
+A limitation of event-triggered execution is that all program behavior must be purely
+reactive, given that no code can execute in the absence of inputs. Tick-triggered execu-
+tion allows for active behavior, since code can execute regulary on every tick. Although
+it supports asynchronous active execution [Sant’Anna et al. 2012], we still consider
+ ́ as purely reactive since its synchronous core cannot express active behavior.
 \item bounded-only reactions,
@@
       library
-    - stack
-        - deterministic intra-reaction execution
-    - imposed intra-reaction execution
-        - for inter-reaction commm
-        - for intra-reaction commm
-        - for stmts
 
@@
 %
-We believe that for software design, this approach is more familiar to 
-programmers and simplifies the reasoning about concurrency.
-For instance, the uniqueness of external events in \CEU is a prerequisite for 
-the temporal analysis that enables safe shared-memory concurrency.
+%We believe that for software design, this approach is more familiar to 
+%programmers and simplifies the reasoning about concurrency.
+%For instance, the uniqueness of external events in \CEU is a prerequisite for 
+%the temporal analysis that enables safe shared-memory concurrency.
 

### 2016-04-02 paper.tex
...

@@
 \begin{abstract}
-\CEU is a synchronous language inspired by Esterel which aims to offer a 
-simpler semantics and more fine-grained control over the program execution.
-%
-\CEU employs an event-triggered notion of time that allows for a simple
-reasoning about reactions in parallel and enables compile-time analysis to 
-produce deterministic and concurrency-safe programs.
+\CEU is a synchronous language inspired by Esterel with support for a simpler 
+semantics and more fine-grained execution control.
+%
+\CEU employs an event-triggered notion of time which enables compile-time 
+analysis for simultaneous reactions to produce deterministic and 
+concurrency-safe programs.
 %
@@
 %
+% TODO: simple simple
 Overall, \CEU has a simple semantics with fine-grained control for 
@@
 microcontroller with 32K of FLASH and 2K of SRAM.
-}, showing the practical aspect of our clean semantics.
+}, showing the practical aspect of our simple semantics.
 %:\url{http://www.atmel.com/devices/atmega328.aspx}}
@@
 An \code{await} is the only statement that halts a running reaction and allows 
-a program to advance in this notion of time.
+a program to advance in this discrete notion of time.
 %
@@
 translated almost identically in the two languages (ln. 5--10, in both 
-implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
-are equivalent.
+implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
+equivalent.
 %
@@
 %
-Based on the synchronous hypothesis, a program takes no time on step 2 and is 
-always idle on step 3.
+Based on the synchronous hypothesis, a program takes a negligible time on step 
+2 and is always idle on step 3.
 In practice, if a new external input event occurs while a reaction chain is 
-running (step 2), it is enqueued to occur in the next reaction.
+running, it is enqueued to occur in the next reaction.
 
@@
 
-Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
-\CEU.
+Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
+of Esterel and in \CEU.
 The box \code{Real World} shows event occurrences over a continuous timeline 
 divided in units of 10 milliseconds.
-Esterel and \CEU use discrete logical units of time in which the occurring 
-events (which are the same in all boxes) have to fit somehow:
+The other boxes define distinct discrete logical units of time in which the 
+same occurring events fit differently.
 %
@@
 ``Esterel with fixed-length ticks'',
-within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
+within which reactions must fit~\cite{esterel.fixed}.
 %
@@
 Finally, one instance of event \code{D} and two instances of event \code{E} 
-occur ``simultaneously'' during the idle \code{tick-3}.
-However, the program can only detect one occurrence of \code{E}, which is 
-considered in \code{R(D+E)}.
+occur during the idle \code{tick-3}.
+However, only one occurrence of \code{E} is considered in \code{R(D+E)}.
 %
@@
 For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
-events \code{B} and \code{C} are simultaneous, even though \code{A} and 
-\code{B} actually happen much closer to one another.
+events \code{B} and \code{C} are considered simultaneous, even though \code{A} 
+and \code{B} ``physically'' happen much closer to one another.
 In Section~\ref{sec.ceu.timers}, after introducing internal events and 
-first-class timers, we show an example that detects simultaneous button clicks.
+first-class timers, we show an example to detect simultaneous button clicks in 
+\CEU.
 %
@@
 %
-Although it supports asynchronous active execution~\cite{ceu.tr}, we still 
-consider \CEU as purely reactive since its synchronous core cannot express 
-active behavior.
+Although \CEU supports asynchronous active execution~\cite{ceu.tr}, we still 
+consider it as purely reactive since its synchronous core cannot express active 
+behavior.
 
@@
 Esterel makes no semantic distinctions between internal and external signals.
-In particular, programs can emit external and internal signals simultaneously, 
-with all coexisting during the entire reaction.
+%In particular, programs can emit external and internal signals simultaneously, 
+%with all coexisting during the entire reaction.
 %
@@
 %
-In \CEU, subsequent external events coming from the environment define a 
-timeline, which programs cannot interfere by emitting new external events.
-%
-To cope with deterministic intra-reaction communication, \CEU supports internal 
-events which programs can \code{emit} and \code{await}.
-In contrast with external events, internal events follow a stack-based 
-execution policy similar to subroutine calls in typical programming languages.
+%In \CEU, subsequent external events coming from the environment define a 
+%timeline, which programs cannot interfere by emitting new external events.
+%
+%To cope with deterministic intra-reaction communication, \CEU supports 
+%internal events which programs can \code{emit} and \code{await}.
+In \CEU, in contrast with queue-based external events, internal events follow a 
+stack-based execution policy similar to subroutine calls in typical programming 
+languages.
 %
@@
 events become active, resulting in the invocation of \code{f()} and \code{g()} 
-in no particular order.
+(ln. 6,9) in no particular order.
 %
@@
 \setlength{\itemsep}{0pt}
-\item 1st trail awakes (ln. 4), broadcasts \code{b}, and pauses.
-\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
+\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
+\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
       \emph{(No other trails awake to \code{b}.)}
 \item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
-    terminates.
+    terminates (ln. 5--6).
 \item Both trails have terminated, so the \code{par/and} rejoins, and the 
@@
 
-Internal events provide a fine-grained intra-reaction control mechanism.
+Internal events define fine-grained micro reactions which stack one on top of 
+another inside the reaction to an external event.
+%Internal events provide a fine-grained intra-reaction control mechanism.
 For instance, it brings a limited form of subroutines, as depicted in 
@@
 Both sides of the \code{par/or} have an \code{await} statement to avoid 
-instantaneous termination (ln. 4,7).
+instantaneous termination (ln. 4,7) of the enclosing \code{loop} (ln 2--9).
 However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
@@
 the \code{loop} instantaneously, resulting in infinite execution.
-
+%
+\begin{comment}
 \begin{figure}
@@
 \end{figure}
-
+\end{comment}
+%
 In atypical scenarios requiring immediate awake, delayed awaits can be 
@@
 %
+\begin{comment}
 For instance, sometimes we need to execute a block of code immediately, and 
@@
 the periodic code to prevent its immediate execution.
-%
-On the one hand, we transfer the burden of dealing with these specific corner 
-cases to the programmer.
+\end{comment}
+%
+On the one hand, we transfer the burden of dealing with these corner cases to 
+the programmer.
 On the other hand, we simplify the semantics of the language and eliminate the 
@@
 %
-Figure~\ref{fig.redraw}.a shows an abstract code to animate and redraw objects 
+Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw objects 
 \code{back} and \code{fore} in the screen in trails in parallel.
@@
 on lexical order.
-Therefore, in Figure~\ref{fig.redraw}, we redefine the redrawing function to an 
+Therefore, in Figure~\ref{lst.redraw}, we redefine the redrawing function to an 
 ``intimidating'' name \code{\_redraw\_non\_commutative} to be explicit about 
@@
 \caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
-\label{lst.c}
+\label{lst.redraw}
 }
@@
 
-The design decisions XXX, raise some important limitations as follows:
+Our synchronous and deterministic approach also leads to some limitations as 
+follows:
 %
 \begin{itemize}
-\item Purely reactive execution.
-    hardware
-A limitation of event-triggered execution is that all program behavior must be purely
-reactive, given that no code can execute in the absence of inputs. Tick-triggered execu-
-tion allows for active behavior, since code can execute regulary on every tick. Although
-it supports asynchronous active execution [Sant’Anna et al. 2012], we still consider
- ́ as purely reactive since its synchronous core cannot express active behavior.
-\item bounded-only reactions,
-\item sequential
+\item Execution is purely reactive as a result of event-triggered reactions.
+Since only occurrences of events can start reactions, programs cannot execute 
+proactively in the absence of events.
+\item Reactions must execute in bounded time due to the synchronous hypothesis.
+As a synchronous language, \CEU requires that CPU times for reactions are 
+negligible in comparison to the rate of incoming events.
+\item Execution is sequential because of intra-reaction determinism.
+The deterministic semantics of \CEU does not make parallelization easy (to be 
+discussed in Section~\ref{sec.impl.single}).
 \end{itemize}
 %
-We believe that a simple-tractable synchronous core
-and extend it with memory-isolated parallelizabe asynchronous execution.
-
-    - permeates all aspects
-    - As far as we know, \CEU is the first Esterel-based language to HAVE?
-      queue-based/unique in the semantics of the language and not in the
-      library
+Nonetheless, we advocate for a tractable synchronous reactive core that 
+supports shared memory with deterministic execution.
+%
+To deal with the limitations, we recommend for memory-isolated parallelizable 
+asynchronous primitives as language extensions, but which are not in the scope 
+of this paper.
 
@@
 static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
+single-threaded dispatching (Section~\ref{sec.impl.single}),
 interaction with the environment (Section~\ref{sec.impl.environment}),
@@
 \subsection{Single-Threaded Dispatching}
+\label{sec.impl.single}
 
@@
 
-TODO: comparison with Esterel
+TODO: most about comparison with Esterel
+    - TODO: related work?
+    - As far as we know, \CEU is the first Esterel-based language to HAVE?
+      queue-based/unique in the semantics of the language and not in the
+      library
 

### 2016-04-02 paper.tex
...

@@
 
-Internal events define fine-grained micro reactions which stack one on top of 
-another inside the reaction to an external event.
+% TODO
+Nested \code{emit} invocations (e.g., suppose we \code{emit c} after the 
+\code{await a} in line 8) results in fine-grained micro reactions stacked one 
+on top of the other, all inside the same reaction to an external event.
 %Internal events provide a fine-grained intra-reaction control mechanism.
@@
 
-Esterel is only deterministic with respect to reactive control: ``the same 
+Esterel is only deterministic with respect to external behavior: ``the same 
 sequence of inputs always produces the same sequence of 
@@
 %
-\CEU also takes the deterministic approach and when multiple trails are active 
-during the same reaction, they are scheduled in the order they appear in the 
-program source code.
+\CEU also takes the deterministic approach and, when multiple trails are active 
+during the same reaction, they are scheduled in lexical order, i.e., in the 
+order they appear in the program source code.
 
@@
 %
-In the example \ax, the two assignments to \code{x} execute from reactions to 
-different events \code{A} and \code{B} which, by definition, cannot occur 
+In the example \ax, the two assignments to \code{x} only execute in reactions 
+to different events \code{A} and \code{B} which, by definition, cannot occur 
 simultaneously (Section~\ref{sec.ceu.exts}).
-Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
-for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
+Hence, for the sequence \code{A->B}, \code{x} becomes \code{4=(1+1)*2}, while 
+for \code{B->A}, \code{x} becomes \code{3=(1*2)+1}.
 %
@@
 Since \CEU employs lexical order for intra-reaction statements, the execution 
-is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
-%
-However, an (apparently innocuous) change in the order of trails changes the 
+is still deterministic, and \code{y} always becomes \code{4=(1+1)*2}.
+%
+However, an (apparently innocuous) change in the order of trails modifies the 
 semantics of the program, which we consider unsafe.
@@
 \item The assignments to \code{x} in lines 2 and 5 in \ax
-      \textbf{cannot} be concurrent because they are not in parallel trails.
+      \textbf{cannot} be concurrent because they are \textbf{not} in parallel trails.
 \item The assignments to \code{x} in lines 5 and 8 in \ax
@@
 
-Note that this analysis is only possible due to the uniqueness of input events 
+Note that such analysis is only possible due to the uniqueness of input events 
 within reactions.
-Otherwise, any two trail segments in parallel can be concurrent, even if they 
+Otherwise, any two trail segments in parallel could be concurrent, even if they 
 react to different input events.
@@
 %
-Similarly to Esterel, which supports the \code{call} primitive for external 
-code, calls are assumed to be instantaneous~\cite{esterel.primer}.
-%
-This way, programs should only resort to \C for asynchronous functionality%
+Similarly to Esterel with the \code{call} primitive, external calls are assumed 
+to be instantaneous~\cite{esterel.primer}.
+%
+This way, programs should only resort to \C for asynchronous functionality, 
+such as non-blocking I/O, or simple \code{struct} accessors, but never for 
+control purposes.%
 \footnote{
@@
 option.
-}, such as non-blocking I/O, or simple \code{struct} accessors, but never for 
-control purposes.
+}
 
@@
 native @const _NUM;
-native @pure  _id;
-native @safe  _f with _g;
+native @pure  _id();
+native @safe  _f() with _g();
 
@@
 
-The temporal analysis of Section~\ref{sec.ceu.analysis} also considers calls 
-and accesses to external symbols in \C.
+\subsubsection{Temporal Analysis}
+
+As a safety measure, the temporal analysis of Section~\ref{sec.ceu.analysis} 
+also considers calls and accesses to external symbols in \C.
 %
@@
 %
-The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
-during the boot reaction (ln. 8,10):
-the first trail calls symbol \code{\_f},
-while the second calls \code{\_g} and \code{\_id}, and also reads \code{\_NUM}.
+During the boot reaction, two trails react concurrently inside the 
+\code{par/and} (ln. 7--11):
+the first trail calls symbol \code{\_f} (ln. 8),
+while the second calls \code{\_g} and \code{\_id}, and also reads \code{\_NUM} 
+(ln. 10).
 %
@@
 
-The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
-compiler about the semantics of the \C symbols in use in \ax.
-With the annotations in \bx, the program in \ax compiles successfully:
+\subsubsection{Annotations}
+
+The annotations in Figure~\ref{lst.c}.b provides hints to the compiler about 
+the semantics of the \C symbols in code \ax, which now compiles successfully:
 %
@@
       concurrently with any other symbol in the program.
-\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
-      conflict and they can be safely called concurrently.
+\item Both \code{\_f} and \code{\_g} are impure, but have non-conflicting 
+      commutative effects, and can be safely called concurrently.
 \end{itemize}
 
-From our experience, however, we find that it is not uncommon to require 
-stateful and non-commutative calls to execute concurrently.
-%
-This is the case for logging (e.g., calls to \code{\_printf} in parallel) or 
-redrawing objects in the screen.
-%
-Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw objects 
-\code{back} and \code{fore} in the screen in trails in parallel.
+From our experience, we find that it is not uncommon to also require 
+non-commutative calls to execute concurrently.
+%
+This is the case for logging (e.g., calls to \code{\_printf} in trails in 
+parallel) and for redrawing objects in the screen.
+%
+Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
+objects \code{background} and \code{foreground} in trails in parallel.
 In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
-conflicting pixels.
-This way, \code{\_redraw} is non commutative and the code does not compile 
-correctly.
-%
-However, since the \code{back} object should be redrawn before the \code{fore} 
-object, we want to rely on \CEU's intra-reaction deterministic execution based 
-on lexical order.
-Therefore, in Figure~\ref{lst.redraw}, we redefine the redrawing function to an 
-``intimidating'' name \code{\_redraw\_non\_commutative} to be explicit about 
-its effect (ln. 2), and annotate it as safe (ln. 4--5) to compile successfully.
+conflicting pixels, which makes it non commutative and prevents the code to 
+compile.
+%
+However, we want to rely on trails lexical order to always redraw the 
+\code{background} object before the \code{foreground} object.
+%
+Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} to an 
+``intimidating'' name \code{\_redraw\_non\_commutative} (ln. 2, to be explicit 
+about its effect), and annotate it as safe (ln. 4--5) to compile successfully.
 
@@
 par do
-    <...> // animate and redraw "back"
-        _redraw(back);
+    <...> // animate and redraw "background"
+        _redraw(background);
 with
-    <...> // animate and redraw "fore"
-        _redraw(fore);
+    <...> // animate and redraw "foreground"
+        _redraw(foreground);
 end
@@
 par do
-    <...> // animate and redraw "back"
-        _redraw_non_commutative(back);
+    <...> // animate and redraw "background"
+        _redraw_non_commutative(background);
 with
     <...> // animate and redraw "fore"
-        _redraw_non_commutative(fore);
+        _redraw_non_commutative(foreground);
 end
@@
 
-Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish 
-synchronous from asynchronous languages with respect to orthogonal abortion of
-lines of execution (i.e., without tweaking them with synchronization 
-primitives)~\cite{esterel.preemption}.
-In contrast, traditional (asynchronous) multi-threaded languages cannot express 
-thread termination safely~\cite{sync_async.threadsstop}.
-%
-Still, aborting lines of execution that deal with external resources is 
-challenging because they may end up in an inconsistent state.
+\subsubsection{Finalization}
+
+Esterel's \code{abort} and \CEU's \code{par/or} statements provide orthogonal 
+abortion of lines of execution, which is a distinctive feature of synchronous 
+language in comparison to asynchronous languages~\cite{esterel.preemption}.
+%In contrast, traditional (asynchronous) multi-threaded languages cannot 
+%express thread termination safely~\cite{sync_async.threadsstop}.
+%
+However, aborting lines of execution that deal with external resources is still 
+error prone because they may end up in an inconsistent state.
 %
@@
 
-The example in Figure~\ref{lst.fin} in Esterel \ax and \CEU \bx uses the 
-\code{lock} and \code{unlock} calls which represent accesses to an external 
-resource.
+In the example in Figure~\ref{lst.fin} in Esterel \ax and \CEU \bx, the calls 
+to \code{lock} and \code{unlock} represent accesses to an external resource.
 The normal behavior is to \code{lock} the resource (ln. \ax:4 and \bx:3), 
@@
 \bx:5).
-However, if the aborting input \code{B} (ln. \ax:12 and \bx:12) occurs after 
+Note that if the aborting input \code{B} (ln. \ax:12 and \bx:12) occurs after 
 the \code{lock} but before the reactions to \code{A}, we still want to call 
@@
 %
-In Esterel and \CEU, the finalize clauses (ln. \ax:10 and \bx:5) are 
-automatically called if the enclosing blocks (ln. \ax:3--1 and \bx:3--10) are 
+In Esterel and \CEU, the finalize clause (ln. \ax:10 and \bx:5) execute
+automatically if the enclosing block (ln. \ax:3--1 and \bx:3--10) is
 externally aborted (ln. \ax:12 and \bx:12).
@@
 %
-In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} 
-returns an external file resource as a pointer.
+In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
+4) returns an external file resource as a pointer.
 If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
@@
 %
-In both cases, the code does not compile if the programmer forgets to use the
-\code{finalize} construct.
+In both cases, the code does not compile without the \code{finalize} construct.
 
@@
 
-Note that the illustrative example of Figure~\ref{lst.fin} does not manipulate 
+Note that the illustrative example in Figure~\ref{lst.fin} does not manipulate 
 pointers (i.e., the resource is a \emph{singleton}).
@@
 automatically, resulting in more robust applications.
-For the example in Figure~\ref{lst.timers}.a, suppose that after the first 
+In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
 \code{await} request, the underlying system gets busy and takes 15ms to check 
@@
 Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
-As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
-the trail is rescheduled for execution, now with \code{delta=4ms}.
+As the current delta is still higher than the requested timeout (i.e. $5ms > 
+1ms$), the trail is rescheduled for execution, now with \code{delta=4ms}.
 
@@
 
-\CEU also takes into account the fact that time is a physical quantity that can 
-be added and compared.
+\CEU also considers that time is a physical quantity that can be added and 
+compared.
 For instance, for the example in Figure~\ref{lst.timers}.b, although the 
@@
 infinitesimal and can never be absolutely simultaneous.
-However, the ``sensation of simultaneity'' is not infinitesimal, but increases 
-with the inaccuracy of the observer (e.g., a human being).
-Therefore, simultaneity should be defined with respect to the observer, which 
-varies from case to case.
-
+However, the ``sensation of simultaneity'' is not infinitesimal, but actually 
+increases with the inaccuracy of the observer (e.g., a human being).
+Therefore, we consider that simultaneity should be defined with respect to the 
+observer, which varies from case to case.
+%
 First-class timers simplify the implementation of application-defined 
@@
 However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
-we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
-with the enclosing \code{loop} (ln. 4).
+we abort the whole behavior with the \code{par/or} (ln. 5) and try again with 
+the enclosing \code{loop} (ln. 4).
 %
 In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
-which is a huge amount of time for a language-defined tick, and which would 
-break the synchronous hypothesis.
-%
-Note that a similar implementation for Esterel would not rely on the tick 
+which is a huge amount of time for a language-defined tick.
+%, and which would break the synchronous hypothesis.
+%
+Therefore, a similar implementation for Esterel would not rely on the tick 
 notion of simultaneity either.
 %
-Conversely, an environment for \CEU could provide \code{TICK} input and 
-broadcast it periodically to allow applications to recover the abstract notion 
-of simultaneity (e.g., changing ln. 10 to \code{await TICK}).
+%Conversely, an environment for \CEU could provide \code{TICK} input and 
+%broadcast it periodically to allow applications to recover the abstract notion 
+%of simultaneity (e.g., changing ln. 10 to \code{await TICK}).
 %the expiring event from \code{200ms} to \code{TICK} in ln.  10).
@@
 
-\CEU aims to offer a deterministic and simpler semantics than Esterel with a 
-more fine-grained control over the program execution.
-%
-The list that follows enumerate the contributions of our design in this 
+\CEU aims to offer a simpler semantics than Esterel with more determinism and 
+fine-grained control over the program execution.
+%
+The list that follows summarizes the contributions of our design in this 
 direction:
@@
 development.
-We believe that programmers are more familiar with this notion of time, which 
-simplifies the reasoning about concurrency.
-For instance, the uniqueness of external events is a prerequisite for the 
+We believe that programmers are more familiar with dealing with events in 
+isolation, which simplifies the reasoning about concurrency.
+In addition, the uniqueness of external events is a prerequisite for the 
 temporal analysis of \CEU.
@@
 It encompasses the whole language, including memory accesses, system calls, and 
-the expressive stack-based internal events.
-For instance, programmers can always figure out which statement executes next, 
-making runtime analysis and debugging easier.
+stack-based internal events.
+Programmers can always figure out which statement executes next, making runtime 
+analysis and debugging easier.
 %
@@
 Although execution is deterministic, the compiler of \CEU still advises about 
-suspicious statements reacting concurrently to the same event.
-This way, the complier forces the programmer to serialize the statements 
-accordingly.
+suspicious statements that can react concurrently to the same event, forcing 
+the to serialize them accordingly.
 %
 \item Safe integration with \C.
-\CEU forces the programmer to deal with pointers representing resources through 
-finalization clauses.
+\CEU enforces the programmer to deal with pointers representing resources 
+through finalization clauses.
 When dealing with concurrent system calls, programmers can provide the compiler 
@@
 Furthermore, automatic synchronization releases the programmer from the burden 
-of accumulating timers in sequence and comparing timers in parallel.
+of adjusting timers in sequence and in parallel.
 %
@@
 
-\begin{figure}[b]
+\begin{figure}
 \begin{minipage}[t]{0.31\linewidth}

### 2016-04-02 paper.tex
...

@@
 for expiring awaits.
-The \CEU scheduler will notice that the \code{await 10ms} has not only already 
-expired, but is delayed with \code{delta=5ms}.
-Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
+The \CEU scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
+already expired, but is delayed with \code{delta=5ms}.
+Then, the awaiting trail awakes, sets \code{v=1} (ln. 3), and invokes 
+\code{await 1ms} (ln. 4).
 As the current delta is still higher than the requested timeout (i.e. $5ms > 
@@
 \begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
 var int v;
@@
 %
-\begin{minipage}[t]{0.51\linewidth}
-\begin{lstlisting}
+\begin{minipage}[t]{0.55\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
 par/or do
@@
 For instance, for the example in Figure~\ref{lst.timers}.b, although the 
-scheduler cannot guarantee that the first trail terminates exactly in 11ms, it 
-can at least ensure that the program always terminates with \code{v=1}.
+scheduler cannot guarantee that the first trail terminates exactly in 11ms (ln.  
+2,4), it can at least ensure that the program always terminates with 
+\code{v=1}.
 %
 Given that any non-awaiting sequence is considered to take no time in the 
-synchronous model, the first trail is guaranteed to terminate before the second 
-trail, because $10+1 < 12$.
+synchronous model, the first trail (ln. 2--5) is guaranteed to terminate before 
+the second trail (ln. 7--8), because $10+1 < 12$.
 %

### 2016-04-04 paper.tex
...

@@
 \newcommand{\bx}{\code{[b]}\xspace}
+\newcommand{\cx}{\code{[c]}\xspace}
 \newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}
@@
 
-% TODO
-Nested \code{emit} invocations (e.g., suppose we \code{emit c} after the 
-\code{await a} in line 8) results in fine-grained micro reactions stacked one 
-on top of the other, all inside the same reaction to an external event.
-%Internal events provide a fine-grained intra-reaction control mechanism.
-For instance, it brings a limited form of subroutines, as depicted in 
+Internal events provide a limited form of subroutines, as depicted in 
 Figure~\ref{lst.sub}.
@@
 
+\CEU also supports nested \code{emit} invocations for internal events.
+For instance, the code that executes after the \code{await inc} (ln. 4) could 
+\code{emit} another event, creating a new level in the stack.
+The runtime stack represents fine-grained micro reactions, one on top of the 
+other, all inside the same reaction to an external event.
+%Internal events provide a fine-grained intra-reaction control mechanism.
+
 \begin{figure}
@@
 program counter, similarly to coroutines~\cite{lua.coroutines}.
+%
+In a previous work, we build on top of internal events other advanced control 
+mechanisms, such as resumable exceptions and reactive 
+variables~\cite{ceu.rem13}.
 %In Section~\ref{sec.adv.excpt} we show how to use them to implement 
@@
 %
-A number of Esterel-based synchronous languages, such as 
-\emph{SOL}~\cite{wsn.sol},
-\emph{SC}~\cite{rp.synchc}, and
-\emph{PRET-C}~\cite{rp.pretc},
-enforce intra-reaction determinism with an arbitrary execution order for 
-statements in multiple lines of execution.
+A number of Esterel-based synchronous languages enforce intra-reaction 
+determinism with an arbitrary execution order for statements in multiple lines 
+of execution (%
+\emph{Reactive C}~\cite{rp.rc},                 % 1991
+\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
+\emph{SOL}~\cite{wsn.sol},                      % 2007
+\emph{SC}~\cite{rp.synchc},                     % 2009
+and
+\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
 %, e.g., based on lexical order or manual priority assignment.
@@
 react to different input events.
+%
+Note also that the analysis is completely optional and does not affect the 
+semantics of the program.
+% TODO-no: static analsysis also simpler
+% highlight that is not only important to have a comptutally-sound algorithm, 
+% but that something that programmers can understand
 
@@
 \item Deterministic intra-reaction execution and communication.
-Determinism in \CEU is all-inclusive and does not depend on additional levels 
-of static analysis.
+Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
+levels of static analysis.
 It encompasses the whole language, including memory accesses, system calls, and 
@@
 \begin{itemize}
-\item Execution is purely reactive as a result of event-triggered reactions.
+\item Execution is purely reactive as result of event-triggered reactions.
 Since only occurrences of events can start reactions, programs cannot execute 
@@
 \item Reactions must execute in bounded time due to the synchronous hypothesis.
-As a synchronous language, \CEU requires that CPU times for reactions are 
+As a synchronous language, \CEU requires CPU times for reactions to be 
 negligible in comparison to the rate of incoming events.
 \item Execution is sequential because of intra-reaction determinism.
-The deterministic semantics of \CEU does not make parallelization easy (to be 
-discussed in Section~\ref{sec.impl.single}).
+%We discuss some implementation considerations in 
+%Section~\ref{sec.impl.single}.
+The deterministic semantics of \CEU does not make implicit parallelization easy 
+(to be discussed in Section~\ref{sec.impl.single}).
 \end{itemize}
 %
-Nonetheless, we advocate for a tractable synchronous reactive core that 
-supports shared memory with deterministic execution.
-%
-To deal with the limitations, we recommend for memory-isolated parallelizable 
-asynchronous primitives as language extensions, but which are not in the scope 
-of this paper.
+Nonetheless, we advocate for keeping a tractable synchronous reactive core 
+which supports shared memory concurrency with deterministic execution.
+%
+To deal with the limitations above, we recommend for memory-isolated 
+parallelizable asynchronous functionality as a separate extension to the 
+synchronous core~\cite{esterel.crp} (which is not in the scope of this paper).
+
+\begin{comment}
+TODO-no: parallel
+On the one hand, enforcing an execution order for concurrent operations may 
+seen arbitrary and also precludes true parallelism.
+On the other hand, it provides a priority scheme for trails, and makes 
+shared-memory concurrency more tractable.
+For constrained embedded development, we believe that deterministic 
+shared-memory concurrency is beneficial, given the extensive use of memory 
+mapped ports for \emph{I/O} and the lack of hardware support for real 
+parallelism.
+\end{comment}
 
@@
 
-The compilation process of a program in \CEU for the original \C back end is 
-composed of three main phases:
-the \emph{parsing phase} converts the source code in \CEU to an \emph{abstract 
-syntax tree (AST)};
+The compilation process of programs in \CEU is composed of three main phases:
+the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
+tree (\emph{AST});
 the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
@@
 In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
-in comparison to hand-crafted event-driven code in \emph{nesC}~\cite{wsn.nesc} 
-(a \C variant) available in a stable codebase%
-\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}.
+in comparison to a hand-crafted event-driven code base%
+\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
+in \emph{nesC} (a \C variant)~\cite{wsn.nesc}.
 %
@@
 network protocols and a radio driver.
-The small increase in resource usage shows that \CEU fits the context of 
-constrained embedded systems which is typically reserved to \C only.
+The small increase in resource usage shows that the gains in productivity and 
+safety with \CEU make it a viable alternative to \C in the context of 
+constrained embedded systems.
+
+Unfortunately, most real applications in Esterel seems to be closed source.
+In a future work, we plan to cooperate with groups that have more experience 
+and access to real code bases in Esterel for a comprehensive comparison with 
+Céu.
+% TODO-no:
+%Still, considering \C as an ``absolute benchmark'', we consider the small 
+%overhead .  solid results
 
@@
 
-- TODO: no available Esterel compiler for the constrained target platform
-    - we compare with C, which is a benchmark, technically unbeatable for 
-      resource efficiency
-
-In the sections that follow, we discuss the implementation highlights related 
-to the peculiarities of \CEU:
+In the sections that follow, we discuss the implementation of subsystems that 
+are specific (or in some way important) to \CEU:
 the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
@@
 
-The compile-time \emph{temporal analysis} phase detects inconsistencies in \CEU 
+The compile-time temporal analysis phase detects inconsistencies in \CEU 
 programs.
@@
 terminate the node.
-
-A node inherits the set $I$ from its direct parent and calculates $O$ according 
-to its type:
+As an example, for the single-statement program \code{await~A}, we have 
+$I=\{boot\}$ and $O=\{A\}$.
+
+A node inherits the set $I$ from its immediate parent and calculates $O$ 
+according to its type, as follows:
 %
@@
 %
-\item An \code{await E} statement has $O=\{E\}$, where \code{E} is an input 
-event (see internal events below).
+\item An \code{await E} statement, where \code{E} is an external input event,
+      has $O=\{E\}$ (see also internal events below).
 %
 \item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
-\code{loop} and never terminates, i.e., never proceeds to the statement 
-immediately following it (see also \code{loop} below);
+\code{loop} and never proceeds to the statement immediately following it (see 
+also \code{loop} below);
 %
 \item A \emph{sequence node (;)} modifies each of its children to have 
-$I_n=O_{n-1}$.
-The first child inherits $I$ from its parent node, while the set $O$ for the 
-whole node is copied from its last child, i.e., $O=O_n$.
+$I_n=O_{n-1}$, except for $n=1$ which inherits $I$ from the parent node.
+The set $O$ for the whole node is copied from its last child, i.e., $O=O_n$.
 %
@@
 O_{body}$), as the loop is also reached from its own body.
-The union of all \code{break} statements' $O$ forms the set $O$ for a 
-\code{loop}.
+The union of all $O$ from nested \code{break} statements forms the set $O$ for 
+a \code{loop}.
 %
@@
 With all sets calculated, we take all pairs of nodes that perform side effects 
-and are in parallel branches, comparing their sets $I$ for intersections.
-For each pair, if the intersection is not the empty set, we mark both nodes as 
+and are in parallel branches, and compare their sets $I$ for intersections.
+For each pair, if the intersection is not empty, we mark both nodes as 
 suspicious.
 
-The example in Figure~\ref{lst.impl.ast}.a has a corresponding $AST$, in 
-Figure~\ref{lst.impl.ast}.b, with the sets $I$ and $O$ for each node.
+The code in Figure~\ref{lst.impl.ast}.a has the $AST$ and sets $I$ and $O$ for 
+each node in Figure~\ref{lst.impl.ast}.b.
 The event $.$ (dot) represents the ``boot'' reaction.
@@
 Note that although the accesses in ln. 5,11 in the code (ln. 6,11 in the AST) 
-do have an intersection, they are not in parallel and are also safe.
+do have an intersection, they are not in parallel branches and are also safe.
 
@@
 \end{lstlisting}
-\centering\small{\ax}
+\centering\small{\ax} A program in \CEU...
 \end{minipage}
@@
 \end{lstlisting}
-\centering\small{\bx}
+\centering\small{\bx} ...with corresponding sets $I$ and $O$.
 \end{minipage}
@@
 \CEU favors a fine-grained use of trails, being common to use trails that await 
-a single event.
+a single event and terminate.
 For this reason, \CEU does not allocate per-trail stacks; instead, all data 
@@
 \CEU reserves a single static block of memory to hold all memory slots, whose 
-size is the maximum the program uses at a given time.
-A given position in the memory may hold different data (with variable sizes) 
-during runtime.
-%
-As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
+size is the maximum the program uses at any given time.
+A position in the memory may hold different data (with variable sizes) during 
+runtime.
+%
+As an example, Figure~\ref{lst.impl.mem} shows a program with its corresponding 
 memory layout.
 %
-Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
-with the same name can be distinguished.
-%
-The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
-their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
-\code{MEM.b\_2} can safely share the same memory slot.
-%
-The example also illustrates the presence of runtime flags related to the 
-parallel composition, which also reside in reusable slots in the static memory.
+Each variable is assigned a unique $id$ so that variables with the same name 
+can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
+\ax:3).
+%
+The \code{do-end} blocks in sequence (ln. \ax:2--7) are packed in a 
+\code{union} (ln. \bx:2--3), given that their variables cannot be in scope at 
+the same time, e.g., \code{MEM.a\_1} and \code{MEM.b\_2} (ln. \bx:2,3) can 
+safely share the same memory slot.
+%
+The example also illustrates the presence of runtime flags (ln. \bx:4--7) 
+related to the \code{par/and} termination (ln. \ax:8--12), which also reside in 
+reusable slots in the static memory.
 
@@
 \begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
 input int A, B, C;
@@
 \end{lstlisting}
+\centering\small{\ax} A program in \CEU...
 \end{minipage}
@@
 \begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
 union {             // sequence
@@
 } MEM ;
-\end{lstlisting}
+
+.
+\end{lstlisting}
+\centering\small{\bx} ...with corresponding memory layout
 \end{minipage}
@@
 Each line of execution in \CEU needs to carry associated data, such as which 
-event it is awaiting and which code to execute when it awakes.
+event it is currently awaiting and which code to execute when it awakes.
 %
@@
 %
-Like normal variables, trails that cannot be active at the same time can share 
+Like normal variables, trails that cannot be active at the same time share 
 slots in the static memory vector.
@@
 %
-A \code{STACKED} trail holds an associated stack level and is delayed until the 
-scheduler runtime reaches that level again.
+A \code{STACKED} trail holds an associated numeric stack level and is delayed 
+until the scheduler runtime reaches that level again.
 %
@@
 A trail waiting for an event stays in the state of the corresponding event, 
-also holding the minimum sequence number (\emph{seqno}) in which it can awake.
+also holding the minimum sequence number (\emph{seqno}) in which it can awake, 
+to prevent from awaking in the same reaction it is reached.
 %
@@
     union {
-        unsigned char seqno;
-        stack_t       stk;
+        unsigned char seqno;  // if evt=EVT_*
+        stack_t       stk;    // if evt=STACKED
     };
@@
 However, given that the scheduler traverses all trails on every reaction, it 
-can adjust them to properly handle overflows (actually, 2 bits to hold the 
+adjusts them to properly handle overflows (actually, 2 bits to hold the 
 \code{seqno} is already enough).
@@
 In the worst case, a trail emits an event that awakes another trail, which 
-emits an event that awakes another trail, and so on.
-The last trail cannot awake any trail, because they are all hanged in the 
+emits an event that awakes another trail, and so on;
+the last trail cannot awake any trail, because they are all hanged in the 
 \code{STACKED} state.
@@
 For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
-simultaneous lines of execution with an overhead of 2\% in comparison to the 
-original version in \C~\cite{ceu.sensys13}.
+simultaneous lines of execution with memory overhead of only 2\% in comparison 
+to the original version in \C~\cite{ceu.sensys13}.
 
@@
 For the program in \ax, the compiler extracts the entry points and associated 
-trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln. 7).
+trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln.  
+\ax:7).
 %
@@
 (\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
-and \code{break}), the compiler splits the trail into segments with associated 
-entry points.
-%
-The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
-in \bx).
+and \code{break}), the compiler splits the trail in two segments with 
+associated entry points.
+%
+The entry points translate to an \code{enum} in the generated code (ln.  
+\bx:1--10).
 The state of trails translates to a vector of type \code{trail\_t} with the 
-maximum number of simultaneous trails (ln. 12--15).
+maximum number of simultaneous trails (ln. \bx:12--15).
 On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
-(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
-one, in ln. 14).
+(ln. \bx:13), while all others are set to \code{INACTIVE} (ln. \bx:14).
 
@@
 \end{lstlisting}
-\centering\small{(c)}
+\centering\small{\cx}
 \end{minipage}
@@
 generated by the compiler.
-(c) Dispatch function: uses a \code{switch} to associate each segment 
+\cx Dispatch function: uses a \code{switch} to associate each segment 
 identifier with the corresponding code to execute.
@@
 in the \emph{dispatch} pass, it executes each trail that is \code{STACKED} to 
-run in the current level, setting it immediately to \code{INACTIVE} (the trail 
-segment may reset it in sequence if it doesn't terminate).
+run in the current level, setting it immediately to \code{INACTIVE}.
+%(the trail segment may reset it in sequence if it doesn't terminate).
 
@@
 level and resumes the previous execution.
-For the first reaction, the scheduler starts from the \emph{dispatch} pass, 
+For the boot reaction, the scheduler starts from the \emph{dispatch} pass, 
 given that the \code{Main} label is the only one that can be active at the 
-stack level 0 (ln. 13, in Figure~\ref{lst.impl.trails}.b).
-
-The code in Figure~\ref{lst.impl.trails}.c dispatches a trail segment according 
-to the current label to execute.
+stack level 0 (ln. \bx: 13).
+
+The code in \cx dispatches a trail segment according to the current label to 
+execute.
 %
@@
 %
-When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
-(ln. 4--7) and proceeds to the code in \code{TRAIL-0} (ln. 9--14), respecting 
-the deterministic execution order.
+When the \code{Main} label reaches the \code{par/and} (ln. \ax:4), it stacks 
+\code{TRAIL-1} (ln. \cx:4--7) and proceeds to the code in \code{TRAIL-0} (ln.  
+\cx:9--14) to respect lexical execution order.
 %
@@
 
-Nonetheless, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
+Note that, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
 analysis of \CEU infers precisely trails that are concurrent yet do not share 
@@
 %
-However, our experiments with a multi-threaded implementation in multi-core 
-CPUs execute slower than the single-threaded implementation in the same CPUs.
+However, our experiments with multi-core execution are actually slower than 
+single-core execution in the same system.
 Considering that we use \CEU primarily in control-dominated applications, this 
-result is not surprising~\cite{esterel.multicore.1,esterel.multicore.2}.
-We believe that
-    the continuous fork-and-rejoin overhead due to small reactions
-as well as
-    the excessive locality of data due to stackless threads sharing contiguous static memory
-seem to prevent any speed-up gains.
+result is not surprising and also appears in related 
+work~\cite{esterel.multicore.1,esterel.multicore.2}.
+One reason is the continuous fork-and-rejoin overhead due to small reactions.
+Another reason is the excessive locality of data due to stackless threads 
+sharing contiguous static memory.
 
@@
 However, data-intensive computations do not typically require a disciplined 
-step-wise execution and can actually execute in asynchronous calls.
+step-wise execution and can actually execute in isolated asynchronous calls.
 Esterel provides a \code{task} primitive for this 
-purpose~\cite{esterel.primer}, and \CEU provides an equivalent 
-\code{async/thread} primitive (which are out of the scope of this paper).
-
-Since loops in \CEU must contain \code{await} statements, reactions run in 
-bounded time, which guarantees that successive calls to \code{dispatch} never 
-block the scheduler for long.
-However, the code generation phase does not inspect C calls and also has no 
-extra analysis such as for worst-case reaction 
-times~\cite{esterel.wcrt,esterel.fixed}.
+purpose~\cite{esterel.primer}, while \CEU provides an equivalent 
+\code{async/thread} primitive.
+Asynchronous execution is out of the scope of this paper.
+
+%Since loops in \CEU must contain \code{await} statements, reactions run in 
+%bounded time, which guarantees that successive calls to \code{dispatch} never 
+%block the scheduler for long.
+%However, the code generation phase does not inspect C calls and also has no 
+%extra analysis such as for worst-case reaction 
+%times~\cite{esterel.wcrt,esterel.fixed}.
 
@@
 deployment.
+For instance, we currently do not perform any worst-case reaction times 
+analysis~\cite{esterel.wcrt,esterel.fixed}.
 
@@
 The binding for a specific platform (environment) is responsible for calling 
-hook functions in the API of the runtime of \CEU whenever an external event 
+hook functions in the API of the \CEU runtime whenever an external event 
 occurs.
@@
 However, the binding must never interleave or run multiple API calls in 
-parallel.
-This would break the \CEU sequential/discrete semantics of time.
+parallel to preserve the sequential/discrete semantics of time in \CEU.
 
@@
 instructions for complex and recurrent operations from the runtime of \CEU, 
-such as handling events and trails.
+such as for handling events and trails.
 
@@
 
-TODO: most about comparison with Esterel
-    - TODO: related work?
-    - As far as we know, \CEU is the first Esterel-based language to HAVE?
-      queue-based/unique in the semantics of the language and not in the
-      library
-
-%\emph{(As subsecoes serao retiradas do texto ao final...)}
-
 %\subsection{Semantics}
 
-\CEU has a strong influence from Esterel, embracing its disciplined 
-synchronous-reactive model with support for lexical composition of lines of 
-execution.
-%
-However, there are fundamental semantic differences that prevents the design of 
-\CEU as pure extensions to Esterel.
-%
-In particular, Esterel has a notion of time similar to digital circuits in 
-which multiple signals can be active at a clock tick.
-%In fact, Esterel is also used in hardware design.
-%
-In \CEU, instead of clock ticks, the occurrence of a single external event that 
-defines a time unit.
-%
-\CEU also distinguishes external events from stack-based internal events, which 
-provide a limited form of coroutines supporting reactive statements.
-
+\CEU has a strong influence from Esterel but differ in the most fundamental 
+aspect of notion of time.
+%
+In \CEU, instead of clock ticks, the occurrence of a single external event 
+defines an unit of time.
+%
 The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
@@
 %
-Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
-through a queue and are dispatched sequentially and atomically to prevent race 
-conditions.
-%
-%We believe that for software design, this approach is more familiar to 
-%programmers and simplifies the reasoning about concurrency.
-%For instance, the uniqueness of external events in \CEU is a prerequisite for 
-%the temporal analysis that enables safe shared-memory concurrency.
-
-Many synchronous languages have been designed to interoperate with \C, such as 
-\emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
-\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
-%
-They offer Esterel-like parallel compositions with communication via shared 
-variables, relying on deterministic scheduling to preserve determinism.
-%
-However, it is the responsibility of the programmer to specify the execution 
-order for threads, based on either explicit priorities, or source code lexical 
-order (similar to \CEU).
-%
-These languages have a tick-based notion of time similar to Esterel, which 
-prevents the event-based temporal analysis of \CEU.
-
-\emph{URBI}~\cite{rp.urbi} is a reactive scripting language with a rich set of 
-control constructs for time management, event-driven communication, and 
-concurrency.
-Concurrency is based on stackful coroutines, diverging from our goals regarding 
-resource efficiency and static bounds for memory and execution time.
+We consider that this approach is more familiar to programmers and simplifies 
+the reasoning about concurrency.
+%
+As far as we know, \CEU is unique in this notion of time encrusted in the core 
+of the language, which is a prerequisite for the temporal analysis that enables 
+safe shared-memory concurrency.
+
+Another unique aspect of \CEU is the distinction between external and internal 
+events.
+Internal events define stack-based micro reactions within external reactions, 
+providing more fine-grained control for intra-reaction execution.
+
+Like \CEU, many synchronous languages rely on deterministic scheduling to 
+preserve intra-reaction determinism (%
+\emph{Reactive C}~\cite{rp.rc},                 % 1991
+\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
+\emph{SOL}~\cite{wsn.sol},                      % 2007
+\emph{SC}~\cite{rp.synchc},                     % 2009
+and
+\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
+%
+%It is the responsibility of the programmer to specify the execution order for 
+%threads, based on either explicit priorities, or source code lexical order 
+%(similar to \CEU).
+%
+\CEU goes one step further and performs a temporal analysis to detect 
+concurrent trails that actually rely on deterministic scheduling, i.e., the 
+observable behavior of the program changes if they are reordered.
+%
+%These languages have a tick-based notion of time similar to Esterel, which 
+%prevents the temporal analysis of \CEU.
+%
+\CEU also guarantees deterministic behavior for composition of timers by 
+adjusting residual awaking times from inaccurate system clocks.
+
+Regarding resource management, Esterel supports a finalization mechanism to 
+unconditionally execute a series of statements on abortion.
+%
+\CEU also tracks pointers representing resources that cross \C boundaries and
+enforce the programmer to provide corresponding finalizers.
+
+%\emph{URBI}~\cite{rp.urbi} is a dynamic reactive scripting language with a 
+%rich set of control constructs for time management, event-driven 
+%communication, and concurrency.
+%Concurrency is based on dynamically created stackful coroutines, diverging 
+%from our goals regarding resource efficiency and static bounds for memory and 
+%execution time.
 
@@
 
-\begin{comment}
-%\subsection{ Discussion }
-\label{sec.ceu.discussion}
-
-TODO: related
-- simplicity
-    - intra-reaction det
-    - internal events?
-        - more fine grained
-
-% SIMPLE
-
-    - flexible, coroutines, resumable exceptions
-    - yet simple, programmers understand stacks
-    - most of the time doesn't matter
-
-Intra-reaction determinism is another
-The temporal analsysi (Section~\ref{sec.ceu.analysis}).
-
-The temporal analsysi (Section~\ref{sec.ceu.analysis}).
-simple, programmers understand
-
-targets control-intensive applications and provides abstractions that can 
-express program flow specifications concisely.
-
-lexical order pragamatics
-    - again, familiar to programmers
-        - register order
-    - implementation aspects in Section XX
-
-TODO: move POE to here, simpler more natural for software dev
-
-- static analsysis also simpler
-    - highlight that is not only important to have a comptutally-sound 
-      algorithm, but that something that programmers can understand
-- same with 1st-class timers, also, too much common to be ignored
-
-- understandable semantics
-    - implementable
-
-TODO add a remark about tasks/async/thread in "Discussion"
-
-TODO Link the sections, connect the dots, draw conclusions, eliminate remaining xxx.
-
-Timers allows for an internal, defined by the programmer in charge of the 
-specification, definition of simultaneity.
-
-On the one hand, enforcing an execution order for concurrent operations may 
-seen arbitrary and also precludes true parallelism.
-On the other hand, it provides a priority scheme for trails, and makes 
-shared-memory concurrency more tractable.
-    - again, simple
-%
-For constrained embedded development, we believe that deterministic 
-shared-memory concurrency is beneficial, given the extensive use of memory 
-mapped ports for \emph{I/O} and the lack of hardware support for real 
-parallelism.
-
-TODO: we could detect that trails do not share memory and execute in multiple 
-threads
-
-%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
-%parallelism is inherent.
-
-Furthermore, our broad safety analysis, encompassing all proposed concurrency 
-mechanisms, ensures that the high degree of concurrency in WSNs does not pose 
-safety threats to applications.
-%
-%Even with the restrictions that were needed to enable this analysis, \CEU was 
-%shown to be expressive enough to imply in a decrease in code complexity for a 
-%range of typical WSN applications.
-%
-%By restricting the expressiveness of the language, we could enable a broad 
-%safety analysis in programs at compile time.
-%The analysis encompasses all proposed concurrency mechanisms, such as parallel 
-%compositions, first-class timers, and communication via events.
-%
-
-% SAFE
-
-Esterel XXX
-thanks to the uniqueness of events, pointer tracking, and first-class timers, 
-\CEU also provides additional safety protections:
-%
-No concurrency in accesses to shared variables and system calls sharing a 
-resource
-Finalization for blocks going out of scope
-Auto-adjustment for timers in sequence and synchronization for timers in parallel
-
-These properties are desirable in any application and are guaranteed as 
-preconditions in \CEU by design.
-Ensuring or even extracting these properties from less restricted languages 
-requires significant manual analysis.
-
-% LIMITATIONS
-
-Even though the achieved expressiveness and overhead of \CEU meet the 
-requirements of WSNs, its design imposes two inherent limitations:
-the lack of dynamic loading which would forbid the static analysis,
-and the lack of hard real-time guarantees.
-
-Regarding the first limitation, dynamic features are already discouraged due to 
-resource constraints.
-For instance, even object-oriented languages targeting WSNs forbid dynamic 
-allocation~\cite{wsn.flowtalk,wsn.virgil}.
-%
-Given that we focus on system-level development which does not require rich 
-dynamic functionality, we leave for a future work an in-depth discussion about 
-this issue.
-Note that queues, stacks, and other simple dynamic data structures for handling 
-message packets can be made available to \CEU through its safe integration with 
-$C$.
-
-To deal with the second limitation, which can be critical in the presence of 
-lengthy computations, we can consider the following approaches:
-(1) manually placing \code{pause} statements in unbounded loops;
-(2) integrating \CEU with a preemptive system.
-%
-The first option requires the lengthy operations to be rewritten in \CEU using
-\code{pause} statements so that other trails can be interleaved with them.
-This option is the one recommended in many related work that provide a similar 
-cooperative primitive (e.g. \code{pause}~\cite{esterel.primer}, 
-\code{PT\_YIELD}~\cite{wsn.protothreads}, \code{yield}~\cite{wsn.sol}, 
-\code{post}~\cite{wsn.nesc}).
-%
-Considering the second option, \CEU and preemptive threads are not mutually 
-exclusive.
-% and the second option is appealing to be explored in a future work.
-For instance, TOSThreads~\cite{wsn.tosthreads} proposes a message-based 
-integration with \emph{nesC} that is safe and matches the semantics of \CEU 
-external events.
-
-TODO: search "limitation" in the paper and condense all
-
-A limitation of event-triggered execution is that all program behavior must be 
-purely reactive, given that no code can execute in the absence of inputs.
-%
-Tick-triggered execution allows for active behavior, since code can execute 
-regulary on every tick.
-%
-Although it supports asynchronous active execution~\cite{ceu.tr}, we still 
-consider \CEU as purely reactive since its synchronous core cannot express 
-active behavior.
-
-\end{comment}
-
 \section{Conclusion}
@@
 
-We presented the design, semantics, and implementation of \CEU, a synchronous 
-reactive language inspired by Esterel targeting constrained embedded systems.
+We present the design and implementation of \CEU, a synchronous reactive 
+language inspired by Esterel with support for a simpler semantics and more 
+fine-grained control for intra-reaction execution.
 
@@
 successfully compile in \CEU:
-time and memory-bounded reactions to the environment (except for external 
-system calls),
+time and memory-bounded reactions to the environment (except for system calls),
 no race conditions in shared memory,
@@
 lexically scoped parallel compositions, convenient first-class timers, and a 
-unique stack-based signaling mechanism.
+unique stack-based internal signaling mechanism.
 %

### 2016-04-04 paper.tex
...

@@
 \begin{abstract}
-\CEU is a synchronous language inspired by Esterel with support for a simpler 
-semantics and more fine-grained execution control.
-%
-\CEU employs an event-triggered notion of time which enables compile-time 
+\CEU is a synchronous language inspired by Esterel with simpler semantics and 
+more fine-grained control over program execution.
+%
+\CEU adopts an event-triggered notion of time which enables compile-time 
 analysis for simultaneous reactions to produce deterministic and 
@@
 %
-We discuss the particularities of our design in comparison to Esterel, such as 
+Using Esterel as a base, we discuss the particularities of our design, such as 
 stack-based internal events, temporal analysis for concurrent statements, safe 
@@
 \subsection{Summary}
-
 \CEU aims to offer a simpler semantics than Esterel with more determinism and 
-fine-grained control over the program execution.
+fine-grained control over program execution.
 %
@@
 \CEU has a strong influence from Esterel but differ in the most fundamental 
-aspect of notion of time.
-%
-In \CEU, instead of clock ticks, the occurrence of a single external event 
-defines an unit of time.
+aspect of the notion of time.
+%
+In \CEU, instead of clock ticks, atomic external event occurrences define time 
+units.
 %
@@
 %
-As far as we know, \CEU is unique in this notion of time encrusted in the core 
-of the language, which is a prerequisite for the temporal analysis that enables 
-safe shared-memory concurrency.
+As far as we know, \CEU is to encrust this event-triggered notion of time in 
+the core of the language, which is a prerequisite for the temporal analysis 
+that enables safe shared-memory concurrency.
 
@@
 events.
-Internal events define stack-based micro reactions within external reactions, 
+Internal events support stack-based micro reactions within external reactions, 
 providing more fine-grained control for intra-reaction execution.
@@
 %
-\CEU goes one step further and performs a temporal analysis to detect 
-concurrent trails that actually rely on deterministic scheduling, i.e., the 
-observable behavior of the program changes if they are reordered.
+\CEU goes one step further and performs a temporal analysis to detect trails 
+that, when reordered, change the observable behavior of the program, i.e., 
+trails that actually rely on deterministic scheduling.
+
 %
@@
 %
-\CEU also guarantees deterministic behavior for composition of timers by 
-adjusting residual awaking times from inaccurate system clocks.
+\CEU also guarantees deterministic behavior for timer compositions by adjusting 
+inaccuracies in the system clock.
 
@@
 \CEU also tracks pointers representing resources that cross \C boundaries and
-enforce the programmer to provide corresponding finalizers.
+enforce the programmer to provide associated finalizers.
 
@@
 We present the design and implementation of \CEU, a synchronous reactive 
-language inspired by Esterel with support for a simpler semantics and more 
-fine-grained control for intra-reaction execution.
+language inspired by Esterel with simpler semantics and more fine-grained 
+control for intra-reaction execution.
 
@@
 lexically scoped parallel compositions, convenient first-class timers, and a 
-unique stack-based internal signaling mechanism.
+unique stack-based mechanism for internal signalling.
 %
@@
 weeks.
+%
+More recently, a company specialized in embedded systems (not related to our 
+research group) released a product based on \CEU.
 

### 2016-04-05 paper.tex
...

@@
 Francisco Sant'Anna
-\affil{Departamento de Inform\'atica, PUC--Rio}
+\affil{Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ}
 Roberto Ierusalimschy
@@
 %
-\CEU adopts an event-triggered notion of time which enables compile-time 
-analysis for simultaneous reactions to produce deterministic and 
-concurrency-safe programs.
-%
-Using Esterel as a base, we discuss the particularities of our design, such as 
-stack-based internal events, temporal analysis for concurrent statements, safe 
-integration with \C, and first-class timers.
+\CEU adopts an event-triggered notion of time that enables compile-time 
+temporal analysis to detect conflicting statements, resulting in deterministic 
+and concurrency-safe programs.
+%
+Using Esterel as a base, we discuss the particularities of our design such as 
+stack-based internal events, the temporal analysis, safe integration with \C, 
+and first-class timers.
 %
@@
 one aiming for resource efficiency and interoperability with \C,
-and another based on a virtual machine that allows remote reprogramming.
+and another as a virtual machine that allows remote reprogramming.
 \end{abstract}
@@
 \emph{causality} and \emph{schizophrenia} problems.
-Both problems have an extensive coverage and debate in the literature 
+These problems have an extensive coverage in the literature 
 ~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
@@
 %
-Another drawback of the Esterel semantics consists of the loose and 
-non-deterministic execution for intra-reaction statements, which prevents 
-threads to interact with stateful system calls safely and makes shared-memory 
-concurrency not as straightforward as reading and writing to shared variables.
+Another drawback of the Esterel semantics consists of loose and 
+non-deterministic execution for intra-reaction statements.
+Non determinism prevents threads to interact with stateful system calls safely 
+and makes shared-memory concurrency not as straightforward as reading and 
+writing to shared variables.
 %
@@
 Overall, \CEU has a simple semantics with fine-grained control for 
-intra-reaction execution which is amenable to a simple temporal analysis that
+intra-reaction execution, and is amenable to a temporal analysis that
 improves safety.
@@
 microcontroller with 32K of FLASH and 2K of SRAM.
-}, showing the practical aspect of our simple semantics.
+}, showing a practical aspect of our simple semantics.
 %:\url{http://www.atmel.com/devices/atmega328.aspx}}
@@
 
-Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
-\CEU \bx for the following control specification:
+Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
+languages, showing side-by-side the implementations in Esterel \ax and \CEU \bx 
+for the following control specification:
 %
@@
 For the second phrase, the reset behavior, the Esterel version uses a 
-\code{abort-when} (ln. 4--11), which, in this case, serves the same purpose of 
+\code{abort-when} (ln. 4--11) which, in this case, serves the same purpose of 
 \CEU's \code{par/or} (ln. 4--13):
@@
     (\ref{sec.ceu.timers}).
-%We finish with a summary of our design
-    %(\ref{sec.ceu.discussion}).
+We finish with a summary of our design
+    (\ref{sec.ceu.discussion}).
 %

### 2016-04-06 paper.tex
...

@@
 %
-Using Esterel as a base, we discuss the particularities of our design such as 
+Using Esterel as a base, we discuss the particularities of our design, such as 
 stack-based internal events, the temporal analysis, safe integration with \C, 
@@
 requires careful static analysis such as to detect and refuse programs with 
-\emph{causality} and \emph{schizophrenia} problems.
-These problems have an extensive coverage in the literature 
+\emph{causality} and \emph{schizophrenia} problems%
+%These problems have an extensive coverage in the literature 
 ~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
@@
 %
-\item Safe integration with \C which enforces finalization for external 
+\item Safe integration with \C, which enforces finalization for external 
 resources.
@@
 We finish with a summary of our design
-    (\ref{sec.ceu.discussion}).
-%
-We provide the formal specification of the semantics of \CEU in a separate 
+    (\ref{sec.ceu.summary}).
+%
+We present the formal specification of the semantics of \CEU in a separate 
 work~\cite{ceu.phd}.
@@
 At each tick, the program reacts to an arbitrary number of simultaneous input 
-events, depending on external stimuli the environment provides.
+events from the environment.
 In contrast, \CEU defines time as a discrete sequence of reactions to
@@
       parallel constructs may create new trails).
-\item Active trails execute until they await or terminate.
+\item Active trails execute, one after the other, until they await or 
+      terminate.
       This step is named a \emph{reaction chain}, and always runs in bounded 
@@
 of Esterel and in \CEU.
-The box \code{Real World} shows event occurrences over a continuous timeline 
-divided in units of 10 milliseconds.
-The other boxes define distinct discrete logical units of time in which the 
-same occurring events fit differently.
+The box \code{Real World} shows event occurrences over a continuous timeline.
+The other boxes show discrete logical units of time in which the same occurring 
+events fit in different ways.
 %
@@
 \item \code{[Box-1]}:
-``Esterel with fixed-length ticks'',
-within which reactions must fit~\cite{esterel.fixed}.
-%
-We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
+\emph{Esterel with fixed-length ticks},
+in which reactions must fit~\cite{esterel.fixed}.
+%
+We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
 input.
@@
 \item \code{[Box-2]}:
-``Esterel with variable-length ticks'', which are adjusted to the CPU time the 
-corresponding reaction takes to complete~\cite{esterel.variable}.
+\emph{Esterel with variable-length ticks}, which are adjusted to the CPU time 
+the corresponding reaction takes to complete~\cite{esterel.variable}.
 This approach avoids the time violation for \code{R(A)} and also results in
@@
 \item \code{[Box-3]}:
-``\CEU with unique and queue-based input events''.
-We also assume a \code{R(boot)} reaction before any input.
+\emph{\CEU with unique and queue-based input events}.
+We also assume a reaction \code{R(boot)} before any input.
 %
 Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
-behavior in \CEU is similar to \code{Box-2} for the first two units of time 
+behavior in \CEU is similar to \code{Box-2} for the first two reactions 
 (\code{boot-0} and \code{A-1}).
@@
 discrete ticks.
-For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
-events \code{B} and \code{C} are considered simultaneous, even though \code{A} 
-and \code{B} ``physically'' happen much closer to one another.
+For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), 
+Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
+though \code{A} and \code{B} ``physically'' happen much closer to one another.
 In Section~\ref{sec.ceu.timers}, after introducing internal events and 
@@
 \item \textbf{Unique input events imply mutual exclusion:}
-Reactions to multiple events never overlap because they are atomic.
-Automated mutual exclusion simplifies the reasoning about the program and is a 
+Reactions to multiple events in \CEU never overlap because they are atomic.
+Automated mutual exclusion simplifies the reasoning about concurrency and is a 
 prerequisite for the temporal analysis to be discussed in 
@@
 events become active, resulting in the invocation of \code{f()} and \code{g()} 
-(ln. 6,9) in no particular order.
+in no particular order (ln. 6,9).
 %
@@
 awaits the next ``call'' (ln. 4, again).
-Only after this sequence the calling trail resumes and passes the assertion 
+Only after this sequence, the calling trail resumes and passes the assertion 
 test (ln. 11).
@@
 \CEU also supports nested \code{emit} invocations for internal events.
-For instance, the code that executes after the \code{await inc} (ln. 4) could 
-\code{emit} another event, creating a new level in the stack.
-The runtime stack represents fine-grained micro reactions, one on top of the 
+For instance, the code in Figure~\ref{lst.sub} that executes after the 
+\code{await inc} (ln.  4) could \code{emit} another event, creating a new level 
+in the stack.
+The runtime stack constitutes fine-grained micro reactions, one on top of the 
 other, all inside the same reaction to an external event.
@@
 prevents infinite execution by design.
-Both sides of the \code{par/or} have an \code{await} statement to avoid 
-instantaneous termination (ln. 4,7) of the enclosing \code{loop} (ln 2--9).
+Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
+makes the enclosing \code{loop} non instantaneous (ln 2--9).
 However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
@@
 %
-A number of Esterel-based synchronous languages enforce intra-reaction 
-determinism with an arbitrary execution order for statements in multiple lines 
-of execution (%
+A number of Esterel-based synchronous languages have intra-reaction determinism 
+by enforcing an arbitrary execution order for statements in multiple lines of 
+execution (%
 \emph{Reactive C}~\cite{rp.rc},                 % 1991
@@
 \caption{ Shared-memory concurrency in \CEU:
-Example \ax is safe because the trails access \code{x} atomically in different 
+example \ax is safe because the trails access \code{x} atomically in different 
 reactions;
-Example \bx is unsafe because both trails access \code{y} in the same reaction.
+example \bx is unsafe because both trails access \code{y} in the same reaction.
 \label{lst.shared}
@@
 %
-For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
-which define a shared variable (ln. 2), and assign to it in parallel trails 
-(ln.  5, 8).
-%
-In the example \ax, the two assignments to \code{x} only execute in reactions 
-to different events \code{A} and \code{B} which, by definition, cannot occur 
+For instance, consider the two examples in Figure~\ref{lst.shared}, both 
+defining a shared variable (ln. 2), and assigning to it in parallel trails (ln.  
+5, 8).
+%
+In the example \ax, the two assignments to \code{x} execute in reactions to 
+different events \code{A} and \code{B} only, which by definition, cannot occur 
 simultaneously (Section~\ref{sec.ceu.exts}).
-Hence, for the sequence \code{A->B}, \code{x} becomes \code{4=(1+1)*2}, while 
-for \code{B->A}, \code{x} becomes \code{3=(1*2)+1}.
+Hence, for the sequence \code{A->B}, \code{x} becomes \code{4} 
+(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes \code{3} 
+(\code{(1*2)+1}).
 %
@@
 Since \CEU employs lexical order for intra-reaction statements, the execution 
-is still deterministic, and \code{y} always becomes \code{4=(1+1)*2}.
+is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
 %
@@
 %
+Concurrency in \CEU is characterized when two or more trail segments in 
+parallel react to the same input event.
 A trail segment is a sequence of statements followed by an \code{await} (or 
 termination).
-Concurrency in \CEU is characterized when two or more trail segments in 
-parallel react to the same input event.
 %
@@
 \begin{itemize}
-\item The assignments to \code{x} in lines 2 and 5 in \ax
+\item The assignments to \code{x} (\ax:2,5)
       \textbf{cannot} be concurrent because they are \textbf{not} in parallel trails.
-\item The assignments to \code{x} in lines 5 and 8 in \ax
+\item The assignments to \code{x} (\ax:5,8)
       \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
       the same reaction.
-\item The assignments to \code{y} in lines 5 and 8 in \bx
+\item The assignments to \code{y} (\bx:5,8)
       \textbf{can} be concurrent because they are in parallel trails and 
@@
 %
-Note also that the analysis is completely optional and does not affect the 
-semantics of the program.
+Note also that the analysis is optional and does not affect the semantics of 
+the program.
 % TODO-no: static analsysis also simpler
@@
 As a safety measure, the temporal analysis of Section~\ref{sec.ceu.analysis} 
-also considers calls and accesses to external symbols in \C.
+also considers concurrent calls and accesses to external symbols in \C.
 %
@@
 %
-Since \CEU does not inspect any code in \C, it complains about suspicious 
-concurrent accesses between \code{\_f} and all symbols in the second trail.
+Since \CEU does not inspect any code in \C, it opts to complain about 
+suspicious concurrent accesses between \code{\_f} and all symbols in the second 
+trail.
 
@@
 
-From our experience, we find that it is not uncommon to also require 
-non-commutative calls to execute concurrently.
+From our experience, however, we find that programs often need non-commutative 
+concurrent calls.
 %
@@
 In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
-conflicting pixels, which makes it non commutative and prevents the code to 
+conflicting pixels, which makes them non commutative and prevent the code to 
 compile.
 %
-However, we want to rely on trails lexical order to always redraw the 
+However, we want to rely on lexical order to always redraw the 
 \code{background} object before the \code{foreground} object.
@@
 ``intimidating'' name \code{\_redraw\_non\_commutative} (ln. 2, to be explicit 
-about its effect), and annotate it as safe (ln. 4--5) to compile successfully.
+about its effect), and annotate it as safe (ln. 4--5) to make the code compile 
+successfully.
 
 \begin{figure}
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.48\linewidth}
 \begin{lstlisting}
@@
 %
-\begin{minipage}[t]{0.55\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+\begin{minipage}[t]{0.52\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
 native do
@@
 abortion of lines of execution, which is a distinctive feature of synchronous 
-language in comparison to asynchronous languages~\cite{esterel.preemption}.
+languages in comparison to asynchronous languages~\cite{esterel.preemption}.
 %In contrast, traditional (asynchronous) multi-threaded languages cannot 
@@
 to \code{lock} and \code{unlock} represent accesses to an external resource.
-The normal behavior is to \code{lock} the resource (ln. \ax:4 and \bx:3), 
-perform some operations in subsequent reactions to input \code{A} (ln.  
-\ax:5--8 and \bx:7--10), and then \code{unlock} the resource (ln.  \ax:10 and 
-\bx:5).
+After we \code{lock} the resource (ln. \ax:4 and \bx:3), we perform some 
+operations in subsequent reactions to input \code{A} (ln.  \ax:5--8 and 
+\bx:7--10), and then we \code{unlock} the resource (ln.  \ax:10 and \bx:5).
 Note that if the aborting input \code{B} (ln. \ax:12 and \bx:12) occurs after 
@@
 
-\CEU goes one step further and enforces the use of \code{finalize} for system 
-calls that deal with pointers representing resources:
+\CEU goes one step further and actually enforces the use of \code{finalize} for 
+system calls that deal with pointers representing resources:
 %
 \begin{itemize}
-\item If a system call \textbf{receives} a pointer from \CEU, the pointer 
+\item If \CEU \textbf{passes} a pointer to a system call, the pointer 
 represents a \textbf{local} resource that requires finalization.
-\item If a system call \textbf{returns} a pointer to \CEU, the pointer 
+\item If \CEU \textbf{receives} a pointer from a system call, the pointer 
 represents an \textbf{external} resource that requires finalization.
@@
     finalize with
-        _unlock();
+        _unlock();  // defer execution
     end
@@
 and sensor samplings.
-However, support for wall-clock time is somewhat low-level in existing 
-languages, usually through timer callbacks or ``sleep'' blocking calls.
-%
-Furthermore, in any concrete timer implementation, a requested timeout does not 
-expire precisely without delays, a fact that is usually ignored in the 
-development process.
-We define the difference between the requested timeout and the actual expiring 
+%
+However, the interaction between system clocks and programs is not 100\% 
+precise, a fact that is usually ignored in the development process.
+We define the difference between a requested timeout and the actual expiring 
 time as the \emph{residual delta time (delta)}.
@@
 compared.
-For instance, for the example in Figure~\ref{lst.timers}.b, although the 
+For instance, in the example in Figure~\ref{lst.timers}.b, although the 
 scheduler cannot guarantee that the first trail terminates exactly in 11ms (ln.  
@@
 
-In Section~\ref{sec.ceu.exts}, we argue that events occurrences are 
+In Section~\ref{sec.ceu.exts}, we argue that event occurrences are 
 infinitesimal and can never be absolutely simultaneous.
@@
 increases with the inaccuracy of the observer (e.g., a human being).
-Therefore, we consider that simultaneity should be defined with respect to the 
-observer, which varies from case to case.
+Therefore, we consider that simultaneity should be defined case by case, and 
+should not be imposed by the language.
 %
@@
 \subsection{Summary}
+\label{sec.ceu.summary}
+
 \CEU aims to offer a simpler semantics than Esterel with more determinism and 
@@
 %
-\item Event-triggered notion of time bound to the semantics of the language.
+\item \emph{Event-triggered notion of time bound to the semantics of the 
+language.}
 Event-driven programming is popular in many domains, such as server and GUI 
@@
 %
-\item Deterministic intra-reaction execution and communication.
+\item \emph{Deterministic intra-reaction execution and communication.}
 Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
@@
 %
-\item Static temporal analysis for concurrent statements.
+\item \emph{Static temporal analysis for concurrent statements.}
 Although execution is deterministic, the compiler of \CEU still advises about 
-suspicious statements that can react concurrently to the same event, forcing 
-the to serialize them accordingly.
-%
-\item Safe integration with \C.
+suspicious statements that can react concurrently to the same event, 
+appropriate serialization.
+%
+\item \emph{Safe integration with \C.}
 \CEU enforces the programmer to deal with pointers representing resources 
@@
 %
-\item First-class timers with dedicated syntax and automatic synchronization.
+\item \emph{First-class timers with dedicated syntax and automatic 
+synchronization.}
 Given the omnipresence of timers in embedded systems, a dedicated syntax can 
@@
 \begin{itemize}
-\item Execution is purely reactive as result of event-triggered reactions.
-Since only occurrences of events can start reactions, programs cannot execute 
+\item \emph{Execution is purely reactive as result of event-triggered 
+reactions.}
+Since only event occurrences can start reactions, programs cannot execute 
 proactively in the absence of events.
-\item Reactions must execute in bounded time due to the synchronous hypothesis.
+In addition, \code{await} statements cannot awake in the same reaction they are
+reached.
+\item \emph{Reactions must execute in bounded time due to the synchronous 
+hypothesis.}
 As a synchronous language, \CEU requires CPU times for reactions to be 
 negligible in comparison to the rate of incoming events.
-\item Execution is sequential because of intra-reaction determinism.
+\item \emph{Execution is sequential because of intra-reaction determinism.}
 %We discuss some implementation considerations in 
@@
 %
-Nonetheless, we advocate for keeping a tractable synchronous reactive core 
-which supports shared memory concurrency with deterministic execution.
-%
-To deal with the limitations above, we recommend for memory-isolated 
-parallelizable asynchronous functionality as a separate extension to the 
-synchronous core~\cite{esterel.crp} (which is not in the scope of this paper).
+Nonetheless, we advocate for keeping a tractable synchronous reactive core with 
+support for shared memory concurrency and deterministic execution.
+%
+To deal with the limitations above (which is not in the scope of this paper), 
+we recommend for memory-isolated parallelizable asynchronous primitives as 
+separate extensions to the synchronous core~\cite{esterel.crp}.
 
@@
 the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
-unbounded loops and suspicious accesses to shared memory;
+unbounded loops and suspicious concurrent statements;
 the \emph{code generation phase} converts the \emph{AST} to standard \C code 
@@
 %
-Figure~\ref{fig.eval} compares source size (\emph{tokens}), binary size 
-(\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
+Figure~\ref{fig.eval} compares source size (\emph{number of tokens}), binary 
+size (\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
 network protocols and a radio driver.
-The small increase in resource usage shows that the gains in productivity and 
+The small overhead in resource usage shows that the gains in productivity and 
 safety with \CEU make it a viable alternative to \C in the context of 
 constrained embedded systems.
-
-Unfortunately, most real applications in Esterel seems to be closed source.
+%
+Unfortunately, most real applications in Esterel seem to be closed source.
 In a future work, we plan to cooperate with groups that have more experience 
 and access to real code bases in Esterel for a comprehensive comparison with 
-Céu.
+\CEU.
 % TODO-no:
@@
 
-In the sections that follow, we discuss the implementation of subsystems that 
-are specific (or in some way important) to \CEU:
-the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
+In the subsections that follow, we discuss \CEU-specific implementation 
+details:
+temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
 static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
@@
 For each node representing a statement in the program AST, we keep the set of 
-input events $I$ (for \emph{incoming}) that can lead to the execution of the 
-node, and also the set of input events $O$ (for \emph{outgoing}) that can 
-terminate the node.
+input events $I$ (\emph{incoming}) that can lead to the execution of the node, 
+and also the set of input events $O$ (\emph{outgoing}) that can terminate the 
+node.
 As an example, for the single-statement program \code{await~A}, we have 
@@
 %
-\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
-O_{body}$), as the loop is also reached from its own body.
+\item A \code{loop} node includes the output of its body on its own $I$ ($I=I 
+\cup O_{body}$), as the loop is also reached from its own body.
 The union of all $O$ from nested \code{break} statements forms the set $O$ for 
@@
 %
-\item An \code{if} node has $O=O_{true} \cup O_{false}$.
+\item An \code{if} node has $O=O_{true} \cup O_{false}$, where \code{true} and 
+\code{false} are the two clauses of the \code{if}.
 %
@@
     \begin{itemize}
-    \item An \code{emit e} terminates in the same reaction, having $O=I$.
     \item An \code{await e} has $O=I_{e1} \cup ... \cup I_{eN}$,
         where $e1...eN$ are \code{emit e} statements in trails in parallel.
+    \item An \code{emit e} terminates in the same reaction, having $O=I$.
     \end{itemize}
@@
 
-The code in Figure~\ref{lst.impl.ast}.a has the $AST$ and sets $I$ and $O$ for 
-each node in Figure~\ref{lst.impl.ast}.b.
+The code in Figure~\ref{lst.impl.ast}.a has its corresponding $AST$ and sets 
+$I$ and $O$ in Figure~\ref{lst.impl.ast}.b.
 The event $.$ (dot) represents the ``boot'' reaction.
-The assignments to \code{y} in parallel (ln. 5,8 in the code) have an empty 
-intersection of $I$ (ln. 6,9 in the AST), hence, they do not conflict.
-Note that although the accesses in ln. 5,11 in the code (ln. 6,11 in the AST) 
-do have an intersection, they are not in parallel branches and are also safe.
+The assignments to \code{y} in parallel (ln. \ax:5,8) have an empty 
+intersection of $I$ (ln. \bx:6,9), hence, they do not conflict.
+Note that, although the accesses to \code{y} in sequence (ln. \ax:5,11)
+do have an intersection (ln. \bx:6,11), they are not in parallel branches and 
+are also safe.
 
@@
 Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
-for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
-in a \code{union}.
+for blocks in sequence are packed in a \code{union}, while blocks in parallel, 
+in a \code{struct}.
 %In the example, the code following the loop (identified as \code{...}) reuses 
@@
 %
-The \code{do-end} blocks in sequence (ln. \ax:2--7) are packed in a 
-\code{union} (ln. \bx:2--3), given that their variables cannot be in scope at 
-the same time, e.g., \code{MEM.a\_1} and \code{MEM.b\_2} (ln. \bx:2,3) can 
-safely share the same memory slot.
+The \code{do-end} blocks and \code{par/and} in sequence
+(ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
+given that their variables cannot be in scope at the same time, e.g., 
+\code{MEM.a\_1} and \code{MEM.b\_2} (ln.  \bx:2,3) can safely share the same 
+memory slot.
 %
@@
 %
-A \code{STACKED} trail holds an associated numeric stack level and is delayed 
-until the scheduler runtime reaches that level again.
-%
-A \code{FINALIZE} trail represents a hanged finalization block which is only 
-scheduled when its corresponding block goes out of scope.
+A \code{STACKED} trail holds an associated numeric stack level and can only 
+execute when scheduler runtime reaches that level.
+%
+A \code{FINALIZE} trail represents a hanged finalization block which is
+scheduled only  when its corresponding block goes out of scope.
 %
 A trail waiting for an event stays in the state of the corresponding event, 
-also holding the minimum sequence number (\emph{seqno}) in which it can awake, 
+also holding the minimum sequence number (\emph{seqno}) in which it can awake 
+(for \emph{delayed awaits}).
+
 to prevent from awaking in the same reaction it is reached.
@@
 struct trail_t {
-    state_t evt;
-    label_t lbl;
+    state_t evt;              // awaiting event
+    label_t lbl;              // awaking execution label
     union {
@@
 field and may hold the \code{seqno} for an event, or the stack level \code{stk} 
-for a
-\code{STACKED} state.
+for a \code{STACKED} trail.
 
@@
 %
-The \code{seqno} could eventually overflow during execution (i.e., every 256 
-reactions).
-However, given that the scheduler traverses all trails on every reaction, it 
-adjusts them to properly handle overflows (actually, 2 bits to hold the 
-\code{seqno} is already enough).
+Since the scheduler traverses all trails on every reaction, it adjusts all 
+\code{seqno} to avoid overflows (actually, 2 bits to hold the \code{seqno} is 
+already enough).
 %
 The size of \code{stack\_t} depends on the maximum depth of nested emissions 
-and is bounded to the maximum number of trails.
-In the worst case, a trail emits an event that awakes another trail, which 
+but is bounded to the maximum number of trails:
+in the worst case, a trail emits an event that awakes another trail, which 
 emits an event that awakes another trail, and so on;
@@
 simultaneous lines of execution with memory overhead of only 2\% in comparison 
-to the original version in \C~\cite{ceu.sensys13}.
+to the original single-threaded version in \C~\cite{ceu.sensys13}.
 
@@
 For the program in \ax, the compiler extracts the entry points and associated 
-trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln.  
-\ax:7).
+trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0}
+(ln. \ax:7 and \bx:3).
 %
@@
 in the \emph{broadcast} pass, it sets all trails that are waiting for the 
-current event to \code{STACKED} in the current stack level;
+current event to the state \code{STACKED} in the current numeric stack level;
 %
@@
 %
-The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
-registered label \code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to 
-await \code{EVT\_A} and also halts.
+Then, \code{dispatch} takes \code{TRAIL-1} and executes its registered label 
+\code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to await 
+\code{EVT\_A} and also halts.
 
@@
 Given that trails in parallel are allocated in subsequent slots in the static 
-vector \code{TRLS}, this pass only aborts the desirable trails.
-The subsequent \emph{dispatch} pass executes the finalization code.
+vector \code{TRLS} (ln. \bx:12--15), this pass only aborts the desirable 
+trails.
+The subsequent \emph{dispatch} pass executes the finalization code properly.
 %Given that finalization blocks cannot contain \code{await} statements, the 
@@
 work~\cite{esterel.multicore.1,esterel.multicore.2}.
-One reason is the continuous fork-and-rejoin overhead due to small reactions.
-Another reason is the excessive locality of data due to stackless threads 
+One reason is due to continuous fork-and-rejoin overhead from small reactions.
+Another reason is due to excessive locality of data from stackless trails 
 sharing contiguous static memory.
@@
 
-Single-thread dispatching may not be suitable for hard real-time activities.
+Single-threaded dispatching may not be suitable for hard real-time activities.
 In a previous work~\cite{ceu.sensys13}, we measure how synchronous lengthy 
-computations in C (e.g., hashing and compression) can block the scheduler and 
+computations in \C (e.g., hashing and compression) can block the scheduler and 
 affect higher-priority activities such as a radio driver.
@@
 %
-The binding for a specific platform (environment) is responsible for calling 
-hook functions in the API of the \CEU runtime whenever an external event 
-occurs.
-%
-However, the binding must never interleave or run multiple API calls in 
-parallel to preserve the sequential/discrete semantics of time in \CEU.
-
-As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
+The binding for a specific platform (environment) calls hook functions in the 
+API of the \CEU runtime whenever an external event occurs.
+%
+However, to preserve the sequential/discrete semantics of time in \CEU, these 
+calls must never interleave or parallelize execution.
+
+Figure~\ref{lst.impl.tinyos} shows our binding for 
 \emph{TinyOS}~\cite{wsn.tos}, which maps system callbacks to input events in 
@@
 If the \CEU program uses timers, we also start a periodic timer (ln. 8--10) 
-that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds 
-and advances the wall-clock time of \CEU (ln. 15)%
+that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds to 
+advance the wall-clock time of \CEU (ln. 15)%
 \footnote{We also offer a mechanism to start the underlying timer on demand to
@@
 %
-The scheduler of the TinyOS is already synchronous by default and always 
-execute \code{event} handlers atomically, hence, the API calls to \CEU are 
-properly serialized.
+The scheduler of TinyOS is already synchronous by default and always execute 
+\code{event} handlers atomically, hence, the API calls to \CEU are properly 
+serialized.
 
@@
 %
-In \CEU, instead of clock ticks, atomic external event occurrences define time 
-units.
+In \CEU, instead of clock ticks, atomic external event occurrences that define 
+time units.
 %
@@
 %
-As far as we know, \CEU is to encrust this event-triggered notion of time in 
-the core of the language, which is a prerequisite for the temporal analysis 
-that enables safe shared-memory concurrency.
+As far as we know, \CEU is the first to encrust this event-triggered notion of 
+time in the core of the language, which is a prerequisite for the temporal 
+analysis that enables safe shared-memory concurrency.
 
@@
 trails that actually rely on deterministic scheduling.
-
 %
@@
 %
-\CEU also tracks pointers representing resources that cross \C boundaries and
+In addition, \CEU also tracks pointers representing resources that cross \C 
+boundaries and
 enforce the programmer to provide associated finalizers.
@@
 
-\CEU is a concurrency-safe language, employing a static analysis that encompass 
-all control constructs and ensures that the high degree of concurrency in 
-embedded systems does not pose safety threats to applications.
+\CEU is a concurrency-safe language, employing static analysis to ensure that 
+the high degree of concurrency in embedded systems does not pose safety threats 
+to applications.
 %

### 2016-04-11 paper.tex
...

@@
 \begin{abstract}
-\CEU is a synchronous language inspired by Esterel with simpler semantics and 
+\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
 more fine-grained control over program execution.
 %
-\CEU adopts an event-triggered notion of time that enables compile-time 
-temporal analysis to detect conflicting statements, resulting in deterministic 
-and concurrency-safe programs.
-%
-Using Esterel as a base, we discuss the particularities of our design, such as 
+\CEU uses an event-triggered notion of time that enables compile-time temporal 
+analysis to detect conflicting statements, resulting in deterministic and 
+concurrency-safe programs.
+%
+Using Esterel as a base, we present the particularities of our design, such as 
 stack-based internal events, the temporal analysis, safe integration with \C, 
@@
 In this work, we present \CEU, a new programming language that inherits the 
-synchronous and imperative mindset of Esterel but diverges in fundamental 
+synchronous and imperative mindset of Esterel but diverges in some fundamental 
 semantic aspects.
 %
-% TODO: simple simple
-Overall, \CEU has a simple semantics with fine-grained control for 
-intra-reaction execution, and is amenable to a temporal analysis that
+\CEU has a simple semantics with fine-grained control for intra-reaction 
+execution, and is amenable to a temporal analysis that
 improves safety.
@@
 For the \VM back end, applications have a bytecode footprint in the order of 
-hundreds of bytes which can be transmitted over the air in a few 
+hundreds of bytes and can be transmitted over the air in a few 
 packets~\cite{ceu.terra}.
@@
 %
-\CEU shares the same limitations with Esterel and synchronous languages in 
-general:
+\CEU shares the same limitations with (core) Esterel and synchronous languages 
+in general~\cite{esterel.preemption}:
 computations that run in unbounded time (e.g., cryptography, image processing) 
@@
 %
-\emph{``Emit an output O as soon as two inputs A and B have occurred.
-Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
+\emph{``Emit an output O as soon as inputs A and B occur.
+Reset this behavior each time input R occurs''}~\cite{esterel.primer}.
 %
@@
 \emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
+A \code{par/and} terminates when both trails in parallel terminate.
+A \code{par/or}  terminates when any trail terminates, aborting the 
+other.
 \label{lst.abro}
@@
 
-In the subsections that follow, we discuss the main differences between \CEU 
-and Esterel:
+In the subsections that follow, we discuss the main differences from \CEU to 
+Esterel:
 %
@@
     (\ref{sec.ceu.timers}).
-We finish with a summary of our design
+We finish the section with a summary of our design
     (\ref{sec.ceu.summary}).
@@
       parallel constructs may create new trails).
-\item Active trails execute, one after the other, until they await or 
-      terminate.
+\item Active trails execute until they await or terminate, one after the other.
       This step is named a \emph{reaction chain}, and always runs in bounded 
@@
 The box \code{Real World} shows event occurrences over a continuous timeline.
-The other boxes show discrete logical units of time in which the same occurring 
-events fit in different ways.
+The other boxes show how the same occurring events fit differently in each 
+logical notion of time.
 %
@@
 \item \code{[Box-1]}:
-\emph{Esterel with fixed-length ticks},
-in which reactions must fit~\cite{esterel.fixed}.
+\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
 %
@@
 %
-Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
-\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
+Note that \code{R(A)} takes more time than \code{tick-1} and invades 
+\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
 %
@@
 occur during the idle \code{tick-3}.
-However, only one occurrence of \code{E} is considered in \code{R(D+E)}.
+However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
 %
 \item \code{[Box-2]}:
-\emph{Esterel with variable-length ticks}, which are adjusted to the CPU time 
-the corresponding reaction takes to complete~\cite{esterel.variable}.
+\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
 This approach avoids the time violation for \code{R(A)} and also results in
-smaller idle periods.
+smaller idle periods because it adjusts the tick lengths to match the CPU times 
+for the reactions.
 For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
-react alone as \code{R(D)} on \code{tick-4}.
+react alone as \code{R(D)} at \code{tick-4}.
 Similarly to the fixed-tick approach, only one of the two simultaneous 
-occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
+occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
 takes too long.
@@
 %
-Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
-behavior in \CEU is similar to \code{Box-2} for the first two reactions 
-(\code{boot-0} and \code{A-1}).
+Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
+behavior in \CEU is similar to \code{Box-2} for its first two reactions 
+(\code{tick-0} and \code{tick-1}).
 %
@@
 independently, which are handled in sequence.
+%
+Note that, by definition, each reaction corresponds to an event occurrence of 
+the same length.
 \end{itemize}
@@
 probability of being simultaneous.
-This way, we believe that the notion of simultaneity should not be imposed by 
+This way, we consider that the notion of simultaneity should not be imposed by 
 the language, but defined explicitly for each use case.
@@
 discrete ticks.
-For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), 
+For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
 Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
@@
 Reactions to multiple events in \CEU never overlap because they are atomic.
-Automated mutual exclusion simplifies the reasoning about concurrency and is a 
+Automatic mutual exclusion simplifies the reasoning about concurrency and is a 
 prerequisite for the temporal analysis to be discussed in 
@@
 For the fixed-length-tick approach of Esterel, a breach in the synchronous 
-hypothesis causes timing violations, which can be avoided with \emph{worst case 
-reaction time} analysis to infer an appropriate value for the tick 
-length~\cite{esterel.fixed}.
+hypothesis causes timing violations, requiring \emph{worst case reaction time} 
+analysis to infer appropriate tick lengths~\cite{esterel.fixed}.
 % TODO-no: Such analsysis is orthogonal
 
-A limitation of event-triggered execution is that all program behavior must be 
+A limitation of event-triggered execution is that all program behavior is 
 purely reactive, given that no code can execute in the absence of inputs.
@@
 %
-Although \CEU supports asynchronous active execution~\cite{ceu.tr}, we still 
-consider it as purely reactive since its synchronous core cannot express active 
-behavior.
+Although \CEU supports active \emph{asynchronous} execution~\cite{ceu.tr}, its 
+synchronous core is still purely reactive.
 
@@
 
-Esterel makes no semantic distinctions between internal and external signals.
+In Esterel, the behavior of internal and external signals is equivalent.
 %In particular, programs can emit external and internal signals simultaneously, 
@@
 %
-In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
-events become active, resulting in the invocation of \code{f()} and \code{g()} 
-in no particular order (ln. 6,9).
-%
-In \CEU, the occurrence of \code{A} makes the program behave as follows:
+In Esterel, when \code{A} occurs, the program emits \code{B} (ln.  4--5) and 
+both events become active, resulting in the invocation of \code{f()} and 
+\code{g()} in no particular order (ln. 6,9).
+%
+In \CEU, when \code{A} occurs, the program behave as follows:
 %
@@
 
-Internal events provide a limited form of subroutines, as depicted in 
-Figure~\ref{lst.sub}.
-The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
+Internal events provide fine-grained execution control and can express a 
+limited form of subroutines, as depicted in Figure~\ref{lst.sub}.
+The ``subroutine'' \code{inc} is defined as a loop (ln. 3--6) that continuously 
 awaits its identifying event (ln. 4), incrementing the value passed as 
 reference (ln. 5).
-A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
-\code{A} through an \code{emit} (ln. 10).
+A trail in parallel (ln. 8--11) invokes the subroutine through an \code{emit 
+inc} (ln. 10) in reaction to some code (ln. 9).
 Given the stacked execution for internal events, the calling trail pauses, the 
@@
 Only after this sequence, the calling trail resumes and passes the assertion 
-test (ln. 11).
+test (ln. 10--11).
 
 \CEU also supports nested \code{emit} invocations for internal events.
-For instance, the code in Figure~\ref{lst.sub} that executes after the 
-\code{await inc} (ln.  4) could \code{emit} another event, creating a new level 
-in the stack.
-The runtime stack constitutes fine-grained micro reactions, one on top of the 
-other, all inside the same reaction to an external event.
+For instance, the body of the subroutine \code{inc} in Figure~\ref{lst.sub} 
+could \code{emit} another event after awaking (ln. 4), creating a new level in 
+the stack.
+The runtime stack constitutes fine-grained micro reactions, with one on top of 
+the other, all inside the same reaction to an external event.
 %Internal events provide a fine-grained intra-reaction control mechanism.
@@
     var int v = 1;
-    await A;
+    <...>
     emit inc => &v; // call `inc'
@@
 %
-In a previous work, we build on top of internal events other advanced control 
-mechanisms, such as resumable exceptions and reactive 
+In a previous work, we build other advanced control mechanisms on top of 
+internal events, such as resumable exceptions and reactive 
 variables~\cite{ceu.rem13}.
@@
 Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
-makes the enclosing \code{loop} non instantaneous (ln 2--9).
-However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
-same reaction that reaches them, the \code{par/or} would terminate and restart 
-the \code{loop} instantaneously, resulting in infinite execution.
+characterizes the enclosing \code{loop} as non instantaneous (ln 2--9).
+However, if the \code{emit e} (ln. 6) could awake the \code{await e} 
+instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
+\code{loop} instantaneously, resulting in infinite execution.
 %
@@
 In atypical scenarios requiring immediate awake, delayed awaits can be 
-circumvented by manually copying or transforming the code to execute on awake.
+circumvented by placing the code to execute before the \code{await}.
 %
@@
 However, the execution order for operations within a reaction is 
-non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
-|| call f2()>>}, the order is unspecified and it would be an error to rely on 
+non-deterministic: ``if there is no control dependency, as in
+\code{(call~f1()~||~call~f2())},
+the order is unspecified and it would be an error to rely on 
 it''~\cite{esterel.primer}.
 %
-A number of Esterel-based synchronous languages have intra-reaction determinism 
-by enforcing an arbitrary execution order for statements in multiple lines of 
-execution (%
+A number of Esterel-inspired synchronous languages enforces an arbitrary 
+execution order for statements in multiple lines of execution to achieve 
+intra-reaction determinism
+(%
 \emph{Reactive C}~\cite{rp.rc},                 % 1991
@@
 and
-\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
+\emph{PRET-C}~\cite{rp.pretc}%                  % 2010
+).
 %, e.g., based on lexical order or manual priority assignment.
@@
 %
-In the example \ax, the two assignments to \code{x} execute in reactions to 
-different events \code{A} and \code{B} only, which by definition, cannot occur 
-simultaneously (Section~\ref{sec.ceu.exts}).
+In the example \ax, the two assignments to \code{x} can only execute in 
+reactions to different events \code{A} and \code{B}, which cannot occur 
+simultaneously by definition (Section~\ref{sec.ceu.exts}).
 Hence, for the sequence \code{A->B}, \code{x} becomes \code{4} 
@@
 native do
-    #define NUM 10
+    ##define NUM 10
     void f  (void)  { <...> }
@@
 %
-Since \CEU does not inspect any code in \C, it opts to complain about 
-suspicious concurrent accesses between \code{\_f} and all symbols in the second 
-trail.
+Since \CEU does not inspect any code in \C, it complains about suspicious 
+concurrent accesses between \code{\_f} and all symbols in the second trail.
 
@@
 
-The annotations in Figure~\ref{lst.c}.b provides hints to the compiler about 
-the semantics of the \C symbols in code \ax, which now compiles successfully:
+The annotations in Figure~\ref{lst.c}.b provide hints to the compiler about the 
+semantics of the \C symbols in program \ax, which now compiles successfully:
 %
@@
 In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
-conflicting pixels, which makes them non commutative and prevent the code to 
-compile.
-%
-However, we want to rely on lexical order to always redraw the 
+conflicting pixels, which makes the calls non commutative and prevent the code 
+to compile.
+%
+However, in this case we want to rely on lexical order to always redraw the 
 \code{background} object before the \code{foreground} object.
@@
 Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} to an 
-``intimidating'' name \code{\_redraw\_non\_commutative} (ln. 2, to be explicit 
+``intimidating name'' \code{\_redraw\_non\_commutative} (ln. 2, to be explicit 
 about its effect), and annotate it as safe (ln. 4--5) to make the code compile 
@@
 %
-However, aborting lines of execution that deal with external resources is still 
-error prone because they may end up in an inconsistent state.
+However, aborting lines of execution that deal with external resources may lead 
+to inconsistencies.
+%is still error prone because they may end up in an inconsistent state.
 %
@@
 represents a \textbf{local} resource that requires finalization.
-\item If \CEU \textbf{receives} a pointer from a system call, the pointer 
-represents an \textbf{external} resource that requires finalization.
+\item If \CEU \textbf{receives} a pointer from a system call return, the 
+pointer represents an \textbf{external} resource that requires finalization.
 \end{itemize}
@@
 %
-However, the interaction between system clocks and programs is not 100\% 
+However, the interaction between system clocks and programs is not absolutely
 precise, a fact that is usually ignored in the development process.
@@
 In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
-\code{await} request, the underlying system gets busy and takes 15ms to check 
-for expiring awaits.
-The \CEU scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
+\code{await} request, the underlying system gets busy and takes 15ms to notify 
+\CEU.
+The scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
 already expired, but is delayed with \code{delta=5ms}.
@@
 %
-Therefore, a similar implementation for Esterel would not rely on the tick 
-notion of simultaneity either.
+For instance, a similar implementation of this specification in Esterel would 
+not rely on the tick notion of simultaneity either.
 %
@@
 Although execution is deterministic, the compiler of \CEU still advises about 
-suspicious statements that can react concurrently to the same event, 
-appropriate serialization.
+suspicious statements that can react concurrently to the same event. 
+%appropriate serialization.
 %
 \item \emph{Safe integration with \C.}
-\CEU enforces the programmer to deal with pointers representing resources 
-through finalization clauses.
-When dealing with concurrent system calls, programmers can provide the compiler 
-with annotations to remove false positives from the temporal analysis or to 
-allow non commutative concurrent behavior.
+When dealing with concurrent system calls, programmers can provide annotations 
+to improve the temporal analysis results, or to enforce non commutative 
+concurrent behavior.
+\CEU also requires finalization clauses to handle pointers representing 
+resources.
 %
@@
 %
-To deal with the limitations above (which is not in the scope of this paper), 
-we recommend for memory-isolated parallelizable asynchronous primitives as 
-separate extensions to the synchronous core~\cite{esterel.crp}.
+To deal with the limitations above, we recommend for memory-isolated 
+parallelizable asynchronous primitives as separate extensions to the 
+synchronous core~\cite{esterel.crp} (which are in the scope of this paper).
 
@@
 
-\begin{figure*}[t]
+\begin{figure*}[b]
 \includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}

### 2016-04-11 paper.tex
...

@@
 %
-We evaluated the expressiveness of \CEU in comparison to hand-crafted 
-event-driven code in \C and attested a reduction in source code size (around 
-25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
+We evaluated the expressiveness of \CEU in comparison to event-driven code in 
+\C and attested a reduction in source code size (around 25\%) with a small 
+increase in memory usage (around 5--10\% for \emph{text} and 
 \emph{data})~\cite{ceu.sensys13}.
@@
 In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
-in comparison to a hand-crafted event-driven code base%
+in comparison to a code base%
 \footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
@@
 
-In the subsections that follow, we discuss \CEU-specific implementation 
-details:
+In the subsections that follow, we discuss implementation details specific to 
+\CEU:
 temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
@@
 For each node representing a statement in the program AST, we keep the set of 
-input events $I$ (\emph{incoming}) that can lead to the execution of the node, 
+input events $I$ (\emph{incoming}) that can start the execution of the node, 
 and also the set of input events $O$ (\emph{outgoing}) that can terminate the 
@@
 \item A \emph{sequence node (;)} modifies each of its children to have 
-$I_n=O_{n-1}$, except for $n=1$ which inherits $I$ from the parent node.
+$I_n=O_{n-1}$, except for $n=1$ (which inherits $I$ from the parent node).
 The set $O$ for the whole node is copied from its last child, i.e., $O=O_n$.
@@
 \item An \code{if} node has $O=O_{true} \cup O_{false}$, where \code{true} and 
-\code{false} are the two clauses of the \code{if}.
+\code{false} are the two \code{if} branches.
 %
@@
 $I$ and $O$ in Figure~\ref{lst.impl.ast}.b.
-The event $.$ (dot) represents the ``boot'' reaction.
+%The event $.$ (dot) represents the ``boot'' reaction.
 The assignments to \code{y} in parallel (ln. \ax:5,8) have an empty 
@@
 \begin{lstlisting}[numbers=left,xleftmargin=2.5em]
-Stmts I={.} O={A}
-    Dcl_y I={.} O={.}
-    ParOr I={.} O={A,B}
-        Stmts I={.} O={A}
-            Await_A I={.} O={A}
+Stmts I={boot} O={A}
+    Dcl_y I={boot} O={boot}
+    ParOr I={boot} O={A,B}
+        Stmts I={boot} O={A}
+            Await_A I={boot} O={A}
             Set_y I={A} O={A}
-        Stmts I={.} O={B}
-            Await_B I={.} O={B}
+        Stmts I={boot} O={B}
+            Await_B I={boot} O={B}
             Set_y I={B} O={B}
@@
 %
-Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
-for blocks in sequence are packed in a \code{union}, while blocks in parallel, 
-in a \code{struct}.
+Translating this idea to \C is straightforward~\cite{wsn.osm}: memory for 
+blocks in sequence are packed in a \code{union}, while blocks in parallel, in a 
+\code{struct}.
 %In the example, the code following the loop (identified as \code{...}) reuses 
@@
 %
-Each variable is assigned a unique $id$ so that variables with the same name 
-can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
-\ax:3).
+%Each variable is assigned a unique $id$ so that variables with the same name 
+%can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
+%\ax:3).
 %
@@
 (ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
-given that their variables cannot be in scope at the same time, e.g., 
-\code{MEM.a\_1} and \code{MEM.b\_2} (ln.  \bx:2,3) can safely share the same 
-memory slot.
+given that their variables cannot be in scope at the same time, e.g., \code{a} 
+and \code{b} can safely share the same memory slot.
 %
@@
 union {             // sequence
-    int a_1;        //   do_1
-    int b_2;        //   do_2
+    int a;          //   do_1
+    int b;          //   do_2
     struct {        //   par/and
-        int _and_3: 1;
-        int _and_4: 1;
+        int _and_1: 1;
+        int _and_2: 1;
     };
@@
 A \code{STACKED} trail holds an associated numeric stack level and can only 
-execute when scheduler runtime reaches that level.
+execute when scheduler runtime drops to that level.
 %
@@
 %
-A trail waiting for an event stays in the state of the corresponding event, 
-also holding the minimum sequence number (\emph{seqno}) in which it can awake 
-(for \emph{delayed awaits}).
-
-to prevent from awaking in the same reaction it is reached.
+A trail waiting for an event stays in that event, also holding the minimum 
+sequence reaction number (\emph{seqno}) in which it can awake (to respect
+\emph{delayed awaits}).
 %
@@
 %
-Since the scheduler traverses all trails on every reaction, it adjusts all 
-\code{seqno} to avoid overflows (actually, 2 bits to hold the \code{seqno} is 
-already enough).
+The fields \code{seqno} requires only 2 bits because the scheduler adjusts them 
+while traversing all trails.
 %
@@
 emits an event that awakes another trail, and so on;
-the last trail cannot awake any trail, because they are all hanged in the 
+the last trail cannot awake any other trail, because they are all hanged in the 
 \code{STACKED} state.
@@
 For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
-simultaneous lines of execution with memory overhead of only 2\% in comparison 
-to the original single-threaded version in \C~\cite{ceu.sensys13}.
+simultaneous lines of execution but has a memory overhead of only 2\% in 
+comparison to the original single-threaded version in \C~\cite{ceu.sensys13}.
 
@@
 %
-For each statement that pauses (\code{emit} and \code{await}), resumes 
-(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
-and \code{break}), the compiler splits the trail in two segments with 
-associated entry points.
+For each yielding statement (e.g., \code{emit}, \code{await}, \code{par/and}, 
+etc.), the compiler splits the trail in two segments with associated entry 
+points.
 %
@@
 %
-For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
-%
-When the \code{Main} label reaches the \code{par/and} (ln. \ax:4), it stacks 
-\code{TRAIL-1} (ln. \cx:4--7) and proceeds to the code in \code{TRAIL-0} (ln.  
-\cx:9--14) to respect lexical execution order.
-%
-The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
+For the first reaction, it executes the \code{Main} label in \code{TRAIL-0} 
+(ln.~3--14).
+%
+When the \code{Main} label reaches the \code{par/and} (ln. \ax:4), it first 
+stacks \code{TRAIL-1} (ln. \cx:4--7) and then executes the \code{await~e} 
+(ln.~\ax:6) in \code{TRAIL-0} (ln.~\cx:9--14), respecting lexical execution 
+order.
+%
+The dispatcher sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
 \code{Awake\_e}, and then halts with a \code{break}.
 %
-Then, \code{dispatch} takes \code{TRAIL-1} and executes its registered label 
-\code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to await 
-\code{EVT\_A} and also halts.
+Then, it switches to \code{TRAIL-1} and executes label \code{And\_sub\_2} 
+(ln.~\cx:6,16--21), which sets \code{TRAIL-1} to await \code{EVT\_A} and also 
+halts.
 
@@
 %
-This decision comes not only from the fact that \CEU targets constrained 
-embedded systems with a single CPU, but also because \CEU imposes deterministic 
+This decision comes not only from the fact that \CEU targets constrained with 
+single-CPU embedded systems, but also because \CEU imposes deterministic 
 execution for intra-reaction statements.
@@
 Note that, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
-analysis of \CEU infers precisely trails that are concurrent yet do not share 
-resources.
+analysis of \CEU infers precisely trails that are concurrent and yet do not 
+share resources.
 Hence, these non-conflicting trails could potentially execute with real 
@@
 work~\cite{esterel.multicore.1,esterel.multicore.2}.
-One reason is due to continuous fork-and-rejoin overhead from small reactions.
-Another reason is due to excessive locality of data from stackless trails 
-sharing contiguous static memory.
+One reason is due to overhead from continuous fork-and-rejoin in small 
+reactions.
+Another reason is due to contention form excessive locality of data in 
+stackless trails sharing contiguous static memory.
 
@@
 %
-However, to preserve the sequential/discrete semantics of time in \CEU, these 
-calls must never interleave or parallelize execution.
+These calls must never interleave or parallelize execution in order to preserve 
+the sequential/discrete notion of time in \CEU.
 
@@
 \emph{code generation phase}, which here outputs assembly instructions for the 
-\VM, instead of statements in \C.
+\VM (instead of statements in \C).
 %

### 2016-04-13 paper.tex
...

@@
 %
-For the second phrase, the reset behavior, the Esterel version uses a 
-\code{abort-when} (ln. 4--11) which, in this case, serves the same purpose of 
-\CEU's \code{par/or} (ln. 4--13):
+For the second phrase, the reset behavior, the Esterel version uses an 
+\code{abort-when} statement (ln. 4--11) which, in this case, serves the same 
+purpose of \CEU's \code{par/or} (ln. 4--13):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
@@
 %
-A program must react to an occurring event completely before handling the next.
+A program must react to an occurring event completely before handling the next 
+one.
 %
@@
 In Section~\ref{sec.ceu.timers}, after introducing internal events and 
-first-class timers, we show an example to detect simultaneous button clicks in 
-\CEU.
+first-class timers, we show an example to illustrate how to detect simultaneous 
+button clicks in \CEU.
 %
@@
 analysis to infer appropriate tick lengths~\cite{esterel.fixed}.
-% TODO-no: Such analsysis is orthogonal
+% TODO-no: Such analysis is orthogonal
 
@@
 Tick-triggered execution allows for active behavior, since code can execute 
-regulary on every tick.
+regularly on every tick.
 %
@@
 %
-In a previous work, we build other advanced control mechanisms on top of 
-internal events, such as resumable exceptions and reactive 
-variables~\cite{ceu.rem13}.
+In previous work, we build other advanced control mechanisms on top of internal 
+events, such as resumable exceptions and reactive variables~\cite{ceu.rem13}.
 %In Section~\ref{sec.adv.excpt} we show how to use them to implement 
@@
 the program.
-% TODO-no: static analsysis also simpler
+% TODO-no: static analysis also simpler
 % highlight that is not only important to have a comptutally-sound algorithm, 
@@
 time as the \emph{residual delta time (delta)}.
-Without explicit manipulation, the recurrent use of timed activities in 
-sequence (or in a loop) may accumulate a considerable amount of deltas that can 
-lead to incorrect behavior in programs.
+Without explicit manipulation, the recurrent use of timed activities in a row 
+(or in a loop) may accumulate a considerable amount of deltas that can lead to 
+incorrect behavior in programs.
 
@@
 %
-Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
+Figure~\ref{lst.simultaneous} emulates a \code{middle\_click} event (ln. 3) in 
 terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
@@
 %
-The list that follows summarizes the contributions of our design in this 
+The following list summarizes the contributions of our design in this 
 direction:
@@
 \item \emph{Static temporal analysis for concurrent statements.}
-Although execution is deterministic, the compiler of \CEU still advises about 
+Although execution is deterministic, the \CEU compiler still advises about 
 suspicious statements that can react concurrently to the same event. 
@@
 When dealing with concurrent system calls, programmers can provide annotations 
-to improve the temporal analysis results, or to enforce non commutative 
+to improve the temporal analysis results, or to force non commutative 
 concurrent behavior.
@@
 %
-Nonetheless, we advocate for keeping a tractable synchronous reactive core with 
+Nonetheless, we advocate keeping a tractable synchronous reactive core with 
 support for shared memory concurrency and deterministic execution.
 %
-To deal with the limitations above, we recommend for memory-isolated 
-parallelizable asynchronous primitives as separate extensions to the 
-synchronous core~\cite{esterel.crp} (which are in the scope of this paper).
+To deal with the limitations above, we recommend memory-isolated parallelizable 
+asynchronous primitives as separate extensions to the synchronous 
+core~\cite{esterel.crp} (which are in the scope of this paper).
 
@@
 
-In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
+In previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
 in comparison to a code base%
@@
 
-\CEU favors a fine-grained use of trails, being common to use trails that await 
+\CEU favors a fine-grained use of trails: it is common to use trails that await 
 a single event and terminate.
@@
 work~\cite{esterel.multicore.1,esterel.multicore.2}.
-One reason is due to overhead from continuous fork-and-rejoin in small 
-reactions.
-Another reason is due to contention form excessive locality of data in 
-stackless trails sharing contiguous static memory.
-
-If we consider data-dominated applications, multi-core implementations can 
-offer considerable speed-up gains.
+One reason is the overhead from continuous fork-and-rejoin in small reactions.
+Another reason is contention from excessive locality of data in stackless 
+trails sharing contiguous static memory.
+
+If we consider data-intensive applications, multi-core implementations can 
+offer considerable speedups.
 However, data-intensive computations do not typically require a disciplined 
@@
 Single-threaded dispatching may not be suitable for hard real-time activities.
-In a previous work~\cite{ceu.sensys13}, we measure how synchronous lengthy 
+In previous work~\cite{ceu.sensys13}, we measure how synchronous lengthy 
 computations in \C (e.g., hashing and compression) can block the scheduler and 
@@
 In addition, \CEU also tracks pointers representing resources that cross \C 
-boundaries and
-enforce the programmer to provide associated finalizers.
-
-%\emph{URBI}~\cite{rp.urbi} is a dynamic reactive scripting language with a 
-%rich set of control constructs for time management, event-driven 
-%communication, and concurrency.
-%Concurrency is based on dynamically created stackful coroutines, diverging 
-%from our goals regarding resource efficiency and static bounds for memory and 
-%execution time.
+boundaries and forces the programmer to provide associated finalizers.
+
+ReactiveML~\cite{rml} and \emph{URBI}~\cite{rp.urbi} extend the synchronous 
+model with dynamic lines of execution.
+%
+The implementations use coroutines or CPS transformations and rely on heap 
+allocation and/or garbage collection, diverging from our goals regarding 
+resource efficiency and static bounds for memory and execution time.
+%
+%These languages and may not be suitable for constrained embedded systems.
+%
+We discuss dynamic abstractions in \CEU in previous work~\cite{ceu.mod15}.
 
@@
 
-For the past three years, we have been teaching \CEU for undergraduate and 
-graduate students in research projects and two hands-on courses on 
-\emph{distributed systems} and \emph{reactive programming}.
-%, in which the studnets designing protocols, embedded systems, and graphical 
+For the past three years, we have been teaching \CEU to undergraduate and 
+graduate students in courses on \emph{distributed systems} and \emph{reactive 
+programming}.
+%, in which the students designing protocols, embedded systems, and graphical 
 %applications.

### 2016-09-19 paper.tex
...
(>) temporal analysis -> concurrency analysis

@@
 %
-\CEU uses an event-triggered notion of time that enables compile-time temporal 
+\CEU uses an event-triggered notion of time that enables compile-time concurrency 
 analysis to detect conflicting statements, resulting in deterministic and 
@@
 Using Esterel as a base, we present the particularities of our design, such as 
-stack-based internal events, the temporal analysis, safe integration with \C, 
+stack-based internal events, the concurrency analysis, safe integration with \C, 
 and first-class timers.
@@
 \CEU has a simple semantics with fine-grained control for intra-reaction 
-execution, and is amenable to a temporal analysis that
+execution, and is amenable to a concurrency analysis that
 improves safety.
@@
 %
-\item Static temporal analysis to detect suspicious concurrent statements.
+\item Static concurrency analysis to detect suspicious concurrent statements.
 %(Section~\ref{sec.ceu.analysis}).
@@
     (\ref{sec.ceu.ints});
-Static temporal analysis for concurrent statements
+Static concurrency analysis
     (\ref{sec.ceu.analysis});
@@
 Automatic mutual exclusion simplifies the reasoning about concurrency and is a 
-prerequisite for the temporal analysis to be discussed in 
+prerequisite for the concurrency analysis to be discussed in 
 Section~\ref{sec.ceu.analysis}.
@@
 
-\subsection{ Static Temporal Analysis for Concurrent Statements }
+\subsection{ Static Concurrency Analysis }
 \label{sec.ceu.analysis}
@@
 
-To mitigate this threat, \CEU performs a temporal analysis at compile time to 
+To mitigate this threat, \CEU performs a concurrency analysis at compile time to 
 detect concurrent accesses to shared variables:
@@
 
-\subsubsection{Temporal Analysis}
-
-As a safety measure, the temporal analysis of Section~\ref{sec.ceu.analysis} 
+\subsubsection{Concurrency Analysis}
+
+As a safety measure, the concurrency analysis of Section~\ref{sec.ceu.analysis} 
 also considers concurrent calls and accesses to external symbols in \C.
@@
 In addition, the uniqueness of external events is a prerequisite for the 
-temporal analysis of \CEU.
+concurrency analysis of \CEU.
 %
@@
 %
-\item \emph{Static temporal analysis for concurrent statements.}
+\item \emph{Static concurrency analysis.}
 Although execution is deterministic, the \CEU compiler still advises about 
@@
 When dealing with concurrent system calls, programmers can provide annotations 
-to improve the temporal analysis results, or to force non commutative 
+to improve the static analysis results, or to force non commutative 
 concurrent behavior.
@@
 tree (\emph{AST});
-the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
+the \emph{concurrency analysis phase} detects inconsistencies in programs, such as 
 unbounded loops and suspicious concurrent statements;
@@
 \CEU:
-temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
+concurrency analysis for determinism (Section~\ref{sec.impl.analysis}),
 static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
@@
 
-\subsection{Temporal Analysis for Concurrent Statements}
+\subsection{Concurrency Analysis}
 \label{sec.impl.analysis}
 
-The compile-time temporal analysis phase detects inconsistencies in \CEU 
+The compile-time concurrency analysis phase detects inconsistencies in \CEU 
 programs.
@@
 
-Note that, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
+Note that, as discussed in Section~\ref{sec.ceu.analysis}, the concurrency 
 analysis of \CEU infers precisely trails that are concurrent and yet do not 
@@
 As far as we know, \CEU is the first to encrust this event-triggered notion of 
-time in the core of the language, which is a prerequisite for the temporal 
+time in the core of the language, which is a prerequisite for the concurrency 
 analysis that enables safe shared-memory concurrency.
-
+%
 Another unique aspect of \CEU is the distinction between external and internal 
@@
 %
-\CEU goes one step further and performs a temporal analysis to detect trails 
+\CEU goes one step further and performs a concurrency analysis to detect trails 
 that, when reordered, change the observable behavior of the program, i.e., 
@@
 %These languages have a tick-based notion of time similar to Esterel, which 
-%prevents the temporal analysis of \CEU.
+%prevents the concurrency analysis of \CEU.
 %

### 2016-09-19 paper.tex
(>) referee 3: minor details

@@
 \usepackage{color}
-\definecolor{light}{gray}{0.87}
+\definecolor{light}{gray}{0.97}
 \definecolor{dark}{gray}{0.30}
@@
 %
-Considering the control-based languages, Esterel~\cite{esterel.ieee91} was the 
+Of the control-based languages, Esterel~\cite{esterel.ieee91} was the 
 first to appear and succeed, influencing a number of embedded languages, such 
@@
 Despite its success and influence, Esterel has an overly complex semantics that 
-requires careful static analysis such as to detect and refuse programs with 
+requires careful static analysis to detect and refuse programs with 
 \emph{causality} and \emph{schizophrenia} problems%
@@
 Above all, it also affects the programmer's understanding about the code, 
-which, ultimately, has to solve the errors when facing corner cases.
+who, ultimately, has to solve the errors when facing corner cases.
 %
@@
 non-deterministic execution for intra-reaction statements.
-Non determinism prevents threads to interact with stateful system calls safely 
+Non determinism prevents threads from interacting with stateful system calls safely 
 and makes shared-memory concurrency not as straightforward as reading and 
@@
 %
-\item Safe integration with \C, which enforces finalization for external 
+\item Safe integration with \C that enforces finalization for external 
 resources.
@@
 \code{abort-when} statement (ln. 4--11) which, in this case, serves the same 
-purpose of \CEU's \code{par/or} (ln. 4--13):
+purpose as \CEU's \code{par/or} (ln. 4--13):
 the occurrence of event \code{R} aborts the awaiting statements in parallel and 
@@
 
-In the subsections that follow, we discuss the main differences from \CEU to 
+In the subsections that follow, we discuss the main differences between \CEU and
 Esterel:
@@
 We present the formal specification of the semantics of \CEU in a separate 
-work~\cite{ceu.phd}.
+paper~\cite{ceu.phd}.
 
@@
 %
-A program must react to an occurring event completely before handling the next 
+A program must react to an event completely before handling the next 
 one.
@@
 In practice, if a new external input event occurs while a reaction chain is 
-running, it is enqueued to occur in the next reaction.
+running, it is enqueued to occur in a subsequent reaction.
 
@@
 \emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
-This approach avoids the time violation for \code{R(A)} and also results in
+This approach avoids the timing violation for \code{R(A)} and also results in
 smaller idle periods because it adjusts the tick lengths to match the CPU times 
@@
 \CEU to be roughly the same.
-This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
+This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
 %
@@
 Reactions to multiple events in \CEU never overlap because they are atomic.
-Automatic mutual exclusion simplifies the reasoning about concurrency and is a 
+Automatic mutual exclusion simplifies reasoning about concurrency and is a 
 prerequisite for the concurrency analysis to be discussed in 
@@
 the more inputs to handle, the longer the reaction takes, and the more inputs 
-accumulates for subsequent ticks.
+accumulate for subsequent ticks.
 %
@@
 %
-In \CEU, when \code{A} occurs, the program behave as follows:
+In \CEU, when \code{A} occurs, the program behaves as follows:
 %
@@
 The ``subroutine'' \code{inc} is defined as a loop (ln. 3--6) that continuously 
-awaits its identifying event (ln. 4), incrementing the value passed as 
+awaits its identifying event (ln. 4), incrementing the value passed by
 reference (ln. 5).
@@
 
-On the one hand, this form of subroutines has a significant limitation that it 
+On the one hand, this form of subroutine has a significant limitation that it 
 cannot express recursive calls: an \code{emit} to itself is always ignored, 
@@
 %
-Also, this form of subroutines can use the other primitives of \CEU, such as 
+Also, this form of subroutine can use the other primitives of \CEU, such as 
 parallel compositions and the \code{await} statement.
@@
 communication~\cite{esterel.compiling}, i.e., two threads can react to one 
-another during the same cycle due to mutual signal dependency.
-%
-\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
+another during the same cycle due to mutual signal dependencies.
+%
+\CEU takes a different approach, posing a tradeoff that an \code{await} is only valid 
 for the next reaction, i.e., if an \code{await} and \code{emit} occur 
@@
 %
-A number of Esterel-inspired synchronous languages enforces an arbitrary 
+A number of Esterel-inspired synchronous languages enforce an arbitrary 
 execution order for statements in multiple lines of execution to achieve 
@@
 
-In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
+In \CEU, any identifier prefixed with an underscore is passed unchanged to 
 the \C compiler that generates the final binary.
@@
 In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
-conflicting pixels, which makes the calls non commutative and prevent the code 
+conflicting pixels, which makes the calls non commutative and prevents the code 
 to compile.
@@
 %
-In Esterel and \CEU, the finalize clause (ln. \ax:10 and \bx:5) execute
+In Esterel and \CEU, the finalize clause (ln. \ax:10 and \bx:5) executes
 automatically if the enclosing block (ln. \ax:3--1 and \bx:3--10) is
@@
 
-\CEU goes one step further and actually enforces the use of \code{finalize} for 
+\CEU goes one step further and enforces the use of \code{finalize} for 
 system calls that deal with pointers representing resources:
@@
 asynchronous primitives as separate extensions to the synchronous 
-core~\cite{esterel.crp} (which are in the scope of this paper).
+core~\cite{esterel.crp} (which are not in the scope of this paper).
 
@@
 the \emph{code generation phase} converts the \emph{AST} to standard \C code 
-and packs it with platform-dependent functionality (e.g., system calls) and the 
+and augments it with platform-dependent functionality (e.g., system calls) and the 
 runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
@@
 Unfortunately, most real applications in Esterel seem to be closed source.
-In a future work, we plan to cooperate with groups that have more experience 
+In future work, we plan to cooperate with groups that have more experience 
 and access to real code bases in Esterel for a comprehensive comparison with 
@@
 Translating this idea to \C is straightforward~\cite{wsn.osm}: memory for 
-blocks in sequence are packed in a \code{union}, while blocks in parallel, in a 
-\code{struct}.
+blocks in sequence are packed into a \code{union}, while blocks in parallel
+are packed into a \code{struct}.
 %In the example, the code following the loop (identified as \code{...}) reuses 
@@
 %
-A \code{FINALIZE} trail represents a hanged finalization block which is
+A \code{FINALIZE} trail represents a pending finalization block which is
 scheduled only when its corresponding block goes out of scope.
@@
 
-The size of \code{state\_t} depends on the number of events in the application;
+The size of \code{state\_t} depends on the number of events in an application;
 for an application with less than 253 events (plus the 3 states), one byte is 
@@
 The size of \code{label\_t} depends primarily on the number of \code{await} 
-statements in the application---each \code{await} splits the code in two 
+statements in the application---each \code{await} splits the code into two 
 segments and requires a unique entry point in the code for its continuation.
@@
 The size of \code{stack\_t} depends on the maximum depth of nested emissions 
-but is bounded to the maximum number of trails:
+but is bounded by the maximum number of trails:
 in the worst case, a trail emits an event that awakes another trail, which 
 emits an event that awakes another trail, and so on;
-the last trail cannot awake any other trail, because they are all hanged in the 
+the last trail cannot awake any other trail, because they are all blocked in the 
 \code{STACKED} state.
@@
 For each yielding statement (e.g., \code{emit}, \code{await}, \code{par/and}, 
-etc.), the compiler splits the trail in two segments with associated entry 
+etc.), the compiler splits the trail into two segments with associated entry 
 points.
@@
 %
-This decision comes not only from the fact that \CEU targets constrained with 
+This decision comes not only from the fact that \CEU targets constrained 
 single-CPU embedded systems, but also because \CEU imposes deterministic 
@@
 %
-The scheduler of TinyOS is already synchronous by default and always execute 
+The scheduler of TinyOS is already synchronous by default and always executes 
 \code{event} handlers atomically, hence, the API calls to \CEU are properly 
@@
 %
-Considering the domain of sensor networks, Terra already provides components 
+In the domain of sensor networks, Terra already provides components 
 organized in four areas: radio communication, group management, data 
@@
 \CEU also guarantees deterministic behavior for timer compositions by adjusting 
-inaccuracies in the system clock.
+for inaccuracies in the system clock.
 

### 2016-09-24 paper.tex
(>) referee 1,3: concurrency analysis -> concurrency checks

@@
 %
-\CEU uses an event-triggered notion of time that enables compile-time concurrency 
-analysis to detect conflicting statements, resulting in deterministic and 
+\CEU uses an event-triggered notion of time that enables compile-time checks to 
+detect conflicting concurrent statements, resulting in deterministic and 
 concurrency-safe programs.
@@
 Using Esterel as a base, we present the particularities of our design, such as 
-stack-based internal events, the concurrency analysis, safe integration with \C, 
+stack-based internal events, concurrency detection, safe integration with \C, 
 and first-class timers.
@@
 \CEU has a simple semantics with fine-grained control for intra-reaction 
-execution, and is amenable to a concurrency analysis that
-improves safety.
+execution, and is amenable to concurrency checks that improve safety.
 %
@@
 %
-\item Static concurrency analysis to detect suspicious concurrent statements.
-%(Section~\ref{sec.ceu.analysis}).
+\item Static concurrency checks to detect suspicious concurrent statements.
+%(Section~\ref{sec.ceu.checks}).
 %
@@
     (\ref{sec.ceu.ints});
-Static concurrency analysis
-    (\ref{sec.ceu.analysis});
+Static concurrency checks
+    (\ref{sec.ceu.checks});
 Safe integration with \C
@@
 Automatic mutual exclusion simplifies reasoning about concurrency and is a 
-prerequisite for the concurrency analysis to be discussed in 
-Section~\ref{sec.ceu.analysis}.
+prerequisite for the concurrency checks to be discussed in 
+Section~\ref{sec.ceu.checks}.
 \end{itemize}
@@
 
-\subsection{ Static Concurrency Analysis }
-\label{sec.ceu.analysis}
+\subsection{ Static Concurrency Checks }
+\label{sec.ceu.checks}
 
@@
 
-To mitigate this threat, \CEU performs a concurrency analysis at compile time to 
-detect concurrent accesses to shared variables:
+To mitigate this threat, \CEU performs concurrency checks at compile time to 
+detect conflicting accesses to shared variables:
 if a variable is written in a trail segment, then a concurrent trail segment 
@@
 %
-The algorithm for the analysis, which is depicted in 
-Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
-that precede a variable access and keep a list with all corresponding awaking 
-events.
+The detection algorithm, which is depicted in Section~\ref{sec.impl.checks},
+inspects all possible \code{await} statements that precede a variable access
+and keeps a list with all corresponding awaking events.
 Then, it checks all accesses in parallel trails to see if they share an awaking 
@@
 %
-Note also that the analysis is optional and does not affect the semantics of 
-the program.
+Note also that the static checks are optional and do not affect the semantics
+of the program.
 % TODO-no: static analysis also simpler
@@
 
-\subsubsection{Concurrency Analysis}
-
-As a safety measure, the concurrency analysis of Section~\ref{sec.ceu.analysis} 
-also considers concurrent calls and accesses to external symbols in \C.
+\subsubsection{Concurrency Checks}
+
+As a safety measure, the concurrency checks of Section~\ref{sec.ceu.checks} 
+also consider concurrent calls and accesses to external symbols in \C.
 %
@@
 In addition, the uniqueness of external events is a prerequisite for the 
-concurrency analysis of \CEU.
+concurrency checks of \CEU.
 %
@@
 %
-\item \emph{Static concurrency analysis.}
+\item \emph{Static concurrency checks.}
 Although execution is deterministic, the \CEU compiler still advises about 
@@
 When dealing with concurrent system calls, programmers can provide annotations 
-to improve the static analysis results, or to force non commutative 
+to reduce false positives in the static checks, or to force non commutative 
 concurrent behavior.
@@
 tree (\emph{AST});
-the \emph{concurrency analysis phase} detects inconsistencies in programs, such as 
+the \emph{concurrency checks phase} detects inconsistencies in programs, such as 
 unbounded loops and suspicious concurrent statements;
@@
 \begin{figure*}[b]
-\includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
+%\includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
 \caption{ Resource usage for \CEU and \emph{nesC} in the domain of sensor 
@@
 \CEU:
-concurrency analysis for determinism (Section~\ref{sec.impl.analysis}),
+concurrency checks for determinism (Section~\ref{sec.impl.checks}),
 static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
@@
 
-\subsection{Concurrency Analysis}
-\label{sec.impl.analysis}
-
-The compile-time concurrency analysis phase detects inconsistencies in \CEU 
+\subsection{Concurrency Checks}
+\label{sec.impl.checks}
+
+The compile-time concurrency checks phase detects inconsistencies in \CEU 
 programs.
@@
 such as accesses to shared variables, as discussed in 
-Section~\ref{sec.ceu.analysis}.
+Section~\ref{sec.ceu.checks}.
 
@@
 
-Note that, as discussed in Section~\ref{sec.ceu.analysis}, the concurrency 
-analysis of \CEU infers precisely trails that are concurrent and yet do not 
+Note that, as discussed in Section~\ref{sec.ceu.checks}, the concurrency 
+checks of \CEU infer precisely trails that are concurrent and yet do not 
 share resources.
@@
  \centering
- \includegraphics[width=0.5\columnwidth]{terra}
+ %\includegraphics[width=0.5\columnwidth]{terra}
 \caption{Terra programming system basic elements.}
@@
 time in the core of the language, which is a prerequisite for the concurrency 
-analysis that enables safe shared-memory concurrency.
+checks that enable safe shared-memory concurrency.
 %
@@
 %
-\CEU goes one step further and performs a concurrency analysis to detect trails 
+\CEU goes one step further and performs concurrency checks to detect trails 
 that, when reordered, change the observable behavior of the program, i.e., 
@@
 %These languages have a tick-based notion of time similar to Esterel, which 
-%prevents the concurrency analysis of \CEU.
+%prevents the concurrency checks of \CEU.
 %
@@
 
-\CEU is a concurrency-safe language, employing static analysis to ensure that 
+\CEU is a concurrency-safe language, employing static checks to ensure that 
 the high degree of concurrency in embedded systems does not pose safety threats 

### 2016-09-24 paper.tex
(*) TODO: data dependency, constructive semantics, internal events
(>) more neutral style

@@
 
-Despite its success and influence, Esterel has an overly complex semantics that 
+Despite its success and influence, Esterel has a complex semantics that
 requires careful static analysis to detect and refuse programs with 
@@
 %
-The complex semantics not only challenges the analysis and compilation of 
+A complex semantics not only challenges the analysis and compilation of 
 programs, but also results in incompatible and non-compliant implementations.
@@
 %
-Another drawback of the Esterel semantics consists of loose and 
-non-deterministic execution for intra-reaction statements.
+Another drawback of the Esterel semantics consists of non-deterministic
+execution for intra-reaction statements.
 Non determinism prevents threads from interacting with stateful system calls safely 
@@
 %
-\CEU has a simple semantics with fine-grained control for intra-reaction 
+\CEU proposes a semantics with fine-grained control for intra-reaction
 execution, and is amenable to concurrency checks that improve safety.
@@
 microcontroller with 32K of FLASH and 2K of SRAM.
-}, showing a practical aspect of our simple semantics.
+}, showing a practical aspect of the proposed semantics.
 %:\url{http://www.atmel.com/devices/atmega328.aspx}}
@@
 %
-\CEU takes a different approach, posing a tradeoff that an \code{await} is only valid 
-for the next reaction, i.e., if an \code{await} and \code{emit} occur 
+\CEU takes a different approach, posing a tradeoff that an \code{await} is only
+valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
 simultaneously in parallel trails, the \code{await} does not awake.
@@
 statements}~\cite{esterel.constructive}).
+
+TODO: mesmo sem mutual dependency, falar de semantica construtiva e lei da
+coerencia
 
@@
 
-Note that such analysis is only possible due to the uniqueness of input events 
-within reactions.
-Otherwise, any two trail segments in parallel could be concurrent, even if they 
+The uniqueness of input events within reactions makes this analysis possible,
+otherwise, any two trail segments in parallel could be concurrent, even if they 
 react to different input events.
 %
-Note also that the static checks are optional and do not affect the semantics
-of the program.
+Note that the static checks are optional and do not affect the semantics of the
+program.
 % TODO-no: static analysis also simpler
@@
 the \C compiler that generates the final binary.
-Therefore, access to \C is seamless and, more importantly, easily trackable.
+Therefore, access to \C is straightforward and syntactically trackable.
 %
@@
 
-\CEU goes one step further and enforces the use of \code{finalize} for 
-system calls that deal with pointers representing resources:
+\CEU also enforces the use of \code{finalize} for system calls that deal with
+pointers representing resources:
 %
@@
 
-\CEU favors a fine-grained use of trails: it is common to use trails that await 
-a single event and terminate.
+\CEU promotes a fine-grained use of trails: it is common to use trails that
+await a single event and terminate.
 For this reason, \CEU does not allocate per-trail stacks; instead, all data 
@@
 %
-Another unique aspect of \CEU is the distinction between external and internal 
-events.
+\CEU also the distinguishes external events from internal events.
 Internal events support stack-based micro reactions within external reactions, 
@@
 %
-\CEU goes one step further and performs concurrency checks to detect trails 
-that, when reordered, change the observable behavior of the program, i.e., 
-trails that actually rely on deterministic scheduling.
+\CEU also performs concurrency checks to detect trails that, when reordered,
+change the observable behavior of the program, i.e., trails that actually rely
+on deterministic scheduling.
 %
@@
 
+TODO: compilação mais sofisticada, tem que lidar com data dependency
+
 %\subsection{VM}

### 2016-09-26 paper.tex
(>) timers

@@
 }
-appear in typical patterns of embedded development, such as timeout watchdogs 
-and sensor samplings.
+appear in typical patterns of embedded development~\cite{esterel.delays}, such
+as timeout watchdogs and sensor samplings.
 %
@@
 automatically, resulting in more robust applications.
+Due to platform overheads and occasional slow reactions, an \code{await 10ms}
+suspends a line of execution for \emph{at least} 10 milliseconds.
 In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
@@
 
-\CEU also considers that time is a physical quantity that can be added and 
-compared.
+Delta compensation also implies that timers can be added and compared when
+reasoning about time as a physical quantity.
 For instance, in the example in Figure~\ref{lst.timers}.b, although the 
@@
 Given that any non-awaiting sequence is considered to take no time in the 
-synchronous model, the first trail (ln. 2--5) is guaranteed to terminate before 
-the second trail (ln. 7--8), because $10+1 < 12$.
+synchronous model, the first trail (ln. 2--5) is guaranteed to awake first and
+terminate the \code{par/or} before the second trail executes (ln. 7--8),
+because $10+1 < 12$.
 %
@@
 %prevents the concurrency checks of \CEU.
-%
-\CEU also guarantees deterministic behavior for timer compositions by adjusting 
-for inaccuracies in the system clock.
+
+\emph{Esterel+Delay}~\cite{esterel.delays} is a non-intrusive extension to
+Esterel to program in terms of physical time with \code{delay} statements.
+The global transformation relies on a \emph{platform statement} the specifies
+the available timers in the system (e.g., sample-driven or interrupt-driven).
+It then expands \code{delay} statements into existing Esterel statements in a
+way that the desired semantics can be realised in the platform.
+%
+\CEU makes physical time a special input event that feeds the runtime with an
+associated time to elapse which is decremented from all awaiting trails.
+The compensation scheme of \CEU guarantees that trails awake in the correct
+order and that errors are not propagated on subsequent awaits.
+Interrupt-driven timers are also supported with a hook callback that the
+runtime calls whenever the program awaits an earlier timer.
 

### 2016-09-27 paper.tex
(>) notion of time

@@
 of Esterel and in \CEU.
-The box \code{Real World} shows event occurrences over a continuous timeline.
+The box \code{Real World} assumes an external observer with an absolute
+reference clock that timestamps event occurrences over a
+continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
+$17ms521us$).
 The other boxes show how the same occurring events fit differently in each 
@@
 \emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
+%
+Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.
 %
@@
 %
-Finally, \CEU recognizes and reacts to the two instances of \code{E} 
-independently, which are handled in sequence.
-%
-Note that, by definition, each reaction corresponds to an event occurrence of 
-the same length.
+Finally, \CEU reacts to the two instances of \code{E} independently, which are
+handled in sequence.
+%
+%Note that, by definition, each reaction corresponds to an event occurrence of 
+%the same length.
 \end{itemize}
 
-We decided for the unique and queue-based semantics in \CEU for the reasons 
+Sample-driven execution maps directly to hardware implementations and
+involve no runtime queues.
+%
+However, considering the soft real-time application domain of \CEU, we
+decided for the unique and queue-based semantics in \CEU for the reasons
 that follow:
@@
 \begin{itemize}
-\item \textbf{A ``tick'' is too abstract and imprecise:}
-Outside the domain of hardware specification, a tick has no natural counterpart 
-in the real world.
-Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
-approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
-different behaviors for the same sequence of inputs.
+\item \textbf{A ``tick'' is implementation dependent:}
+Tick lengths are not part of the program specification.
+For instance, if we log the \code{Real World} box timeline and reproduce it in
+implementations with different tick lengths, the behaviors might diverge.
 %
 \item \textbf{Events are never absolutely simultaneous:}
-From a rigorous point of view, event occurrences are infinitesimal, having zero 
-probability of being simultaneous.
-This way, we consider that the notion of simultaneity should not be imposed by 
-the language, but defined explicitly for each use case.
-In the case of Esterel, simultaneity depends on the imprecise length of 
-discrete ticks.
+We consider that the notion of simultaneity should not be imposed by the
+language, but defined explicitly for each use case (to be discussed in
+Section~\ref{sec.ceu.timers}).
+%In Section~\ref{sec.ceu.timers}, after introducing internal events and 
+%first-class timers, we show an example to illustrate how to detect simultaneous 
+%button clicks in \CEU.
+In tick-based approaches, simultaneity depends on the length of discrete ticks.
 For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
-Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
-though \code{A} and \code{B} ``physically'' happen much closer to one another.
-In Section~\ref{sec.ceu.timers}, after introducing internal events and 
-first-class timers, we show an example to illustrate how to detect simultaneous 
-button clicks in \CEU.
+events \code{B} and \code{C} are simultaneous, even though \code{A} and
+\code{B} ``physically'' happen much closer to one another.
 %
 \item \textbf{Unique input events imply mutual exclusion:}
-Reactions to multiple events in \CEU never overlap because they are atomic.
+Reactions to exclusive events are atomic and never overlap.
 Automatic mutual exclusion simplifies reasoning about concurrency and is a 
@@
 \end{itemize}
+
+Note that Esterel supports declarations for mutually exclusive inputs which
+cannot be simultaneous in the same reaction
+(e.g., \code{relation A\#B\#C\#D\#E;})~\cite{esterel.primer}.
+Making all inputs mutually exclusive and adopting the variable-length-tick
+approach is equivalent to \CEU's notion of time, which can be seen as an
+imposed restriction on the semantics of Esterel.
 

### 2016-09-27 paper.tex
(>) feasibility of the synchronous hypothesis

@@
 cannot be simultaneous in the same reaction
-(e.g., \code{relation A\#B\#C\#D\#E;})~\cite{esterel.primer}.
+(e.g., ``\code{relation A\#B\#C\#D\#E;}'')~\cite{esterel.primer}.
 Making all inputs mutually exclusive and adopting the variable-length-tick
@@
 Otherwise, the application continuously accumulates delays between the real 
-occurrence and actual reaction of a given event.
+occurrence and actual reaction of a given event (we discuss implementation
+considerations in Sections~\ref{sec.impl.single}~and~\ref{sec.impl.environment}).
 %

### 2016-09-30 paper.tex
(>) quantitative comparison with Esterel

@@
 
-In previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
-in comparison to a code base%
-\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
-in \emph{nesC} (a \C variant)~\cite{wsn.nesc}.
-%
-Figure~\ref{fig.eval} compares source size (\emph{number of tokens}), binary 
-size (\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
-network protocols and a radio driver.
-The small overhead in resource usage shows that the gains in productivity and 
-safety with \CEU make it a viable alternative to \C in the context of 
-constrained embedded systems.
-%
-Unfortunately, most real applications in Esterel seem to be closed source.
-In future work, we plan to cooperate with groups that have more experience 
-and access to real code bases in Esterel for a comprehensive comparison with 
-\CEU.
-% TODO-no:
-%Still, considering \C as an ``absolute benchmark'', we consider the small 
-%overhead .  solid results
-
-\begin{figure*}[b]
-%\includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
-\caption{ Resource usage for \CEU and \emph{nesC} in the domain of sensor 
-networks.
-\label{fig.eval}
-}
-\end{figure*}
-
 In the subsections that follow, we discuss implementation details specific to 
@@
 
+In previous work~\cite{ceu.sensys13}, we evaluate our implementation by
+rewriting some components of the \emph{TinyOS} code base%
+\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
+from \emph{nesC} (a \C variant)~\cite{wsn.tos,wsn.nesc} to \CEU.
+%
+TinyOS provides open-source industrial-level applications and serve as basis
+for a considerable amount of research in the Wireless Sensor Network academic
+community.
+%
+The generated code has been tested in real hardware and also in congested
+networks simulated in sofware~\cite{ceu.sensys13}.
+%
+Figure~\ref{fig.eval} compares source size (\emph{number of tokens}), binary 
+size (\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
+network protocols and a radio driver.
+The small overhead in resource usage shows that the gains in productivity and 
+safety with \CEU make it a viable alternative to \C in the context of 
+constrained embedded systems.
+%
 Single-threaded dispatching may not be suitable for hard real-time activities.
-In previous work~\cite{ceu.sensys13}, we measure how synchronous lengthy 
+We also measure how synchronous lengthy 
 computations in \C (e.g., hashing and compression) can block the scheduler and 
@@
 analysis~\cite{esterel.wcrt,esterel.fixed}.
+
+In future work, we plan to perform an equivalent throughout analysis in
+comparison to industrial-quality code bases in Esterel.
+%
+An Esterel implementation that compiles to sequential C code with similar
+techniques also generates ``code less than $10\%$ larger the original
+manually-sequenced code'' for an industrial-size
+application~\cite{esterel.efficient}.
+%
+%Unfortunately, most real applications in Esterel seem to be closed source.
+%In future work, we plan to cooperate with groups that have more experience 
+%and access to real code bases in Esterel for a comprehensive comparison with 
+%\CEU.
+% TODO-no:
+%Still, considering \C as an ``absolute benchmark'', we consider the small 
+%overhead .  solid results
+
+\begin{figure*}[b]
+\includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
+\caption{ Resource usage for \CEU and \emph{nesC} in the domain of sensor 
+networks.
+\label{fig.eval}
+}
+\end{figure*}
 
@@
 also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
-Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
-the closest to our implementation with respect to trail allocation and 
-scheduling:
+Among the software-based approaches,
+\emph{SAXO--RT}~\cite{esterel.saxort,esterel.efficient} is the closest to our
+implementation with respect to trail allocation and scheduling:
 the compiler slices programs into ``control points'' (analogous to our ``entry 

### 2016-09-30 paper.tex
(>) overly complex semantics

@@
 A complex semantics not only challenges the analysis and compilation of 
-programs, but also results in incompatible and non-compliant implementations.
-%
-Above all, it also affects the programmer's understanding about the code, 
-who, ultimately, has to solve the errors when facing corner cases.
+programs, but also affects the programmer's understanding about the code, who,
+ultimately, has to solve the errors when facing corner cases.
 %

### 2016-09-30 paper.tex
(>) related work for stack-based internal events

@@
 checks that enable safe shared-memory concurrency.
-%
-\CEU also the distinguishes external events from internal events.
-Internal events support stack-based micro reactions within external reactions, 
-providing more fine-grained control for intra-reaction execution.
+
+In \CEU internal events support stack-based micro reactions within external
+reactions, providing more fine-grained control for intra-reaction execution.
+%
+Some variants of Statecharts~\cite{statecharts} also distinguish internal from
+external events~\cite{statecharts.variants}.
+%Some variants of Statecharts support local events which imply handling intra
+%reactions like Céu.
+In Statemate, ``reactions to external and internal events (...) can be sensed only
+after completion of the step''~\cite{statecharts.statemate}, implying queue-based
+execution.
+In Stateflow, ``the receiving state (of the event) acts here as a
+function''~\cite{statecharts.stateflow}, which is similar to Céu's stack-based
+execution.
+To avoid recursion, Céu adopts delayed awaits, while in Statemate, ``loops in
+the broadcasting of events (between states) are forbidden''.
 

### 2016-09-30 paper.tex
(>) instantaneous dialogue (>) shared-mem-concurrency

@@
 communication~\cite{esterel.compiling}, i.e., two threads can react to one 
-another during the same cycle due to mutual signal dependencies.
+another during the same cycle due to mutual signal dependencies
+(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
 %
@@
 %
+For this reason, Esterel, does not support shared-memory concurrency:
+\emph{``if a variable is written by some thread, then it can neither be read
+nor be written by concurrent threads''}~\cite{esterel.primer}.
+%
 A number of Esterel-inspired synchronous languages enforce an arbitrary 

### 2016-10-01 paper.tex
(>) soft real-time system

@@
 \CEU is a synchronous language inspired by Esterel with a simpler semantics and 
-more fine-grained control over program execution.
+more fine-grained control over program execution targeting soft real-time
+systems.
 %
@@
 %
-Another drawback of the Esterel semantics consists of non-deterministic
-execution for intra-reaction statements.
-Non determinism prevents threads from interacting with stateful system calls safely 
-and makes shared-memory concurrency not as straightforward as reading and 
-writing to shared variables.
+Furthermore, Esterel's semantics is non-deterministic for intra-reaction
+statements, which prevents threads from interacting with stateful system calls
+safely and makes shared-memory concurrency not as straightforward as reading
+and writing to shared variables.
 %
@@
 %
-\CEU proposes a semantics with fine-grained control for intra-reaction
-execution, and is amenable to concurrency checks that improve safety.
+\CEU proposes a semantics for soft real-time systems with fine-grained control
+for intra-reaction execution, which is amenable to concurrency checks that
+improve safety.
 %
@@
 packets~\cite{ceu.terra}.
+
+The semantics of \CEU implies an implementation with a runtime stack and a
+sequential scheduler.
+In contrast with Esterel's semantics, this prevents programs to be compiled
+directly and efficiently to hardware~\cite{esterel.constructive}.
+It also makes difficult to statically determine worst-case reaction times
+for hard real-time systems~\cite{esterel.fixed}.
 
@@
 % TODO-no: Such analysis is orthogonal
+%
+For soft real-time systems, accumulating delays and occasionally postponing
+reactions might be enough.
+However, hard real-time systems require a more robust approach such as
+determining fixed-length ticks that can be statically verified.
 
@@
 %
-We consider that this approach is more familiar to programmers and simplifies 
-the reasoning about concurrency.
-%
 As far as we know, \CEU is the first to encrust this event-triggered notion of 
@@
 
-In \CEU internal events support stack-based micro reactions within external
+In \CEU, internal events support stack-based micro reactions within external
 reactions, providing more fine-grained control for intra-reaction execution.
@@
 We present the design and implementation of \CEU, a synchronous reactive 
-language inspired by Esterel with simpler semantics and more fine-grained 
+language inspired by Esterel with event-driven semantics and more fine-grained 
 control for intra-reaction execution.
+%
+On the one hand, this approach is familiar to programmers in general,
+abstracting tick sampling with reactions to unique events from their
+application domain.
+On the other hand, this level of abstraction does not suit systems with
+hard-real time requirements in which interacting directly with a concrete
+notion of tick is more robust.
 

### 2016-10-01 paper.tex
(>) syntax

@@
 
+Figure~\ref{lst.syn} presents the subset of the concrete syntax of \CEU used in
+the paper as a quick reference.
+
+\begin{figure}
+\begin{lstlisting}
+// DECLARATIONS
+
+input  <type> <ids>;                        // external input  events
+output <type> <ids>;                        // external output events
+event  <type> <ids>;                        // internal events
+var    <type> <id> = <exp>;                 // a variable with initial value
+
+// EVENT HANDLING
+
+<id> = await <id>;                          // await an event and assign received value
+<id> = await <time>;                        // await time and assign delayed delta
+emit <id> => <exp>;                         // emit an event passing a value
+
+// CONTROL FLOW
+
+<stmt> ; <stmt>                             // sequence
+if <exp> then <stmts> else <stmts> end      // conditional
+loop do <stmts> end                         // repetition
+do <stmts> end                              // lexical block
+
+par/or  do <stmts> with <stmts> end         // aborts when one side terminates
+par/and do <stmts> with <stmts> end         // terminates when both sides terminate
+par     do <stmts> with <stmts> end         // never terminates
+
+finalize <stmts> with <stmts> end           // block finalization
+
+// INTEGRATION WITH C
+
+<_id>(<exps>)                               // C call (identifier starts with `_')
+native do <stmts> end                       // declarations in C
+native @const <_ids>                        // annotations: constant symbols
+native @pure  <_ids>                        //              pure functions
+native @safe  <_ids> with <_ids>            //              non-conflicting symbols
+\end{lstlisting}
+\caption{ Subset of the concrete syntax of \CEU used in the paper.
+\label{lst.syn}
+}
+\end{figure}
+
 In the subsections that follow, we discuss the main differences between \CEU and
@@
 %
-We present the formal specification of the semantics of \CEU in a separate 
-paper~\cite{ceu.phd}.
+We present the formal specification of the semantics of \CEU with a simplified
+abstract syntax in a separate paper~\cite{ceu.phd}.
 
@@
 %Some variants of Statecharts support local events which imply handling intra
-%reactions like Céu.
+%reactions like \CEU.
 In Statemate, ``reactions to external and internal events (...) can be sensed only
@@
 In Stateflow, ``the receiving state (of the event) acts here as a
-function''~\cite{statecharts.stateflow}, which is similar to Céu's stack-based
+function''~\cite{statecharts.stateflow}, which is similar to \CEU's stack-based
 execution.
-To avoid recursion, Céu adopts delayed awaits, while in Statemate, ``loops in
+To avoid recursion, \CEU adopts delayed awaits, while in Statemate, ``loops in
 the broadcasting of events (between states) are forbidden''.

### 2016-10-01 paper.tex
(>) @safe (>) finalize

@@
 %
-Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} to an 
-``intimidating name'' \code{\_redraw\_non\_commutative} (ln. 2, to be explicit 
-about its effect), and annotate it as safe (ln. 4--5) to make the code compile 
-successfully.
+Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
+\code{\_redraw\_non\_commutative} (ln. 2) and annotate it as safe (ln. 4--5) to
+make the code compile successfully.
+The new function name is explicit about its effect, improving code readability.
 
@@
 end
-@safe _redraw_non_commutative with
+native @safe _redraw_non_commutative with
              _redraw_non_commutative;
@@
 aborted and does not terminate normally.
+
+\CEU also enforces the use of \code{finalize} for system calls that deal with
+pointers representing resources, as illustrated in the two examples of
+Figure~\ref{lst.fin.ceu}:
+%
+\begin{itemize}
+\item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
+pointer represents a \textbf{local} resource (ln. \ax:2) that requires finalization
+(ln. \ax:7).
+\item If \CEU \textbf{receives} a pointer from a system call return (ln. \bx:4),
+the pointer represents an \textbf{external} resource (ln. \bx:2) that requires
+finalization (ln. \bx:6).
+\end{itemize}
+%
+\CEU tracks the interaction of system calls with pointers and requires 
+finalization clauses to accompany them.
+%
+In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
+(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
+(ln. 5), which is an asynchronous call that transmits the buffer in the 
+background.
+If the block aborts (ln. 11) before receiving an acknowledge from the 
+environment (ln. 9), the local \code{msg} goes out of scope and the external 
+transmission now holds a \emph{dangling pointer}.
+The finalization ensures that the transmission also aborts (ln. 7).
+%
+In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
+4) returns an external file resource as a pointer.
+If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
+remains open as a \emph{memory leak}.
+The finalization ensures that the file closes properly (ln. 6).
+%
+In both cases, the code does not compile without the \code{finalize} construct.
+
+\begin{figure}
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
+par/or do
+   var _buffer_t msg;
+   <...> // prepare msg
+   finalize
+      _send_request(&msg);
+   with
+      _send_cancel(&msg);
+   end
+   await SEND_ACK;
+with
+   <...>
+end
+.
+\end{lstlisting}
+\centering\small{\ax Local resource finalization}
+\end{minipage}
+%
+\begin{minipage}[t]{0.48\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+par/or do
+   var _FILE* f;
+   finalize
+      f = _fopen(...);
+   with
+      _fclose(f);
+   end
+   _fwrite(..., f);
+   await A;
+   _fwrite(..., f);
+with
+   <...>
+end
+\end{lstlisting}
+\centering\small{\bx External resource finalization}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{
+\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
+local resources and \emph{memory leaks} for external resources.
+\label{lst.fin.ceu}
+}
+\end{figure}
+
+\begin{comment}
 
@@
 externally aborted (ln. \ax:12 and \bx:12).
-
-\CEU also enforces the use of \code{finalize} for system calls that deal with
-pointers representing resources:
-%
-\begin{itemize}
-\item If \CEU \textbf{passes} a pointer to a system call, the pointer 
-represents a \textbf{local} resource that requires finalization.
-\item If \CEU \textbf{receives} a pointer from a system call return, the 
-pointer represents an \textbf{external} resource that requires finalization.
-\end{itemize}
-%
-\CEU tracks the interaction of system calls with pointers and requires 
-finalization clauses to accompany them.
-%
-In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
-(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
-(ln. 5), which is an asynchronous call that transmits the buffer in the 
-background.
-If the block aborts (ln. 11) before receiving an acknowledge from the 
-environment (ln. 9), the local \code{msg} goes out of scope and the external 
-transmission now holds a \emph{dangling pointer}.
-The finalization ensures that the transmission also aborts (ln. 7).
-%
-In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
-4) returns an external file resource as a pointer.
-If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
-remains open as a \emph{memory leak}.
-The finalization ensures that the file closes properly (ln. 6).
-%
-In both cases, the code does not compile without the \code{finalize} construct.
 
@@
 
-\begin{figure}[b]
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
-par/or do
-   var _buffer_t msg;
-   <...> // prepare msg
-   finalize
-      _send_request(&msg);
-   with
-      _send_cancel(&msg);
-   end
-   await SEND_ACK;
-with
-   <...>
-end
-.
-\end{lstlisting}
-\centering\small{\ax Local resource finalization}
-\end{minipage}
-%
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
-par/or do
-   var _FILE* f;
-   finalize
-      f = _fopen(...);
-   with
-      _fclose(f);
-   end
-   _fwrite(..., f);
-   await A;
-   _fwrite(..., f);
-with
-   <...>
-end
-\end{lstlisting}
-\centering\small{\bx External resource finalization}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{
-\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
-local resources and \emph{memory leaks} for external resources.
-\label{lst.fin.ceu}
-}
-\end{figure}
+\end{comment}
 

### 2016-10-01 paper.tex
(>) 200ms huge tick

@@
 In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
-which is a huge amount of time for a language-defined tick.
+which might be a huge amount of time for a language-defined tick.
 %, and which would break the synchronous hypothesis.
 %
-For instance, a similar implementation of this specification in Esterel would 
-not rely on the tick notion of simultaneity either.
+For instance, an implementation of this specification in Esterel would be
+similar, since it would not rely on the tick notion of simultaneity either.
 %

### 2016-10-01 paper.tex
(>) stack/lexical/deterministic

@@
 \item \emph{Deterministic intra-reaction execution and communication.}
-Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
-levels of static analysis.
+Determinism in \CEU does not depend on additional levels of static analysis.
 It encompasses the whole language, including memory accesses, system calls, and 

### 2016-10-01 paper.tex
(>) static checks algorithm (>) flat structure

@@
 stack-based internal events.
-Programmers can always figure out which statement executes next, making runtime 
-analysis and debugging easier.
+Programmers can figure out which statement executes next, helping runtime 
+analysis and debugging.
 %
@@
 %(the trail segment may reset it in sequence if it doesn't terminate).
+%
+Since the trail space is flat and carries no information about the program
+hierarchy (ln. \bx:12--15), all trails (active or inactive) are traversed in
+all passes.
+We believe that this makes reaction times more predictable and improves
+worst-case reaction times (e.g., with all trails active).
+However, we did not investigate these assumptions.
 

### 2016-10-02 paper.tex
(>) data dependency

@@
 %
-These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
-re-execution of statements in the same reaction (known as \emph{schizophrenic 
-statements}~\cite{esterel.constructive}).
-
-TODO: mesmo sem mutual dependency, falar de semantica construtiva e lei da
-coerencia
+This approach avoids any form of graph construction or data dependency analysis
+as required in Esterel, e.g.:
+``set or reset of a signal must always precede any test of this signal (signal
+coherence law and constructivity''~\cite{esterel.saxort}.
+%
+Also, these \emph{delayed awaits} avoid corner cases of instantaneous
+termination and re-execution of statements in the same reaction (known as
+\emph{schizophrenic statements}~\cite{esterel.constructive}).
 
@@
 the compiler slices programs into ``control points'' (analogous to our ``entry 
-points'') and rearranges them into a directed acyclic graph respecting the 
-constructive semantics of Esterel.
+points'') and rearranges them into a directed acyclic graph.
 Then, it flattens the graph into sequential code in \C suitable for static 
 scheduling.
-
-TODO: compilação mais sofisticada, tem que lidar com data dependency
+Esterel compilation is more sophisticated since it involves solving data
+dependency relations to comply with its constructive semantics.
 

### 2016-10-03 paper.tex
(>) sccharts

@@
 
+TODO:
 \begin{figure}[!htb]
@@
 %prevents the concurrency checks of \CEU.
+\emph{SCCharts}~\cite{sccharts.1} proposes \emph{sequentially constructive
+programs} that allow ``shared variables to have multiple values per tick as
+long as these values are explicitly ordered by sequential
+statements''~\cite{sccharts.2}.
+This relaxes Esterel's restriction for \emph{read-after-write} accesses only
+(e.g., a signal must be emitted to be present) and permits
+\emph{write-after-read} accesses typical in imperative programs (e.g., test and
+set).
+Although this approach still restricts concurrent accesses, it does not impose
+strict deterministic scheduling such as in \CEU.
 

### 2016-10-04 paper.tex
(>) Figura "terra"
(>) \emph all quotes

@@
 as required in Esterel, e.g.:
-``set or reset of a signal must always precede any test of this signal (signal
-coherence law and constructivity''~\cite{esterel.saxort}.
+\emph{``set or reset of a signal must always precede any test of this signal
+(signal coherence law and constructivity''}~\cite{esterel.saxort}.
 %
@@
 
-Esterel is only deterministic with respect to external behavior: ``the same 
-sequence of inputs always produces the same sequence of 
-outputs''~\cite{esterel.primer}.
+Esterel is only deterministic with respect to external behavior: \emph{``the
+same sequence of inputs always produces the same sequence of 
+outputs''}~\cite{esterel.primer}.
 %
 However, the execution order for operations within a reaction is 
-non-deterministic: ``if there is no control dependency, as in
+non-deterministic: \emph{``if there is no control dependency, as in
 \code{(call~f1()~||~call~f2())},
 the order is unspecified and it would be an error to rely on 
-it''~\cite{esterel.primer}.
+it''}~\cite{esterel.primer}.
 %
@@
 An Esterel implementation that compiles to sequential C code with similar
-techniques also generates ``code less than $10\%$ larger the original
-manually-sequenced code'' for an industrial-size
+techniques also generates \emph{``code less than $10\%$ larger the original
+manually-sequenced code''} for an industrial-size
 application~\cite{esterel.efficient}.
@@
 
-TODO:
 \begin{figure}[!htb]
     \centering
- %\includegraphics[width=0.5\columnwidth]{terra}
+    \includegraphics[width=0.5\columnwidth]{terra}
     \caption{Terra programming system basic elements.}
@@
 %reactions like \CEU.
-In Statemate, ``reactions to external and internal events (...) can be sensed only
-after completion of the step''~\cite{statecharts.statemate}, implying queue-based
-execution.
-In Stateflow, ``the receiving state (of the event) acts here as a
-function''~\cite{statecharts.stateflow}, which is similar to \CEU's stack-based
-execution.
-To avoid recursion, \CEU adopts delayed awaits, while in Statemate, ``loops in
-the broadcasting of events (between states) are forbidden''.
+In Statemate, \emph{``reactions to external and internal events (...) can be
+sensed only after completion of the step''}~\cite{statecharts.statemate},
+implying queue-based execution.
+In Stateflow, \emph{``the receiving state (of the event) acts here as a
+function''}~\cite{statecharts.stateflow}, which is similar to \CEU's
+stack-based execution.
+To avoid recursion, \CEU adopts delayed awaits, while in Statemate,
+\emph{``loops in the broadcasting of events (between states) are forbidden''}.
 
@@
 \emph{SCCharts}~\cite{sccharts.1} proposes \emph{sequentially constructive
-programs} that allow ``shared variables to have multiple values per tick as
-long as these values are explicitly ordered by sequential
-statements''~\cite{sccharts.2}.
+programs} that allow \emph{``shared variables to have multiple values per tick
+as long as these values are explicitly ordered by sequential
+statements''}~\cite{sccharts.2}.
 This relaxes Esterel's restriction for \emph{read-after-write} accesses only

### 2016-10-04 paper.tex
(>) many fixes

@@
 \begin{abstract}
-\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
-more fine-grained control over program execution targeting soft real-time
-systems.
+\CEU is a synchronous language targeting soft real-time systems.
+It is inspired by Esterel and has a simple semantics with fine-grained control
+over program execution.
 %
@@
 %
-Using Esterel as a base, we present the particularities of our design, such as 
-stack-based internal events, concurrency detection, safe integration with \C, 
+We present the particularities of our design in comparison to Esterel, such as 
+stack-based internal events, concurrency checks, safe integration with \C, 
 and first-class timers.
@@
 \CEU proposes a semantics for soft real-time systems with fine-grained control
-for intra-reaction execution, which is amenable to concurrency checks that
+for intra-reaction execution which is amenable to concurrency checks that
 improve safety.
@@
 It also makes difficult to statically determine worst-case reaction times
-for hard real-time systems~\cite{esterel.fixed}.
+towards hard real-time systems~\cite{esterel.fixed}.
 
@@
 
-Figure~\ref{lst.syn} presents the subset of the concrete syntax of \CEU used in
-the paper as a quick reference.
+Figure~\ref{lst.syn} shows the subset of the concrete syntax of \CEU used in
+this paper as a quick reference.
+%
+In a separate document, we present the formal semantics with a simplified
+abstract syntax~\cite{ceu.phd}.
 
@@
 event  <type> <ids>;                        // internal events
-var    <type> <id> = <exp>;                 // a variable with initial value
+var    <type> <id> = <exp>;                 // a variable with an initial value
 
@@
 
-<id> = await <id>;                          // await an event and assign received value
-<id> = await <time>;                        // await time and assign delayed delta
+<id> = await <id>;                          // await an event and assign the received value
+<id> = await <time>;                        // await time and assign the delayed delta
 emit <id> => <exp>;                         // emit an event passing a value
@@
 loop do <stmts> end                         // repetition
-do <stmts> end                              // lexical block
+do <stmts> end                              // explicit block
 
@@
 \end{lstlisting}
-\caption{ Subset of the concrete syntax of \CEU used in the paper.
+\caption{ Subset of the concrete syntax of \CEU used in this paper.
 \label{lst.syn}
@@
     (\ref{sec.ceu.summary}).
-%
-We present the formal specification of the semantics of \CEU with a simplified
-abstract syntax in a separate paper~\cite{ceu.phd}.
 
@@
 The box \code{Real World} assumes an external observer with an absolute
-reference clock that timestamps event occurrences over a
+reference clock which timestamps event occurrences over a
 continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
@@
 %
-Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.
+(Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.)
 %
@@
 \item \textbf{Events are never absolutely simultaneous:}
-We consider that the notion of simultaneity should not be imposed by the
-language, but defined explicitly for each use case (to be discussed in
-Section~\ref{sec.ceu.timers}).
+We consider that the notion of simultaneity should be defined explicitly for
+each use case (to be discussed in Section~\ref{sec.ceu.timers}).
 %In Section~\ref{sec.ceu.timers}, after introducing internal events and 
@@
 
-Note that Esterel supports declarations for mutually exclusive inputs which
+Note that Esterel supports declarations for mutually exclusive inputs that
 cannot be simultaneous in the same reaction
@@
 %
-\CEU takes a different approach, posing a tradeoff that an \code{await} is only
-valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
-simultaneously in parallel trails, the \code{await} does not awake.
-%
-This approach avoids any form of graph construction or data dependency analysis
-as required in Esterel, e.g.:
-\emph{``set or reset of a signal must always precede any test of this signal
-(signal coherence law and constructivity''}~\cite{esterel.saxort}.
-%
-Also, these \emph{delayed awaits} avoid corner cases of instantaneous
+\CEU takes a different approach, posing a restriction that an \code{await} is
+only valid for the next reaction, i.e., if an \code{await} and \code{emit}
+occur simultaneously in parallel trails, the \code{await} does not awake.
+%
+These \emph{delayed awaits} avoid corner cases of instantaneous
 termination and re-execution of statements in the same reaction (known as
 \emph{schizophrenic statements}~\cite{esterel.constructive}).
+%
+Also, this approach does not require data dependency analysis like Esterel
+(e.g., \emph{``(set or reset of a signal must always precede any test of this
+signal (signal coherence law and constructivity)''}~\cite{esterel.saxort}).
 
@@
 The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
-prevents infinite execution by design.
+prevents infinite execution in the loop.
 Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
@@
 instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
-\code{loop} instantaneously, resulting in infinite execution.
+\code{loop} also instantaneously, resulting in infinite execution.
 %
@@
 native do
-    ##define NUM 10
+    #define NUM 10
     void f  (void)  { <...> }
@@
 Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
-\code{\_redraw\_non\_commutative} (ln. 2) and annotate it as safe (ln. 4--5) to
-make the code compile successfully.
-The new function name is explicit about its effect, improving code readability.
+\code{\_redraw\_non\_commutative} (ln. 2), communicating its effect explicitly,
+and annotate it as safe (ln. 4--5) to make the code compile successfully.
 
@@
 %
-In both cases, the code does not compile without the \code{finalize} construct.
+In both cases, the code does not compile without the \code{finalize}
+construct.%
+\footnote{
+The compiler only enforces the programmer to write the finalization clause, but
+cannot check if it actually handles the resource properly.
+}
 
@@
 automatically, resulting in more robust applications.
-Due to platform overheads and occasional slow reactions, an \code{await 10ms}
-suspends a line of execution for \emph{at least} 10 milliseconds.
 In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
-\code{await} request, the underlying system gets busy and takes 15ms to notify 
+\code{await} request, the underlying system gets busy%
+\footnote{Due to platform overheads (e.g., OS scheduling) and occasional slow
+reactions, \code{await~10ms} means suspending a line of execution for
+\emph{at least} 10 milliseconds.
+}
+and takes 15ms to notify 
 \CEU.
-The scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
+The scheduler will notice that the \code{await~10ms} (ln. 2) has not only 
 already expired, but is delayed with \code{delta=5ms}.
@@
 infinitesimal and can never be absolutely simultaneous.
-However, the ``sensation of simultaneity'' is not infinitesimal, but actually 
-increases with the inaccuracy of the observer (e.g., a human being).
-Therefore, we consider that simultaneity should be defined case by case, and 
-should not be imposed by the language.
+%However, the ``sensation of simultaneity'' is not infinitesimal, but actually 
+%increases with the inaccuracy of the observer (e.g., a human being).
+%Therefore, we consider that simultaneity should be defined case by case, and 
+%should not be imposed by the language.
 %
@@
 \code{RIGHT\_CLICK} (ln. 1--2).
-If both events occur, we emit the internal event \code{middle\_click} (ln.  
-6--7).
+After both events occur in any order, we emit the internal event
+\code{middle\_click} (ln.  6--7).
 However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
@@
 We believe that programmers are more familiar with dealing with events in 
-isolation, which simplifies the reasoning about concurrency.
+isolation, simplifying the reasoning about concurrency.
 In addition, the uniqueness of external events is a prerequisite for the 
@@
 stack-based internal events.
-Programmers can figure out which statement executes next, helping runtime 
-analysis and debugging.
+Programmers can figure out which statement executes next following clear linear
+rules.
 %
@@
 
-The compile-time concurrency checks phase detects inconsistencies in \CEU 
-programs.
+The compile-time concurrency checks detect inconsistencies in \CEU programs.
 Here, we focus on the algorithm that detects suspicious concurrent statements, 
@@
 %
-\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
-\code{loop} and never proceeds to the statement immediately following it (see 
-also \code{loop} below);
-%
 \item A \emph{sequence node (;)} modifies each of its children to have 
@@
 %
+\item An \code{if} node has $O=O_{true} \cup O_{false}$, where \code{true} and 
+\code{false} are the two \code{if} branches.
+%
 \item A \code{loop} node includes the output of its body on its own $I$ ($I=I 
 \cup O_{body}$), as the loop is also reached from its own body.
-The union of all $O$ from nested \code{break} statements forms the set $O$ for 
-a \code{loop}.
-%
-\item An \code{if} node has $O=O_{true} \cup O_{false}$, where \code{true} and 
-\code{false} are the two \code{if} branches.
+The union of all $I$ from nested \code{break} statements forms the set $O$ for 
+a \code{loop}, i.e., entering a \code{break} means exiting a \code{loop}.
+%
+A \code{break} statement has $O=\{\}$ as it never proceeds to the statement
+immediately following it.
 %
@@
 %
-Single-threaded dispatching may not be suitable for hard real-time activities.
-We also measure how synchronous lengthy 
-computations in \C (e.g., hashing and compression) can block the scheduler and 
-affect higher-priority activities such as a radio driver.
+However, single-threaded dispatching may not be suitable for hard real-time
+activities.
+We also measure how synchronous lengthy computations in \C (e.g., hashing and
+compression) can block the scheduler and affect higher-priority activities such
+as a radio driver.
+Currently, we do not perform any worst-case reaction times 
+analysis~\cite{esterel.wcrt,esterel.fixed}.
 In such cases, the system requires careful testing to avoid undersized hardware 
 deployment.
-For instance, we currently do not perform any worst-case reaction times 
-analysis~\cite{esterel.wcrt,esterel.fixed}.
 
@@
 %
-An Esterel implementation that compiles to sequential C code with similar
-techniques also generates \emph{``code less than $10\%$ larger the original
-manually-sequenced code''} for an industrial-size
+An existing Esterel implementation compiles to sequential C code with similar
+techniques and also generates \emph{``code less than $10\%$ larger the
+original manually-sequenced code''} for an industrial-size
 application~\cite{esterel.efficient}.
@@
 %
-As far as we know, \CEU is the first to encrust this event-triggered notion of 
-time in the core of the language, which is a prerequisite for the concurrency 
-checks that enable safe shared-memory concurrency.
+\CEU encrusts this event-triggered notion of time in the core of the language,
+which is a prerequisite for the concurrency checks that enable safe
+shared-memory concurrency.
 
@@
 %
-Some variants of Statecharts~\cite{statecharts} also distinguish internal from
-external events~\cite{statecharts.variants}.
+Some variants of Statecharts also distinguish internal from external
+events~\cite{statecharts.variants}.
 %Some variants of Statecharts support local events which imply handling intra
@@
 %
-\CEU also performs concurrency checks to detect trails that, when reordered,
-change the observable behavior of the program, i.e., trails that actually rely
-on deterministic scheduling.
+In addition, \CEU also performs concurrency checks to detect trails that, when
+reordered, change the observable behavior of the program, i.e., trails that
+actually rely on deterministic scheduling.
 %
@@
 Although this approach still restricts concurrent accesses, it does not impose
-strict deterministic scheduling such as in \CEU.
+strict deterministic scheduling like in \CEU.
 
@@
 Esterel to program in terms of physical time with \code{delay} statements.
-The global transformation relies on a \emph{platform statement} the specifies
+The global transformation relies on a special platform statement the describes
 the available timers in the system (e.g., sample-driven or interrupt-driven).
@@
 \CEU makes physical time a special input event that feeds the runtime with an
-associated time to elapse which is decremented from all awaiting trails.
+associated time to elapse, which is decremented from all awaiting trails.
 The compensation scheme of \CEU guarantees that trails awake in the correct
@@
 %
-Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
-design choices to our work.
+An existing Esterel-based VM~\cite{esterel.vm} makes similar design choices to
+our work.
 To reduce code size, the \VM has a specialized instruction set to deal with 
 events and concurrency constructs that are particular to Esterel.
-However, the proposed \VM is only a proof of concept, with no support for 
+The proposed \VM is only a proof of concept, with no support for 
 arithmetic operations, external system calls, or remote reprogramming.
@@
 We present the design and implementation of \CEU, a synchronous reactive 
-language inspired by Esterel with event-driven semantics and more fine-grained 
+language inspired by Esterel with event-driven semantics and fine-grained 
 control for intra-reaction execution.
@@
 These properties are usually desirable in embedded applications and are 
-guaranteed as preconditions in \CEU by design.
+guaranteed by design.
 
@@
 no special requirements for OS threads or per-trail data stacks.
-The \VM implementation uses the same front end and imposes no extra 
+The \VM implementation uses the same front end and imposes no semantic 
 restrictions, being equally suitable for constrained systems.
@@
 lexically scoped parallel compositions, convenient first-class timers, and a 
-unique stack-based mechanism for internal signalling.
+stack-based mechanism for internal signalling.
 %

### 2016-10-06 paper.tex
(>) Noemi corrections.

@@
 sequential scheduler.
-In contrast with Esterel's semantics, this prevents programs to be compiled
-directly and efficiently to hardware~\cite{esterel.constructive}.
+Unlike Esterel's semantics, this prevents programs to be compiled directly and
+efficiently to hardware~\cite{esterel.constructive}.
 It also makes difficult to statically determine worst-case reaction times
@@
 %
-Also, this approach does not require data dependency analysis like Esterel
+Also, unlike Esterel, this approach does not require data dependency analysis
 (e.g., \emph{``(set or reset of a signal must always precede any test of this
@@
 tree (\emph{AST});
-the \emph{concurrency checks phase} detects inconsistencies in programs, such as 
-unbounded loops and suspicious concurrent statements;
+the \emph{concurrency checks phase} detects inconsistencies in programs, such
+as unbounded loops and suspicious concurrent statements;
 the \emph{code generation phase} converts the \emph{AST} to standard \C code 
-and augments it with platform-dependent functionality (e.g., system calls) and the 
-runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
+and augments it with platform-dependent functionality (e.g., system calls) and
+the runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
 binary.
+%
+An alternative implementation generates bytecode for an embedded virtual
+machine, which can disseminate and install programs dynamically.
 
@@
 Note that, as discussed in Section~\ref{sec.ceu.checks}, the concurrency 
-checks of \CEU infer precisely trails that are concurrent and yet do not 
-share resources.
+checks of \CEU detect trails that are concurrent and yet do not share
+resources.
 Hence, these non-conflicting trails could potentially execute with real 
@@
 However, our experiments with multi-core execution are actually slower than 
-single-core execution in the same system.
+single-core execution on the same system.
 Considering that we use \CEU primarily in control-dominated applications, this 
@@
 %
+This approach aims to combine the flexibility of remotely uploading code with 
+the expressiveness and safety guarantees of \CEU.
+%
 Figure \ref{fig.terra} shows the three basic elements of Terra:
@@
 bytecode images dynamically.
-%
-This approach aims to combine the flexibility of remotely uploading code with 
-the expressiveness and safety guarantees of \CEU.
 
@@
 
-\CEU has a strong influence from Esterel but differ in the most fundamental 
+\CEU was strongly influenced by Esterel but they differ in the most fundamental 
 aspect of the notion of time.
 %
-In \CEU, instead of clock ticks, atomic external event occurrences that define 
-time units.
+In \CEU, instead of clock ticks, atomic external event occurrences define time
+units.
 %
@@
 \CEU encrusts this event-triggered notion of time in the core of the language,
-which is a prerequisite for the concurrency checks that enable safe
-shared-memory concurrency.
+and is a prerequisite for the concurrency checks that enable safe shared-memory
+concurrency.
 
@@
 %
-Some variants of Statecharts also distinguish internal from external
-events~\cite{statecharts.variants}.
+Some variants of the Statecharts synchronous visual language also distinguish
+internal from external events~\cite{statecharts.variants}.
 %Some variants of Statecharts support local events which imply handling intra
 %reactions like \CEU.
-In Statemate, \emph{``reactions to external and internal events (...) can be
-sensed only after completion of the step''}~\cite{statecharts.statemate},
+In Statemate~\cite{statecharts.statemate}, \emph{``reactions to external and
+internal events (...) can be sensed only after completion of the step''},
 implying queue-based execution.
-In Stateflow, \emph{``the receiving state (of the event) acts here as a
-function''}~\cite{statecharts.stateflow}, which is similar to \CEU's
+In Stateflow~\cite{statecharts.stateflow}, \emph{``the receiving state (of the
+event) acts here as a function''}, which is similar to \CEU's
 stack-based execution.
@@
 
-Like \CEU, many synchronous languages rely on deterministic scheduling to 
-preserve intra-reaction determinism (%
+Like other synchronous languages
+(%
     \emph{Reactive C}~\cite{rp.rc},                 % 1991
@@
     and
-\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
+    \emph{PRET-C}~\cite{rp.pretc}%                  % 2010
+),
+\CEU also relies on deterministic scheduling to preserve intra-reaction
+determinism.
 %
@@
 %
-In addition, \CEU also performs concurrency checks to detect trails that, when
+In addition, it also performs concurrency checks to detect trails that, when
 reordered, change the observable behavior of the program, i.e., trails that
@@
 %prevents the concurrency checks of \CEU.
-\emph{SCCharts}~\cite{sccharts.1} proposes \emph{sequentially constructive
-programs} that allow \emph{``shared variables to have multiple values per tick
-as long as these values are explicitly ordered by sequential
+The \emph{SCCharts} formalism~\cite{sccharts.1} proposes \emph{sequentially
+constructive programs} that allow \emph{``shared variables to have multiple
+values per tick as long as these values are explicitly ordered by sequential
 statements''}~\cite{sccharts.2}.
@@
 set).
-Although this approach still restricts concurrent accesses, it does not impose
-strict deterministic scheduling like in \CEU.
+This approach still restricts concurrent accesses but does not impose
+strict deterministic scheduling as \CEU does.
 
@@
 Esterel to program in terms of physical time with \code{delay} statements.
-The global transformation relies on a special platform statement the describes
-the available timers in the system (e.g., sample-driven or interrupt-driven).
+The global transformation relies on a special platform statement that describes
+available timers in the system (e.g., sample-driven or interrupt-driven).
 It then expands \code{delay} statements into existing Esterel statements in a
@@
 %
-\CEU makes physical time a special input event that feeds the runtime with an
-associated time to elapse, which is decremented from all awaiting trails.
+In contrast, \CEU makes physical time a special input event that feeds the
+runtime with an associated time to elapse, which is decremented from all
+awaiting trails.
 The compensation scheme of \CEU guarantees that trails awake in the correct
@@
 Interrupt-driven timers are also supported with a hook callback that the
-runtime calls whenever the program awaits an earlier timer.
+runtime calls whenever the program awaits on an earlier timer.
 
@@
 %
-In addition, \CEU also tracks pointers representing resources that cross \C 
-boundaries and forces the programmer to provide associated finalizers.
+As a complement to finalization, \CEU also tracks pointers representing
+resources that cross \C boundaries and forces the programmer to provide
+associated finalizers.
 
@@
 More recently, a company specialized in embedded systems (not related to our 
-research group) released a product based on \CEU.
+research group) has released a product based on \CEU.
 

### 2016-10-06 paper.tex
(>) Noemi corrections.

@@
 \footnote{
-The compiler only enforces the programmer to write the finalization clause, but
+The compiler only forces the programmer to write the finalization clause, but
 cannot check if it actually handles the resource properly.
@@
 stack-based internal events.
-Programmers can figure out which statement executes next following clear linear
-rules.
+Programmers can figure out which statement executes next following clear
+sequential rules.
 %

### 2016-10-07 paper.tex
(>) Comments Noemi/Adriano

@@
 
-\CEU aims to offer a simpler semantics than Esterel with more determinism and 
-fine-grained control over program execution.
+\CEU aims to offer a simple semantics with internal determinism and
+fine-grained control over execution.
 %
@@
 synchronization.}
-Given the omnipresence of timers in embedded systems, a dedicated syntax can 
-simplify the development and readability of programs.
+Given the demand of timers in embedded systems, a dedicated syntax can simplify
+the development and readability of programs.
 Furthermore, automatic synchronization releases the programmer from the burden 
@@
 
-Our synchronous and deterministic approach also leads to some limitations as 
-follows:
+Our approach for synchronous and deterministic execution also leads to some
+limitations as follows:
 %

### 2017-01-08 paper.tex
(>) typo
(>) images
(>) apply minor suggestions

@@
 Despite its success and influence, Esterel has a complex semantics that
-requires careful static analysis to detect and refuse programs with 
+requires careful static analysis to detect and reject programs with 
 \emph{causality} and \emph{schizophrenia} problems%
@@
 programs, but also affects the programmer's understanding about the code, who,
-ultimately, has to solve the errors when facing corner cases.
+ultimately, has to solve the errors arising from corner cases.
 %
@@
 Our implementations target resource-constrained devices, such as \emph{Arduino} 
-and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
+and \emph{MICAz} sensor nodes based on 8-bit microcontrollers,%
 \footnote{
@@
 microcontroller with 32K of FLASH and 2K of SRAM.
-}, showing a practical aspect of the proposed semantics.
+} showing a practical aspect of the proposed semantics.
 %:\url{http://www.atmel.com/devices/atmega328.aspx}}
@@
 sequential scheduler.
-Unlike Esterel's semantics, this prevents programs to be compiled directly and
+Unlike Esterel's semantics, this precludes programs being compiled directly and
 efficiently to hardware~\cite{esterel.constructive}.
-It also makes difficult to statically determine worst-case reaction times
-towards hard real-time systems~\cite{esterel.fixed}.
+It also makes difficult to statically determine worst-case reaction times for
+hard real-time systems~\cite{esterel.fixed}.
 
@@
 %
-Internal computations within a reaction (e.g. expressions, assignments, and 
+Internal computations within a reaction (e.g., expressions, assignments, and 
 system calls) are considered to take no time in accordance with the synchronous 
@@
 loops are statically required to contain at least one \code{await} statement in 
-all possible paths~\cite{ceu.sensys13,esterel.primer}.
+each possible path~\cite{ceu.sensys13,esterel.primer}.
 %
@@
 Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
-languages, showing side-by-side the implementations in Esterel \ax and \CEU \bx 
-for the following control specification:
+languages, showing side-by-side the implementations in Esterel \ax and \CEU~\bx 
+of the following control specification:
 %
@@
 implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
-equivalent.
+analogous.
 %
@@
       parallel constructs may create new trails).
-\item Active trails execute until they await or terminate, one after the other.
+\item Active trails execute until they await or terminate, one after another.
       This step is named a \emph{reaction chain}, and always runs in bounded 
@@
 %
-Based on the synchronous hypothesis, a program takes a negligible time on step 
+Based on the synchronous hypothesis, a program takes negligible time on step 
 2 and is always idle on step 3.
@@
 
-\begin{figure}[!htb]
+\begin{figure}
 \centering
@@
 %
-(Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.)
+(Usually referred to as \emph{sample-driven execution}~\cite{rp.twelve}.)
 %
@@
 
-Sample-driven execution maps directly to hardware implementations and
-involve no runtime queues.
+Sample-driven execution maps directly to hardware implementations and does not
+require runtime queues.
 %
@@
 For instance, if we log the \code{Real World} box timeline and reproduce it in
-implementations with different tick lengths, the behaviors might diverge.
+implementations with different tick lengths, the behaviors might differ.
 %
@@
 Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
-Esterel \ax and \CEU \bx.
+Esterel \ax and \CEU~\bx.
 %
@@
 Internal events provide fine-grained execution control and can express a 
-limited form of subroutines, as depicted in Figure~\ref{lst.sub}.
+limited form of subroutines, as described in Figure~\ref{lst.sub}.
 The ``subroutine'' \code{inc} is defined as a loop (ln. 3--6) that continuously 
@@
 awaits the next ``call'' (ln. 4, again).
-Only after this sequence, the calling trail resumes and passes the assertion 
+Only after this sequence does the calling trail resumes and passes the assertion 
 test (ln. 10--11).
@@
 characterizes the enclosing \code{loop} as non instantaneous (ln 2--9).
-However, if the \code{emit e} (ln. 6) could awake the \code{await e} 
+However, if the \code{emit~e} (ln. 6) could awake the \code{await e} 
 instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
@@
 %
-The detection algorithm, which is depicted in Section~\ref{sec.impl.checks},
+The detection algorithm, which is described in Section~\ref{sec.impl.checks},
 inspects all possible \code{await} statements that precede a variable access
@@
 This is the case for logging (e.g., calls to \code{\_printf} in trails in 
-parallel) and for redrawing objects in the screen.
+parallel) and for redrawing objects on the screen.
 %
@@
 objects \code{background} and \code{foreground} in trails in parallel.
-In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
-conflicting pixels, which makes the calls non commutative and prevents the code 
-to compile.
+In typical graphical APIs, consecutive calls to \code{\_redraw} overwrite 
+conflicting pixels, which makes the calls non-commutative and prevents the code 
+from compiling.
 %
@@
 When dealing with concurrent system calls, programmers can provide annotations 
-to reduce false positives in the static checks, or to force non commutative 
+to reduce false positives in the static checks, or to force non-commutative 
 concurrent behavior.
@@
 synchronization.}
-Given the demand of timers in embedded systems, a dedicated syntax can simplify
+Given the need for timers in embedded systems, a dedicated syntax can simplify
 the development and readability of programs.
@@
 
-The compilation process of programs in \CEU is composed of three main phases:
+The compilation of \CEU programs consists of three main phases:
 the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
@@
 With all sets calculated, we take all pairs of nodes that perform side effects 
-and are in parallel branches, and compare their sets $I$ for intersections.
+and are in parallel branches, and compare their sets $I$ for non-empty
+intersections.
 For each pair, if the intersection is not empty, we mark both nodes as 
@@
 The \code{do-end} blocks and \code{par/and} in sequence
-(ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
+(ln. \ax:2--4,5--7,8--12) are packed into a \code{union} (ln. \bx:2,3,4--7), 
 given that their variables cannot be in scope at the same time, e.g., \code{a} 
@@
 they have to be aborted.
-Given that trails in parallel are allocated in subsequent slots in the static 
-vector \code{TRLS} (ln. \bx:12--15), this pass only aborts the desirable 
+Given that trails in parallel are allocated in successive slots in the static 
+vector \code{TRLS} (ln. \bx:12--15), this pass only aborts the appropriate 
 trails.
@@
 %
-TinyOS provides open-source industrial-level applications and serve as basis
+TinyOS provides open-source industrial-level applications and serves as basis
 for a considerable amount of research in the Wireless Sensor Network academic
@@
 The generated code has been tested in real hardware and also in congested
-networks simulated in sofware~\cite{ceu.sensys13}.
+networks simulated in software~\cite{ceu.sensys13}.
 %
@@
 network protocols and a radio driver.
-The small overhead in resource usage shows that the gains in productivity and 
-safety with \CEU make it a viable alternative to \C in the context of 
+The small overhead in resource usage suggests that the gains in productivity
+and safety with \CEU make it a viable alternative to \C in the context of 
 constrained embedded systems.
@@
 
-In future work, we plan to perform an equivalent throughout analysis in
+In future work, we plan to perform an equivalent thorough analysis in
 comparison to industrial-quality code bases in Esterel.
@@
 
-\begin{figure*}[b]
+\begin{figure*}
 \includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
@@
 that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds to 
-advance the wall-clock time of \CEU (ln. 15)%
+advance the wall-clock time of \CEU (ln. 15).%
 \footnote{We also offer a mechanism to start the underlying timer on demand to
-avoid the ``battery unfriendly'' 10ms polling.}.
+avoid the ``battery unfriendly'' 10ms polling.}
 The remaining lines map pre-defined TinyOS events that can be used in \CEU 
@@
 
-\begin{figure}[!htb]
+\begin{figure}
     \centering
@@
 
-In Terra, \CEU scripts cannot execute arbitrary \C code, instead, they rely on 
+In Terra, \CEU scripts cannot execute arbitrary \C code, instead they rely on 
 pre-built components that can be customized for different application domains.
@@
 In contrast, the template-based \C back end (illustrated in 
-Figure~\ref{lst.impl.tinyos}) allows applications to choose all possible 
+Figure~\ref{lst.impl.tinyos}) allows applications to choose possible 
 combinations of functionalities from the underlying platform at compile time.
@@
 %
-\CEU encrusts this event-triggered notion of time in the core of the language,
-and is a prerequisite for the concurrency checks that enable safe shared-memory
-concurrency.
+\CEU incorporates this event-triggered notion of time into the core of the
+language, and is a prerequisite for the concurrency checks that enable safe
+shared-memory concurrency.
 
@@
 Esterel has different compilation back ends that synthesizes to software and 
-also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
+also to hardware circuits~\cite{esterel.constructive,esterel.emperor,esterel.tutorial}.
 Among the software-based approaches,
@@
 %
+The \emph{The real-time specification for Java}~\cite{java.rtsj,java.rtsj.ovm}
+provides region-based memory that is not subject to garbage collection, and
+threads that are not preemptable by the collector.
+%
 Java has antagonistic design choices in comparison to \CEU:
@@
 abstracting tick sampling with reactions to unique events from their
-application domain.
+application domain (e.g., a button click or an expiring timer).
 On the other hand, this level of abstraction does not suit systems with
-hard-real time requirements in which interacting directly with a concrete
+hard real-time requirements in which interacting directly with a concrete
 notion of tick is more robust.

### 2017-01-08 paper.tex
(>) timers

@@
 The \code{await} statement of \CEU supports wall-clock time and handles deltas 
-automatically, resulting in more robust applications.
+automatically. %, resulting in more robust applications.
 In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
@@
 reasoning about the execution behavior more difficult.
+
+The \code{await} statement for timers returns the current \emph{delta}, as
+illustrated in Figure~\ref{lst.delta}.
+Figure~\ref{lst.delta}.a is a variation of Figure~\ref{lst.timers}.a in which
+the second \code{await} adds the current delta to annul it and guarantee that
+\code{v} remains set to \code{1} during 1 millisecond.
+Figure~\ref{lst.delta}.b deals with the possible case in which a small $1us$
+timer in a loop never ``catches up'' with the external clock, resulting in a
+\emph{delta} that increases indefinitely.
+
+\begin{figure}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}
+var int v;
+var int late = await 10ms;
+v = 1;
+await (late+1000)us;
+v = 2;
+
+.
+\end{lstlisting}
+\centering\small{\ax}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}
+loop do
+    var int late = await 1us;
+    if late > 1000 then     // I am late 1000x 1us?
+        <...>               // abnormal behavior
+    else
+        <...>               // normal behavior
+    end
+end
+\end{lstlisting}
+\centering\small{\bx}
+\end{minipage}
+%\rule{8.5cm}{0.37pt}
+\caption{ The \code{await} statement for timers returns the current
+\emph{delta} in microseconds.
+\label{lst.delta}
+}
+\end{figure}
 
